### 2023-11-01 t1_analysis.Rmd
initial commit

---
title: "Varying paths to building the social state in the progressive era"
author: "Frank Edwards"
date: "2023-08-15"
output: bookdown::html_document2
---

```{r setup, include = F, message = F}
knitr::opts_chunk$set(message = F, echo = F, warning = F)
library(tidyverse)
library(lme4)
library(maps)
library(usmap)
library(grid)
library(gridExtra)
options(scipen = 999)
options(dplyr.print_max = 1e9)
map<-map_data("state")

theme_set(theme_minimal())
```

# Theoretical framework

(GENERAL FRAMEWORK: Racial capitalism / settler colonization) MATERIAL CONDITIONS FOR SETTLER COLONIZATION STRUCTURE POLICY DESIGN. socio-political processes generate diverse white elite interests

1. Frontier settler polity -> elite interest in disposession / assimiatilion -> aggressive paternalistic welfare state for AIAN, social democracy for settlers
2. Jim Crow -> elite interest in exploitation and exclusion -> elite paternalism, malign neglect, hard coercion
3. Metropole settler polity -> elite interests in labor discipline / immig assimilation -> progressive paternalism, strong coercive welfare state for W, hard coercion and contested welfare for Blk

THESE ARE MULTIVARIATE INDICATORS, CAN USE CLUSTERING TO CHECK IT AND MEASURE IT.

1: MEASURE VARIATION IN CONTEXT, ASSUME CONTEXT PRODUCES INTERESTS. 2: MEASURE VARIATION IN POLICY DEVELOPMENT.

3 dimensions: frontier, antiBlack, white immig assim (urban)

frontier: {homestead claims, tribal density, https://www.census.gov/dataviz/visualizations/001/}
railrooad (corporate) + homestaed (individual) + morrill (state) - nice three piece approach, more complex and complete story.

antiblack: {jim crow polit econ (sharecropping), prop black}

urban assim: {urbanization, industrial employment?, urban white immig, log black pop}

Following Wolfe, DuBois, etc, variation in the interests of elites across places (material + cultural) drives variation in processes of racialization across places. The state is a key vector through which racialization is implemented. Early 20th c. saw rapid development and expansion of social state. Can we observe differences in social state developmental trajectories as a function of variation in elite interests in particular racial projects? Are regional differences in social policy regimes a function of the social, demographic, and economic interests of regional elites?

Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native. Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion. Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines.

Interests of social policy actors defined within these fields. Bourdieusian historical stuff?

Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development. This can lock in particular racial regimes over time as social policy does work on social groups.

interests can be observed indirectly sometimes, right? homesteading is direct evidence of interest. tenant farming (or land ownership) is direct evidence of interest. industrial urban is tougher, but maybe there we just go indirect and use urbanization + racial demography.

I THINK THIS WORKS. CLUSTER ON INTERESTS + DEMOGRAPHY. OUTCOMES ARE POLICY MEASURES: POLICY OUTCOMES (EG EMPLOYMENT, SPENDING, GROUP QUARTERS POPS, ENROLLMENTS).

TRY TO DO ONLY A FEW BIG NATIONAL PARAMETRIC TESTS. LOTS OF DESCRIPTIVE SUBSET WORK

TEST 1: clustering, spatial clustering

then TONS of descriptives to paint pictures of regional social / political contexts

TEST 2: regression on outcomes

negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1

positive case: white immigrants -> more social workers in t1

```{r read, cache = T}
### read employment data
### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
dat_occ<-read_csv("./data/usa_00071.csv") %>% 
  rename_all(tolower) %>% 
  mutate(occ1950 = 
           case_when(
             occ1950 == 79 ~ "socwork",
             occ1950 == 773 ~ "police",
             occ1950 == 771 ~ "police",
             occ1950 == 782 ~ "police",
             occ1950 == 93 ~ "teachers"
           )) %>% 
  group_by(year, statefip, occ1950) %>% 
  summarize(emp = sum(perwt)) %>% 
  pivot_wider(names_from = occ1950, values_from = emp,
              values_fill = 0)
### AK OR HI HERE? check that at IPUMS

# pop data (total, age, race, nativitiy, urban)
dat_immig<-read_csv("./data/usa_00070.csv") %>% 
  rename_all(tolower) %>% 
  group_by(year, statefip) %>% 
  summarize(tot_pop = sum(perwt),
            children = sum((age<18) * perwt),
            white_immig = 
              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
            blk = 
              sum((race == 2) * perwt),
            urban = sum(hhwt * (urban ==2)),
            tot_hh = sum(hhwt)) 
### AK OR HI HERE? check that at IPUMS

### group quarters data, 100% 
gq<-read_csv("./data/usa_00069.csv") %>% 
  rename_all(tolower) %>% 
  filter(age<18) %>% 
  group_by(year, statefip, gqtype) %>% 
  summarize(gq_pop = sum(perwt, na.rm = T)) %>% 
  mutate(gqtype = case_when(
    gqtype == 2 ~ "child_inst_correctional",
    gqtype == 3 ~ "child_inst_mental",
    gqtype == 4 ~ "child_inst_poor_disabled")) %>% 
  filter(gqtype!="other") %>% 
  pivot_wider(names_from = gqtype, values_from = gq_pop,
              values_fill = 0)
### AK OR HI HERE? check that at IPUMS

### join
dat <- data.frame(state = state.abb, 
               region = state.region, 
               division = state.division) %>% 
  left_join(state.fips %>% 
              select(fips, abb) %>% 
              distinct() %>% 
              rename(state = abb, statefip = fips)) %>% 
  left_join(dat_immig) %>% 
  left_join(dat_occ) %>% 
  ungroup() %>% 
  left_join(gq)

### homestead act data
#### load in homestead data and boarding school data
### state measure can be cumulative sum of homesteads, that gets at polity
### homesteading data
y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
  left_join(dat %>% 
              select(state, region) %>% 
              distinct())

hs<-read_csv("./data/blm_homesteads_clean.csv") %>% 
  pivot_longer(claims1868:acres1961,
               names_to = "name",
               values_to = "value") %>% 
  mutate(year = str_sub(
    name, start = -4, end = -1),
    type = str_sub(
      name, start = 1, end = -5)) %>% 
  select(-name) %>% 
  mutate(year = as.numeric(year)) %>% 
  mutate(value = ifelse(is.na(value),
                        0,
                        value))

### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
hs_dakota<-hs %>% 
  filter(state == "Dakota Territory") %>% 
  mutate(value = value / 2) %>% 
  select(-state)

hs<-hs %>% 
  bind_rows(hs_dakota %>% 
              mutate(state = "North Dakota")) %>% 
  bind_rows(hs_dakota %>% 
              mutate( state = "South Dakota")) %>% 
  filter(state != "Dakota Territory") %>% 
  group_by(state, year, type) %>% 
  summarize(value = sum(value))

hs<-hs %>% 
  arrange(state, year) %>% 
  group_by(state, type) %>% 
  mutate(cum_value = cumsum(value)) %>% 
  ungroup()

## expand_grid to add zeroes
### use y_o_s for index
state<-unique(y_o_s$statename)
year<-unique(hs$year)
type<-unique(hs$type)
temp<-expand_grid(state, year, type)

hs<-temp %>% 
  left_join(hs) %>% 
  replace_na(list(value = 0,
                  cum_value = 0))

hs <- hs %>% 
  select(-value) %>% 
  pivot_wider(names_from = type,
              values_from = cum_value,
              names_prefix = "hs_")

### add ID and pop vars back to hs

y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
  left_join(dat %>% 
              select(state, region) %>% 
              distinct()) %>% 
  filter(!(is.na(region)))

hs<-hs %>% 
  rename(statename = state) %>% 
  left_join(y_o_s)

#### join cumulative acres / claims onto dat for the valid census years, then carried forward

dat<-dat %>% 
  left_join(hs)
```

# methods

here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.

## NOTES ON GEOGRAPHIES and MEASUREMENT

### STATES, TERRITORIES, AND NATIVE LANDS

TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION. CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION. SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.

STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).

### DIRECT AND PROXY MEASURES FOR ELITE SETTLER INTERESTS

We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states. In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity.

The works of settler colonial and racial politics historical and theoretical scholars suggest a straightforward approach to selecting measures that are either a) direct, or b) proxy measures for elite white settler interests given a few assumptions about relationships between population distributions and political processes.

#### Measuring anti-Blackness and post-Reconstruction racial politics

1) POP COMP (BLACK) - VERY CLEAR

  Explain

2) jim crow salience

3)

#### Measuring the size, power, and interests of the settler polity in active conflict zones

2) HOMESTEAD CLAIMS - VERY CLEAR

  Explain

3) PRESENCE OF NATIVE LAND HOLDINGS - VERY CLEAR

  Explain

#### Measuring the magnitude of social problems Progressive reformers would seek to 'fix'

4) URBANIZATION + IMMIGRANT POP - CLEAR-ISH. HOW DOES THIS RELATE TO SETTLER INTERESTS?

  Explain

# Findings

1) walk through the context variables descriptively
2) walk through the state formation variables descriptively
3) estimate a few multivariate models
4) do I need a DAG + causal ID?

## The distributions of Black populations in the US

```{r blk-ts, fig.cap = "Black population by state 1870-1940"}
p1<-ggplot(dat,
       aes(x = year, y = blk / tot_pop,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  geom_text(data = dat %>% 
              filter(!(is.na(region))) %>% 
              filter(year == 1930),
            aes(x = year,
                y = blk / tot_pop,
                label = state),
            position = position_dodge(width = 8),
            size = 2,
            alpha = 0.7) + 
  labs(subtitle = "Panel 2. Proportion Black",
       y = "Proportion",
       x = "Year") + 
  facet_wrap(~region)

p2<-ggplot(dat,
       aes(x = year, y = blk,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  geom_text(data = dat %>% 
              filter(!(is.na(region))) %>% 
              filter(year == 1930),
            aes(x = year,
                y = blk,
                label = state),
            position = position_dodge(width = 8),
            size = 2,
            alpha = 0.7) + 
  # scale_y_log10() + 
  labs(subtitle = "Panel 1. Black population",
       y = "Count",
       x = "Year") + 
  facet_wrap(~region)

grid.arrange(p2, p1, nrow = 1)
```

```{r}
dat %>% 
  filter(year == 1920) %>% 
  mutate(state = tolower(statename)) %>% 
  ggplot(aes(fill = log(blk))) + 
  geom_map(aes(map_id = state), map = map) + 
  expand_limits(x = map$long, y = map$lat) + 
  scale_fill_continuous(type = "viridis") + 
  theme_void() + 
  theme(legend.position = "bottom")
```

Figure \@ref(fig:blk-ts) shows two sets of time series describing the Black population in US States between 1870 and 1930 using the US Census. Panel 1 shows the size of the Black population in counts, represented on a logarithmic scale to allow for easier comparisons of population order of magnitude. Panel 2 shows the proportion of a state's population reported as Black on the US Census, to allow for comparisons of population composition across places.

Black populations increased uniformly across all US states between 1870 and 1930 (see panel 1), though there are clear differences in rates of increase and absolute levels across states. For example, in the Northeast region, the primarily rural states of Maine, Vermont, and New Hampshire had relatively small Black populations across the time period. In Maine, reported Black populations were between 300 and 2400 during this period (1% sample data, NO standard errors REPORTED HERE because it's a pain but I'll keep working on it). By contrast, states with large and growing cities, like New York, New Jersey, and Pennsylvania all had large Black populations that grew substantially between 1870 and 1930. For example, Pennsylvania's Black population was recorded at about 70,000 in 1870, and 430,000 in 1930.

Note that some places described in these plots were not admitted as states at the beginning of this time period (CHECK US CENSUS QUALITY IN TERRITORIES), and were intensely contested in military and non-military conflicts between the US government and tribal nations. MORE HERE. HOW DOES CONTESTATION ENTER INTO THE STAT PROCESS? CENSUS IS COUNTING SETTLERS + ENSLAVED, SO THATS STILL VALID FOR CONTEXTS OF SET COL ASSUMING GOOD TERRITORIAL MEASUREMENT (CHECK THIS!).

Similar population trends are evident in the North Central region (commonly referred to as the Midwest). States with large cities, like Illinois, Ohio, Missouri, and Michigan had both relatively high Black populations at the beginning of the period and saw substantial growth in these populations over time. For example, the US Census recorded about 31,000 Black residents in Illinois in 1870. By 1930, the Census recorded about 330,000 Black residents of Illinois. Note that some of these places were not admitted as states at the beginning of this time period. (CHECK US CENSUS QUALITY IN TERRITORIES), and were sites of ongoing military conflicts between the US government and tribal nations.

In the Western Region, states reported uniformly small Black populations at the beginning of this period, with varying (but positive) rates of growth. Note that many of these places were not yet states at the beginning of the time series; only California, Oregon, and Nevada had been admitted as states by 1870. California recorded the largest Black population by the end of the period. The Census records an increase in Black population in California from about 3000 in 1870 to about 84,000 in 1930. Arizona, Colorado, and Washington had smaller but substantial Black populations by the end of the time period, at around 10,000 persons.

In the Southern Region (which includes all states that seceded to join the former Confederate States of America in addition to Delaware, Maryland, West Virginia, Kentucky, and Oklahoma), Black populations were consistently large during this period, with the largest populations in states in which white elites had enslaved millions of Black people for work in plantation-driven agriculture, like Georgia, Mississippi, and South Carolina. For example, the Census recorded 540,000 Black residents of Georgia in 1870, increasing to 1,100,000 Black residents in 1930.

Note though, that these increases in population size do not always mirror trends in relative population composition (see Panel 2). During this period, even as Black populations grew dramatically in some Northeastern and Midwestern states during the first Great Migration, proportional population shares of Black residents remained low and relatively stable. In the South, even as Black populations continued to grow in absolute terms, the proportional population share of Black residents was decreasing consistently throughout the region.

This divergence between absolute and relative population trends illustrates that Black population share likely fails to adequately capture the salience of race for local political contexts.

IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?

## The spatial distribution of the frontier settler polity

### HOMESTEAD ACT

We use data on Homestead Act claims as an indicator of the volume of settlers at the time and place-level seeking to actively dispossess Native Nations of their lands. Data on homestead claims provide useful indicators of both the demographic size of the settler polity (measured as counts of claimants) and the magnitude of their resources (measured as acres claimed). COUPLED WITH WHAT ELSE?

Figure \@ref(fig:ts-homestead-claims) shows the distribution of Homestead Act claims made by settlers between 1868 and 1961. No homestead claims were completed in the states in the Northeast region. In general, states in the North Central (Midwest) and Western regions had higher volumes of claims, though Oklahoma and Arkansas in the South also saw relatively large numbers of claims. By the end of the Homestead era, Montana had seen the largest number of claimants (151,500), followed by North Dakota (135,448), South Dakota (114,172), Colorado (107,618), and Nebraska (104,260).

```{r ts-homestead-claims, fig.cap = "Cumulative Homestead act claims by state and year"}
ggplot(hs %>% 
         filter(year < 1940),
       aes(x = year,
           y = hs_claims,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  geom_text(data = hs %>% 
              filter(year == 1935),
            aes(x = year,
                y = hs_claims,
                label = state),
            position = position_nudge(x = 8),
            alpha = 0.7) + 
facet_wrap(~region)  + 
  labs(y = "Number of claims",
       x = "Year",
       subtitle = "") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

### ALTERNATIVE?
# 
# ggplot(hs %>% 
#          filter(year < 1940),
#        aes(x = year,
#            y = hs_claims,
#            group = state,
#            color = region)) + 
#   geom_line(alpha = 0.3) + 
#   geom_text(data = hs %>%
#               filter(year == 1940,
#                      hs_claims > 10000),
#             aes(x = year,
#                 y = hs_claims,
#                 label = state),
#             position = position_nudge(x = 2),
#             size = 3) +
#   geom_line(data = hs %>% 
#               group_by(year, region) %>%
#               filter(year < 1940) %>% 
#               summarize(hs_claims = mean(hs_claims)),
#             aes(x = year, y = hs_claims, color = region,
#                 group = region), linewidth = 2, lty = 2, alpha = 0.7) +
#   labs(y = "Number of claims",
#        x = "Year",
#        subtitle = "Panel 1.") + 
#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

THIS IS COOL! Look at homestead (as settler polity) plus mode (as boarding school on / off)

LOOK AT ACRES PER CLAIM? GET AT BOTH POP SIZE AND RELATIVE RESOURCES?

```{r}
ggplot(hs %>% 
         filter(year == 1961) %>% 
         mutate(region = factor(region, 
                                levels = c("West", 
                                           "North Central",
                                           "South",
                                           "Northeast"))),
       aes(x = hs_acres / 1e6,
           y = fct_reorder(state, hs_acres))) + 
  geom_point() + 
  labs(x = "Acres claimed (millions)",
       y = "",
       color = "Census Region") + 
  facet_wrap(~region, 
             scales = "free_y")
```

pop / claims ratio is a little funky, but probably useful. can peg to 1940 census.

WOWZERS. 
20+pct of pop likely had HS claim land in 3 states.

Wow, let's pull through this. ONE INDICATOR OF SETTLER POLITY. DEmographic base is one dimension - cultures / institutions/ resources also key. but useful!

```{r}
dat %>% 
  filter(year == 1940) %>% 
  mutate(region = factor(region, 
                         levels = c("West", 
                                    "North Central",
                                    "South",
                                    "Northeast"))) %>% 
  mutate(hs_pop = hs_claims / tot_pop) %>% 
  ggplot(aes(x = hs_pop,
             y = fct_reorder(state, hs_pop))) + 
  geom_point() + 
  labs(x = "Cumulative homestead claims per state population, 1940",
       y = "",
       color = "Census Region") + 
  facet_wrap(~region, 
             scales = "free_y")
```

```{r}
dat %>% 
  filter(year == 1940) %>% 
  mutate(hs_pop = hs_claims / tot_pop) %>% 
  mutate(state = tolower(statename)) %>% 
  ggplot(aes(fill = hs_pop)) + 
  geom_map(aes(map_id = state), map = map) + 
  expand_limits(x = map$long, y = map$lat) + 
  scale_fill_continuous(type = "viridis") + 
  theme_void() + 
  theme(legend.position = "bottom")
```

### white immigrants and cities

WRITE IT UP

```{r immig-ts, fig.cap = "White immigrant population by state 1870-1940"}
p1<-ggplot(dat,
       aes(x = year, y = white_immig / tot_pop,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Proportion white immigrant",
       y = "Proportion",
       x = "Year") + 
  facet_wrap(~region)

p2<-ggplot(dat%>% 
  filter(!(is.na(region))),
       aes(x = year, y = white_immig,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  scale_y_log10() + 
  labs(subtitle = "White immigrant population",
       y = "Count (note log10 axis)",
       x = "Year") + 
  facet_wrap(~region)

grid.arrange(p2, p1, nrow = 1)
```

Relative population share as first generation immigrant declines in NC across period uniformly. In NE, increase until 1920 for most states, then monotonic decline. Low levels in S. Declining levels in W.

Absolute white immigrant pop numbers increasing in NE, Increasing, then plateau in NC, increasing then plateau in W, and stable and low in S.

### urban

```{r urban-ts, fig.cap = "Urban population by state 1870-1940"}
p1<-ggplot(dat%>% 
  filter(!(is.na(region))),
       aes(x = year, y = urban / tot_hh,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Proportion of households urban",
       y = "Proportion",
       x = "Year") + 
  facet_wrap(~region)

p2<-ggplot(dat%>% 
  filter(!(is.na(region))),
       aes(x = year, y = urban,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Number of households urban",
       y = "Count",
       x = "Year")+ 
  facet_wrap(~region)

grid.arrange(p2, p1, nrow = 1)
```

increase in proportion of households in urban across regions

Also clear increase in absolute n of households in urban settings.

# maps of focal measures

# distributions of policy outcomes

### the distribution of social state capacity building

CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho...

we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.

let's look at how it played out in the progressive era across states and regions. we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.

First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita?

We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states. A census taker is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.). We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.

```{r emp-ts-plot, fig.cap="Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}
dat %>% 
  filter(year > 1880) %>% 
  group_by(state, year) %>% 
  summarize(police = sum(police) / sum(tot_pop) * 1e3,
            teachers = sum(teachers) / sum(tot_pop) * 1e3,
            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
  pivot_longer(cols = police:socworkers,
               names_to = "name",
               values_to = "employment")  %>%  
  filter(year < 1940) %>% 
  ggplot(aes(x = year, y = employment)) + 
  geom_line(aes(group = state), alpha = 0.2) + 
  geom_line(data = dat %>% ### add national line
              group_by(year, region) %>%
              filter(year > 1880, year < 1940) %>% 
              summarize(police = sum(police, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3,
                        teachers = sum(teachers, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3,
                        socworkers = sum(socwork, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3) %>%
              pivot_longer(cols = police:socworkers,
                           names_to = "name",
                           values_to = "employment"),
  aes(x = year, y = employment, color = region), linewidth = 1.2) +
  facet_wrap(~name) + 
  scale_y_log10() + 
  labs(subtitle = "note variable y axis scales, national average dashed")
```

During this period, we see very clear upward trajectories in employment in each of three categories: police, social workers, and teachers. At the national level, police employment nearly doubled from about 0.7 per 1,000 in 1900 to 1.3 per 1,000 population in 1930. 
Data on social workers shows an employment rate of 5 per 10,000 in 1900 nationally doubling to a rate of 1 per 1,000 by 1930. Teachers shows 1900 rate of 5.5 per 1,000, 1930 of 8.6 per 1,000.

Figure 1 shows substantial variation in state employment initial values and trajectories during this period (thin solid lines in fig 1). There is a clear upward trajectory in employment rates for each class of worker. Social workers experienced the largest relative rate of growth (a feature of the near-zero initial rate of employment in most states), while teachers experience the largest absolute rate of growth nationally.

```{r tab1}
### tabular mean / variance
t<-dat %>% 
  group_by(state, year) %>% 
  summarize(police = sum(police) / sum(tot_pop) * 1e3,
            teachers = sum(teachers) / sum(tot_pop) * 1e3,
            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
  pivot_longer(cols = police:socworkers,
               names_to = "name",
               values_to = "employment")  %>% 
  group_by(year, name) %>% 
  summarize(mean = mean(employment, na.rm=T),
                        sd = sd(employment, na.rm=T)) %>% 
  filter(year == 1900 | year == 1930)

knitr::kable(t, 
             caption = "Table 1. Mean and variance of state employment rates 1900 - 1930",
             digits = 3)
```

There is substantial heterogeneity in 1900 values for social policy employment across states. The standard deviation of social work employment is greater than the mean value of social work employment. For police, the sd is half the value of the mean, and for teachers it is one third of the value of the mean. At 1930, the sd for social work employment had increased dramatically, as had the sd for police employment. The sd for teacher employment was relatively stable over the period. These features show clear variation in starting conditions of employment during the period and suggest variation in trajectories of growth over the period.

Some of this heterogeneity in starting conditions and trajectories is likely a feature of regional variation in policy configurations (which we will seek to explain variation in racial political economy as we proceed).

```{r emp-ts-region, fig.cap = "Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}

### CUT

dat %>% 
  filter(!(is.na(region))) %>% 
  group_by(region, year) %>% 
  summarize(police = sum(police) / sum(tot_pop) * 1e3,
            teachers = sum(teachers) / sum(tot_pop) * 1e3,
            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
  pivot_longer(cols = police:socworkers,
               names_to = "name",
               values_to = "employment")  %>%  
  ggplot(aes(x = year, y = employment, 
             color = region)) + 
  geom_line() + 
  facet_wrap(~name, scales = "free", ncol = 3) + 
  labs(subtitle = "note variable y axis scales")
```

In figure 2 we subset the data from figure 1 to regional averages using the four-region US Census classification for states. Clear patterns are visible in the data here. On average, states in the US South had lower employment rates for all categories of social policy system employees. Northeastern states had the highest levels of police and social work employment across the period. And states in the West and North Central (Midwest) had the highest average rates of teacher employment. However, there is also substantial sub-regional variation in employment rates, as shown in Figure 3.

```{r emp-ts-region2, fig.cap = "Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}
dat %>% 
  group_by(state, region, year) %>% 
  filter(!is.na(region)) %>% 
  summarize(police = sum(police) / sum(tot_pop) * 1e3,
            teachers = sum(teachers) / sum(tot_pop) * 1e3,
            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
  pivot_longer(cols = police:socworkers,
               names_to = "name",
               values_to = "employment")  %>%  
  ggplot(aes(x = year, y = employment, group = state)) + 
  geom_line(alpha = 0.3) + 
  facet_wrap(region~name, scales = "free", ncol = 3) + 
  labs(subtitle = "note variable y axis scales")
```

Figure 3 shows how states within regions still exhibit substantial variation in both initial employment rates and growth trajectories between 1900 and 1930 for social policy employment. Again, note that social work employment began at or near zero in 1900 in all states as the social work was built as a concrete profession. In the South a cluster of low-employment states drive the regional trend for social work and police, while a cluster of high employment states appear to drive the trend for social work employment in the Northeast, North Central, and West.

```{r emp-map, fig.cap = "Employment rates per 1,000 1930"}
t<-dat %>% 
  group_by(state, region, year) %>% 
  summarize(police = sum(police) / sum(tot_pop) * 1e3,
            teachers = sum(teachers) / sum(tot_pop) * 1e3,
            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
  pivot_longer(cols = police:socworkers,
               names_to = "name",
               values_to = "employment")  %>%  
  filter(year == 1930)

p1<-plot_usmap(data = t %>% 
             filter(name == "socworkers"),
           values = "employment")+ 
  scale_fill_viridis_c() + 
  labs(title = "social workers",
       fill = "") + 
  theme(legend.position = "bottom")

p2<-plot_usmap(data = t %>% 
             filter(name == "teachers"),
           values = "employment")+ 
  scale_fill_viridis_c() + 
  labs(title = "teachers",
       fill = "") + 
  theme(legend.position = "bottom")

p3<-plot_usmap(data = t %>% 
             filter(name == "police"),
           values = "employment")+ 
  scale_fill_viridis_c() + 
  labs(title = "police per 1,000",
       fill = "") + 
  theme(legend.position = "bottom")

grid.arrange(p1, p2, p3, ncol = 3)

```

Figure 4 shows these distributions as state choropleth maps for 1930 employment per 1,000. It clearly shows the geographic distribution of clusters of high and low employment states. Northeastern states like New York, New Jersey, and Massachusetts had very high levels of police and social work employment. Midwestern states like Minnesota, Michigan, and Illinois also had high social work and police employment, as did California. Midwestern and Plains states like North and South Dakota, Wyoming, and Iowa had especially high levels of teacher employment.

Employment provides a measure of the magnitude of state and private investment in social control and social services (aside from transfers - think about distinction b/w hands on and redistribution). But this is one dimension of infrastructure. Given the ambiguity in 'social state capacity', we are working multiple measures to think this through.

#### kids in institutional group quarters

Next, let's turn to reach of social state policy systems. One way to get at that is through the existence and utilization of era-specific policy institutions. Here, we will look at the distribution of children recorded as being housed in 'institutional group quarters' on the 1900 - 1930 decennial censuses. To improve the quality of measurement, I use the 100% decennial data for 1900, 1920, 1930 and 1940. The 1910 census data available through IPUMS only includes detailed group quarters information on the 1% and 1.4% samples and after exploratory work, they do not appear to be continuous with the 100% data. I exclude 1910 data for now.

Group quarters indicate non-traditional housing situations (e.g. not in a house or apartment). Here, we subset this group to include only children recorded as residents of 'correctional institutions', 'mental institutions', or 'institutions for the elderly, handicapped, and poor'. This excludes children recorded as living in rooming houses, military housing, other non-institutional settings. Our goal here is to capture the magnitude of efforts by both public and private actors to coercively or therapeutically detain children for purposes of social control or social welfare.

```{r gq-ts, fig.cap = "Children in institutional group quarters 1900 - 1930"}
dat %>% 
  filter(year!=1910,
         !(is.na(region))) %>% 
  mutate(across(child_inst_correctional:child_inst_poor_disabled, 
                ~ .x/children * 1e3)) %>% 
  pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
               names_to = "type",
               values_to = "pop") %>% 
  ggplot(aes(x = year, y = pop)) + 
  geom_line(aes(group = state),
            alpha = 0.2) + 
  geom_line(data = dat %>% ### add national line
              filter(year!=1910,
                     !(is.na(region))) %>%
              group_by(year) %>% 
              summarize(across(child_inst_correctional:child_inst_poor_disabled, 
                               ~ sum(.x, na.rm=T)/sum(children, na.rm=T) * 1e3)) %>% 
              pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
                           names_to = "type",
                           values_to = "pop"),            
            aes(x = year, y = pop), linewidth = 2, lty = 2) + 
  facet_wrap(~type, scales = "free") + 
  labs(y = "institutional population per 1,000 children",
       subtitle = "note varying y axes. National average dashed, state values solid")

```

These plots strongly suggest that there are measurement artifacts present in the census data for these measures. Many states have broken time series with zeroes observed in 1920 between non-zero rates in 1900 and 1930, or large jumps between periods that defy expected linear growth patterns that revert in later periods. I think it makes sense to treat these with caution.

I'd like to use total group quarters, but it includes rooming houses, which are a huge proportion of the count. COULD JUST USE GQ == 3 (Institutions). That may be the ticket.

```{r gq-box, fig.cap = "Children in institutional group quarters 1930"}
dat %>% 
  filter(year == 1930) %>% 
  mutate(across(child_inst_correctional:child_inst_poor_disabled, 
          ~ .x/children * 1e3)) %>% 
  pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
               names_to = "type",
               values_to = "pop") %>% 
  ggplot(aes(x = type, y = pop, 
             color = region)) + 
  geom_boxplot()

```

### agents data eda

```{r}
agents<-read_csv("./data/AGENTS_DATA_COLLECTION.csv")

agents1895<-agents %>% 
  filter(YEAR == 1895)

state_counts95<-agents1895 %>% 
  group_by(`STATE/SUPERINTENDENCY`) %>% 
  summarize(enrolled = sum(STUDENTS_ENROLL, na.rm=T))
```

NEED A STATE MAPPING ST SOME POINT
I WANT TO GET STUDENT POPS BY STATE...

USE BIA DATA FOR NOW FOR RESIDENTIAL / NON

```{r}
bs<-read_csv("./data/Boarding_Schools.csv")

bs <- bs %>% 
  group_by(State, ST_Reservation) %>% 
  summarize(n = n()) %>% 
  filter(!(is.na(ST_Reservation)))
```

### 2023-11-02 t1_analysis.Rmd
modeling beginnings

@@
 author: "Frank Edwards"
-date: "2023-08-15"
+date: "last updated 11/1/23"
 output: bookdown::html_document2
@@
 ```
+
+# TO DO
+
+1. pull immig + urban data
+2. pull tenant farmer data
+3. model employment -> each, consider loglinear process?
+
+theory re: white immigs: coercion / rehab in NE, yeoman uplift in W, frontier as panacea for urban, lots of hist on that, gets at SocWork/Police vs Teachers employment
 
@@
 ```{r read, cache = T}
-### read employment data
-### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
-dat_occ<-read_csv("./data/usa_00071.csv") %>% 
-  rename_all(tolower) %>% 
-  mutate(occ1950 = 
-           case_when(
-             occ1950 == 79 ~ "socwork",
-             occ1950 == 773 ~ "police",
-             occ1950 == 771 ~ "police",
-             occ1950 == 782 ~ "police",
-             occ1950 == 93 ~ "teachers"
-           )) %>% 
-  group_by(year, statefip, occ1950) %>% 
-  summarize(emp = sum(perwt)) %>% 
-  pivot_wider(names_from = occ1950, values_from = emp,
-              values_fill = 0)
-### AK OR HI HERE? check that at IPUMS
+# ### on libra, read raw
+# ### read employment data
+# ### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
+# dat_occ<-read_csv("./data/usa_00071.csv") %>% 
+#   rename_all(tolower) %>% 
+#   mutate(occ1950 = 
+#            case_when(
+#              occ1950 == 79 ~ "socwork",
+#              occ1950 == 773 ~ "police",
+#              occ1950 == 771 ~ "police",
+#              occ1950 == 782 ~ "police",
+#              occ1950 == 93 ~ "teachers"
+#            )) %>% 
+#   group_by(year, statefip, occ1950) %>% 
+#   summarize(emp = sum(perwt)) %>% 
+#   pivot_wider(names_from = occ1950, values_from = emp,
+#               values_fill = 0)
+# ### AK OR HI HERE? check that at IPUMS
+# 
+# # pop data (total, age, race, nativitiy, urban)
+# dat_immig<-read_csv("./data/usa_00070.csv") %>% 
+#   rename_all(tolower) %>% 
+#   group_by(year, statefip) %>% 
+#   summarize(tot_pop = sum(perwt),
+#             children = sum((age<18) * perwt),
+#             white_immig = 
+#               sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+#             blk = 
+#               sum((race == 2) * perwt),
+#             urban = sum(hhwt * (urban ==2)),
+#             tot_hh = sum(hhwt)) 
+# ### AK OR HI HERE? check that at IPUMS
+# 
+# 
+# ### group quarters data, 100% 
+# gq<-read_csv("./data/usa_00069.csv") %>% 
+#   rename_all(tolower) %>% 
+#   filter(age<18) %>% 
+#   group_by(year, statefip, gqtype) %>% 
+#   summarize(gq_pop = sum(perwt, na.rm = T)) %>% 
+#   mutate(gqtype = case_when(
+#     gqtype == 2 ~ "child_inst_correctional",
+#     gqtype == 3 ~ "child_inst_mental",
+#     gqtype == 4 ~ "child_inst_poor_disabled")) %>% 
+#   filter(gqtype!="other") %>% 
+#   pivot_wider(names_from = gqtype, values_from = gq_pop,
+#               values_fill = 0)
+# ### AK OR HI HERE? check that at IPUMS
+# 
+#  
+# ### join
+# dat <- data.frame(state = state.abb, 
+#                region = state.region, 
+#                division = state.division) %>% 
+#   left_join(state.fips %>% 
+#               select(fips, abb) %>% 
+#               distinct() %>% 
+#               rename(state = abb, statefip = fips)) %>% 
+#   left_join(dat_immig) %>% 
+#   left_join(dat_occ) %>% 
+#   ungroup() %>% 
+#   left_join(gq)
+# 
+# ### homestead act data
+# #### load in homestead data and boarding school data
+# ### state measure can be cumulative sum of homesteads, that gets at polity
+# ### homesteading data
+# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
+#   left_join(dat %>% 
+#               select(state, region) %>% 
+#               distinct()) 
+# 
+# hs<-read_csv("./data/blm_homesteads_clean.csv") %>% 
+#   pivot_longer(claims1868:acres1961,
+#                names_to = "name",
+#                values_to = "value") %>% 
+#   mutate(year = str_sub(
+#     name, start = -4, end = -1),
+#     type = str_sub(
+#       name, start = 1, end = -5)) %>% 
+#   select(-name) %>% 
+#   mutate(year = as.numeric(year)) %>% 
+#   mutate(value = ifelse(is.na(value),
+#                         0,
+#                         value))
+# 
+# ### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
+# hs_dakota<-hs %>% 
+#   filter(state == "Dakota Territory") %>% 
+#   mutate(value = value / 2) %>% 
+#   select(-state)
+# 
+# hs<-hs %>% 
+#   bind_rows(hs_dakota %>% 
+#               mutate(state = "North Dakota")) %>% 
+#   bind_rows(hs_dakota %>% 
+#               mutate( state = "South Dakota")) %>% 
+#   filter(state != "Dakota Territory") %>% 
+#   group_by(state, year, type) %>% 
+#   summarize(value = sum(value))
+#   
+# hs<-hs %>% 
+#   arrange(state, year) %>% 
+#   group_by(state, type) %>% 
+#   mutate(cum_value = cumsum(value)) %>% 
+#   ungroup()
+# 
+# ## expand_grid to add zeroes
+# ### use y_o_s for index
+# state<-unique(y_o_s$statename)
+# year<-unique(hs$year)
+# type<-unique(hs$type)
+# temp<-expand_grid(state, year, type)
+# 
+# hs<-temp %>% 
+#   left_join(hs) %>% 
+#   replace_na(list(value = 0,
+#                   cum_value = 0))
+# 
+# hs <- hs %>% 
+#   select(-value) %>% 
+#   pivot_wider(names_from = type,
+#               values_from = cum_value,
+#               names_prefix = "hs_") 
+# 
+# ### add ID and pop vars back to hs
+# 
+# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
+#   left_join(dat %>% 
+#               select(state, region) %>% 
+#               distinct()) %>% 
+#   filter(!(is.na(region)))
+# 
+# hs<-hs %>% 
+#   rename(statename = state) %>% 
+#   left_join(y_o_s)
+# 
+# #### join cumulative acres / claims onto dat for the valid census years, then carried forward
+# 
+# dat<-dat %>% 
+#   left_join(hs)
 
-# pop data (total, age, race, nativitiy, urban)
-dat_immig<-read_csv("./data/usa_00070.csv") %>% 
-  rename_all(tolower) %>% 
-  group_by(year, statefip) %>% 
-  summarize(tot_pop = sum(perwt),
-            children = sum((age<18) * perwt),
-            white_immig = 
-              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
-            blk = 
-              sum((race == 2) * perwt),
-            urban = sum(hhwt * (urban ==2)),
-            tot_hh = sum(hhwt)) 
-### AK OR HI HERE? check that at IPUMS
-
-### group quarters data, 100% 
-gq<-read_csv("./data/usa_00069.csv") %>% 
-  rename_all(tolower) %>% 
-  filter(age<18) %>% 
-  group_by(year, statefip, gqtype) %>% 
-  summarize(gq_pop = sum(perwt, na.rm = T)) %>% 
-  mutate(gqtype = case_when(
-    gqtype == 2 ~ "child_inst_correctional",
-    gqtype == 3 ~ "child_inst_mental",
-    gqtype == 4 ~ "child_inst_poor_disabled")) %>% 
-  filter(gqtype!="other") %>% 
-  pivot_wider(names_from = gqtype, values_from = gq_pop,
-              values_fill = 0)
-### AK OR HI HERE? check that at IPUMS
-
-### join
-dat <- data.frame(state = state.abb, 
-               region = state.region, 
-               division = state.division) %>% 
-  left_join(state.fips %>% 
-              select(fips, abb) %>% 
-              distinct() %>% 
-              rename(state = abb, statefip = fips)) %>% 
-  left_join(dat_immig) %>% 
-  left_join(dat_occ) %>% 
-  ungroup() %>% 
-  left_join(gq)
-
-### homestead act data
-#### load in homestead data and boarding school data
-### state measure can be cumulative sum of homesteads, that gets at polity
-### homesteading data
-y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
-  left_join(dat %>% 
-              select(state, region) %>% 
-              distinct())
-
-hs<-read_csv("./data/blm_homesteads_clean.csv") %>% 
-  pivot_longer(claims1868:acres1961,
-               names_to = "name",
-               values_to = "value") %>% 
-  mutate(year = str_sub(
-    name, start = -4, end = -1),
-    type = str_sub(
-      name, start = 1, end = -5)) %>% 
-  select(-name) %>% 
-  mutate(year = as.numeric(year)) %>% 
-  mutate(value = ifelse(is.na(value),
-                        0,
-                        value))
-
-### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
-hs_dakota<-hs %>% 
-  filter(state == "Dakota Territory") %>% 
-  mutate(value = value / 2) %>% 
-  select(-state)
-
-hs<-hs %>% 
-  bind_rows(hs_dakota %>% 
-              mutate(state = "North Dakota")) %>% 
-  bind_rows(hs_dakota %>% 
-              mutate( state = "South Dakota")) %>% 
-  filter(state != "Dakota Territory") %>% 
-  group_by(state, year, type) %>% 
-  summarize(value = sum(value))
-
-hs<-hs %>% 
-  arrange(state, year) %>% 
-  group_by(state, type) %>% 
-  mutate(cum_value = cumsum(value)) %>% 
-  ungroup()
-
-## expand_grid to add zeroes
-### use y_o_s for index
-state<-unique(y_o_s$statename)
-year<-unique(hs$year)
-type<-unique(hs$type)
-temp<-expand_grid(state, year, type)
-
-hs<-temp %>% 
-  left_join(hs) %>% 
-  replace_na(list(value = 0,
-                  cum_value = 0))
-
-hs <- hs %>% 
-  select(-value) %>% 
-  pivot_wider(names_from = type,
-              values_from = cum_value,
-              names_prefix = "hs_")
-
-### add ID and pop vars back to hs
-
-y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
-  left_join(dat %>% 
-              select(state, region) %>% 
-              distinct()) %>% 
-  filter(!(is.na(region)))
-
-hs<-hs %>% 
-  rename(statename = state) %>% 
-  left_join(y_o_s)
-
-#### join cumulative acres / claims onto dat for the valid census years, then carried forward
-
-dat<-dat %>% 
-  left_join(hs)
+# on local read temp
+dat<-read_csv("./data/temp_data.csv")
+hs<-read_csv("./data/temp_hs.csv")
 ```
@@
 
-#### Measuring anti-Blackness and post-Reconstruction racial politics
+### Measuring anti-Blackness and post-Reconstruction racial politics
 
-1) POP COMP (BLACK) - VERY CLEAR
+IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
 
-  Explain
+1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
+2) exclusion / formal soc control salience in urban non-S
 
-2) jim crow salience
-
-3)
-
-#### Measuring the size, power, and interests of the settler polity in active conflict zones
-
-2) HOMESTEAD CLAIMS - VERY CLEAR
-
-  Explain
-
-3) PRESENCE OF NATIVE LAND HOLDINGS - VERY CLEAR
-
-  Explain
-
-#### Measuring the magnitude of social problems Progressive reformers would seek to 'fix'
-
-4) URBANIZATION + IMMIGRANT POP - CLEAR-ISH. HOW DOES THIS RELATE TO SETTLER INTERESTS?
-
-  Explain
-
-# Findings
-
-1) walk through the context variables descriptively
-2) walk through the state formation variables descriptively
-3) estimate a few multivariate models
-4) do I need a DAG + causal ID?
-
-## The distributions of Black populations in the US
+POPULATION ALONE DESCRIPTIVES
 
@@
 
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = log(blk))) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() + 
-  theme(legend.position = "bottom")
-```
-
 Figure \@ref(fig:blk-ts) shows two sets of time series describing the Black population in US States between 1870 and 1930 using the US Census.
@@
 
-IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
+```{r}
+dat %>% 
+  filter(year == 1930) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = log(blk))) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis") + 
+  theme_void() + 
+  theme(legend.position = "bottom")
+```
 
-## The spatial distribution of the frontier settler polity
+### Measuring the size, power, and interests of the expropriative settler polity
 
-### HOMESTEAD ACT
+2) HOMESTEAD CLAIMS - VERY CLEAR
+
+  Explain
+
+3) PRESENCE OF NATIVE LAND HOLDINGS - VERY CLEAR
+
+  Explain
 
@@
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
-
-### ALTERNATIVE?
-# 
-# ggplot(hs %>% 
-#          filter(year < 1940),
-#        aes(x = year,
-#            y = hs_claims,
-#            group = state,
-#            color = region)) + 
-#   geom_line(alpha = 0.3) + 
-#   geom_text(data = hs %>%
-#               filter(year == 1940,
-#                      hs_claims > 10000),
-#             aes(x = year,
-#                 y = hs_claims,
-#                 label = state),
-#             position = position_nudge(x = 2),
-#             size = 3) +
-#   geom_line(data = hs %>% 
-#               group_by(year, region) %>%
-#               filter(year < 1940) %>% 
-#               summarize(hs_claims = mean(hs_claims)),
-#             aes(x = year, y = hs_claims, color = region,
-#                 group = region), linewidth = 2, lty = 2, alpha = 0.7) +
-#   labs(y = "Number of claims",
-#        x = "Year",
-#        subtitle = "Panel 1.") + 
-#   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
-
 ```
@@
 dat %>% 
-  filter(year == 1940) %>% 
+  filter(year == 1920) %>% 
   mutate(region = factor(region, 
@@
   geom_point() + 
-  labs(x = "Cumulative homestead claims per state population, 1940",
+  labs(x = "Cumulative homestead claims per state population, 1920",
        y = "",
@@
 dat %>% 
-  filter(year == 1940) %>% 
+  filter(year == 1920) %>% 
   mutate(hs_pop = hs_claims / tot_pop) %>% 
@@
 
-### white immigrants and cities
+### Measuring the magnitude of social problems Progressive reformers would seek to 'fix'
 
-WRITE IT UP
+PULL IMMIG + URBAN - UNIQUE COMBO THAT PUSHES FORWARD WELFARE STATE
 
@@
 
-# maps of focal measures
+```{r}
+dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = log(white_immig))) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis") + 
+  theme_void() + 
+  theme(legend.position = "bottom")
+```
 
@@
   ggplot(aes(x = year, y = employment)) + 
-  geom_line(aes(group = state), alpha = 0.2) + 
+  geom_line(aes(group = state), alpha = 0.1) + 
   geom_line(data = dat %>% ### add national line
@@
                            values_to = "employment"),
-  aes(x = year, y = employment, color = region), linewidth = 1.2) +
+  aes(x = year, y = employment, color = region), linewidth = 1) +
   facet_wrap(~name) + 
   scale_y_log10() + 
-  labs(subtitle = "note variable y axis scales, national average dashed")
+  labs(subtitle = "note variable y axis scales, national average dashed",
+       y = "Employment per 1,000 population (note log scale)") + 
+  theme(legend.position = "bottom")
 ```
@@
   geom_line() + 
-  facet_wrap(~name, scales = "free", ncol = 3) + 
+  scale_y_log10() + 
+  facet_wrap(~name, ncol = 3) + 
   labs(subtitle = "note variable y axis scales")
@@
 
-### agents data eda
+## PRELIMINARY MODELS
+
+employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.
+
+q1: salience of settler expropriotion -> state formation
+
+four models, each employment outcome + children in poverty institutions
+
+start with lm for log(rates)
+
+trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.
+
+switch to brms for estimation on tricky ones
 
 ```{r}
-agents<-read_csv("./data/AGENTS_DATA_COLLECTION.csv")
+### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE
+model_dat <- dat %>% 
+  filter(year>=1880) %>% 
+  mutate(police_rate = (police + 1) / tot_pop * 1e5,
+         socwork_rate = (socwork + 1) / tot_pop * 1e5,
+         teacher_rate = (teachers + 1) / tot_pop * 1e5,
+         child_inst_rate = (child_inst_poor_disabled + 0.1) / 
+           children * 1e5,
+         hsclaim_prop = hs_claims / tot_pop,
+         year = scale(year))
 
-agents1895<-agents %>% 
-  filter(YEAR == 1895)
+sc_police<-lmer(log(police_rate) ~ 
+              scale(hsclaim_prop) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
 
-state_counts95<-agents1895 %>% 
-  group_by(`STATE/SUPERINTENDENCY`) %>% 
-  summarize(enrolled = sum(STUDENTS_ENROLL, na.rm=T))
+sc_teacher<-lmer(log(teacher_rate) ~ 
+              scale(hsclaim_prop) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
+
+sc_socwork<-brm(log(socwork_rate) ~ 
+              scale(hsclaim_prop) +
+              year + 
+              (1 + year|state),
+            data = model_dat)
+
+sc_gq<-brm(log(child_inst_rate) ~ 
+              scale(hsclaim_prop) +
+              year + 
+              (1 + year|state),
+            data = model_dat)
 ```
 
-NEED A STATE MAPPING ST SOME POINT
-I WANT TO GET STUDENT POPS BY STATE...
+This all looks fine. How do I want to visualize this?
 
-USE BIA DATA FOR NOW FOR RESIDENTIAL / NON
+we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930?
+Think about how to handle to intercepts / slopes.
+What would a zero mean?
+pulling out everything not directly in national trend and regime measures.
+is that what I'm after?
+I guess I really do just want the regime-specific trend lines, in which case E[rate | regime] should zero out the individual idioscyncracies.
+it's oversimplified, but that should be ok.
+Just be super clear on what the iterpretation is on prediction.
+it's not a hypothetical state, its the mean expectation for the slope of the regime var.
+
+model is
+
+\[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\]
+
+Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.
+
+Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?
 
 ```{r}
-bs<-read_csv("./data/Boarding_Schools.csv")
+new_dat<-data.frame(year =  c(1.3, 1.3, 1.3),
+                    hsclaim_prop = c(0, 0.05, 0.15))
+t<-predict(sc_police, 
+           newdata = new_dat,
+           re.form = ~0,
+           type = "link")
 
-bs <- bs %>% 
-  group_by(State, ST_Reservation) %>% 
-  summarize(n = n()) %>% 
-  filter(!(is.na(ST_Reservation)))
+t1<-predict(sc_teacher, 
+           newdata = new_dat,
+           re.form = ~0,
+           type = "link")
+
+exp(t1)
 ```

### 2023-11-03 t1_analysis.Rmd
add models

@@
            children * 1e5,
+         prop_blk = (blk + 1) / tot_pop,
          hsclaim_prop = hs_claims / tot_pop,
@@
             data = model_dat)
+
+jc_police<-lmer(log(police_rate) ~ 
+              scale(prop_blk) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
+
+jc_teacher<-lmer(log(teacher_rate) ~ 
+              scale(prop_blk) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
+
+jc_socwork<-lmer(log(socwork_rate) ~ 
+              scale(prop_blk) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
+
+jc_gq<-lmer(log(child_inst_rate) ~ 
+              scale(prop_blk) +
+              year + 
+              (1 + year|state),
+            data = model_dat,
+            control=lmerControl(optimizer="nloptwrap",
+                                optCtrl=list(maxfun=2e5)))
 ```
@@
 exp(t1)
+
 ```

### 2023-11-03 t1_analysis.Rmd
a ddsharecropping and urban immig pop data

@@
 (GENERAL FRAMEWORK: Racial capitalism / settler colonization) MATERIAL CONDITIONS FOR SETTLER COLONIZATION STRUCTURE POLICY DESIGN. socio-political processes generate diverse white elite interests
+
+SETTLER COLONIALISM AS STRUCTURE - PROCESS. BUT SPATIALLY VARIABLE DEPENDING ON LOCAL INTERESTS
 
@@
 
-```{r read, cache = T}
+```{r read}
 # ### on libra, read raw
@@
 # ### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
-# dat_occ<-read_csv("./data/usa_00071.csv") %>% 
-#   rename_all(tolower) %>% 
-#   mutate(occ1950 = 
-#            case_when(
-#              occ1950 == 79 ~ "socwork",
-#              occ1950 == 773 ~ "police",
-#              occ1950 == 771 ~ "police",
-#              occ1950 == 782 ~ "police",
-#              occ1950 == 93 ~ "teachers"
-#            )) %>% 
-#   group_by(year, statefip, occ1950) %>% 
-#   summarize(emp = sum(perwt)) %>% 
-#   pivot_wider(names_from = occ1950, values_from = emp,
-#               values_fill = 0)
-# ### AK OR HI HERE? check that at IPUMS
-# 
-# # pop data (total, age, race, nativitiy, urban)
-# dat_immig<-read_csv("./data/usa_00070.csv") %>% 
-#   rename_all(tolower) %>% 
-#   group_by(year, statefip) %>% 
-#   summarize(tot_pop = sum(perwt),
-#             children = sum((age<18) * perwt),
-#             white_immig = 
-#               sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
-#             blk = 
-#               sum((race == 2) * perwt),
-#             urban = sum(hhwt * (urban ==2)),
-#             tot_hh = sum(hhwt)) 
-# ### AK OR HI HERE? check that at IPUMS
-# 
-# 
-# ### group quarters data, 100% 
-# gq<-read_csv("./data/usa_00069.csv") %>% 
-#   rename_all(tolower) %>% 
-#   filter(age<18) %>% 
-#   group_by(year, statefip, gqtype) %>% 
-#   summarize(gq_pop = sum(perwt, na.rm = T)) %>% 
-#   mutate(gqtype = case_when(
-#     gqtype == 2 ~ "child_inst_correctional",
-#     gqtype == 3 ~ "child_inst_mental",
-#     gqtype == 4 ~ "child_inst_poor_disabled")) %>% 
-#   filter(gqtype!="other") %>% 
-#   pivot_wider(names_from = gqtype, values_from = gq_pop,
-#               values_fill = 0)
-# ### AK OR HI HERE? check that at IPUMS
-# 
-#  
-# ### join
-# dat <- data.frame(state = state.abb, 
-#                region = state.region, 
-#                division = state.division) %>% 
-#   left_join(state.fips %>% 
-#               select(fips, abb) %>% 
-#               distinct() %>% 
-#               rename(state = abb, statefip = fips)) %>% 
-#   left_join(dat_immig) %>% 
-#   left_join(dat_occ) %>% 
-#   ungroup() %>% 
-#   left_join(gq)
-# 
-# ### homestead act data
-# #### load in homestead data and boarding school data
-# ### state measure can be cumulative sum of homesteads, that gets at polity
-# ### homesteading data
-# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
-#   left_join(dat %>% 
-#               select(state, region) %>% 
-#               distinct()) 
-# 
-# hs<-read_csv("./data/blm_homesteads_clean.csv") %>% 
-#   pivot_longer(claims1868:acres1961,
-#                names_to = "name",
-#                values_to = "value") %>% 
-#   mutate(year = str_sub(
-#     name, start = -4, end = -1),
-#     type = str_sub(
-#       name, start = 1, end = -5)) %>% 
-#   select(-name) %>% 
-#   mutate(year = as.numeric(year)) %>% 
-#   mutate(value = ifelse(is.na(value),
-#                         0,
-#                         value))
-# 
-# ### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
-# hs_dakota<-hs %>% 
-#   filter(state == "Dakota Territory") %>% 
-#   mutate(value = value / 2) %>% 
-#   select(-state)
-# 
-# hs<-hs %>% 
-#   bind_rows(hs_dakota %>% 
-#               mutate(state = "North Dakota")) %>% 
-#   bind_rows(hs_dakota %>% 
-#               mutate( state = "South Dakota")) %>% 
-#   filter(state != "Dakota Territory") %>% 
-#   group_by(state, year, type) %>% 
-#   summarize(value = sum(value))
-#   
-# hs<-hs %>% 
-#   arrange(state, year) %>% 
-#   group_by(state, type) %>% 
-#   mutate(cum_value = cumsum(value)) %>% 
-#   ungroup()
-# 
-# ## expand_grid to add zeroes
-# ### use y_o_s for index
-# state<-unique(y_o_s$statename)
-# year<-unique(hs$year)
-# type<-unique(hs$type)
-# temp<-expand_grid(state, year, type)
-# 
-# hs<-temp %>% 
-#   left_join(hs) %>% 
-#   replace_na(list(value = 0,
-#                   cum_value = 0))
-# 
-# hs <- hs %>% 
-#   select(-value) %>% 
-#   pivot_wider(names_from = type,
-#               values_from = cum_value,
-#               names_prefix = "hs_") 
-# 
-# ### add ID and pop vars back to hs
-# 
-# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
-#   left_join(dat %>% 
-#               select(state, region) %>% 
-#               distinct()) %>% 
-#   filter(!(is.na(region)))
-# 
-# hs<-hs %>% 
-#   rename(statename = state) %>% 
-#   left_join(y_o_s)
-# 
-# #### join cumulative acres / claims onto dat for the valid census years, then carried forward
-# 
-# dat<-dat %>% 
-#   left_join(hs)
+dat_occ<-read_csv("./data/usa_00071.csv") %>%
+  rename_all(tolower) %>%
+  mutate(occ1950 =
+           case_when(
+             occ1950 == 79 ~ "socwork",
+             occ1950 == 773 ~ "police",
+             occ1950 == 771 ~ "police",
+             occ1950 == 782 ~ "police",
+             occ1950 == 93 ~ "teachers"
+           )) %>%
+  group_by(year, statefip, occ1950) %>%
+  summarize(emp = sum(perwt)) %>%
+  pivot_wider(names_from = occ1950, values_from = emp,
+              values_fill = 0)
+### AK OR HI HERE? check that at IPUMS
 
-# on local read temp
-dat<-read_csv("./data/temp_data.csv")
-hs<-read_csv("./data/temp_hs.csv")
+# pop data (total, age, race, nativitiy, urban, farm, landown)
+dat_immig<-read_csv("./data/usa_00072.csv") %>%
+  rename_all(tolower) %>%
+  group_by(year, statefip) %>%
+  summarize(tot_pop = sum(perwt),
+            children = sum((age<18) * perwt),
+            white_immig =
+              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+            blk =
+              sum((race == 2) * perwt),
+            blk_sharecrop = 
+              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
+            blk_farms = sum((race==2) * (farm==2) * perwt),
+            urban_pop = sum(hhwt * (urban ==2)),
+            urban_immig_pop = 
+              sum((urban==2) * 
+                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+            tot_hh = sum(hhwt))
+### AK OR HI HERE? check that at IPUMS
+
+### group quarters data, 100%
+gq<-read_csv("./data/usa_00069.csv") %>%
+  rename_all(tolower) %>%
+  filter(age<18) %>%
+  group_by(year, statefip, gqtype) %>%
+  summarize(gq_pop = sum(perwt, na.rm = T)) %>%
+  mutate(gqtype = case_when(
+    gqtype == 2 ~ "child_inst_correctional",
+    gqtype == 3 ~ "child_inst_mental",
+    gqtype == 4 ~ "child_inst_poor_disabled")) %>%
+  filter(gqtype!="other") %>%
+  pivot_wider(names_from = gqtype, values_from = gq_pop,
+              values_fill = 0)
+### AK OR HI HERE? check that at IPUMS
+
+### join
+dat <- data.frame(state = state.abb,
+               region = state.region,
+               division = state.division) %>%
+  left_join(state.fips %>%
+              select(fips, abb) %>%
+              distinct() %>%
+              rename(state = abb, statefip = fips)) %>%
+  left_join(dat_immig) %>%
+  left_join(dat_occ) %>%
+  ungroup() %>%
+  left_join(gq)
+
+### homestead act data
+#### load in homestead data and boarding school data
+### state measure can be cumulative sum of homesteads, that gets at polity
+### homesteading data
+y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
+  left_join(dat %>%
+              select(state, region) %>%
+              distinct())
+
+hs<-read_csv("./data/blm_homesteads_clean.csv") %>%
+  pivot_longer(claims1868:acres1961,
+               names_to = "name",
+               values_to = "value") %>%
+  mutate(year = str_sub(
+    name, start = -4, end = -1),
+    type = str_sub(
+      name, start = 1, end = -5)) %>%
+  select(-name) %>%
+  mutate(year = as.numeric(year)) %>%
+  mutate(value = ifelse(is.na(value),
+                        0,
+                        value))
+
+### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
+hs_dakota<-hs %>%
+  filter(state == "Dakota Territory") %>%
+  mutate(value = value / 2) %>%
+  select(-state)
+
+hs<-hs %>%
+  bind_rows(hs_dakota %>%
+              mutate(state = "North Dakota")) %>%
+  bind_rows(hs_dakota %>%
+              mutate( state = "South Dakota")) %>%
+  filter(state != "Dakota Territory") %>%
+  group_by(state, year, type) %>%
+  summarize(value = sum(value))
+
+hs<-hs %>%
+  arrange(state, year) %>%
+  group_by(state, type) %>%
+  mutate(cum_value = cumsum(value)) %>%
+  ungroup()
+
+## expand_grid to add zeroes
+### use y_o_s for index
+state<-unique(y_o_s$statename)
+year<-unique(hs$year)
+type<-unique(hs$type)
+temp<-expand_grid(state, year, type)
+
+hs<-temp %>%
+  left_join(hs) %>%
+  replace_na(list(value = 0,
+                  cum_value = 0))
+
+hs <- hs %>%
+  select(-value) %>%
+  pivot_wider(names_from = type,
+              values_from = cum_value,
+              names_prefix = "hs_")
+
+### add ID and pop vars back to hs
+
+y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
+  left_join(dat %>%
+              select(state, region) %>%
+              distinct()) %>%
+  filter(!(is.na(region)))
+
+hs<-hs %>%
+  rename(statename = state) %>%
+  left_join(y_o_s)
+
+#### join cumulative acres / claims onto dat for the valid census years, then carried forward
+
+dat<-dat %>%
+  left_join(hs)
+
+# # on local read temp
+# dat<-read_csv("./data/temp_data.csv")
+# hs<-read_csv("./data/temp_hs.csv")
 ```
@@
 ```
+
+next let's look at sharecropping
+
+```{r}
+ggplot(dat %>% 
+             filter(year == 1920,
+                    blk_farms >= 1e4),
+       aes(x = blk_farms, y = blk_sharecrop,
+           label = state)) + 
+  geom_text(alpha = 0.5) + 
+  geom_abline(slope = 1) + 
+  labs(x = "Black people living on farms",
+       y = "Black people living on tenant farms",
+       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
+```
+
+And the bivar with sharecropping / blk pop comp
+
+```{r}
+ggplot(dat %>% 
+         filter(year == 1920),
+       aes(x = blk / tot_pop, y = blk_sharecrop,
+           label = state,
+           color = region)) + 
+  #geom_point(alpha = 0.5) + 
+  geom_text(alpha = 0.5) + 
+  scale_y_log10() + 
+  labs(x = "Proportion Black pop",
+       y = "Black people living on tenant farms",
+       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
+```
+
+Places with large Black pops in 1920 also had very high sharecropping rates.
+Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though.
+Likely makese sense to use sharecropping as log count rather than as proportion.
 
@@
 
-```{r urban-ts, fig.cap = "Urban population by state 1870-1940"}
+```{r urban-ts, fig.cap = "Urban white immigrant population by state 1870-1940"}
 p1<-ggplot(dat%>% 
+             filter(year<=1930) %>% 
              filter(!(is.na(region))),
-       aes(x = year, y = urban / tot_hh,
+           aes(x = year, y = urban_immig_pop / tot_pop,
                group = state)) + 
   geom_line(alpha = 0.5) + 
-  labs(subtitle = "Proportion of households urban",
+  labs(subtitle = "Proportion of population white immigrant urban",
        y = "Proportion",
@@
 p2<-ggplot(dat%>% 
+             filter(year<=1930) %>% 
              filter(!(is.na(region))),
-       aes(x = year, y = urban,
+           aes(x = year, y = urban_immig_pop,
                group = state)) + 
   geom_line(alpha = 0.5) + 
-  labs(subtitle = "Number of households urban",
+  labs(subtitle = "White immigrant urban population",
        y = "Count",

### 2023-11-21 t1_analysis.Rmd
ff

@@
 ---
-title: "Varying paths to building the social state in the progressive era"
+title: "Settler colonial projects and variation in paths to building the welfare and carceral state"
 author: "Frank Edwards"
-date: "last updated 11/1/23"
+date: "last updated `r Sys.Date()`"
 output: bookdown::html_document2
+abstract: "Place and time variation in problems of settler colonization are related to the kinds of governments that local settler elites build. "
 ---
@@
 library(gridExtra)
+library(brms)
 options(scipen = 999)
@@
 
-1. pull immig + urban data
-2. pull tenant farmer data
-3. model employment -> each, consider loglinear process?
+leaving out railroad grants induces bias if morill or homestead grants are negatively correlated with railroad grants.
 
-theory re: white immigs: coercion / rehab in NE, yeoman uplift in W, frontier as panacea for urban, lots of hist on that, gets at SocWork/Police vs Teachers employment
+2. Lit review
+3. writeup demographics and bivars
+4. models
 
@@
 
-(GENERAL FRAMEWORK: Racial capitalism / settler colonization) MATERIAL CONDITIONS FOR SETTLER COLONIZATION STRUCTURE POLICY DESIGN. socio-political processes generate diverse white elite interests
+HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?
 
-SETTLER COLONIALISM AS STRUCTURE - PROCESS. BUT SPATIALLY VARIABLE DEPENDING ON LOCAL INTERESTS
+Settler colonial regimes seek to resolve different problems across places and within places over time.
+This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places.
+We show that a series of historical economic and demographic measures (proxies for variation in modes of settler colonial extraction and exploitation) are closely related to trajectories of state formation across places, measured as employment levels in core social policy system sectors.
 
-1. Frontier settler polity -> elite interest in disposession / assimiatilion -> aggressive paternalistic welfare state for AIAN, social democracy for settlers
-2. Jim Crow -> elite interest in exploitation and exclusion -> elite paternalism, malign neglect, hard coercion
-3. Metropole settler polity -> elite interests in labor discipline / immig assimilation -> progressive paternalism, strong coercive welfare state for W, hard coercion and contested welfare for Blk
+At a sub-national level, settler colonial regimes seek to address a range of problems inherent to their nation-building project: the dispossession of indigenous lands, the erasure of indigenous peoples, the exploitation of racially marginalized laborers, and the reproduction and expansion of the settler polity.
+We show that the historical geography of these problems is closely related to the policy infrastructure that governments develop.
+The consequences of this variation in trajectories are likely broad and varied.
 
-THESE ARE MULTIVARIATE INDICATORS, CAN USE CLUSTERING TO CHECK IT AND MEASURE IT.
-
-1: MEASURE VARIATION IN CONTEXT, ASSUME CONTEXT PRODUCES INTERESTS. 2: MEASURE VARIATION IN POLICY DEVELOPMENT.
-
-3 dimensions: frontier, antiBlack, white immig assim (urban)
-
-frontier: {homestead claims, tribal density, https://www.census.gov/dataviz/visualizations/001/}
-railrooad (corporate) + homestaed (individual) + morrill (state) - nice three piece approach, more complex and complete story.
-
-antiblack: {jim crow polit econ (sharecropping), prop black}
-
-urban assim: {urbanization, industrial employment?, urban white immig, log black pop}
+Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it?
 
@@
 
-Interests of social policy actors defined within these fields. Bourdieusian historical stuff?
-
 Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development.
@@
 
-interests can be observed indirectly sometimes, right?
-homesteading is direct evidence of interest.
-tenant farming (or land ownership) is direct evidence of interest.
-industrial urban is tougher, but maybe there we just go indirect and use urbanization + racial demography.
+<!-- 1. Frontier settler polity -> elite interest in disposession / assimiatilion -> aggressive paternalistic welfare state for AIAN, social democracy for settlers -->
+<!-- 2. Jim Crow -> elite interest in exploitation and exclusion -> elite paternalism, malign neglect, hard coercion -->
+<!-- 3.
+Metropole settler polity -> elite interests in labor discipline / immig assimilation -> progressive paternalism, strong coercive welfare state for W, hard coercion and contested welfare for Blk -->
 
-I THINK THIS WORKS. CLUSTER ON INTERESTS + DEMOGRAPHY. OUTCOMES ARE POLICY MEASURES: POLICY OUTCOMES (EG EMPLOYMENT, SPENDING, GROUP QUARTERS POPS, ENROLLMENTS).
+IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.
 
-TRY TO DO ONLY A FEW BIG NATIONAL PARAMETRIC TESTS. LOTS OF DESCRIPTIVE SUBSET WORK
+# What is already known about settler colonization and state formation in late 19th and Early 20th c.
 
-TEST 1: clustering, spatial clustering
+## Frontier
 
-then TONS of descriptives to paint pictures of regional social / political contexts
+review here
 
-TEST 2: regression on outcomes
+## Jim Crow
 
-negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1
+review here
 
-positive case: white immigrants -> more social workers in t1
+## Metropole
+
+review here
 
@@
 dat<-dat %>%
-  left_join(hs)
+  left_join(hs) %>% 
+  filter(year>=1880, year<=1920)
 
@@
 # hs<-read_csv("./data/temp_hs.csv")
+
+### land grab u data
+# lgu_cessions<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Cessions.csv")
+### this is cessions from the Royce maps, overlaps with morrill, but includes prior cessions as well
+### would need to use shapefiles to portion out by state
+lgu_parcels<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Parcels.csv")
+lgu_unis<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Universities.csv")
 ```
@@
 
+We are generally interested in evaluating whether a set of policy outcomes {employment, infrastructure} are related to a set of demographic and economic predictors {population composition, economic activity, ownership regime}.
+
+To do so, we rely on historical data from the US Census accessed via IPUMS (Ruggles et al), from the Bureau of Land Management (Archived by the University of Nebraska), from the US Department of the Interior (CITE BS REPORT), and from records our team digitized from the Commissioner on Indian Affairs (CITE).
+
+Less interested in causal identification, more interested in understanding spatial patterns and trajectories.
+We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data.
+Move from specific to general.
+
+demonstrate var(infrastructure), show e[infrastructure] = f(settler interests)
+
+negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1
+
+positive case: white immigrants -> more social workers in t1
+
 here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.
@@
 ## NOTES ON GEOGRAPHIES and MEASUREMENT
-
-### STATES, TERRITORIES, AND NATIVE LANDS
 
@@
 
-### DIRECT AND PROXY MEASURES FOR ELITE SETTLER INTERESTS
+geography itself endogenous to settler colonization, this is tricky!
+
+## sources of data
+
+Census; ARCIA; Interior; etc
+
+CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho...
+
+we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.
+
+let's look at how it played out in the progressive era across states and regions.
+we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.
+
+First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita?
+
+We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states.
+A census taker is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.).
+We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.
+
+## statistical methods and models
+
+EDA; multilevel models on outcome rates (state intercepts, state x year slopes, log outcomes for nonlinear process, etc) estimated with hmc, weak priors
+
+# Findings
 
@@
 
-### Measuring anti-Blackness and post-Reconstruction racial politics
+## Geographic contexts for settler colonial policymaking across US states 1880 - 1920
 
-IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
+### Jim Crow and the salience of anti-Blackness for local social and carceral policymaking
 
-1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
-2) exclusion / formal soc control salience in urban non-S
+Figure \@ref(fig:blk-map) shows the distribution of the recorded Black population across US states in 1920.
+Panel 1 displays this data as the relative population composition of Black residents relative to to all residents, and panel 2 displays the absolute size of the Black population across US states.
 
-POPULATION ALONE DESCRIPTIVES
+DISCUSS THE DISTRIBUTION
+
+```{r blk-map, fig.cap = "Black population, US States 1920 (note log scale)"}
+p1<-dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = blk / tot_pop * 100)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis") + 
+  theme_void() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "Percent of total",
+       subtitle = "Panel 1. Population composition")
+
+p2<-dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = blk + 100)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis", trans = "log10") + 
+  theme_void() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "Population\n(note logarithmic scale)",
+       subtitle = "Panel 2. Total population")
+
+grid.arrange(p1, p2, nrow = 1)
+```
 
@@
 
-```{r}
-dat %>% 
-  filter(year == 1930) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = log(blk))) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() + 
-  theme(legend.position = "bottom")
-```
-
 next let's look at sharecropping
@@
 
+```{r}
+dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = blk_sharecrop + 10)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis",
+                        trans = "log10") + 
+  theme_void() + 
+  labs(fill = "Number of Black residents\non tenant farms\n(note logarithmic scale)")
+```
+
 Places with large Black pops in 1920 also had very high sharecropping rates.
@@
 
-### Measuring the size, power, and interests of the expropriative settler polity
+### Settlers and indigenous disposession
 
@@
 
-3) PRESENCE OF NATIVE LAND HOLDINGS - VERY CLEAR
+```{r}
+p1<-dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = hs_acres / 1e6)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis") + 
+  theme_void() + 
+  labs(fill = "Cumulative acres claimed\n(millions), 1920") + 
+  theme(legend.position = "bottom")
 
-  Explain
+p2<-dat %>% 
+  filter(year == 1920) %>% 
+  mutate(hs_pop = hs_claims / tot_pop) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = hs_pop)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis") + 
+  theme_void() +
+  labs(fill = "Claims per resident\n1920")  +
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, nrow = 1)
+```
+
+```{r}
+dat %>% 
+  filter(year == 1920) %>% 
+  mutate(region = factor(region, 
+                         levels = c("West", 
+                                    "North Central",
+                                    "South",
+                                    "Northeast"))) %>% 
+  mutate(hs_pop = hs_claims / tot_pop) %>% 
+  ggplot(aes(x = hs_pop,
+             y = fct_reorder(state, hs_pop))) + 
+  geom_point() + 
+  labs(x = "Cumulative homestead claims per state population, 1920",
+       y = "",
+       color = "Census Region") + 
+  facet_wrap(~region, 
+             scales = "free_y")
+```
 
@@
             position = position_nudge(x = 8),
-            alpha = 0.7) + 
+            alpha = 0.7,
+            size = 2.5) + 
 facet_wrap(~region)  + 
@@
 
-THIS IS COOL! Look at homestead (as settler polity) plus mode (as boarding school on / off)
+NATIVE LANDHOLDINGS OVER TIME ARE DIRECTLY RELATED TO HOMESTEAD CLAIMS, MECHANISM OF DISPOSESSION.
 
-LOOK AT ACRES PER CLAIM? GET AT BOTH POP SIZE AND RELATIVE RESOURCES?
+PULL AGENTS DATA FOR LANDHOLDINGS?
 
-```{r}
-ggplot(hs %>% 
-         filter(year == 1961) %>% 
-         mutate(region = factor(region, 
-                                levels = c("West", 
-                                           "North Central",
-                                           "South",
-                                           "Northeast"))),
-       aes(x = hs_acres / 1e6,
-           y = fct_reorder(state, hs_acres))) + 
-  geom_point() + 
-  labs(x = "Acres claimed (millions)",
-       y = "",
-       color = "Census Region") + 
-  facet_wrap(~region, 
-             scales = "free_y")
+### Urbanization and European immigration
+
+```{r immig-map, fig.cap = "White immigrant population in cities 1920"}
+dat %>% 
+  filter(year == 1920) %>% 
+  mutate(state = tolower(statename)) %>% 
+  ggplot(aes(fill = urban_immig_pop)) + 
+  geom_map(aes(map_id = state), map = map) + 
+  expand_limits(x = map$long, y = map$lat) + 
+  scale_fill_continuous(type = "viridis",
+                        trans = "log10") + 
+  theme_void() + 
+  labs(fill = "Population\n(note logarithmic scale)")
 ```
 
-pop / claims ratio is a little funky, but probably useful. can peg to 1940 census.
-
-WOWZERS. 
-20+pct of pop likely had HS claim land in 3 states.
-
-Wow, let's pull through this. ONE INDICATOR OF SETTLER POLITY. DEmographic base is one dimension - cultures / institutions/ resources also key. but useful!
-
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(region = factor(region, 
-                         levels = c("West", 
-                                    "North Central",
-                                    "South",
-                                    "Northeast"))) %>% 
-  mutate(hs_pop = hs_claims / tot_pop) %>% 
-  ggplot(aes(x = hs_pop,
-             y = fct_reorder(state, hs_pop))) + 
-  geom_point() + 
-  labs(x = "Cumulative homestead claims per state population, 1920",
-       y = "",
-       color = "Census Region") + 
-  facet_wrap(~region, 
-             scales = "free_y")
-```
-
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(hs_pop = hs_claims / tot_pop) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = hs_pop)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() + 
-  theme(legend.position = "bottom")
-```
-
-### Measuring the magnitude of social problems Progressive reformers would seek to 'fix'
-
-PULL IMMIG + URBAN - UNIQUE COMBO THAT PUSHES FORWARD WELFARE STATE
-
-```{r immig-ts, fig.cap = "White immigrant population by state 1870-1940"}
-p1<-ggplot(dat,
-       aes(x = year, y = white_immig / tot_pop,
-           group = state)) + 
-  geom_line(alpha = 0.5) + 
-  labs(subtitle = "Proportion white immigrant",
-       y = "Proportion",
-       x = "Year") + 
-  facet_wrap(~region)
-
-p2<-ggplot(dat%>% 
-  filter(!(is.na(region))),
-       aes(x = year, y = white_immig,
-           group = state)) + 
-  geom_line(alpha = 0.5) + 
-  scale_y_log10() + 
-  labs(subtitle = "White immigrant population",
-       y = "Count (note log10 axis)",
-       x = "Year") + 
-  facet_wrap(~region)
-
-grid.arrange(p2, p1, nrow = 1)
-```
-
-Relative population share as first generation immigrant declines in NC across period uniformly.
-In NE, increase until 1920 for most states, then monotonic decline.
-Low levels in S.
-Declining levels in W.
-
-Absolute white immigrant pop numbers increasing in NE, Increasing, then plateau in NC, increasing then plateau in W, and stable and low in S.
-
-### urban
-
-```{r urban-ts, fig.cap = "Urban white immigrant population by state 1870-1940"}
+```{r urban-ts, fig.cap = "Urban white immigrant population by state 1880-1920"}
 p1<-ggplot(dat%>% 
-             filter(year<=1930) %>% 
              filter(!(is.na(region))),
@@
 p2<-ggplot(dat%>% 
-             filter(year<=1930) %>% 
              filter(!(is.na(region))),
@@
 
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = log(white_immig))) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() + 
-  theme(legend.position = "bottom")
-```
+MAYBE ALSO NOTE WHITE IMMIG POP IN FRONTIER STATES, DIFFERENT REGIMES FOR INCORPORATION, DIFFERENT RACIALIZATION PROCESSES FOR NORTH / W EURO IMMIGS VS SOUTH / EAST / IRISH?
 
-# distributions of policy outcomes
+## Geographic variation in social and carceral state capacities 1880 - 1920
 
-### the distribution of social state capacity building
+what are states doing? how does it vary across places and over time?
 
-CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho...
-
-we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.
-
-let's look at how it played out in the progressive era across states and regions.
-we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.
-
-First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita?
-
-We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states.
-A census taker is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.).
-We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.
+### Employment in core roles for the emergent welfare and carceral states
 
@@
 
-#### kids in institutional group quarters
+### capacity for therapeutic coercion: kids in institutional group quarters
 
@@
   geom_boxplot()
-
 ```
@@
 
-employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.
+IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
 
-q1: salience of settler expropriotion -> state formation
+1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
+2) exclusion / formal soc control salience in urban non-S
 
-four models, each employment outcome + children in poverty institutions
+<!-- employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.  -->
 
-start with lm for log(rates)
+<!-- q1: salience of settler expropriotion -> state formation -->
 
-trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.
+<!-- four models, each employment outcome + children in poverty institutions -->
 
-switch to brms for estimation on tricky ones
+<!-- start with lm for log(rates) -->
 
-```{r}
-### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE
-model_dat <- dat %>% 
-  filter(year>=1880) %>% 
-  mutate(police_rate = (police + 1) / tot_pop * 1e5,
-         socwork_rate = (socwork + 1) / tot_pop * 1e5,
-         teacher_rate = (teachers + 1) / tot_pop * 1e5,
-         child_inst_rate = (child_inst_poor_disabled + 0.1) / 
-           children * 1e5,
-         prop_blk = (blk + 1) / tot_pop,
-         hsclaim_prop = hs_claims / tot_pop,
-         year = scale(year))
+<!-- trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.  -->
 
-sc_police<-lmer(log(police_rate) ~ 
-              scale(hsclaim_prop) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+<!-- switch to brms for estimation on tricky ones -->
 
-sc_teacher<-lmer(log(teacher_rate) ~ 
-              scale(hsclaim_prop) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+<!-- ```{r} -->
+<!-- ### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE -->
+<!-- model_dat <- dat %>%  -->
+<!--   filter(year>=1880) %>%  -->
+<!--   mutate(police_rate = (police + 1) / tot_pop * 1e5, -->
+<!--          socwork_rate = (socwork + 1) / tot_pop * 1e5, -->
+<!--          teacher_rate = (teachers + 1) / tot_pop * 1e5, -->
+<!--          child_inst_rate = (child_inst_poor_disabled + 0.1) /  -->
+<!--            children * 1e5, -->
+<!--          prop_blk = (blk + 1) / tot_pop, -->
+<!--          prop_urban_immig = urban_immig_pop / tot_pop, -->
+<!--          hsclaim_prop = hs_claims / tot_pop, -->
+<!--          year = scale(year)) -->
 
-sc_socwork<-brm(log(socwork_rate) ~ 
-              scale(hsclaim_prop) +
-              year + 
-              (1 + year|state),
-            data = model_dat)
+<!-- m_police<-brm(log(police_rate) ~  -->
+<!--                  scale(hsclaim_prop) + -->
+<!--                  scale(prop_urban_immig) + -->
+<!--                  scale(prop_blk) +  -->
+<!--                  scale(log(blk_sharecrop + 1)) +  -->
+<!--                  year +  -->
+<!--                  (1 + year|state), -->
+<!--                data = model_dat) -->
 
-sc_gq<-brm(log(child_inst_rate) ~ 
-              scale(hsclaim_prop) +
-              year + 
-              (1 + year|state),
-            data = model_dat)
+<!-- m_teacher<-brm(log(teacher_rate) ~  -->
+<!--                  scale(hsclaim_prop) + -->
+<!--                  scale(prop_urban_immig) + -->
+<!--                  scale(prop_blk) +  -->
+<!--                  scale(log(blk_sharecrop + 1)) +  -->
+<!--                  year +  -->
+<!--                  (1 + year|state), -->
+<!--                data = model_dat) -->
 
-jc_police<-lmer(log(police_rate) ~ 
-              scale(prop_blk) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+<!-- m_socwork<-brm(log(socwork_rate) ~  -->
+<!--                  scale(hsclaim_prop) + -->
+<!--                  scale(prop_urban_immig) + -->
+<!--                  scale(prop_blk) +  -->
+<!--                  scale(log(blk_sharecrop + 1)) +  -->
+<!--                  year +  -->
+<!--                  (1 + year|state), -->
+<!--                data = model_dat) -->
 
-jc_teacher<-lmer(log(teacher_rate) ~ 
-              scale(prop_blk) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+<!-- sc_teacher<-lmer(log(teacher_rate) ~  -->
+<!--               scale(hsclaim_prop) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat, -->
+<!--             control=lmerControl(optimizer="nloptwrap", -->
+<!--                                 optCtrl=list(maxfun=2e5))) -->
 
-jc_socwork<-lmer(log(socwork_rate) ~ 
-              scale(prop_blk) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+<!-- sc_socwork<-brm(log(socwork_rate) ~  -->
+<!--               scale(hsclaim_prop) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat) -->
 
-jc_gq<-lmer(log(child_inst_rate) ~ 
-              scale(prop_blk) +
-              year + 
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
-```
+<!-- sc_gq<-brm(log(child_inst_rate) ~  -->
+<!--               scale(hsclaim_prop) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat) -->
 
-This all looks fine. How do I want to visualize this?
+<!-- jc_police<-lmer(log(police_rate) ~  -->
+<!--               scale(prop_blk) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat, -->
+<!--             control=lmerControl(optimizer="nloptwrap", -->
+<!--                                 optCtrl=list(maxfun=2e5))) -->
 
-we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930?
+<!-- jc_teacher<-lmer(log(teacher_rate) ~  -->
+<!--               scale(prop_blk) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat, -->
+<!--             control=lmerControl(optimizer="nloptwrap", -->
+<!--                                 optCtrl=list(maxfun=2e5))) -->
+
+<!-- jc_socwork<-lmer(log(socwork_rate) ~  -->
+<!--               scale(prop_blk) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat, -->
+<!--             control=lmerControl(optimizer="nloptwrap", -->
+<!--                                 optCtrl=list(maxfun=2e5))) -->
+
+<!-- jc_gq<-lmer(log(child_inst_rate) ~  -->
+<!--               scale(prop_blk) + -->
+<!--               year +  -->
+<!--               (1 + year|state), -->
+<!--             data = model_dat, -->
+<!--             control=lmerControl(optimizer="nloptwrap", -->
+<!--                                 optCtrl=list(maxfun=2e5))) -->
+<!-- ``` -->
+
+<!-- This all looks fine. How do I want to visualize this?  -->
+
+<!-- we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930?
 Think about how to handle to intercepts / slopes.
@@
 it's not a hypothetical state, its the mean expectation for the slope of the regime var.
+-->
 
-model is
+<!-- model is  -->
 
-\[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\]
+<!-- \[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\] -->
 
-Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.
+<!-- Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.  -->
 
-Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?
+<!-- Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?  -->
 
-```{r}
-new_dat<-data.frame(year =  c(1.3, 1.3, 1.3),
-                    hsclaim_prop = c(0, 0.05, 0.15))
-t<-predict(sc_police, 
-           newdata = new_dat,
-           re.form = ~0,
-           type = "link")
+<!-- ```{r} -->
+<!-- new_dat<-data.frame(year =  c(1.3, 1.3, 1.3), -->
+<!--                     hsclaim_prop = c(0, 0.05, 0.15)) -->
+<!-- t<-predict(sc_police,  -->
+<!--            newdata = new_dat, -->
+<!--            re.form = ~0, -->
+<!--            type = "link") -->
 
-t1<-predict(sc_teacher, 
-           newdata = new_dat,
-           re.form = ~0,
-           type = "link")
+<!-- t1<-predict(sc_teacher,  -->
+<!--            newdata = new_dat, -->
+<!--            re.form = ~0, -->
+<!--            type = "link") -->
 
-exp(t1)
+<!-- exp(t1) -->
 
-```
+<!-- ``` -->

### 2023-11-21 agents_eda.R
add eda file

@@ -1,0 +1,14 @@
+library(tidyverse)
+
+### load in agents data
+agents<-read_csv("./data/AGENTS_DATA_COLLECTION.csv")
+### agency means tribe totals are contained in the single agency row, we can treat tribe numbers 
+### zeroes for the purpose of agency / state totals
+agents<-agents %>% 
+  filter(YEAR >= 1880) %>% 
+  mutate(N_STUDENTS = ifelse(N_STUDENTS == "Agency" | N_STUDENTS == "Reserve"| 
+                               N_STUDENTS == "as above", NA, N_STUDENTS))
+
+
+sum_students <- agents %>% 
+  mutate(n_students = eval(parse(text = N_STUDENTS)))

### 2023-11-21 agents_eda.R
ff

@@ -5,10 +5,28 @@ agents<-read_csv("./data/AGENTS_DATA_COLLECTION.csv")
 ### agency means tribe totals are contained in the single agency row, we can treat tribe numbers 
 ### zeroes for the purpose of agency / state totals
 agents<-agents %>% 
-  filter(YEAR >= 1880) %>% 
-  mutate(N_STUDENTS = ifelse(N_STUDENTS == "Agency" | N_STUDENTS == "Reserve"| 
-                               N_STUDENTS == "as above", NA, N_STUDENTS))
+  filter(YEAR >= 1895) %>% 
+  mutate(AVG_ATTENDANCE = ifelse(AVG_ATTENDANCE == "Agency" | AVG_ATTENDANCE == "Reserve"| 
+                               AVG_ATTENDANCE == "as above", NA, AVG_ATTENDANCE))
 
 
 sum_students <- agents %>% 
-  mutate(n_students = eval(parse(text = N_STUDENTS)))
+  mutate(AVG_ATTENDANCE = eval(parse(text = AVG_ATTENDANCE)))
+
+
+agents_1890<-agents %>% 
+  filter(YEAR>=1890) %>% 
+  select(`STATE/SUPERINTENDENCY`,
+         TRIBE,
+         AGENCY,
+         YEAR,
+         `DAY Y/N/B`,
+         AVG_ATTENDANCE) 
+
+
+students_ts<-agents_1890 %>% 
+  group_by(`STATE/SUPERINTENDENCY`, YEAR) %>% 
+  summarize(students = sum(AVG_ATTENDANCE))
+
+### use the flag on day school to sort into attendance categories
+### create AVG_ATTENDANCE, AVG_ATTENDANCE - DAY !="Y"

### 2023-11-21 t1_analysis.Rmd
work on shapefiles and on morill data

@@
 lgu_parcels<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Parcels.csv")
-lgu_unis<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Universities.csv")
 ```

### 2023-11-21 scratch_land_measures.r

@@ -1,0 +1,271 @@
+library(tidyverse)
+library(tidycensus)
+library(tigris)
+library(units)
+library(sf)
+
+options(tigris_use_cache = T)
+
+# 1. geodata merge in shape_merge.r. 
+# - land grab u shapes for morill parcels https://github.com/HCN-Digital-Projects/landgrabu-data
+# - royce map shapes from forest service 'Indian Land Cessions in US' https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
+# - contemporary (or 1920) state shape files
+# - Pacific railroad act shapes, need to figure this out (WAITING ON BLM FOR MORE INFO)
+# 
+# active.digitsposession = (morill act + homestead act + railroad grants) / total state area
+# 
+# given that railroad acreage is unknown, then 
+# 
+# \[Disp_{s} \propto \frac{MorillAcres_s + HomesteadAcres_s}{Acreage_s}\]
+# 
+# If
+# 
+# \[cor(MorillAcres_s + HomesteadAcres_s, RailroadAcres_s)\] is constant across states
+# 
+# Remaining native lands as of 1892 can be computed with Royce data as 
+# 
+# \[RemainingLands_s = RoyceAcres_s - Acreage_s\]
+# 
+# Should be a pretty high negative correlation between Disp_s and RemainingLands_s
+
+### First obtain Acreage_s
+
+st<-states(cb = T) %>%
+  st_transform(8528) %>% 
+  filter(!STUSPS%in%
+           c("HI", "AK", 
+             "PR", "GU", "AS","VI",
+             "MP"))
+aianh<-native_areas(cb = T) %>%
+  st_transform(8528)  
+
+### lower 48
+ggplot(st%>% 
+         filter(!STUSPS%in%
+                  c("HI", "AK", 
+                    "PR", "GU", "AS","VI",
+                    "MP"))) + 
+  geom_sf() + 
+  theme_void()
+
+### aianh
+ggplot(aianh) + 
+  geom_sf() + 
+  theme_void()
+
+### overlay
+ggplot(st) + 
+  geom_sf() + 
+  geom_sf(data = aianh,
+          fill = "dodgerblue") + 
+  theme_void()
+
+aianh_acres<-data.frame(name = aianh[[6]],
[…]

### 2023-12-06 t1_analysis.Rmd
ff

@@
 
-# TO DO
+# Introduction
 
-leaving out railroad grants induces bias if morill or homestead grants are negatively correlated with railroad grants.
-
-2. Lit review
-3. writeup demographics and bivars
-4. models
-
-# Theoretical framework
-
-HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?
+<!-- HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?  -->
 
 Settler colonial regimes seek to resolve different problems across places and within places over time.
-This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places.
+This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects).
 We show that a series of historical economic and demographic measures (proxies for variation in modes of settler colonial extraction and exploitation) are closely related to trajectories of state formation across places, measured as employment levels in core social policy system sectors.
@@
 
-Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it?
+<!-- Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it? -->
 
@@
 
-Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native.
+<!-- Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native.
 Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion.
 Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines.
+-->
 
-Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development.
+<!-- Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development.
 This can lock in particular racial regimes over time as social policy does work on social groups.
+-->
 
@@
 
-IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.
+<!-- IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.
+-->
 
-# What is already known about settler colonization and state formation in late 19th and Early 20th c.
+<!-- # What is already known about settler colonization and state formation in late 19th and Early 20th c.  -->
 
-## Frontier
+<!-- ## Frontier -->
 
-review here
+<!-- review here -->
 
-## Jim Crow
+<!-- ## Jim Crow -->
 
-review here
+<!-- review here -->
 
-## Metropole
+<!-- ## Metropole  -->
 
-review here
+<!-- review here -->
 
@@
 
-# methods
+# Methods
 
@@
 
-Less interested in causal identification, more interested in understanding spatial patterns and trajectories.
+<!-- Less interested in causal identification, more interested in understanding spatial patterns and trajectories.
 We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data.
 Move from specific to general.
+-->
 
-demonstrate var(infrastructure), show e[infrastructure] = f(settler interests)
+Generally our approach demonstrates suggests that state formation in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.
 
-negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1
+\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]
 
-positive case: white immigrants -> more social workers in t1
+<!-- negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1 -->
 
-here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.
+<!-- positive case: white immigrants -> more social workers in t1 -->
 
-## NOTES ON GEOGRAPHIES and MEASUREMENT
+<!-- here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.  -->
 
-TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION.
+<!-- ## NOTES ON GEOGRAPHIES and MEASUREMENT -->
+
+<!-- TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION.
 CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION.
 SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.
+-->
 
-STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).
+<!-- STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).  -->
 
-geography itself endogenous to settler colonization, this is tricky!
+<!-- geography itself endogenous to settler colonization, this is tricky! -->
 
-## sources of data
+## Sources of data
 
-Census; ARCIA; Interior; etc
+Census (pop, employment, group quarters); ARCIA; Interior; BLM; US Forest Service.
 
-CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho...
+<!-- CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho... -->
 
-we treat police, education, and social work employment as core areas of the domestic social state. what is the name for that? the welfare state? the carceral state? the therapeutic state? that thing.
+<!-- we treat police, education, and social work employment as core areas of the domestic social state.
+what is the name for that?
+the welfare state?
+the carceral state?
+the therapeutic state?
+that thing.
+-->
 
-let's look at how it played out in the progressive era across states and regions.
+<!-- let's look at how it played out in the progressive era across states and regions.
 we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.
+-->
 
-First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita?
+<!-- First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita? -->
 
@@
 Panel 1 displays this data as the relative population composition of Black residents relative to to all residents, and panel 2 displays the absolute size of the Black population across US states.
-
-DISCUSS THE DISTRIBUTION
+We display panel 2 on a logarithmic scale to enable closer examination of the first stage of the Great Migration.
+While Black populations were a low relative share of the population in places like New York, Pennsylvania and Illinois in 1920, there were substantial and growing Black populations in these states, as shown on the count, rather than proportion, scale.
 
@@
 ```
+
+Figure \@ref(fig:blk-ts) shows two sets of time series describing the Black population in US States between 1870 and 1930 using the US Census.
+Panel 1 shows the size of the Black population in counts, represented on a logarithmic scale to allow for easier comparisons of population order of magnitude.
+Panel 2 shows the proportion of a state's population reported as Black on the US Census, to allow for comparisons of population composition across places.
+
+Black populations increased uniformly across all US states between 1870 and 1930 (see panel 1), though there are clear differences in rates of increase and absolute levels across states.
+For example, in the Northeast region, the primarily rural states of Maine, Vermont, and New Hampshire had relatively small Black populations across the time period.
+In Maine, reported Black populations were between 300 and 2400 during this period (1% sample data, NO standard errors REPORTED HERE because it's a pain but I'll keep working on it).
+By contrast, states with large and growing cities, like New York, New Jersey, and Pennsylvania all had large Black populations that grew substantially between 1870 and 1930.
+For example, Pennsylvania's Black population was recorded at about 70,000 in 1870, and 430,000 in 1930.
+
+Note that some places described in these plots were not admitted as states at the beginning of this time period (CHECK US CENSUS QUALITY IN TERRITORIES), and were intensely contested in military and non-military conflicts between the US government and tribal nations.
+MORE HERE.
+HOW DOES CONTESTATION ENTER INTO THE STAT PROCESS?
+CENSUS IS COUNTING SETTLERS + ENSLAVED, SO THATS STILL VALID FOR CONTEXTS OF SET COL ASSUMING GOOD TERRITORIAL MEASUREMENT (CHECK THIS!).
+
+Similar population trends are evident in the North Central region (commonly referred to as the Midwest).
+States with large cities, like Illinois, Ohio, Missouri, and Michigan had both relatively high Black populations at the beginning of the period and saw substantial growth in these populations over time.
+For example, the US Census recorded about 31,000 Black residents in Illinois in 1870.
+By 1930, the Census recorded about 330,000 Black residents of Illinois.
+Note that some of these places were not admitted as states at the beginning of this time period.
+(CHECK US CENSUS QUALITY IN TERRITORIES), and were sites of ongoing military conflicts between the US government and tribal nations.
+
+In the Western Region, states reported uniformly small Black populations at the beginning of this period, with varying (but positive) rates of growth.
+Note that many of these places were not yet states at the beginning of the time series; only California, Oregon, and Nevada had been admitted as states by 1870.
+California recorded the largest Black population by the end of the period.
+The Census records an increase in Black population in California from about 3000 in 1870 to about 84,000 in 1930.
+Arizona, Colorado, and Washington had smaller but substantial Black populations by the end of the time period, at around 10,000 persons.
+
+In the Southern Region (which includes all states that seceded to join the former Confederate States of America in addition to Delaware, Maryland, West Virginia, Kentucky, and Oklahoma), Black populations were consistently large during this period, with the largest populations in states in which white elites had enslaved millions of Black people for work in plantation-driven agriculture, like Georgia, Mississippi, and South Carolina.
+For example, the Census recorded 540,000 Black residents of Georgia in 1870, increasing to 1,100,000 Black residents in 1930.
+
+Note though, that these increases in population size do not always mirror trends in relative population composition (see Panel 2).
+During this period, even as Black populations grew dramatically in some Northeastern and Midwestern states during the first Great Migration, proportional population shares of Black residents remained low and relatively stable.
+In the South, even as Black populations continued to grow in absolute terms, the proportional population share of Black residents was decreasing consistently throughout the region.
+
+This divergence between absolute and relative population trends illustrates that Black population share likely fails to adequately capture the salience of race for local political contexts.
+
+Next let's look at sharecropping
 
@@
 
-Figure \@ref(fig:blk-ts) shows two sets of time series describing the Black population in US States between 1870 and 1930 using the US Census.
-Panel 1 shows the size of the Black population in counts, represented on a logarithmic scale to allow for easier comparisons of population order of magnitude.
-Panel 2 shows the proportion of a state's population reported as Black on the US Census, to allow for comparisons of population composition across places.
-
-Black populations increased uniformly across all US states between 1870 and 1930 (see panel 1), though there are clear differences in rates of increase and absolute levels across states.
-For example, in the Northeast region, the primarily rural states of Maine, Vermont, and New Hampshire had relatively small Black populations across the time period.
-In Maine, reported Black populations were between 300 and 2400 during this period (1% sample data, NO standard errors REPORTED HERE because it's a pain but I'll keep working on it).
-By contrast, states with large and growing cities, like New York, New Jersey, and Pennsylvania all had large Black populations that grew substantially between 1870 and 1930.
-For example, Pennsylvania's Black population was recorded at about 70,000 in 1870, and 430,000 in 1930.
-
-Note that some places described in these plots were not admitted as states at the beginning of this time period (CHECK US CENSUS QUALITY IN TERRITORIES), and were intensely contested in military and non-military conflicts between the US government and tribal nations.
-MORE HERE.
-HOW DOES CONTESTATION ENTER INTO THE STAT PROCESS?
-CENSUS IS COUNTING SETTLERS + ENSLAVED, SO THATS STILL VALID FOR CONTEXTS OF SET COL ASSUMING GOOD TERRITORIAL MEASUREMENT (CHECK THIS!).
-
-Similar population trends are evident in the North Central region (commonly referred to as the Midwest).
-States with large cities, like Illinois, Ohio, Missouri, and Michigan had both relatively high Black populations at the beginning of the period and saw substantial growth in these populations over time.
-For example, the US Census recorded about 31,000 Black residents in Illinois in 1870.
-By 1930, the Census recorded about 330,000 Black residents of Illinois.
-Note that some of these places were not admitted as states at the beginning of this time period.
-(CHECK US CENSUS QUALITY IN TERRITORIES), and were sites of ongoing military conflicts between the US government and tribal nations.
-
-In the Western Region, states reported uniformly small Black populations at the beginning of this period, with varying (but positive) rates of growth.
-Note that many of these places were not yet states at the beginning of the time series; only California, Oregon, and Nevada had been admitted as states by 1870.
-California recorded the largest Black population by the end of the period.
-The Census records an increase in Black population in California from about 3000 in 1870 to about 84,000 in 1930.
-Arizona, Colorado, and Washington had smaller but substantial Black populations by the end of the time period, at around 10,000 persons.
-
-In the Southern Region (which includes all states that seceded to join the former Confederate States of America in addition to Delaware, Maryland, West Virginia, Kentucky, and Oklahoma), Black populations were consistently large during this period, with the largest populations in states in which white elites had enslaved millions of Black people for work in plantation-driven agriculture, like Georgia, Mississippi, and South Carolina.
-For example, the Census recorded 540,000 Black residents of Georgia in 1870, increasing to 1,100,000 Black residents in 1930.
-
-Note though, that these increases in population size do not always mirror trends in relative population composition (see Panel 2).
-During this period, even as Black populations grew dramatically in some Northeastern and Midwestern states during the first Great Migration, proportional population shares of Black residents remained low and relatively stable.
-In the South, even as Black populations continued to grow in absolute terms, the proportional population share of Black residents was decreasing consistently throughout the region.
-
-This divergence between absolute and relative population trends illustrates that Black population share likely fails to adequately capture the salience of race for local political contexts.
-
-next let's look at sharecropping
+First, I evaluate the bivariate relationship between Black farm residency and Black tenant farm residency.
+The gap between the line y=x and plotted values indicates the number of Black people living on farms owned by their family.
+Tenant farming was the dominant mode of farming for Black farmers in all states with a substantial population of Black farmers in 1920.
+Virginia appears to be the only state with a substantial share of Black farm owners as of 1920.
+In states like Georgia, Mississippi, and South Carolina, these data clearly illustrate the dominance of the tenant-farming based sharecropping system.
 
@@
 Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though.
-Likely makese sense to use sharecropping as log count rather than as proportion.
+Likely makes sense to use sharecropping as log count rather than as proportion.
 

### 2023-12-06 scratch_land_measures.r

@@ -132,21 +132,19 @@ st_1796<-c("ME", "NH", "VT", "NY", "MA",
            "MD", "VA", "NC", "SC", "GA",
            "WV", "KY", "TN")
 
+cessions_diff<-cessions_diff %>% 
+  mutate(st_1796 = STUSPS%in%st_1796)
+
 ggplot(cessions_diff,
        aes(fill = st_1796)) + 
   geom_sf() + 
   theme_void()
 
-cessions_diff<-cessions_diff %>% 
-  mutate(st_1796 = STUSPS%in%st_1796)
-
-
 ### unceded light blue, reservation dark blue
 ggplot(st) + 
   geom_sf() + 
-  geom_sf(data = cessions_diff %>% 
-            filter(st_1796==F), fill = "dodgerblue") +
-  # geom_sf(data = aianh, fill = "blue") + 
+  geom_sf(data = cessions_diff, fill = "dodgerblue") +
+  geom_sf(data = aianh, fill = "blue") + 
   theme_void()
 
 ### layer state boundaries
@@ -187,21 +185,54 @@ ggplot(ts_full,
             lty = 2) + 
   facet_wrap(~STUSPS)
 
+### top code cessions to total acres
+
+ts_full<-ts_full %>% 
+  filter(!is.na(year)) %>% 
+  mutate(prop_ceded = cumulative_ceded / total_acres_2021,
+         prop_ceded = ifelse(prop_ceded>1, 1, prop_ceded))
+
 ggplot(ts_full,
-       aes(x = year, y = cumulative_ceded / total_acres_2021)) + 
+       aes(x = year, y = prop_ceded)) + 
   geom_line() +
-  geom_hline(yintercept = 1, lty = 2) + 
   facet_wrap(~STUSPS)
 
-#### THIS CAPTURES DISPOSESSION GEOGRAPHY AND PROCESS. WHAT ABOUT RESERVATION LANDS?
+#### THIS CAPTURES DISPOSESSION GEOGRAPHY AND PROCESS. OUTPUT
 
+ts_full %>% 
+  select(year, STUSPS,
+         ceded_acres, cumulative_ceded,
+         total_acres_2021, prop_ceded) %>% 
+  rename(state = STUSPS) %>% 
+  write_csv("./data/disposession_ts.csv")
+
+##############################################################
 ### REDISTRIBUTION.
+##############################################################
 
 ### MORRILL ACT DATA
[…]

### 2023-12-06 make_native_land_theft_state_data.r
add script to make time series for disposession / distribution

@@ -1,0 +1,195 @@
+### edited 12/6/23; FE
+#### read royce data, morill data, homestead data
+#### make state time series for native land disposession up to 1920
+#### make state time series for settler distribution up to 1920
+# - land grab u shapes for morill parcels https://github.com/HCN-Digital-Projects/landgrabu-data
+# - royce map shapes from forest service 'Indian Land Cessions in US' https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
+# - homestead data from http://homestead.unl.edu/projects/homesteading-the-plains/data/blm_homesteads.xlsx
+
+library(tidyverse)
+library(tidycensus)
+library(tigris)
+library(units)
+library(sf)
+
+options(tigris_use_cache = T)
+
+##############################################################################
+# pull state shape file --------------------------------------------------
+##############################################################################
+
+st<-states(cb = T) %>%
+  st_transform(8528) %>% 
+  filter(!STUSPS%in%
+           c("HI", "AK", 
+             "PR", "GU", "AS","VI",
+             "MP"))
+
+##############################################################################
+# read and format land dispossession data /shapes ----------------------------
+##############################################################################
+
+cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
+
+cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
+  mutate(time = as_date(TribalCededLandsTableCessDate1),
+         year = year(time)) %>% 
+  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
+  select(CESSNUM, year)
+
+cessions<-cessions %>% 
+  left_join(cessions_dat)
+
+cessions<-cessions %>% 
+  st_transform(8528)
+
+cessions_diff<-st_difference(st, st_union(cessions))
+
+### cessions_diff is land not ceded prior to 1893. 
+### needs manual coding for pre 1784 claims
+st_1796<-c("ME", "NH", "VT", "NY", "MA",
+           "CT", "RI", "DE", "NJ", "PA",
+           "MD", "VA", "NC", "SC", "GA",
[…]

### 2023-12-06 make_native_land_theft_state_data.r
format for MS

@@ -25,6 +25,10 @@ st<-states(cb = T) %>%
              "PR", "GU", "AS","VI",
              "MP"))
 
+
+st_acres<-data.frame(state = st$STUSPS,
+                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
+
 ##############################################################################
 # read and format land dispossession data /shapes ----------------------------
 ##############################################################################
@@ -94,7 +98,7 @@ ts_full %>%
          ceded_acres, cumulative_ceded,
          total_acres_2021, prop_ceded) %>% 
   rename(state = STUSPS) %>% 
-  write_csv("./data/disposession_ts.csv")
+  write_csv("./data/temp_disposession_ts.csv")
 
 ##############################################################
 ### REDISTRIBUTION.
@@ -122,7 +126,9 @@ morill<-morill %>%
   arrange(year, Loc_State)
 
 ### homestead act data
-### read in, pivot, and select acreage data by state / year
+### modified raw data by cleaning columns / headers for read_csv() otherwise unchanged
+### http://homestead.unl.edu/projects/homesteading-the-plains/data/blm_homesteads.xlsx
+
 hs<-read_csv("./data/blm_homesteads_clean.csv") %>%
   pivot_longer(claims1868:acres1961,
                names_to = "name",
@@ -192,4 +198,4 @@ land_distrib <- land_distrib %>%
 
 ### output
 land_distrib %>% 
-  write_csv("./data/distribution_ts.csv")
+  write_csv("./data/temp_distribution_ts.csv")

### 2023-12-06 make_pop_data_ipumsrmod.R
begin ipumrs/srvyr mod

@@ -1,0 +1,91 @@
+## ----setup, include = F, message = F----------------------------------
+library(tidyverse)
+library(ipumsr)
+library(srvyr)
+library(maps)
+
+## ----read IPUMS tables--------------------------------------------------
+# ### occupations 
+
+dat_occ<-read_ipums_micro("./data/usa_00071.xml")
+dat_occ<-dat_occ %>%
+  rename_all(tolower) %>%
+  mutate(occ1950 =
+           case_when(
+             occ1950 == 79 ~ "socwork",
+             occ1950 == 773 ~ "police",
+             occ1950 == 771 ~ "police",
+             occ1950 == 782 ~ "police",
+             occ1950 == 93 ~ "teachers"
+           )) %>%
+  group_by(year, statefip, occ1950) %>%
+  summarize(emp = sum(perwt)) %>%
+  pivot_wider(names_from = occ1950, values_from = emp,
+              values_fill = 0)
+# pop (total, age, race, nativitiy, urban, farm, landown)
+
+# srvyr mod ---------------------------------------------------------------
+dat_immig<-read_ipums_micro("./data/usa_00073.xml")
+
+dat_immig<-dat_immig %>% 
+  filter(YEAR>=1900) %>% 
+  as_survey_design(
+    ids = CLUSTER,
+    weights = PERWT,
+    strata = STRATA,
+    nest = T
+  )
+
+temp<-dat_immig %>% 
+  filter(YEAR == 1920) %>% 
+  group_by(STATEFIP) %>% 
+  summarize(pop = survey_total())
+
+###
+
+dat_immig<-dat_immig %>% 
+  rename_all(tolower) %>%
+  group_by(year, statefip) %>%
+  summarize(tot_pop = sum(perwt),
+            children = sum((age<18) * perwt),
+            white_immig =
+              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+            blk =
+              sum((race == 2) * perwt),
+            blk_sharecrop = 
+              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
+            blk_farms = sum((race==2) * (farm==2) * perwt),
+            urban_pop = sum(hhwt * (urban ==2)),
+            urban_immig_pop = 
+              sum((urban==2) * 
+                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+            tot_hh = sum(hhwt))
+### group quarters, 100%
[…]

### 2023-12-13 ms.Rmd
begin ipumrs/srvyr mod
add model script, update ms

--- 2023-12-06 t1_analysis.Rmd
@@
 output: bookdown::html_document2
-abstract: "Place and time variation in problems of settler colonization are related to the kinds of governments that local settler elites build. "
+abstract: "Place and time variation in settler colonial projects are related to the kinds of states that settler elites build."
 ---
-
-```{r setup, include = F, message = F}
+```{r setup, include = F, message = F, cache = F}
 knitr::opts_chunk$set(message = F, echo = F, warning = F)
@@
 library(brms)
+library(tidycensus)
+library(tigris)
+library(units)
+library(sf)
+
+options(tigris_use_cache = T)
 options(scipen = 999)
-options(dplyr.print_max = 1e9)
+options(dplyr.print_max = 30)
 map<-map_data("state")
+theme_set(theme_minimal())
+```
 
-theme_set(theme_minimal())
+```{r read, include = F, message = F, cache = T}
+
+### state shape file
+st<-states(cb = T) %>%
+  st_transform(8528) %>% 
+  filter(!STUSPS%in%
+           c("HI", "AK", 
+             "PR", "GU", "AS","VI",
+             "MP"))
+
+# read data from make_* scripts
+pop<-read_csv("./data/temp_population_ts.csv")%>%
+  filter(STUSPS!="AK")
+disp<-read_csv("./data/temp_disposession_ts.csv")%>%
+  filter(STUSPS!="AK")
+dist<-read_csv("./data/temp_distribution_ts.csv")%>%
+  filter(STUSPS!="AK")
+
+## add region to all
+disp<-disp %>%
+  left_join(pop %>%
+              select(STUSPS, region, division) %>%
+  distinct())
+
+dist<-dist %>%
+  left_join(pop %>%
+              select(STUSPS, region, division) %>%
+  distinct())
+
+source("models.R")
 ```
@@
 
-<!-- HOW ARE PROJECTS / PROBLEMS OF SETTLER COLONIAL GOVERNANCE RESOLVED? WHAT ORGS AND INSTITUTIONS ARE BUILT?  -->
-
-Settler colonial regimes seek to resolve different problems across places and within places over time.
-This paper proposes a quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects).
-We show that a series of historical economic and demographic measures (proxies for variation in modes of settler colonial extraction and exploitation) are closely related to trajectories of state formation across places, measured as employment levels in core social policy system sectors.
+This paper proposes a comparative quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects).
+We show that spatial and temporal variation in modes of settler colonial extraction and exploitation are closely related to trajectories of state formation across places.
 
 At a sub-national level, settler colonial regimes seek to address a range of problems inherent to their nation-building project: the dispossession of indigenous lands, the erasure of indigenous peoples, the exploitation of racially marginalized laborers, and the reproduction and expansion of the settler polity.
-We show that the historical geography of these problems is closely related to the policy infrastructure that governments develop.
-The consequences of this variation in trajectories are likely broad and varied.
+We show that the historical geography of these problems of settler colonial governance were closely related to the policy infrastructure to manage social life that governments develop.
 
@@
 
-```{r read}
-# ### on libra, read raw
-# ### read employment data
-# ### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
-dat_occ<-read_csv("./data/usa_00071.csv") %>%
-  rename_all(tolower) %>%
-  mutate(occ1950 =
-           case_when(
-             occ1950 == 79 ~ "socwork",
-             occ1950 == 773 ~ "police",
-             occ1950 == 771 ~ "police",
-             occ1950 == 782 ~ "police",
-             occ1950 == 93 ~ "teachers"
-           )) %>%
-  group_by(year, statefip, occ1950) %>%
-  summarize(emp = sum(perwt)) %>%
-  pivot_wider(names_from = occ1950, values_from = emp,
-              values_fill = 0)
-### AK OR HI HERE? check that at IPUMS
-
-# pop data (total, age, race, nativitiy, urban, farm, landown)
-dat_immig<-read_csv("./data/usa_00072.csv") %>%
-  rename_all(tolower) %>%
-  group_by(year, statefip) %>%
-  summarize(tot_pop = sum(perwt),
-            children = sum((age<18) * perwt),
-            white_immig =
-              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
-            blk =
-              sum((race == 2) * perwt),
-            blk_sharecrop = 
-              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
-            blk_farms = sum((race==2) * (farm==2) * perwt),
-            urban_pop = sum(hhwt * (urban ==2)),
-            urban_immig_pop = 
-              sum((urban==2) * 
-                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
-            tot_hh = sum(hhwt))
-### AK OR HI HERE? check that at IPUMS
-
-### group quarters data, 100%
-gq<-read_csv("./data/usa_00069.csv") %>%
-  rename_all(tolower) %>%
-  filter(age<18) %>%
-  group_by(year, statefip, gqtype) %>%
-  summarize(gq_pop = sum(perwt, na.rm = T)) %>%
-  mutate(gqtype = case_when(
-    gqtype == 2 ~ "child_inst_correctional",
-    gqtype == 3 ~ "child_inst_mental",
-    gqtype == 4 ~ "child_inst_poor_disabled")) %>%
-  filter(gqtype!="other") %>%
-  pivot_wider(names_from = gqtype, values_from = gq_pop,
-              values_fill = 0)
-### AK OR HI HERE? check that at IPUMS
-
-### join
-dat <- data.frame(state = state.abb,
-               region = state.region,
-               division = state.division) %>%
-  left_join(state.fips %>%
-              select(fips, abb) %>%
-              distinct() %>%
-              rename(state = abb, statefip = fips)) %>%
-  left_join(dat_immig) %>%
-  left_join(dat_occ) %>%
-  ungroup() %>%
-  left_join(gq)
-
-### homestead act data
-#### load in homestead data and boarding school data
-### state measure can be cumulative sum of homesteads, that gets at polity
-### homesteading data
-y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
-  left_join(dat %>%
-              select(state, region) %>%
-              distinct())
-
-hs<-read_csv("./data/blm_homesteads_clean.csv") %>%
-  pivot_longer(claims1868:acres1961,
-               names_to = "name",
-               values_to = "value") %>%
-  mutate(year = str_sub(
-    name, start = -4, end = -1),
-    type = str_sub(
-      name, start = 1, end = -5)) %>%
-  select(-name) %>%
-  mutate(year = as.numeric(year)) %>%
-  mutate(value = ifelse(is.na(value),
-                        0,
-                        value))
-
-### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
-hs_dakota<-hs %>%
-  filter(state == "Dakota Territory") %>%
-  mutate(value = value / 2) %>%
-  select(-state)
-
-hs<-hs %>%
-  bind_rows(hs_dakota %>%
-              mutate(state = "North Dakota")) %>%
-  bind_rows(hs_dakota %>%
-              mutate( state = "South Dakota")) %>%
-  filter(state != "Dakota Territory") %>%
-  group_by(state, year, type) %>%
-  summarize(value = sum(value))
-
-hs<-hs %>%
-  arrange(state, year) %>%
-  group_by(state, type) %>%
-  mutate(cum_value = cumsum(value)) %>%
-  ungroup()
-
-## expand_grid to add zeroes
-### use y_o_s for index
-state<-unique(y_o_s$statename)
-year<-unique(hs$year)
-type<-unique(hs$type)
-temp<-expand_grid(state, year, type)
-
-hs<-temp %>%
-  left_join(hs) %>%
-  replace_na(list(value = 0,
-                  cum_value = 0))
-
-hs <- hs %>%
-  select(-value) %>%
-  pivot_wider(names_from = type,
-              values_from = cum_value,
-              names_prefix = "hs_")
-
-### add ID and pop vars back to hs
-
-y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
-  left_join(dat %>%
-              select(state, region) %>%
-              distinct()) %>%
-  filter(!(is.na(region)))
-
-hs<-hs %>%
-  rename(statename = state) %>%
-  left_join(y_o_s)
-
-#### join cumulative acres / claims onto dat for the valid census years, then carried forward
-
-dat<-dat %>%
-  left_join(hs) %>% 
-  filter(year>=1880, year<=1920)
-
-# # on local read temp
-# dat<-read_csv("./data/temp_data.csv")
-# hs<-read_csv("./data/temp_hs.csv")
-
-### land grab u data
-# lgu_cessions<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Cessions.csv")
-### this is cessions from the Royce maps, overlaps with morrill, but includes prior cessions as well
-### would need to use shapefiles to portion out by state
-lgu_parcels<-read_csv("./data/landgrabu-data/Morrill_Act_of_1862_Indigenous_Land_Parcels_Database/CSVs/Parcels.csv")
-```
-
 # Methods
 
-We are generally interested in evaluating whether a set of policy outcomes {employment, infrastructure} are related to a set of demographic and economic predictors {population composition, economic activity, ownership regime}.
+*We evaluate whether a set of policy outcomes {employment, infrastructure} are related to a set of demographic and economic predictors {population composition, economic activity, ownership regime} at the state and year-levels.*
 
@@
 We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states.
-A census taker is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.).
+A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.).
 We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.
+
+### measures
+
+All measures are computed at the state-year level.
+
+EMPLOYMENT:
+
+- teachers (1900 - 1930 100%)
+- cops (1900 - 1930 100%)
+- social workers (1900 - 1930 100%)
+
+POPULATION:
+
+- total (1900 - 1930 1%)
+- children (1900 - 1930 1%)
+- white immigrants in *urban households* (1900 - 1930 1%)
+- Black population living in tenant farm households (1900 - 1930 1%)
+- Population in group quarters (correctional institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
+- Population in group quarters (mental institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
+- Population in group quarters (institutions for the elderly, handicapped, and poor) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
+
+LAND:
+
+- cumulative acres parceled through the Morill Act (Robert Lee, ``Morrill Act of 1862 Indigenous Land Parcels Database,'' High Country News, 2020)
+- cumulative acres parceled through the Homestead Act (``Homesteads,'' Bureau of Land Management of the U.S. Department of the Interior, 1962. digitized by Edwards, Friefeld and and Wingo 2017)
 
@@
 
-EDA; multilevel models on outcome rates (state intercepts, state x year slopes, log outcomes for nonlinear process, etc) estimated with hmc, weak priors
+Model is
+
+\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]
+
+\[ \lambda = \textrm{E}[y_{st}] \]
+
+\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \delta_t + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} \]
+
+\[\gamma_s \sim N(0, \sigma^2_\gamma)\]
+
+With weakly informative priors
+
+\[\beta \sim N(0, 20)\]
+\[\sigma \sim t(3, 0, 2.5)\]
+\[\alpha \sim \textrm{Gamma}(0.01, 0.01)\]
 
@@
 
-## Geographic contexts for settler colonial policymaking across US states 1880 - 1920
+```{r fig.cap = "demographic map of US in 1900 - Black, White immig, white non-immig"}
+plot_dat<-pop %>% 
+  filter(year == 1900) %>% 
+  mutate(white_non_immig = white_non_immig / (white_non_immig + white_immig),
+         white_immig = 1 - white_non_immig,
+         black = blk / tot_pop) %>% 
+  select(STUSPS, white_immig, white_non_immig, black) %>% 
+  pivot_longer(cols = white_immig:black) %>% 
+  mutate(name = case_when(
+    name == "black" ~ "Black / \ntotal population",
+    name == "white_immig" ~ "White immigrant/\ntotal white population",
+    name == "white_non_immig" ~ "White non-immigrant/\ntotal white population"
+  ))
 
-### Jim Crow and the salience of anti-Blackness for local social and carceral policymaking
-
-Figure \@ref(fig:blk-map) shows the distribution of the recorded Black population across US states in 1920.
-Panel 1 displays this data as the relative population composition of Black residents relative to to all residents, and panel 2 displays the absolute size of the Black population across US states.
-We display panel 2 on a logarithmic scale to enable closer examination of the first stage of the Great Migration.
-While Black populations were a low relative share of the population in places like New York, Pennsylvania and Illinois in 1920, there were substantial and growing Black populations in these states, as shown on the count, rather than proportion, scale.
-
-```{r blk-map, fig.cap = "Black population, US States 1920 (note log scale)"}
-p1<-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = blk / tot_pop * 100)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
+st %>% 
+  left_join(plot_dat) %>% 
+  filter(!is.na(value)) %>% 
+  ggplot(aes(fill = value)) + 
+  geom_sf() + 
+  facet_wrap(~name, ncol = 3) + 
   theme_void() + 
+  scale_fill_viridis_c() + 
   theme(legend.position = "bottom") + 
-  labs(fill = "Percent of total",
-       subtitle = "Panel 1. Population composition")
-
-p2<-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = blk + 100)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis", trans = "log10") + 
-  theme_void() + 
-  theme(legend.position = "bottom") + 
-  labs(fill = "Population\n(note logarithmic scale)",
-       subtitle = "Panel 2. Total population")
-
-grid.arrange(p1, p2, nrow = 1)
+  labs(fill = "")
 ```
 
-Figure \@ref(fig:blk-ts) shows two sets of time series describing the Black population in US States between 1870 and 1930 using the US Census.
-Panel 1 shows the size of the Black population in counts, represented on a logarithmic scale to allow for easier comparisons of population order of magnitude.
-Panel 2 shows the proportion of a state's population reported as Black on the US Census, to allow for comparisons of population composition across places.
+This map shows key features of the demography of the US in 1900 that are closely related to the geography of settler colonialism in the early 20th century US.
 
-Black populations increased uniformly across all US states between 1870 and 1930 (see panel 1), though there are clear differences in rates of increase and absolute levels across states.
-For example, in the Northeast region, the primarily rural states of Maine, Vermont, and New Hampshire had relatively small Black populations across the time period.
-In Maine, reported Black populations were between 300 and 2400 during this period (1% sample data, NO standard errors REPORTED HERE because it's a pain but I'll keep working on it).
-By contrast, states with large and growing cities, like New York, New Jersey, and Pennsylvania all had large Black populations that grew substantially between 1870 and 1930.
-For example, Pennsylvania's Black population was recorded at about 70,000 in 1870, and 430,000 in 1930.
+BLACK POPS IN JIM CROW SOUTH
 
-Note that some places described in these plots were not admitted as states at the beginning of this time period (CHECK US CENSUS QUALITY IN TERRITORIES), and were intensely contested in military and non-military conflicts between the US government and tribal nations.
-MORE HERE.
-HOW DOES CONTESTATION ENTER INTO THE STAT PROCESS?
-CENSUS IS COUNTING SETTLERS + ENSLAVED, SO THATS STILL VALID FOR CONTEXTS OF SET COL ASSUMING GOOD TERRITORIAL MEASUREMENT (CHECK THIS!).
+US BORN WHITE POPS IN SOUTH, NEW ENGLAND, RURAL PARTS OF INNER MIDWEST
 
-Similar population trends are evident in the North Central region (commonly referred to as the Midwest).
-States with large cities, like Illinois, Ohio, Missouri, and Michigan had both relatively high Black populations at the beginning of the period and saw substantial growth in these populations over time.
-For example, the US Census recorded about 31,000 Black residents in Illinois in 1870.
-By 1930, the Census recorded about 330,000 Black residents of Illinois.
-Note that some of these places were not admitted as states at the beginning of this time period.
-(CHECK US CENSUS QUALITY IN TERRITORIES), and were sites of ongoing military conflicts between the US government and tribal nations.
+FOREIGN BORN WHITE POPS THROUGHOUT CITIES, WESTERN FRONTIER
 
-In the Western Region, states reported uniformly small Black populations at the beginning of this period, with varying (but positive) rates of growth.
-Note that many of these places were not yet states at the beginning of the time series; only California, Oregon, and Nevada had been admitted as states by 1870.
-California recorded the largest Black population by the end of the period.
-The Census records an increase in Black population in California from about 3000 in 1870 to about 84,000 in 1930.
-Arizona, Colorado, and Washington had smaller but substantial Black populations by the end of the time period, at around 10,000 persons.
+FIG 2, 3 - native disposession
 
-In the Southern Region (which includes all states that seceded to join the former Confederate States of America in addition to Delaware, Maryland, West Virginia, Kentucky, and Oklahoma), Black populations were consistently large during this period, with the largest populations in states in which white elites had enslaved millions of Black people for work in plantation-driven agriculture, like Georgia, Mississippi, and South Carolina.
-For example, the Census recorded 540,000 Black residents of Georgia in 1870, increasing to 1,100,000 Black residents in 1930.
+```{r}
+st<-states(cb = T) %>%
+  st_transform(8528) %>% 
+  filter(!STUSPS%in%
+           c("HI", "AK", 
+             "PR", "GU", "AS","VI",
+             "MP"))
 
-Note though, that these increases in population size do not always mirror trends in relative population composition (see Panel 2).
-During this period, even as Black populations grew dramatically in some Northeastern and Midwestern states during the first Great Migration, proportional population shares of Black residents remained low and relatively stable.
-In the South, even as Black populations continued to grow in absolute terms, the proportional population share of Black residents was decreasing consistently throughout the region.
+st_acres<-data.frame(state = st$STUSPS,
+                     total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
 
-This divergence between absolute and relative population trends illustrates that Black population share likely fails to adequately capture the salience of race for local political contexts.
+### royce map data
+## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
+## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
+cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
 
-Next let's look at sharecropping
+cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
+  mutate(time = as_date(TribalCededLandsTableCessDate1),
+         year = year(time)) %>% 
+  rename(CESSNUM = TribalCededLandsNewCESSNUM) %>% 
+  select(CESSNUM, year)
 
-```{r blk-ts, fig.cap = "Black population by state 1870-1940"}
-p1<-ggplot(dat,
-       aes(x = year, y = blk / tot_pop,
-           group = state)) + 
-  geom_line(alpha = 0.5) + 
-  geom_text(data = dat %>% 
-              filter(!(is.na(region))) %>% 
-              filter(year == 1930),
-            aes(x = year,
-                y = blk / tot_pop,
-                label = state),
-            position = position_dodge(width = 8),
-            size = 2,
-            alpha = 0.7) + 
-  labs(subtitle = "Panel 2. Proportion Black",
-       y = "Proportion",
-       x = "Year") + 
-  facet_wrap(~region)
+cessions<-cessions %>% 
+  left_join(cessions_dat)
 
-p2<-ggplot(dat,
-       aes(x = year, y = blk,
-           group = state)) + 
-  geom_line(alpha = 0.5) + 
-  geom_text(data = dat %>% 
-              filter(!(is.na(region))) %>% 
-              filter(year == 1930),
-            aes(x = year,
-                y = blk,
-                label = state),
-            position = position_dodge(width = 8),
-            size = 2,
-            alpha = 0.7) + 
-  # scale_y_log10() + 
-  labs(subtitle = "Panel 1. Black population",
-       y = "Count",
-       x = "Year") + 
-  facet_wrap(~region)
+cessions<-cessions %>% 
+  st_transform(8528)
 
-grid.arrange(p2, p1, nrow = 1)
+cessions_diff<-st_difference(st, st_union(cessions))
+
+cessions_diff<-cessions_diff %>% 
+  mutate(year = NA)
+
+### layer state boundaries
+cessions<-st_intersection(cessions, st)
+### add in settler claimed territories as of 1796
+cessions2 <- cessions %>%  
+  bind_rows(cessions_diff %>%  
+              mutate(year = NA))
+
+ggplot(cessions2,
+       aes(fill = year)) + 
+  geom_sf() + 
+  theme_void() + 
+  scale_fill_viridis_c() + 
+  labs(fill = "Year",
+       title = "Post-1784 Indigenous disposession",
+       subtitle = "Cessions to US federal government through 1894") 
 ```
 
-First, I evaluate the bivariate relationship between Black farm residency and Black tenant farm residency.
-The gap between the line y=x and plotted values indicates the number of Black people living on farms owned by their family.
-Tenant farming was the dominant mode of farming for Black farmers in all states with a substantial population of Black farmers in 1920.
-Virginia appears to be the only state with a substantial share of Black farm owners as of 1920.
-In states like Georgia, Mississippi, and South Carolina, these data clearly illustrate the dominance of the tenant-farming based sharecropping system.
+MANY OF THE LOCATIONS THESE FIRST GENERATION SETTLERS WERE GOING TO WERE RECENT DISPOSESSIONS, PARTICULARLY CIVIL WAR ERA.
+IN 1900, MANY OF THESE LANDS HAD BEEN CLAIMED BY US IN VERY RECENT PAST, WITHIN 50 YEARS.
+SECOND / THIRD WAVE OF WESTERN SETTLERS.
+BUILDING THE SETTLER COLONIAL POLITY / NATION STATE AS KEY AGENDAS.
+
+```{r fig.cap = "Disposession of Native lands in the 19th c."}
+p1<-ggplot(disp,
+       aes(x = year,
+           group = STUSPS)) + 
+  geom_line(aes(y = cumulative_ceded / 1e6)) + 
+  labs(x = "Year",
+       y = "Acres (millions)") + 
+  lims(y = c(0, 150)) + 
+  facet_wrap(region~division) + 
+  labs(subtitle = "1. Total land cessions")
+
+p2<-ggplot(disp,
+       aes(x = year,
+           group = STUSPS)) + 
+  geom_line(aes(y = prop_ceded)) + 
+  labs(x = "Year",
+       y = "Proportion") + 
+  facet_wrap(region~division) + 
+  labs(subtitle = "2. Proportion of state area")
+
+grid.arrange(p1, p2, ncol = 2)
+```
+
+This shows the timing and spatial distribution in cumulative land disposession by state and region.
+Panel 1 shows the lands claimed by the US in millions of acres by state from 1796 - 1894.
+Panel 2 shows these patterns of Native land loss as a proportion of the contemporary areas of US states.
+
+Clear regional patterns here.
+Land loss occurred earlier in states closer to the Atlantic coast.
+Dispossession accelerated in the North Central and Western regions throughout the middle portion of the 19th c.
+
+```{r fig.cap = "Two mechanisms for the distribution of disposessed Native Lands to settlers"}
+pdat<-disp %>% 
+  select(STUSPS, total_acres_2021) %>% 
+  distinct() %>% 
+  left_join(dist %>% 
+              filter(year == 1900) %>% 
+              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
+  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
+  left_join(st)
+
+p1<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = morill_acres_cum / 1e6)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Acres (millions)",
+       subtitle = "1. Lands patented under Morill Act (1862) by 1900") + 
+  theme(legend.position = "bottom")
+
+p2<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = homestead_acres_cum / 1e6)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Acres (millions)",
+       subtitle = "2. Lands patented under Homestead Act (1862) by 1900") + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, ncol = 2)
+```
+
+FIGURE - by employment type, facet ~ year
 
 ```{r}
-ggplot(dat %>% 
-             filter(year == 1920,
-                    blk_farms >= 1e4),
-       aes(x = blk_farms, y = blk_sharecrop,
-           label = state)) + 
-  geom_text(alpha = 0.5) + 
-  geom_abline(slope = 1) + 
-  labs(x = "Black people living on farms",
-       y = "Black people living on tenant farms",
-       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
-```
+pdat<-pop %>% 
+  left_join(st)
+# could use children in denominator here
+# keeping as total population for comparison 
+ggplot(pdat,
+       aes(fill = teachers / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Teachers per\n1,000 population")
 
-And the bivar with sharecropping / blk pop comp
+ggplot(pdat,
+       aes(fill = police / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Police per\n1,000 population")
 
-```{r}
-ggplot(dat %>% 
-         filter(year == 1920),
-       aes(x = blk / tot_pop, y = blk_sharecrop,
-           label = state,
-           color = region)) + 
-  #geom_point(alpha = 0.5) + 
-  geom_text(alpha = 0.5) + 
-  scale_y_log10() + 
-  labs(x = "Proportion Black pop",
-       y = "Black people living on tenant farms",
-       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
-```
-
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = blk_sharecrop + 10)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis",
-                        trans = "log10") + 
+ggplot(pdat,
+       aes(fill = socwork / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  scale_fill_viridis_c() + 
   theme_void() + 
-  labs(fill = "Number of Black residents\non tenant farms\n(note logarithmic scale)")
-```
-
-Places with large Black pops in 1920 also had very high sharecropping rates.
-Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though.
-Likely makes sense to use sharecropping as log count rather than as proportion.
-
-### Settlers and indigenous disposession
-
-2) HOMESTEAD CLAIMS - VERY CLEAR
-
-  Explain
-
-```{r}
-p1<-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = hs_acres / 1e6)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() + 
-  labs(fill = "Cumulative acres claimed\n(millions), 1920") + 
-  theme(legend.position = "bottom")
-
-p2<-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(hs_pop = hs_claims / tot_pop) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = hs_pop)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis") + 
-  theme_void() +
-  labs(fill = "Claims per resident\n1920")  +
-  theme(legend.position = "bottom")
-
-grid.arrange(p1, p2, nrow = 1)
-```
-
-```{r}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(region = factor(region, 
-                         levels = c("West", 
-                                    "North Central",
-                                    "South",
-                                    "Northeast"))) %>% 
-  mutate(hs_pop = hs_claims / tot_pop) %>% 
-  ggplot(aes(x = hs_pop,
-             y = fct_reorder(state, hs_pop))) + 
-  geom_point() + 
-  labs(x = "Cumulative homestead claims per state population, 1920",
-       y = "",
-       color = "Census Region") + 
-  facet_wrap(~region, 
-             scales = "free_y")
-```
-
-We use data on Homestead Act claims as an indicator of the volume of settlers at the time and place-level seeking to actively dispossess Native Nations of their lands.
-Data on homestead claims provide useful indicators of both the demographic size of the settler polity (measured as counts of claimants) and the magnitude of their resources (measured as acres claimed).
-COUPLED WITH WHAT ELSE?
-
-Figure \@ref(fig:ts-homestead-claims) shows the distribution of Homestead Act claims made by settlers between 1868 and 1961.
-No homestead claims were completed in the states in the Northeast region.
-In general, states in the North Central (Midwest) and Western regions had higher volumes of claims, though Oklahoma and Arkansas in the South also saw relatively large numbers of claims.
-By the end of the Homestead era, Montana had seen the largest number of claimants (151,500), followed by North Dakota (135,448), South Dakota (114,172), Colorado (107,618), and Nebraska (104,260).
-
-```{r ts-homestead-claims, fig.cap = "Cumulative Homestead act claims by state and year"}
-ggplot(hs %>% 
-         filter(year < 1940),
-       aes(x = year,
-           y = hs_claims,
-           group = state)) + 
-  geom_line(alpha = 0.5) + 
-  geom_text(data = hs %>% 
-              filter(year == 1935),
-            aes(x = year,
-                y = hs_claims,
-                label = state),
-            position = position_nudge(x = 8),
-            alpha = 0.7,
-            size = 2.5) + 
-facet_wrap(~region)  + 
-  labs(y = "Number of claims",
-       x = "Year",
-       subtitle = "") + 
-  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
-```
-
-NATIVE LANDHOLDINGS OVER TIME ARE DIRECTLY RELATED TO HOMESTEAD CLAIMS, MECHANISM OF DISPOSESSION.
-
-PULL AGENTS DATA FOR LANDHOLDINGS?
-
-### Urbanization and European immigration
-
-```{r immig-map, fig.cap = "White immigrant population in cities 1920"}
-dat %>% 
-  filter(year == 1920) %>% 
-  mutate(state = tolower(statename)) %>% 
-  ggplot(aes(fill = urban_immig_pop)) + 
-  geom_map(aes(map_id = state), map = map) + 
-  expand_limits(x = map$long, y = map$lat) + 
-  scale_fill_continuous(type = "viridis",
-                        trans = "log10") + 
-  theme_void() + 
-  labs(fill = "Population\n(note logarithmic scale)")
-```
-
-```{r urban-ts, fig.cap = "Urban white immigrant population by state 1880-1920"}
-p1<-ggplot(dat%>% 
-             filter(!(is.na(region))),
-           aes(x = year, y = urban_immig_pop / tot_pop,
-               group = state)) + 
-  geom_line(alpha = 0.5) + 
-  labs(subtitle = "Proportion of population white immigrant urban",
-       y = "Proportion",
-       x = "Year") + 
-  facet_wrap(~region)
-
-p2<-ggplot(dat%>% 
-             filter(!(is.na(region))),
-           aes(x = year, y = urban_immig_pop,
-               group = state)) + 
-  geom_line(alpha = 0.5) + 
-  labs(subtitle = "White immigrant urban population",
-       y = "Count",
-       x = "Year")+ 
-  facet_wrap(~region)
-
-grid.arrange(p2, p1, nrow = 1)
-```
-
-increase in proportion of households in urban across regions
-
-Also clear increase in absolute n of households in urban settings.
-
-MAYBE ALSO NOTE WHITE IMMIG POP IN FRONTIER STATES, DIFFERENT REGIMES FOR INCORPORATION, DIFFERENT RACIALIZATION PROCESSES FOR NORTH / W EURO IMMIGS VS SOUTH / EAST / IRISH?
-
-## Geographic variation in social and carceral state capacities 1880 - 1920
-
-what are states doing? how does it vary across places and over time?
-
-### Employment in core roles for the emergent welfare and carceral states
-
-```{r emp-ts-plot, fig.cap="Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}
-dat %>% 
-  filter(year > 1880) %>% 
-  group_by(state, year) %>% 
-  summarize(police = sum(police) / sum(tot_pop) * 1e3,
-            teachers = sum(teachers) / sum(tot_pop) * 1e3,
-            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
-  pivot_longer(cols = police:socworkers,
-               names_to = "name",
-               values_to = "employment")  %>%  
-  filter(year < 1940) %>% 
-  ggplot(aes(x = year, y = employment)) + 
-  geom_line(aes(group = state), alpha = 0.1) + 
-  geom_line(data = dat %>% ### add national line
-              group_by(year, region) %>%
-              filter(year > 1880, year < 1940) %>% 
-              summarize(police = sum(police, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3,
-                        teachers = sum(teachers, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3,
-                        socworkers = sum(socwork, na.rm=T) / sum(tot_pop, na.rm=T) * 1e3) %>%
-              pivot_longer(cols = police:socworkers,
-                           names_to = "name",
-                           values_to = "employment"),
-  aes(x = year, y = employment, color = region), linewidth = 1) +
-  facet_wrap(~name) + 
-  scale_y_log10() + 
-  labs(subtitle = "note variable y axis scales, national average dashed",
-       y = "Employment per 1,000 population (note log scale)") + 
-  theme(legend.position = "bottom")
-```
-
-During this period, we see very clear upward trajectories in employment in each of three categories: police, social workers, and teachers.
-At the national level, police employment nearly doubled from about 0.7 per 1,000 in 1900 to 1.3 per 1,000 population in 1930.
-Data on social workers shows an employment rate of 5 per 10,000 in 1900 nationally doubling to a rate of 1 per 1,000 by 1930. Teachers shows 1900 rate of 5.5 per 1,000, 1930 of 8.6 per 1,000.
-
-Figure 1 shows substantial variation in state employment initial values and trajectories during this period (thin solid lines in fig 1).
-There is a clear upward trajectory in employment rates for each class of worker.
-Social workers experienced the largest relative rate of growth (a feature of the near-zero initial rate of employment in most states), while teachers experience the largest absolute rate of growth nationally.
-
-```{r tab1}
-### tabular mean / variance
-t<-dat %>% 
-  group_by(state, year) %>% 
-  summarize(police = sum(police) / sum(tot_pop) * 1e3,
-            teachers = sum(teachers) / sum(tot_pop) * 1e3,
-            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
-  pivot_longer(cols = police:socworkers,
-               names_to = "name",
-               values_to = "employment")  %>% 
-  group_by(year, name) %>% 
-  summarize(mean = mean(employment, na.rm=T),
-                        sd = sd(employment, na.rm=T)) %>% 
-  filter(year == 1900 | year == 1930)
-
-knitr::kable(t, 
-             caption = "Table 1. Mean and variance of state employment rates 1900 - 1930",
-             digits = 3)
-```
-
-There is substantial heterogeneity in 1900 values for social policy employment across states.
-The standard deviation of social work employment is greater than the mean value of social work employment.
-For police, the sd is half the value of the mean, and for teachers it is one third of the value of the mean.
-At 1930, the sd for social work employment had increased dramatically, as had the sd for police employment.
-The sd for teacher employment was relatively stable over the period.
-These features show clear variation in starting conditions of employment during the period and suggest variation in trajectories of growth over the period.
-
-Some of this heterogeneity in starting conditions and trajectories is likely a feature of regional variation in policy configurations (which we will seek to explain variation in racial political economy as we proceed).
-
-```{r emp-ts-region, fig.cap = "Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}
-
-### CUT
-
-dat %>% 
-  filter(!(is.na(region))) %>% 
-  group_by(region, year) %>% 
-  summarize(police = sum(police) / sum(tot_pop) * 1e3,
-            teachers = sum(teachers) / sum(tot_pop) * 1e3,
-            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
-  pivot_longer(cols = police:socworkers,
-               names_to = "name",
-               values_to = "employment")  %>%  
-  ggplot(aes(x = year, y = employment, 
-             color = region)) + 
-  geom_line() + 
-  scale_y_log10() + 
-  facet_wrap(~name, ncol = 3) + 
-  labs(subtitle = "note variable y axis scales")
-```
-
-In figure 2 we subset the data from figure 1 to regional averages using the four-region US Census classification for states.
-Clear patterns are visible in the data here.
-On average, states in the US South had lower employment rates for all categories of social policy system employees.
-Northeastern states had the highest levels of police and social work employment across the period.
-And states in the West and North Central (Midwest) had the highest average rates of teacher employment.
-However, there is also substantial sub-regional variation in employment rates, as shown in Figure 3.
-
-```{r emp-ts-region2, fig.cap = "Employment rates per 1000 population 1900 - 1930 (US States, PUMS data)"}
-dat %>% 
-  group_by(state, region, year) %>% 
-  filter(!is.na(region)) %>% 
-  summarize(police = sum(police) / sum(tot_pop) * 1e3,
-            teachers = sum(teachers) / sum(tot_pop) * 1e3,
-            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
-  pivot_longer(cols = police:socworkers,
-               names_to = "name",
-               values_to = "employment")  %>%  
-  ggplot(aes(x = year, y = employment, group = state)) + 
-  geom_line(alpha = 0.3) + 
-  facet_wrap(region~name, scales = "free", ncol = 3) + 
-  labs(subtitle = "note variable y axis scales")
-```
-
-Figure 3 shows how states within regions still exhibit substantial variation in both initial employment rates and growth trajectories between 1900 and 1930 for social policy employment.
-Again, note that social work employment began at or near zero in 1900 in all states as the social work was built as a concrete profession.
-In the South a cluster of low-employment states drive the regional trend for social work and police, while a cluster of high employment states appear to drive the trend for social work employment in the Northeast, North Central, and West.
-
-```{r emp-map, fig.cap = "Employment rates per 1,000 1930"}
-t<-dat %>% 
-  group_by(state, region, year) %>% 
-  summarize(police = sum(police) / sum(tot_pop) * 1e3,
-            teachers = sum(teachers) / sum(tot_pop) * 1e3,
-            socworkers = sum(socwork) / sum(tot_pop) * 1e3) %>% 
-  pivot_longer(cols = police:socworkers,
-               names_to = "name",
-               values_to = "employment")  %>%  
-  filter(year == 1930)
-
-p1<-plot_usmap(data = t %>% 
-             filter(name == "socworkers"),
-           values = "employment")+ 
-  scale_fill_viridis_c() + 
-  labs(title = "social workers",
-       fill = "") + 
-  theme(legend.position = "bottom")
-
-p2<-plot_usmap(data = t %>% 
-             filter(name == "teachers"),
-           values = "employment")+ 
-  scale_fill_viridis_c() + 
-  labs(title = "teachers",
-       fill = "") + 
-  theme(legend.position = "bottom")
-
-p3<-plot_usmap(data = t %>% 
-             filter(name == "police"),
-           values = "employment")+ 
-  scale_fill_viridis_c() + 
-  labs(title = "police per 1,000",
-       fill = "") + 
-  theme(legend.position = "bottom")
-
-grid.arrange(p1, p2, p3, ncol = 3)
+  labs(fill = "Social Workers per\n1,000 population")
 
@@
 
-Figure 4 shows these distributions as state choropleth maps for 1930 employment per 1,000.
-It clearly shows the geographic distribution of clusters of high and low employment states.
-Northeastern states like New York, New Jersey, and Massachusetts had very high levels of police and social work employment.
-Midwestern states like Minnesota, Michigan, and Illinois also had high social work and police employment, as did California.
-Midwestern and Plains states like North and South Dakota, Wyoming, and Iowa had especially high levels of teacher employment.
+FIG 2 - employment maps.
 
-Employment provides a measure of the magnitude of state and private investment in social control and social services (aside from transfers - think about distinction b/w hands on and redistribution).
-But this is one dimension of infrastructure.
-Given the ambiguity in 'social state capacity', we are working multiple measures to think this through.
+INST MAPS
 
-### capacity for therapeutic coercion: kids in institutional group quarters
+```{r}
+ggplot(pdat, aes(fill = inst_correctional / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  theme_void() + 
+  scale_fill_viridis_c()
 
-Next, let's turn to reach of social state policy systems.
-One way to get at that is through the existence and utilization of era-specific policy institutions.
-Here, we will look at the distribution of children recorded as being housed in 'institutional group quarters' on the 1900 - 1930 decennial censuses.
-To improve the quality of measurement, I use the 100% decennial data for 1900, 1920, 1930 and 1940.
-The 1910 census data available through IPUMS only includes detailed group quarters information on the 1% and 1.4% samples and after exploratory work, they do not appear to be continuous with the 100% data.
-I exclude 1910 data for now.
+ggplot(pdat, aes(fill = inst_mental / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  theme_void() + 
+  scale_fill_viridis_c()
 
-Group quarters indicate non-traditional housing situations (e.g.
-not in a house or apartment).
-Here, we subset this group to include only children recorded as residents of 'correctional institutions', 'mental institutions', or 'institutions for the elderly, handicapped, and poor'.
-This excludes children recorded as living in rooming houses, military housing, other non-institutional settings.
-Our goal here is to capture the magnitude of efforts by both public and private actors to coercively or therapeutically detain children for purposes of social control or social welfare.
-
-```{r gq-ts, fig.cap = "Children in institutional group quarters 1900 - 1930"}
-dat %>% 
-  filter(year!=1910,
-         !(is.na(region))) %>% 
-  mutate(across(child_inst_correctional:child_inst_poor_disabled, 
-                ~ .x/children * 1e3)) %>% 
-  pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
-               names_to = "type",
-               values_to = "pop") %>% 
-  ggplot(aes(x = year, y = pop)) + 
-  geom_line(aes(group = state),
-            alpha = 0.2) + 
-  geom_line(data = dat %>% ### add national line
-              filter(year!=1910,
-                     !(is.na(region))) %>%
-              group_by(year) %>% 
-              summarize(across(child_inst_correctional:child_inst_poor_disabled, 
-                               ~ sum(.x, na.rm=T)/sum(children, na.rm=T) * 1e3)) %>% 
-              pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
-                           names_to = "type",
-                           values_to = "pop"),            
-            aes(x = year, y = pop), linewidth = 2, lty = 2) + 
-  facet_wrap(~type, scales = "free") + 
-  labs(y = "institutional population per 1,000 children",
-       subtitle = "note varying y axes. National average dashed, state values solid")
-
+ggplot(pdat, aes(fill = inst_poor_disabled / tot_pop * 1e3)) + 
+  geom_sf(aes(geometry = geometry)) + 
+  facet_wrap(~year) + 
+  theme_void() + 
+  scale_fill_viridis_c()
 ```
 
-These plots strongly suggest that there are measurement artifacts present in the census data for these measures.
-Many states have broken time series with zeroes observed in 1920 between non-zero rates in 1900 and 1930, or large jumps between periods that defy expected linear growth patterns that revert in later periods.
-I think it makes sense to treat these with caution.
+THEN MODELS
 
-I'd like to use total group quarters, but it includes rooming houses, which are a huge proportion of the count. COULD JUST USE GQ == 3 (Institutions). That may be the ticket.
+DISPOSESSION MEASURES - YEAR OF LAST TREATY SIGNED, ACRES PATENTED UNDER MORILL AND HOMESTEAD
 
-```{r gq-box, fig.cap = "Children in institutional group quarters 1930"}
-dat %>% 
-  filter(year == 1930) %>% 
-  mutate(across(child_inst_correctional:child_inst_poor_disabled, 
-          ~ .x/children * 1e3)) %>% 
-  pivot_longer(cols = child_inst_correctional:child_inst_poor_disabled,
-               names_to = "type",
-               values_to = "pop") %>% 
-  ggplot(aes(x = type, y = pop, 
-             color = region)) + 
-  geom_boxplot()
-```
-
-## PRELIMINARY MODELS
-
-IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
-
-1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
-2) exclusion / formal soc control salience in urban non-S
-
-<!-- employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.  -->
-
-<!-- q1: salience of settler expropriotion -> state formation -->
-
-<!-- four models, each employment outcome + children in poverty institutions -->
-
-<!-- start with lm for log(rates) -->
-
-<!-- trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.  -->
-
-<!-- switch to brms for estimation on tricky ones -->
-
-<!-- ```{r} -->
-<!-- ### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE -->
-<!-- model_dat <- dat %>%  -->
-<!--   filter(year>=1880) %>%  -->
-<!--   mutate(police_rate = (police + 1) / tot_pop * 1e5, -->
-<!--          socwork_rate = (socwork + 1) / tot_pop * 1e5, -->
-<!--          teacher_rate = (teachers + 1) / tot_pop * 1e5, -->
-<!--          child_inst_rate = (child_inst_poor_disabled + 0.1) /  -->
-<!--            children * 1e5, -->
-<!--          prop_blk = (blk + 1) / tot_pop, -->
-<!--          prop_urban_immig = urban_immig_pop / tot_pop, -->
-<!--          hsclaim_prop = hs_claims / tot_pop, -->
-<!--          year = scale(year)) -->
-
-<!-- m_police<-brm(log(police_rate) ~  -->
-<!--                  scale(hsclaim_prop) + -->
-<!--                  scale(prop_urban_immig) + -->
-<!--                  scale(prop_blk) +  -->
-<!--                  scale(log(blk_sharecrop + 1)) +  -->
-<!--                  year +  -->
-<!--                  (1 + year|state), -->
-<!--                data = model_dat) -->
-
-<!-- m_teacher<-brm(log(teacher_rate) ~  -->
-<!--                  scale(hsclaim_prop) + -->
-<!--                  scale(prop_urban_immig) + -->
-<!--                  scale(prop_blk) +  -->
-<!--                  scale(log(blk_sharecrop + 1)) +  -->
-<!--                  year +  -->
-<!--                  (1 + year|state), -->
-<!--                data = model_dat) -->
-
-<!-- m_socwork<-brm(log(socwork_rate) ~  -->
-<!--                  scale(hsclaim_prop) + -->
-<!--                  scale(prop_urban_immig) + -->
-<!--                  scale(prop_blk) +  -->
-<!--                  scale(log(blk_sharecrop + 1)) +  -->
-<!--                  year +  -->
-<!--                  (1 + year|state), -->
-<!--                data = model_dat) -->
-
-<!-- sc_teacher<-lmer(log(teacher_rate) ~  -->
-<!--               scale(hsclaim_prop) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat, -->
-<!--             control=lmerControl(optimizer="nloptwrap", -->
-<!--                                 optCtrl=list(maxfun=2e5))) -->
-
-<!-- sc_socwork<-brm(log(socwork_rate) ~  -->
-<!--               scale(hsclaim_prop) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat) -->
-
-<!-- sc_gq<-brm(log(child_inst_rate) ~  -->
-<!--               scale(hsclaim_prop) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat) -->
-
-<!-- jc_police<-lmer(log(police_rate) ~  -->
-<!--               scale(prop_blk) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat, -->
-<!--             control=lmerControl(optimizer="nloptwrap", -->
-<!--                                 optCtrl=list(maxfun=2e5))) -->
-
-<!-- jc_teacher<-lmer(log(teacher_rate) ~  -->
-<!--               scale(prop_blk) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat, -->
-<!--             control=lmerControl(optimizer="nloptwrap", -->
-<!--                                 optCtrl=list(maxfun=2e5))) -->
-
-<!-- jc_socwork<-lmer(log(socwork_rate) ~  -->
-<!--               scale(prop_blk) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat, -->
-<!--             control=lmerControl(optimizer="nloptwrap", -->
-<!--                                 optCtrl=list(maxfun=2e5))) -->
-
-<!-- jc_gq<-lmer(log(child_inst_rate) ~  -->
-<!--               scale(prop_blk) + -->
-<!--               year +  -->
-<!--               (1 + year|state), -->
-<!--             data = model_dat, -->
-<!--             control=lmerControl(optimizer="nloptwrap", -->
-<!--                                 optCtrl=list(maxfun=2e5))) -->
-<!-- ``` -->
-
-<!-- This all looks fine. How do I want to visualize this?  -->
-
-<!-- we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930?
-Think about how to handle to intercepts / slopes.
-What would a zero mean?
-pulling out everything not directly in national trend and regime measures.
-is that what I'm after?
-I guess I really do just want the regime-specific trend lines, in which case E[rate | regime] should zero out the individual idioscyncracies.
-it's oversimplified, but that should be ok.
-Just be super clear on what the iterpretation is on prediction.
-it's not a hypothetical state, its the mean expectation for the slope of the regime var.
--->
-
-<!-- model is  -->
-
-<!-- \[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\] -->
-
-<!-- Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.  -->
-
-<!-- Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?  -->
-
-<!-- ```{r} -->
-<!-- new_dat<-data.frame(year =  c(1.3, 1.3, 1.3), -->
-<!--                     hsclaim_prop = c(0, 0.05, 0.15)) -->
-<!-- t<-predict(sc_police,  -->
-<!--            newdata = new_dat, -->
-<!--            re.form = ~0, -->
-<!--            type = "link") -->
-
-<!-- t1<-predict(sc_teacher,  -->
-<!--            newdata = new_dat, -->
-<!--            re.form = ~0, -->
-<!--            type = "link") -->
-
-<!-- exp(t1) -->
-
-<!-- ``` -->
+model 1: teachers ~ jim crow, urbanization, disposession (child control?)

### 2023-12-13 make_native_land_theft_state_data.r

@@ -3,7 +3,8 @@
 #### make state time series for native land disposession up to 1920
 #### make state time series for settler distribution up to 1920
 # - land grab u shapes for morill parcels https://github.com/HCN-Digital-Projects/landgrabu-data
-# - royce map shapes from forest service 'Indian Land Cessions in US' https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
+# - royce map shapes from forest service 
+# 'Indian Land Cessions in US' https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
 # - homestead data from http://homestead.unl.edu/projects/homesteading-the-plains/data/blm_homesteads.xlsx
 
 library(tidyverse)
@@ -64,7 +65,7 @@ cessions<-st_intersection(cessions, st)
 ### add in settler claimed territories as of 1796
 cessions2 <- cessions %>%  
   bind_rows(cessions_diff %>% filter(st_1796==T) %>% 
-              mutate(year = 1784))
+              mutate(year = 1795))
 
 ts<-cessions2 %>% 
   select(year, STUSPS) %>% 
@@ -97,7 +98,6 @@ ts_full %>%
   select(year, STUSPS,
          ceded_acres, cumulative_ceded,
          total_acres_2021, prop_ceded) %>% 
-  rename(state = STUSPS) %>% 
   write_csv("./data/temp_disposession_ts.csv")
 
 ##############################################################
@@ -112,10 +112,11 @@ morill <- morill %>%
   mutate(Yr_ST_Accept = as.numeric(Yr_ST_Accept),
          Yr_Uni_Assign = as.numeric(Yr_Uni_Assign),
          Yr_Patent = as.numeric(Yr_Patent))
-### rowwise minima of year measures
+### rowwise maxima of year measures
+### aiming to capture salience of transactions
 morill<-morill %>% 
   rowwise() %>% 
-  mutate(year = min(c(Yr_ST_Accept, 
+  mutate(year = max(c(Yr_ST_Accept, 
                       Yr_Uni_Assign, 
                       Yr_Patent),
                     na.rm = T)) 
@@ -166,32 +167,33 @@ state_xwalk <- data.frame(state = state.name,
 
 hs<-hs %>% 
   left_join(state_xwalk) %>% 
-  rename(homestead_acres = value)
+  rename(homestead_acres = value) %>% 
[…]

### 2023-12-13 maps.r

@@ -37,50 +37,9 @@ st<-states(cb = T) %>%
              "PR", "GU", "AS","VI",
              "MP"))
 
-aianh<-native_areas(cb = T) %>%
-  st_transform(8528)  
-
-### lower 48
-ggplot(st%>% 
-         filter(!STUSPS%in%
-                  c("HI", "AK", 
-                    "PR", "GU", "AS","VI",
-                    "MP"))) + 
-  geom_sf() + 
-  theme_void()
-
-### aianh
-ggplot(aianh) + 
-  geom_sf() + 
-  theme_void()
-
-### overlay
-ggplot(st) + 
-  geom_sf() + 
-  geom_sf(data = aianh,
-          fill = "dodgerblue") + 
-  theme_void()
-
-aianh_acres<-data.frame(name = aianh[[6]],
-                        aianh_acres_2021 = as.numeric(set_units(st_area(aianh), "acre")))
-
 st_acres<-data.frame(state = st$STUSPS,
                      total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
 
-### subtract out aianh areas from states to compute (state - aianh) / state
-t<-st_intersection(st, aianh)
-#ggplot(t) + geom_sf()
-# area intersection
-state_overlap_aian<-data.frame(state = t$STUSPS,
-                            area = st_area(t))
-
-state_overlap_aian<-state_overlap_aian %>% 
-  group_by(state) %>% 
-  summarize(acres_aianh = as.numeric(set_units(sum(area), "acre"))) %>% 
-  left_join(st_acres) %>% 
-  mutate(prop_aianh = acres_aianh / total_acres_2021)
-
-## (st_area(st) - st_area(t)) / st_area(st)
 ### royce map data
 ## available here: https://data-usfs.hub.arcgis.com/datasets/usfs::tribal-lands-ceded-to-the-united-states-feature-layer/about
 ## and archived here: https://data.nativeland.info/dataset/indian-land-cessions-in-the-united-states-1784-1894/resource/e2b874d5-3c39-4287-bf89-d4ef14b87c83
@@ -98,56 +57,10 @@ cessions<-cessions %>%
 cessions<-cessions %>% 
   st_transform(8528)
 
-ggplot(cessions,
-       aes(fill = year)) + 
-  geom_sf() + 
-  theme_void() 
-
[…]

### 2023-12-13 models.R

@@ -1,0 +1,129 @@
+library(lme4)
+library(brms)
+
+#DISPOSESSION MEASURES - YEAR OF LAST TREATY SIGNED, ACRES PATENTED UNDER MORILL AND HOMESTEAD
+#model 1: teachers ~ jim crow, urbanization, disposession (child control?)
+
+mdat<-pop %>% 
+  left_join(disp %>% 
+              group_by(STUSPS) %>% 
+              filter(ceded_acres!=0) %>% 
+              summarize(last_treaty = max(year, na.rm = T))) %>% 
+  left_join(dist)
+
+#### set weakly informative priors for model parameters
+### b ~ N(0, 20)
+### use brms defaults for variance params (PEG THEM DOWN)
+
+priors_m1<-c(
+  prior(normal(0, 20), class = b),
+  prior(normal(0, 20), class = Intercept),
+  prior(student_t(3, 0, 2.5), class = sd),
+  prior(gamma(0.01, 0.01), class = shape))
+
+
+# #### lme4 frequentist to check fit
+# teachers_m1_nb<-glmer.nb(
+#   teachers ~ 
+#     scale(I(blk_sharecrop / tot_pop)) +
+#     scale(I(urban_immig_pop / tot_pop)) +
+#     scale(morill_acres_cum + homestead_acres_cum) +
+#     log(children) + 
+#     factor(year) + 
+#     (1|STUSPS) + 
+#     offset(log(tot_pop)),
+#   data = mdat
+# )
+### fit HMC model with brms
+teachers_m1_bayes<-brm(
+  teachers ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  prior = priors_m1,
+  data = mdat
+)
+
+# cops_m1_nb<-glmer.nb(
+#   police ~ 
+#     scale(I(blk_sharecrop / tot_pop)) +
+#     scale(I(urban_immig_pop / tot_pop)) +
+#     scale(morill_acres_cum + homestead_acres_cum) +
+#     log(children) + 
+#     factor(year) + 
+#     (1|STUSPS) + 
+#     offset(log(tot_pop)),
[…]

### 2023-12-13 make_pop_data.R
update pop script

@@ -1,0 +1,73 @@
+## ----setup, include = F, message = F----------------------------------
+library(tidyverse)
+library(maps)
+
+## ----read IPUMS tables--------------------------------------------------
+# ### occupations 
+dat_occ<-read_csv("./data/usa_00071.csv") %>%
+  rename_all(tolower) %>%
+  mutate(occ1950 =
+           case_when(
+             occ1950 == 79 ~ "socwork",
+             occ1950 == 773 ~ "police",
+             occ1950 == 771 ~ "police",
+             occ1950 == 782 ~ "police",
+             occ1950 == 93 ~ "teachers"
+           )) %>%
+  group_by(year, statefip, occ1950) %>%
+  summarize(emp = sum(perwt)) %>%
+  pivot_wider(names_from = occ1950, values_from = emp,
+              values_fill = 0)
+
+# pop (total, age, race, nativitiy, urban, farm, landown)
+dat_immig<-read_csv("./data/usa_00072.csv") %>%
+  rename_all(tolower) %>%
+  group_by(year, statefip) %>%
+  summarize(tot_pop = sum(perwt),
+            children = sum((age<18) * perwt),
+            white_immig =
+              sum((nativity > 1) * (race == 1) * (hispan == 0) * perwt),
+            white_non_immig = 
+              sum((nativity == 1) * (race == 1) * (hispan == 0) * perwt),
+            blk =
+              sum((race == 2) * perwt),
+            blk_sharecrop = 
+              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
+            blk_farms = sum((race==2) * (farm==2) * perwt),
+            urban_pop = sum(hhwt * (urban ==2)),
+            urban_immig_pop = 
+              sum((urban==2) * 
+                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
+            tot_hh = sum(hhwt))
+### group quarters, 100%
+gq<-read_csv("./data/usa_00069.csv") %>%
+  rename_all(tolower) %>%
+  group_by(year, statefip, gqtype) %>%
+  summarize(gq_pop = sum(perwt, na.rm = T)) %>%
+  mutate(gqtype = case_when(
+    gqtype == 2 ~ "inst_correctional",
+    gqtype == 3 ~ "inst_mental",
+    gqtype == 4 ~ "inst_poor_disabled")) %>%
+  filter(gqtype!="other") %>%
+  pivot_wider(names_from = gqtype, values_from = gq_pop,
[…]

### 2023-12-13 make_pop_data.R
update dat_immig with 100% census

@@ -20,7 +20,7 @@ dat_occ<-read_csv("./data/usa_00071.csv") %>%
               values_fill = 0)
 
 # pop (total, age, race, nativitiy, urban, farm, landown)
-dat_immig<-read_csv("./data/usa_00072.csv") %>%
+dat_immig<-read_csv("./data/usa_00076.csv") %>%
   rename_all(tolower) %>%
   group_by(year, statefip) %>%
   summarize(tot_pop = sum(perwt),

### 2024-01-04 models_appx.Rmd
ff

---
title: "Untitled"
author: "Frank Edwards"
date: "`r Sys.Date()`"
output: html_document
---

## PRELIMINARY MODELS

IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?

1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
2) exclusion / formal soc control salience in urban non-S

employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.

q1: salience of settler expropriotion -> state formation

four models, each employment outcome + children in poverty institutions

start with lm for log(rates)

trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.

switch to brms for estimation on tricky ones

```{r}
### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE
model_dat <- dat %>%
  filter(year>=1880) %>%
  mutate(police_rate = (police + 1) / tot_pop * 1e5,
         socwork_rate = (socwork + 1) / tot_pop * 1e5,
         teacher_rate = (teachers + 1) / tot_pop * 1e5,
         child_inst_rate = (child_inst_poor_disabled + 0.1) /
           children * 1e5,
         prop_blk = (blk + 1) / tot_pop,
         prop_urban_immig = urban_immig_pop / tot_pop,
         hsclaim_prop = hs_claims / tot_pop,
         year = scale(year))

m_police<-brm(log(police_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

m_teacher<-brm(log(teacher_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

m_socwork<-brm(log(socwork_rate) ~
                 scale(hsclaim_prop) +
                 scale(prop_urban_immig) +
                 scale(prop_blk) +
                 scale(log(blk_sharecrop + 1)) +
                 year +
                 (1 + year|state),
               data = model_dat)

sc_teacher<-lmer(log(teacher_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

sc_socwork<-brm(log(socwork_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat)

sc_gq<-brm(log(child_inst_rate) ~
              scale(hsclaim_prop) +
              year +
              (1 + year|state),
            data = model_dat)

jc_police<-lmer(log(police_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_teacher<-lmer(log(teacher_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_socwork<-lmer(log(socwork_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))

jc_gq<-lmer(log(child_inst_rate) ~
              scale(prop_blk) +
              year +
              (1 + year|state),
            data = model_dat,
            control=lmerControl(optimizer="nloptwrap",
                                optCtrl=list(maxfun=2e5)))
```

This all looks fine. How do I want to visualize this?

we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930? Think about how to handle to intercepts / slopes. What would a zero mean? pulling out everything not directly in national trend and regime measures. is that what I'm after? I guess I really do just want the regime-specific trend lines, in which case E[rate | regime] should zero out the individual idioscyncracies. it's oversimplified, but that should be ok. Just be super clear on what the iterpretation is on prediction. it's not a hypothetical state, its the mean expectation for the slope of the regime var.

model is

\[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\]

Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.

Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?

```{r}
new_dat<-data.frame(year =  c(1.3, 1.3, 1.3),
                    hsclaim_prop = c(0, 0.05, 0.15))
t<-predict(sc_police,
           newdata = new_dat,
           re.form = ~0,
           type = "link")

t1<-predict(sc_teacher,
           newdata = new_dat,
           re.form = ~0,
           type = "link")

exp(t1)

```

### 2024-01-04 ms.Rmd
ff

@@
 ---
-title: "Settler colonial projects and variation in paths to building the welfare and carceral state"
+title: "Settler colonial projects explain variation in US state formation in the early 20th c."
 author: "Frank Edwards"
@@
 output: bookdown::html_document2
-abstract: "Place and time variation in settler colonial projects are related to the kinds of states that settler elites build."
 ---
@@
 library(sf)
+library(tidybayes)
+library(rstan)
 
@@
 theme_set(theme_minimal())
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
 ```
 
-```{r read, include = F, message = F, cache = T}
-
+```{r read, include = F, message = F}
 ### state shape file
@@
   filter(STUSPS!="AK")
+bs<-read_csv("./data/temp_bs_ts.csv")
+agents<-read_csv("./data/temp_agents.csv")
 
@@
               select(STUSPS, region, division) %>%
-  distinct())
+  distinct()) %>% 
+  left_join(bs) %>% 
+  left_join(agents)
 
@@
               select(STUSPS, region, division) %>%
-  distinct())
+  distinct()) %>% 
+  left_join(bs) %>% 
+  left_join(agents)
+```
 
-source("models.R")
+```{r run-models, cache = T, include = F, message = F}
+#source("models.R")
 ```
@@
 
-This paper proposes a comparative quantitative framework for understanding how the geographic process of settler colonization results in different styles of governance across places (simultaneous settler projects).
-We show that spatial and temporal variation in modes of settler colonial extraction and exploitation are closely related to trajectories of state formation across places.
+Settler colonialism deeply and unevenly impacts the development of social policy systems.
+Settler colonial regimes engage in diverse forms of dispossession, extraction, exploitation, and social reproduction.
+We show that spatial and temporal variation in the salience of settler projects is closely related to the kinds of infrastructure that settler colonial states develop for social control and social reproduction.
+We show that the formation of the social state in the early 20th century in the US varied dramatically across places as a function of geographic variation in the salience of a set of interests of the settler colonial project: land dispossession, labor exploitation, and reproduction of the settler polity.
 
-At a sub-national level, settler colonial regimes seek to address a range of problems inherent to their nation-building project: the dispossession of indigenous lands, the erasure of indigenous peoples, the exploitation of racially marginalized laborers, and the reproduction and expansion of the settler polity.
-We show that the historical geography of these problems of settler colonial governance were closely related to the policy infrastructure to manage social life that governments develop.
+We examine three deeply intertwined, but spatially and temporally variable, contexts for local policymaking in the Progressive Era (1900 - 1930): the Jim Crow agricultural economy; mass urbanization and industrialization; and the rapid distribution of Native lands to white settlers.
+Each of these contexts presented different sets of problems to local and national settler policy elites.
 
-<!-- Historians tell us these projects were active locally. 1. Frontier; 2. Jim Crow; 3. Metropole. Can we use demographic, legal, and employment data to map it out / model it? -->
+Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests.
+Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital.
+In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers.
+White settlers and settler corporate interests flooded recently disposessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.
 
-Following Wolfe, DuBois, etc, variation in the interests of elites across places (material + cultural) drives variation in processes of racialization across places.
-The state is a key vector through which racialization is implemented.
-Early 20th c.
-saw rapid development and expansion of social state.
-Can we observe differences in social state developmental trajectories as a function of variation in elite interests in particular racial projects?
-Are regional differences in social policy regimes a function of the social, demographic, and economic interests of regional elites?
+We suggest that each of these contexts yielded different trajectories for the formation of US state infrastructure.
+In the analysis that follows, we leverage data from a range of archival sources to map contexts for early 20th century settler colonization across the US, then build statistical models to evaluate whether these contexts were related to the kinds of capacities that local governments developed for contemporaneous social control and social reproduction.
 
-<!-- Stratification in frontier was settler social democracy; flat social relations among settlers, extreme hostility to Native.
-Stratification in Jim Crow was rigid racial + class hierarchy; extreme wealth concentration and rigid racialized exploitation + social exclusion.
-Stratification in metropole was more fluid racial + class hierarchy; extreme concentration but porous boundaries, fewer bright lines.
--->
+Consistent with the historigraphy on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure.
+State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects.
+In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding shcools.
+In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies.
+Social control and reproduction instead relied on a coordinated formal and informal system of racial terror that developed alongside a relatively weak set of state social policy institutions.
 
-<!-- Basic theory though is that interests of elites are racialized as a function of local social relations, and inform social policy development.
-This can lock in particular racial regimes over time as social policy does work on social groups.
--->
+The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US.
+States developed more or less robust institutions in various domains to address the problems of governance posed by the local projects of settler colonization that both elites and ordinary settlers sought to advance.
 
-<!-- 1. Frontier settler polity -> elite interest in disposession / assimiatilion -> aggressive paternalistic welfare state for AIAN, social democracy for settlers -->
-<!-- 2. Jim Crow -> elite interest in exploitation and exclusion -> elite paternalism, malign neglect, hard coercion -->
-<!-- 3.
-Metropole settler polity -> elite interests in labor discipline / immig assimilation -> progressive paternalism, strong coercive welfare state for W, hard coercion and contested welfare for Blk -->
+We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity.
+We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation.
+Or, put more directly, that THE LOCAL SALIENCE OF PARTICULAR CONSITUENTS / STRUCTURES IS RELATED TO STATE FORMATION.
 
-<!-- IMPLICATIONS - a) settler colonization as structure / structuring: context is key, and when SC is active, likely can be predictive; b) durable consequences for policy institutions (legacies, etc); c) comparative state formation / comparative welfare / carceral states.
--->
+These are three of the projects that were critical to settler colonialism in the early 20th c US.
+Not the only ones of course!
+Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILLIPINES COLONIZATION, AND OTHERS).
+*Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*
 
-<!-- # What is already known about settler colonization and state formation in late 19th and Early 20th c.  -->
+# Three settler colonial projects in the early 20th c. US
 
-<!-- ## Frontier -->
+The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps).
+The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast.
+While Native land disposession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries.
+In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.
 
-<!-- review here -->
+The Jim Crow system of apartheid was developed and becoming deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people.
+As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.
 
-<!-- ## Jim Crow -->
+And the Industrial Revolution, coupled with a wave of crises across Eastern and Southern Europe, created both strong push and pull forces for large populations of immigrants to travel to North America.
+These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts.
+Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US.
+As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becomingly increasingly urban and foreign born.
 
-<!-- review here -->
+Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'.
+During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels.
+Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.
 
-<!-- ## Metropole  -->
-
-<!-- review here -->
-
-# Methods
-
-*We evaluate whether a set of policy outcomes {employment, infrastructure} are related to a set of demographic and economic predictors {population composition, economic activity, ownership regime} at the state and year-levels.*
-
-To do so, we rely on historical data from the US Census accessed via IPUMS (Ruggles et al), from the Bureau of Land Management (Archived by the University of Nebraska), from the US Department of the Interior (CITE BS REPORT), and from records our team digitized from the Commissioner on Indian Affairs (CITE).
-
-<!-- Less interested in causal identification, more interested in understanding spatial patterns and trajectories.
-We already know there was spatial variation in settler colonial contexts and policy systems, trying to see if we can capture it systematically with national quant data.
-Move from specific to general.
--->
-
-Generally our approach demonstrates suggests that state formation in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.
-
-\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]
-
-<!-- negative case re: black underdevelopment h1: no social workers in places with large black pops at period of emergence in t1 -->
-
-<!-- positive case: white immigrants -> more social workers in t1 -->
-
-<!-- here we are looking at how contexts in post-civil war -> progressive era US related to the emergence of the social state in the us.  -->
-
-<!-- ## NOTES ON GEOGRAPHIES and MEASUREMENT -->
-
-<!-- TALK ABOUT STATES, TERRITORIES, WESTERN EXPANSION.
-CONTESTATION AND TRANSFORMATION ARE KEY PROCESSES OF COLONIZATION.
-SETTLER COLONIALISM IS A SPATIAL STRUCTURE, NOT A SETTLED FACT, ESP VISIBLE IN THIS PERIOD.
--->
-
-<!-- STATES ARE WEIRD UNITS, BUT MAP CLEANLY ONTO ADMIN / LEGAL UNITS. TRIBAL NATIONS BOUNDARIES ARE ALSO V USEFUL, AS ARE SUB-STATE ANALYSES (EG CITIES COUNTIES ETC).  -->
-
-<!-- geography itself endogenous to settler colonization, this is tricky! -->
-
-## Sources of data
-
-Census (pop, employment, group quarters); ARCIA; Interior; BLM; US Forest Service.
-
-<!-- CONSIDER GOING 1880 - 1920 to capture period more effectively? no social workers will be in there tho... -->
-
-<!-- we treat police, education, and social work employment as core areas of the domestic social state.
-what is the name for that?
-the welfare state?
-the carceral state?
-the therapeutic state?
-that thing.
--->
-
-<!-- let's look at how it played out in the progressive era across states and regions.
-we are cutting the data at 1900 - 1930, wanting to capture the boom in progressive reforms that started in the late 1890s and continued until the great depression.
--->
-
-<!-- First, the full country. How many people were employed in these social regulation / reproduction functions over time per capita? -->
-
-We use IPUMS 100% census data for 1900 - 1930 to count employment levels nationally and across states.
-A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 Marshals and constables, 773 Policemen and detectives, 782 Sheriffs and bailiffs}; coded social worker if OCC1950 is coded 079 Social and welfare worker; and coded teacher if OCC1950 is equal to 093 Teachers (n.e.c.).
-We convert these into employment rates per 1,000 population using the full population at the state (and nation) level.
-
-### measures
-
-All measures are computed at the state-year level.
-
-EMPLOYMENT:
-
-- teachers (1900 - 1930 100%)
-- cops (1900 - 1930 100%)
-- social workers (1900 - 1930 100%)
-
-POPULATION:
-
-- total (1900 - 1930 1%)
-- children (1900 - 1930 1%)
-- white immigrants in *urban households* (1900 - 1930 1%)
-- Black population living in tenant farm households (1900 - 1930 1%)
-- Population in group quarters (correctional institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
-- Population in group quarters (mental institutions) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
-- Population in group quarters (institutions for the elderly, handicapped, and poor) (1900 100%, 1910 1.4%, 1920 - 1930 100%)
-
-LAND:
-
-- cumulative acres parceled through the Morill Act (Robert Lee, ``Morrill Act of 1862 Indigenous Land Parcels Database,'' High Country News, 2020)
-- cumulative acres parceled through the Homestead Act (``Homesteads,'' Bureau of Land Management of the U.S. Department of the Interior, 1962. digitized by Edwards, Friefeld and and Wingo 2017)
-
-## statistical methods and models
-
-Model is
-
-\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]
-
-\[ \lambda = \textrm{E}[y_{st}] \]
-
-\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \delta_t + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} \]
-
-\[\gamma_s \sim N(0, \sigma^2_\gamma)\]
-
-With weakly informative priors
-
-\[\beta \sim N(0, 20)\]
-\[\sigma \sim t(3, 0, 2.5)\]
-\[\alpha \sim \textrm{Gamma}(0.01, 0.01)\]
-
-# Findings
-
-We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states.
-In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity.
-
-The works of settler colonial and racial politics historical and theoretical scholars suggest a straightforward approach to selecting measures that are either a) direct, or b) proxy measures for elite white settler interests given a few assumptions about relationships between population distributions and political processes.
-
-```{r fig.cap = "demographic map of US in 1900 - Black, White immig, white non-immig"}
-plot_dat<-pop %>% 
-  filter(year == 1900) %>% 
-  mutate(white_non_immig = white_non_immig / (white_non_immig + white_immig),
-         white_immig = 1 - white_non_immig,
-         black = blk / tot_pop) %>% 
-  select(STUSPS, white_immig, white_non_immig, black) %>% 
-  pivot_longer(cols = white_immig:black) %>% 
-  mutate(name = case_when(
-    name == "black" ~ "Black / \ntotal population",
-    name == "white_immig" ~ "White immigrant/\ntotal white population",
-    name == "white_non_immig" ~ "White non-immigrant/\ntotal white population"
-  ))
-
-st %>% 
-  left_join(plot_dat) %>% 
-  filter(!is.na(value)) %>% 
-  ggplot(aes(fill = value)) + 
+```{r three-maps, fig.cap = "Geographic and demographic contexts in 1910 for settler colonial governance"}
+### ADD DISPOSESSION
+p1 <- dist %>% 
+  select(STUSPS, year,
+         morill_acres_cum, homestead_acres_cum,
+         total_acres_2021) %>% 
+  filter(year == 1910) %>% 
+  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
+  left_join(st) %>% 
+  ggplot(aes(fill = dist / 1e6,
+             geometry = geometry)) + 
   geom_sf() + 
-  facet_wrap(~name, ncol = 3) + 
   theme_void() + 
@@
   theme(legend.position = "bottom") + 
-  labs(fill = "")
+  labs(fill = "Acres\n(millions)",
+       subtitle = "1. Lands patented through\nMorill and Homestead acts")
+
+p2 <- st %>% 
+  left_join(pop %>% 
+              filter(year == 1910)) %>% 
+  ggplot(aes(fill = blk / tot_pop)) + 
+  geom_sf() + 
+  theme_void() + 
+  scale_fill_viridis_c() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "Proportion",
+       subtitle = "2. Black population")
+
+p3 <- st %>% 
+  left_join(pop %>% 
+              filter(year == 1910)) %>% 
+  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
+  geom_sf() + 
+  theme_void() + 
+  scale_fill_viridis_c() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "Proportion",
+       subtitle = "3. Urban immigrant population")
+
+grid.arrange(p1, p2, p3, nrow = 1)
 ```
 
-This map shows key features of the demography of the US in 1900 that are closely related to the geography of settler colonialism in the early 20th century US.
+## Settling the frontier
 
-BLACK POPS IN JIM CROW SOUTH
+OVERVIEW: BUILDING THE SETTLER NATION STATE, ERASING NATIVE NATIONS
 
-US BORN WHITE POPS IN SOUTH, NEW ENGLAND, RURAL PARTS OF INNER MIDWEST
+Dispossession
 
-FOREIGN BORN WHITE POPS THROUGHOUT CITIES, WESTERN FRONTIER
+Redistribution
 
-FIG 2, 3 - native disposession
+THE MORILL, HOMESTEAD AND RR ACTS
 
-```{r}
+Nation building
+
+Forced assimilation
+
+```{r fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
 st<-states(cb = T) %>%
@@
   scale_fill_viridis_c() + 
-  labs(fill = "Year",
-       title = "Post-1784 Indigenous disposession",
-       subtitle = "Cessions to US federal government through 1894") 
+  labs(fill = "Year") 
 ```
@@
   facet_wrap(region~division) + 
-  labs(subtitle = "1. Total land cessions")
+  labs(subtitle = "1. Total area")
 
@@
 
-```{r fig.cap = "Two mechanisms for the distribution of disposessed Native Lands to settlers"}
+```{r fig.cap = "Lands patented under two legislative mechanisms for the distribution of disposessed Native Lands to settlers, cuulative total 1930"}
 pdat<-disp %>% 
@@
   left_join(dist %>% 
-              filter(year == 1900) %>% 
+              filter(year == 1930) %>% 
               select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
@@
   labs(fill = "Acres (millions)",
-       subtitle = "1. Lands patented under Morill Act (1862) by 1900") + 
+       subtitle = "1. Morill Act") + 
   theme(legend.position = "bottom")
@@
   labs(fill = "Acres (millions)",
-       subtitle = "2. Lands patented under Homestead Act (1862) by 1900") + 
+       subtitle = "2. Homestead Act") + 
   theme(legend.position = "bottom")
@@
 
-FIGURE - by employment type, facet ~ year
+## Entrenching the exploitation of Black labor and marginalization of Black social and political power
 
-```{r}
-pdat<-pop %>% 
-  left_join(st)
-# could use children in denominator here
-# keeping as total population for comparison 
-ggplot(pdat,
-       aes(fill = teachers / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
+Between 1900 and 1930, settler elites in the the US South were busy entrenching Jim Crow and profiting from the deeply exploitative sharecropping system of agricultural production.
+
+Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period.
+In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agrictultural production.
+In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.
+
+### The sharecropping economy
+
+DEFINE SYSTEM
+
+SUMMARIZE LITERATURE
+
+### The white southern polity
+
+SUMMARIZE LIT ON SOUTHERN WHITENESS AND COHERENCE OF WHITE SETTLER POLITY IN US SOUTH
+
+### The early and mid-career of Jim Crow
+
+Racial order, law, and terrorism
+
+ESP THINK ON BUREAUCRACY, INFORMALITY OF SOCIAL CONTROL, COLLABORATION BETWEEN FORMAL AND INFORMAL, VIGILANTISM AND LAW
+
+EDUCATION AND JIM CROW
+
+SOCIAL WORK AND JIM CROW
+
+```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
+pdat<-st %>% 
+  left_join(pop) %>% 
+  filter(year == 1910)
+
+p1<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = blk_sharecrop / 1e5)) + 
   scale_fill_viridis_c() + 
   theme_void() + 
-  labs(fill = "Teachers per\n1,000 population")
+  labs(fill = "Population\n(hundred thousands)",
+       subtitle = "1. Black population living on tenant farms") + 
+  theme(legend.position = "bottom")
 
-ggplot(pdat,
-       aes(fill = police / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
+p2<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = white_immig / (white_immig + white_non_immig))) + 
   scale_fill_viridis_c() + 
   theme_void() + 
-  labs(fill = "Police per\n1,000 population")
+  labs(fill = "Proportion",
+        subtitle = "2. Proportion of white poulation foreign born") + 
+  theme(legend.position = "bottom")
 
-ggplot(pdat,
-       aes(fill = socwork / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Social Workers per\n1,000 population")
+grid.arrange(p1, p2, nrow = 1)
 
@@
 
-FIG 2 - employment maps.
+## Incorporating urban white immigrants into the settler polity and industrial working class
 
-INST MAPS
+The industrialization of the US Northeast and Midwest during the late 19th and early 20th centuries spurred two related demographic processes: rapid urbanization and mass European immigration.
+Figure \@ref(fig:immig-maps) displays the distribution of both the urban population and the white immigrant population in the US in 1910.
+States like New York, Massachusetts, Pennsylvania, Ohio, and Illinois saw dramatic growth in their populations, with much of that growth composed of European immigrants and concentrated in cities like Chicago, Boston, Cleveland, Philadelphia, and New York.
+
+```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
+pdat<-st %>% 
+  left_join(pop) %>% 
+  filter(year == 1910)
+
+p1<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = white_immig / 1e6)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Population\n(millions)",
+       subtitle = "1. White, foreign-born") + 
+  theme(legend.position = "bottom")
+
+p2<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = urban_pop / 1e6)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Population\n(millions)",
+        subtitle = "2. Urban") + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, nrow = 1)
+```
+
+SUMMARIZE DEMOGRAPHY LIT
+
+### The rise of progressive domestic governance in Northern US cities
+
+social work
+
+policing
+
+## Patterns in the development of state infrastructure 1900 - 1930
+
+state capacity is hard to measure, but lets try!
+Here we are going to use employment measures for three roles that came to play critical roles in the developing practices of social welfare and social control: teachers, police, and social workers.
+
+We also think about physical infrastructure for similar sets of institutions.
+We look at various forms of institutional group quarters: correctional institutions, mental institutions, and institutions for the poor and disabled.
+
+Indian Boarding schools are another key piece of physical infrastructure. We look at those too, but I am still waiting on validating some of that data.
+
+### Employment in education, policing, and social welfare: 1900 - 1930
+
+There was dramatic growth in public sector employment in general (CHECK THIS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era.
+*These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (CITES).*
+
+In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930.
+Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase.
+Employment of police doubled between during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930.
+Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900).
+However, by 1930, the Census recorded about 17,000 social workers nationally, and increase of over 40 fold in terms of employment rates per capita.
+
+```{r emp-ts, fig.cap = "Employment in three core social policy professions over time, US rates per 1,000 population, 1900 - 1930"}
+emp_pdat<-pop %>% 
+  group_by(year) %>% 
+  summarize(pop = sum(tot_pop),
+            Teachers = sum(teachers),
+            Police = sum(police),
+            social_workers = sum(socwork)) %>% 
+  pivot_longer(cols = Teachers:social_workers) %>% 
+  mutate(name = case_when(
+    name == "Teachers"~ "1. Teachers",
+    name == "Police" ~ "2. Police",
+    name == "social_workers" ~ "3. Social Workers")) %>% 
+  mutate(rate = value / pop * 1e3)
+
+emp_pdat  %>% 
+  ggplot(aes(x = year, y = rate)) + 
+  geom_line() + 
+  facet_wrap(~name, scales = "free_y") + 
+  labs(x = "Year", y = "Employment rate")
+
+```
+
+However, this expansion of the reach of the state did not occur uniformly across the country.
+Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States.
+Employment rates for teachers were lowest-in-the nation in the US South.
+Employment rates for teachers were highest in a cluster of Great Plains states, like the Dakotas, Wyoming, Nebraska, Iowa, and Kansas.
+Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California.
+Police employment rates were relatively lower in places like Arkansas, Missisippi, and North Carolina.
+The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Missisippi and South Dakota recorded very few social workers in the 1930 Census.
+
+```{r emp-maps, fig.cap = "Employment in three core social policy professions across US states in 1930, rates per 1,000 population"}
+pdat<-st %>% 
+  left_join(pop) %>% 
+  filter(year == 1930)
+
+p1<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = teachers / tot_pop * 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Rate",
+       subtitle = "1. Teachers") + 
+  theme(legend.position = "bottom")
+
+p2<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = police / tot_pop * 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Rate",
+        subtitle = "2. Police") + 
+  theme(legend.position = "bottom")
+
+p3<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = socwork / tot_pop * 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Rate",
+        subtitle = "3. Social Workers") + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, p3, nrow = 1)
+```
+
+### Institutional capacity: prisons, jails, asylums, orphanages and poorhouses
+
+State capacity could also be measured through the ability of states to engage in therapeutic, correctional, or paternalistic confinement.
+In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states.
+Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3.
+Using data from the Department of the Interior's ``Federal Indian Boarding School Initiative'' (2022), we display a count of Federal Indian boarding schools operating in 1930 in panel 4.
+I MAY ADD ENROLLMENT DATA FROM ARCIA REPORTS ONCE WE VALIDATE IT.
+
+Data on institutional group quarters populations are less reliable than employment data for constructing a full 1900 - 1930 Census time series.
+
+shows Prisons in NE, S, CA. Mental and poverty instutitons mostly in Urban NE, MW, CA
+
+```{r gq-map, fig.cap = "Institutional group quarters, US 1930"}
+pdat<-st %>% 
+  left_join(pop) %>% 
+  left_join(bs) %>% 
+  filter(year == 1930)
+
+p1<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = inst_correctional / 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Population\n(thousands)",
+       subtitle = "1. Correctional institutions") + 
+  theme(legend.position = "bottom")
+
+p2<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = inst_mental / 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Population\n(thousands)",
+        subtitle = "2. Mental institutions") + 
+  theme(legend.position = "bottom")
+
+p3<-ggplot(pdat) + 
+  geom_sf(aes(geometry = geometry,
+              fill = inst_poor_disabled / 1e3)) + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Population\n(thousands)",
+        subtitle = "3. Institutions for the\npoor and disabled") + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, p3, nrow = 2)
+```
+
+### Institutional capacity: Indian boarding schools
+
+FROM DOI REPORTS
+
+Figure \@ref{fig:bsts} displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference.
+The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905.
+After 1905, a slow decline in the number of operating schools began continued until the end of our focal period, 1930.
+At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850).
+During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173.
+The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as had existed at the beginning of the period.
+
+```{r bsts, fig.cap = "Indian boarding schools operating annually, US 1830 - 1930"}
+cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+t<-bs %>% 
+         filter(year<=1930,
+                year>=1830) %>% 
+         group_by(year) %>% 
+         summarize(boarding_schools = sum(boarding_schools))
+
+ggplot(t) +
+       aes(y = boarding_schools,
+           x = year) + 
+  geom_line() + 
+  annotate("rect",  # add removal era shading
+           ymin = 0, 
+           ymax = 250,
+           xmin = 1830,
+           xmax = 1850,
+           color = NA,
+           fill = cbPalette[1],
+           alpha = 0.2) + 
+  annotate("rect",  # add reservation era shading
+           ymin = 0, 
+           ymax = 250,
+           xmin = 1850,
+           xmax = 1887,
+           color = NA,
+           fill = cbPalette[2],
+           alpha = 0.2) + 
+    annotate("rect",  # add reservation era shading
+           ymin = 0, 
+           ymax = 250,
+           xmin = 1887,
+           xmax = 1930,
+           color = NA,
+           fill = cbPalette[3],
+           alpha = 0.2) + 
+  annotate("text",
+           x = c(1840, 1860, 1910),
+           y = c(60, 90, 160),
+           color = cbPalette[1:3],
+           label = c("Removal era", "Reservation era", "Assimilation era")) +
+  labs(x = "Year", y = "Number of facilities") + 
+  scale_x_continuous(breaks = seq(1830, 1930, by = 20),
+                     limits = c(1830, 1930))
+
+```
+
+In Figure \@ref{fig:bsts2} we zoom in on our focal period. This uses two sources of data, p1 is DoI, p2 is ARCIA. EXPLAIN THAT THESE ARE LIKELY UNDERCOUNTS AGAIN.
+
+Expansion of program via enrollment even as N facilities is declining and capacity is declining. surprise surprise.
+
+```{r bsts2, fig.cap = "Boarding schools and recorded student enrollment and capacity, US totals 1890 - 1920"}
+## map out the enrollment and capacity data
+bs_ts<-agents %>% 
+  left_join(bs) %>% 
+  left_join(st)
+
+t<-bs_ts %>% 
+  group_by(year) %>% 
+  summarize(Capacity = sum(boarding_n_acc),
+            Enrollment = sum(students_enroll)) %>% 
+  pivot_longer(cols = Capacity:Enrollment)
+
+### for in-text callouts
+p1<-ggplot(t, 
+           aes(x = year, y = value, color = name)) + 
+  geom_line() + 
+  labs(x = "Year", y = "Children",
+       color = "",
+       subtitle = "2. Reported capacity and enrollment (ARCIA)")
+
+p2<-ggplot(bs_ts %>% 
+  group_by(year) %>% 
+  summarize(boarding_schools = sum(boarding_schools)),
+           aes(x = year, y = boarding_schools)) + 
+  geom_line() + 
+  labs(x = "Year", y = "Count",
+       subtitle = "1. Facilities operating annually (Dept. of Interior)")
+
+grid.arrange(p2, p1, nrow = 1) 
+```
+
+Figure \@ref{fig:bsmap} shows the geographic distribution of boarding school infrastructure and utilization when enrollments were at a 1890 - 1920 period maximum in 1917. DISCUSS
+
+```{r bsmap, fig.cap = "Boarding schools in operation and reported enrollment and capacity at peak enrollment (1917)"}
+p3<-ggplot(bs_ts %>% 
+         filter(year == 1917),
+       aes(fill = students_enroll / 1e3,
+           geometry = geometry)) + 
+  geom_sf() + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Children\n(thousands)",
+       subtitle = "3. Reported enrollment") + 
+  theme(legend.position = "bottom")
+
+p2<-ggplot(bs_ts %>% 
+         filter(year == 1917),
+       aes(fill = boarding_n_acc / 1e3,
+           geometry = geometry)) + 
+  geom_sf() + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Children\n(thousands)",
+       subtitle = "3. Reported capacity") + 
+  theme(legend.position = "bottom")
+
+p1<-ggplot(bs_ts %>% 
+         filter(year == 1917),
+       aes(fill = boarding_schools,
+           geometry = geometry)) + 
+  geom_sf() + 
+  scale_fill_viridis_c() + 
+  theme_void() + 
+  labs(fill = "Facilities",
+       subtitle = "1. Boarding schools operating") + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, p3, nrow = 1)
+```
+
+# Analytic strategy
+
+The works of settler colonial and racial politics historical and theoretical scholars suggest a straightforward approach to selecting measures that are either a) direct, or b) proxy measures for elite white settler interests given a few assumptions about relationships between population distributions and political processes.
+
+To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of policy outcomes are related to a set of demographic and economic predictors at the state and year-levels.
+We conduct exploratory and inferential analysis using data on US states between 1900 and 1930.
+
+DESCRIBE THE MEASURES IN GENERAL TERMS HERE
+
+Generally our approach suggests that state formation in a particular place *s* and time *t* is a function of settler interests specific to that time and place.
+
+\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]
+
+We proceed by producing and evaluating descriptive data visuals for our operational measures of 'settler interests' and comparing distributions of these measures across states and over time.
+We then produce and evaluate a similar set of visuals to evaluate the distribution of our 'infrastructure' measures.
+Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national temporal trends, and basic characteristics of the population in the state and year.
+
+## Data
+
+All measures on features of the US population are derived from the 1900, 1910, 1920, and 1930 decennial US Census 100% samples, accessed through IPUMS (Ruggles et al.).
+For each census year and for each of the lower 48 US states, we count the total population, the child population (as the population under 18 years old), the white urban immigrant population (as the population identified as foreign born, white, and non-Hispanic living in an urban household), and the Black population living in tenant farm households.
+
+State-year employment measures are computed from the 1900 - 1930 Census.
+A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 'Marshals and constables', 773 'Policemen and detectives', 782 'Sheriffs and bailiffs'}; is counted as a social worker if OCC1950 is coded 079 'Social and welfare worker'; and counted as a teacher if OCC1950 is equal to 093 'Teachers (n.e.c)', which excludes music and art teachers but includes other primary and secondary teachers.
+
+We also compute measures of populations in institutional group quarters, but rely on a slightly different set of samples.
+The 1910 census did not collect information on group quarters type.
+We rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type.
+From these samples, we derive state-year measures of the population housed in correctional instiuttions, the population housed in mental institutions, and the population housed in institutions for the elderly, handicapped, and poor.
+
+Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress (CITE).
+The distribution of Native lands to settlers is measured through data on the implementation of the Homestead Act and Morill Act of 1862.
+We use data on parcels of land transferred to land-grant universities under the Morill Act as documented by Robert Lee an High Country News as part of their 'Land Grab University' project.
+These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available.
+Data on parcels of land distributed to settlers through the Homestead Act are obtained from  the report ``Homesteads,'' by the Bureau of Land Management of the U.S.
+Department of the Interior, 1962, and digitized and distributed by Edwards, Friefeld and and Wingo 2017.
+For both Homestead and Morill act data, we measure cumulative acres parceled since 1862 under each act within the state by year.
+
+## A note on settler colonial geography
+
+geography itself endogenous to settler colonization, this is tricky!
+we are trying to understand the emergence of the state as bureaucratic entity in this period, so it makes sense to do this, but it is messy!
+we don't want to imply a teleological reading here.
+borders are unstable and contested.
+
+## statistical methods and models
+
+Nearly of our 'infrastructure' outcomes are measured as counts of persons.
+As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes.
+For facility counts (boarding schools), we employ a similar model that removes population offsets from the specification.
+
+DISCUSSIONS OF MODEL SELECTION IN APPX
+
+Model 1. is
+
+\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]
+
+\[ \lambda = \textrm{E}[y_{st}] \]
+
+\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} + \beta_4 \log(\textrm{child population}_{st}) \]
+
+\[\gamma_s \sim N(0, \sigma^2_\gamma)\]
+
+With weakly informative priors
+
+\[\beta \sim N(0, 20)\]
+\[\sigma \sim t(3, 0, 2.5)\]
+\[\alpha \sim \textrm{Gamma}(0.01, 0.01)\]
+
+Model 2 (for zero inflated outcomes) is
+
+XXXXX
+
+# Findings
+
+We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states.
+In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity.
+
+## Employment models
 
 ```{r}
-ggplot(pdat, aes(fill = inst_correctional / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
-  theme_void() + 
-  scale_fill_viridis_c()
-
-ggplot(pdat, aes(fill = inst_mental / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
-  theme_void() + 
-  scale_fill_viridis_c()
-
-ggplot(pdat, aes(fill = inst_poor_disabled / tot_pop * 1e3)) + 
-  geom_sf(aes(geometry = geometry)) + 
-  facet_wrap(~year) + 
-  theme_void() + 
-  scale_fill_viridis_c()
+# # ### extract posterior draws for b1, b2, b3, set up for plotting
+# # pdat<-teachers_m1_bayes %>% 
+# #   gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
+# #                  b_scaleIurban_immig_popDtot_pop,             
+# #                  b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
+# #   mutate(model = "1. Teachers") %>% 
+# #   bind_rows(
+# #     cops_m1_bayes %>% 
+# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
+# #                      b_scaleIurban_immig_popDtot_pop,             
+# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
+# #       mutate(model = "2. Police")) %>% 
+# #   bind_rows(
+# #     socwork_m1_bayes %>% 
+# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
+# #                      b_scaleIurban_immig_popDtot_pop,             
+# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>%
+# #       mutate(model = "3. Social workers")) %>% 
+# #   mutate(.variable = case_when(
+# #     .variable == "b_scaleIblk_sharecropDtot_pop" ~ 
+# #       "β1: Black tenant\nfarmers per capita",
+# #     .variable == "b_scaleIurban_immig_popDtot_pop" ~ 
+# #       "β2: Urban white\nimmigrants\nper capita",
+# #     .variable == "b_scalemorill_acres_cumPhomestead_acres_cum" ~ 
+# #       "β3: Cumulative \nMorill +\nHomestead acres"))
+# 
+# 
+# ggplot(pdat,
+#        aes(x = .value, 
+#            y = .variable)) +
+#   stat_halfeye() + 
+#   geom_vline(aes(xintercept = 0), lty = 2) + 
+#   labs(x = "Estimate",
+#        y = "") + 
+#   facet_wrap(~model) + 
+#   theme_bw()
 ```
 
-THEN MODELS
+## Group quarters models
 
-DISPOSESSION MEASURES - YEAR OF LAST TREATY SIGNED, ACRES PATENTED UNDER MORILL AND HOMESTEAD
-
-model 1: teachers ~ jim crow, urbanization, disposession (child control?)
+RAN BOARDING SCHOOL MODEL WITH DISCONTINUITY ON ACREAGE, IT WORKS WELL.

### 2024-01-04 make_pop_data.R

@@ -42,13 +42,13 @@ dat_immig<-read_csv("./data/usa_00076.csv") %>%
 ### group quarters, 100%
 gq<-read_csv("./data/usa_00069.csv") %>%
   rename_all(tolower) %>%
+  filter(year!=1910) %>% 
   group_by(year, statefip, gqtype) %>%
-  summarize(gq_pop = sum(perwt, na.rm = T)) %>%
+  summarize(gq_pop = sum(perwt)) %>%
   mutate(gqtype = case_when(
     gqtype == 2 ~ "inst_correctional",
     gqtype == 3 ~ "inst_mental",
-    gqtype == 4 ~ "inst_poor_disabled")) %>%
-  filter(gqtype!="other") %>%
+    gqtype == 4 ~ "inst_poor_disabled")) %>% 
   pivot_wider(names_from = gqtype, values_from = gq_pop,
               values_fill = 0)
 ### join, format for output

### 2024-01-04 model_fit_diagnostics.R

@@ -1,23 +1,16 @@
-## sourced by ms.Rmd
+## Diagnostics for model fit
+# focus on zero inflated processes
+# and parsimony / convergence trade offs
+# using default brms priors for ease
 
 mdat<-pop %>% 
   left_join(dist) %>% 
   left_join(bs)
 
-#### set weakly informative priors for model parameters
-### b ~ N(0, 20)
-### use brms defaults for variance params (PEG THEM DOWN)
 
-priors_m1<-c(
-  prior(normal(0, 10), class = b),
-  prior(normal(0, 10), class = Intercept),
-  prior(student_t(3, 0, 2.5), class = sd),
-  prior(gamma(0.01, 0.01), class = shape))
-
-
-
-### fit HMC model with brms
-teachers_m1_bayes<-brm(
+###### teachers - easiest outcome given power
+### baseline model as reported in eq (1) in ms
+t_m1<-brm(
   teachers ~ 
     scale(I(blk_sharecrop / tot_pop)) +
     scale(I(urban_immig_pop / tot_pop)) +
@@ -27,19 +20,31 @@ teachers_m1_bayes<-brm(
     (1|STUSPS) + 
     offset(log(tot_pop)),
   family = negbinomial(),
-  prior = priors_m1,
   data = mdat
 ) 
 
+### add regime zero inflation / threshold adjustments
+t_m2<-brm(
+  teachers ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    I(blk / tot_pop > 0.3) +  
+    scale(I(urban_immig_pop / tot_pop)) +
+    I(urban_pop>2e5) + 
+    scale(morill_acres_cum + homestead_acres_cum) +
+    I((morill_acres_cum + homestead_acres_cum)>0) + 
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+) 
 
-### prior visual
-# teachers_m1_bayes %>% 
-#   prior_summary() %>% 
-#   parse_dist(prior) %>% 
-#   ggplot(aes(y = class, dist = .dist, args = .args)) + 
-#   stat_dist_halfeye()
+loo_compare(loo(t_m1), loo(t_m2))
 
-cops_m1_bayes<-brm(
+#### cops
+
+p_m1<-brm(
   police ~ 
[…]

### 2024-01-04 read_agents.R

@@ -6,11 +6,12 @@ library(tidyverse)
 # eval all addition statements
 # replace NA with 0 when appropriate
 # code through 1930
+# check zeroes / fluctuations on time series with ts plot below
+# 1910 looks like a gap
+# some weird single years in AZ, MT, MN
 
 ### load in agents data
 agents<-read_csv("./data/AGENTS_DATA_COLLECTION1890_1920.csv") 
-
-
 
 agents<-agents %>% 
   mutate(across(BOARDING_N_ACC:N_MISSIONARIES, as.numeric)) %>% 
@@ -45,11 +46,36 @@ year<-1890:1920
 state<-unique(state.name)
 
 xwalk<-expand_grid(year, state)
+xwalk<-xwalk %>% 
+  left_join(data.frame(state = state.name, STUSPS = state.abb))
 agents<-xwalk %>% 
   left_join(agents) %>% 
-  mutate(across(boarding_n_acc:n_missionaries, ~replace_na(.x, 0)))
+  mutate(across(boarding_n_acc:n_missionaries, ~replace_na(.x, 0))) %>% 
+  filter(year != 1910)
 
-agents %>% 
-  ggplot(aes(x = year, y = boarding_n_acc)) + 
-  geom_line() + 
-  facet_wrap(~state)
+#### VALIDATION PLOTS
+# ## ts validation plot
+# agents %>%
+#   ggplot(aes(x = year, y = boarding_n_acc)) +
+#   geom_line() +
+#   facet_wrap(~STUSPS)
+# 
+# ## 1920 validation map
+# agents %>% 
+#   filter(year == 1920) %>% 
+#   left_join(st) %>% 
+#   ggplot(aes(fill = boarding_n_acc, geometry = geometry)) + 
+#   geom_sf() + 
+#   theme_void()
+# 
+# ### scatterplot with bs data: looks pretty good. some obs where boarding_schools = 0, n_acc !=0
+# ### should check them
+# source("read_bs.R")
+# agents %>% 
+#   left_join(bs_out) %>% 
+#   ggplot(aes(x = boarding_schools,
+#              y = students_enroll)) + 
+#   geom_point()
+
+write_csv(agents, "./data/temp_agents.csv")
+

### 2024-01-04 read_bs.R

@@ -1,0 +1,53 @@
+library(tidyverse)
+
+bs<-read_csv("./data/Boarding_Schools.csv")
+
+### create row for each boarding-school-year
+### using start end-pair
+
+### make period of interest
+year_prog <- data.frame(year = 1900:1930)
+
+dat_out<-list()
+
+for(i in 1:nrow(bs)){
+  
+  temp<-bs[i,]
+  
+  temp<-temp %>% 
+    mutate(`End Date` = ifelse(is.na(`End Date`),
+                               2022,
+                               `End Date`)) %>% 
+    select(Name, State, `Start Date`, `End Date`) %>% 
+    distinct()
+  
+  ### top code still operating schools
+
+  
+  years<-seq(temp$`Start Date`, temp$`End Date`)
+  
+  out<-data.frame(year = years,
+                  bs_name = temp$Name,
+                  state = temp$State)
+  
+  dat_out[[i]]<-out
+  }
+
+bs_ts<-bind_rows(dat_out)
+
+bs_state_ts<-bs_ts %>% 
+  group_by(state, year) %>% 
+  summarize(n = n()) 
+
+### make full 50 state ts for full period
+state<-state.abb
+year<-seq(min(bs_state_ts$year), max(bs_state_ts$year))
+xwalk<-expand_grid(state, year) 
+
+bs_out<-xwalk %>% 
+  left_join(bs_state_ts) %>% 
+  mutate(boarding_schools = ifelse(is.na(n), 0, n)) %>% 
+  select(-n) %>% 
+  rename(STUSPS = state)
+
+write_csv(bs_out, "./data/temp_bs_ts.csv")

### 2024-03-06 models_appx.Rmd
random updates that I don't remember

@@
 
-## PRELIMINARY MODELS
+explore model fit under alternate specifications. look at bivars and functional relationships b/w outcomes and focal predictors
 
-IMPORTANCE OF URBAN? ADD EXTRA PLOT HERE? BLK / URBAN?
-
-1) jim crow salience (can we define as brutal coupling of formal / informal? less state capacity built for maintenance / strategic underinvestment)
-2) exclusion / formal soc control salience in urban non-S
-
-employment trajectories look kinda log-linear. almost certainly non-linear, so need a curve on trajectories regardless. log probably solves it well enough.
-
-q1: salience of settler expropriotion -> state formation
-
-four models, each employment outcome + children in poverty institutions
-
-start with lm for log(rates)
-
-trouble fitting models for sparse data (e.g. social workers, children in gq), but works fine and in expected directions for teachers, police.
-
-switch to brms for estimation on tricky ones
+describe prior selection
 
 ```{r}
-### FIT PROBLEM WAS WITH EARLY YEARS NEAR ZERO AND LOW VARIANCE
-model_dat <- dat %>%
-  filter(year>=1880) %>%
-  mutate(police_rate = (police + 1) / tot_pop * 1e5,
-         socwork_rate = (socwork + 1) / tot_pop * 1e5,
-         teacher_rate = (teachers + 1) / tot_pop * 1e5,
-         child_inst_rate = (child_inst_poor_disabled + 0.1) /
-           children * 1e5,
-         prop_blk = (blk + 1) / tot_pop,
-         prop_urban_immig = urban_immig_pop / tot_pop,
-         hsclaim_prop = hs_claims / tot_pop,
-         year = scale(year))
+## Diagnostics for model fit
+# focus on zero inflated processes
+# and parsimony / convergence trade offs
+# using default brms priors for ease
+# log(children) not improving teacher fit. leaving it out
 
-m_police<-brm(log(police_rate) ~
-                 scale(hsclaim_prop) +
-                 scale(prop_urban_immig) +
-                 scale(prop_blk) +
-                 scale(log(blk_sharecrop + 1)) +
-                 year +
-                 (1 + year|state),
-               data = model_dat)
+mdat<-pop %>% 
+  left_join(dist) %>% 
+  left_join(bs)
 
-m_teacher<-brm(log(teacher_rate) ~
-                 scale(hsclaim_prop) +
-                 scale(prop_urban_immig) +
-                 scale(prop_blk) +
-                 scale(log(blk_sharecrop + 1)) +
-                 year +
-                 (1 + year|state),
-               data = model_dat)
+###### teachers - easiest outcome given power
+### baseline model as reported in eq (1) in ms
+t_m1<-brm(
+  teachers ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+)
 
-m_socwork<-brm(log(socwork_rate) ~
-                 scale(hsclaim_prop) +
-                 scale(prop_urban_immig) +
-                 scale(prop_blk) +
-                 scale(log(blk_sharecrop + 1)) +
-                 year +
-                 (1 + year|state),
-               data = model_dat)
+### prior visual
+# teachers_m1 %>% 
+#   prior_summary() %>% 
+#   parse_dist(prior) %>% 
+#   ggplot(aes(y = class, dist = .dist, args = .args)) + 
+#   stat_dist_halfeye()
 
-sc_teacher<-lmer(log(teacher_rate) ~
-              scale(hsclaim_prop) +
-              year +
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+### add regime zero inflation / threshold adjustments
+t_m2<-brm(
+  teachers ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    I(blk / tot_pop > 0.3) +  
+    scale(I(urban_immig_pop / tot_pop)) +
+    I(urban_pop>2e5) + 
+    scale(morill_acres_cum + homestead_acres_cum) +
+    I((morill_acres_cum + homestead_acres_cum)>0) + 
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+)
 
-sc_socwork<-brm(log(socwork_rate) ~
-              scale(hsclaim_prop) +
-              year +
-              (1 + year|state),
-            data = model_dat)
+t_m3<-brm(
+  teachers ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+)
 
-sc_gq<-brm(log(child_inst_rate) ~
-              scale(hsclaim_prop) +
-              year +
-              (1 + year|state),
-            data = model_dat)
+loo_compare(loo(t_m1), loo(t_m2))
 
-jc_police<-lmer(log(police_rate) ~
-              scale(prop_blk) +
-              year +
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+#### cops
 
-jc_teacher<-lmer(log(teacher_rate) ~
-              scale(prop_blk) +
-              year +
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+p_m1<-brm(
+  police ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+)
 
-jc_socwork<-lmer(log(socwork_rate) ~
-              scale(prop_blk) +
-              year +
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
+### add regime zero inflation / threshold adjustments
+p_m2<-brm(
+  police ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    I(blk / tot_pop > 0.3) +  
+    scale(I(urban_immig_pop / tot_pop)) +
+    I(urban_pop>2e5) + 
+    scale(morill_acres_cum + homestead_acres_cum) +
+    I((morill_acres_cum + homestead_acres_cum)>0) + 
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat
+)
 
-jc_gq<-lmer(log(child_inst_rate) ~
-              scale(prop_blk) +
-              year +
-              (1 + year|state),
-            data = model_dat,
-            control=lmerControl(optimizer="nloptwrap",
-                                optCtrl=list(maxfun=2e5)))
-```
+loo_compare(loo(p_m1), loo(p_m2))
 
-This all looks fine. How do I want to visualize this?
+#### loo shows no meaningful difference bw m1, m2 for both teachers and cops
 
-we'll eventually do trajectory visuals for each regime type - what is the expected trajectory under different regimes, where do we end up in 1930?
-Think about how to handle to intercepts / slopes.
-What would a zero mean?
-pulling out everything not directly in national trend and regime measures.
-is that what I'm after?
-I guess I really do just want the regime-specific trend lines, in which case E[rate | regime] should zero out the individual idioscyncracies.
-it's oversimplified, but that should be ok.
-Just be super clear on what the iterpretation is on prediction.
-it's not a hypothetical state, its the mean expectation for the slope of the regime var.
+###### GQ MODELS
 
-model is
+mdat <- mdat %>% 
+  mutate(inst_gq = inst_mental + 
+           inst_poor_disabled)
 
-\[\hat{Y} = \alpha + \beta \textrm{Regime} + \delta year + \epsilon\]
+prison_m1<-brm(
+  floor(inst_correctional) ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat %>% 
+    filter(year!=1910))
 
-Here we are very much interested just in $\beta$, the rest are nuissance parameters. that works. can vis e[state type] under diffferent REGIME sets.
+prison_m2<-brm(
+  floor(inst_correctional) ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    I(blk / tot_pop > 0.3) +  
+    scale(I(urban_immig_pop / tot_pop)) +
+    I(urban_pop>2e5) + 
+    scale(morill_acres_cum + homestead_acres_cum) +
+    I((morill_acres_cum + homestead_acres_cum)>0) + 
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat %>% 
+    filter(year!=1910))
 
-Q: are there places with competing regime sets? CA, WA, others? where we need a multivariate regime model?
+loo_compare(loo(prison_m1), loo(prison_m2))
 
-```{r}
-new_dat<-data.frame(year =  c(1.3, 1.3, 1.3),
-                    hsclaim_prop = c(0, 0.05, 0.15))
-t<-predict(sc_police,
-           newdata = new_dat,
-           re.form = ~0,
-           type = "link")
+### no meaningful difference in models according to loo
+### NJ 1900 shows NA
+### looks like failed joins. investigate!
+### ballparking numbers for prisoners against https://bjs.ojp.gov/content/pub/pdf/p2581.pdf
+### not too bad! 1925 n is 91,689; 1930 is 129,453. I report 215682 for 1930, 79073 for 1920 
+# year gqtype  gq_pop
+# 1  1880      2  49272 
+# 2  1900      2  71491 
+# 3  1910      2 135323.
+# 4  1920      2  79073 
+# 5  1930      2 215682 
+### 1910 is discontinuous - that's the sample year, not full census. check docs again
+# From IPUMS
+# User Note: The Census Bureau did not collect information on the type of group quarters in the 1910 census. 
+# This is reflected in the 1910 complete count file, where all group-quarters units have been assigned a GQ code of 4 (other group quarters). 
+# Values for GQ, GQTYPE, and GQFUNDS in the 1910 1% and 1.4% samples have been imputed.
 
-t1<-predict(sc_teacher,
-           newdata = new_dat,
-           re.form = ~0,
-           type = "link")
+welfgq_m1<-brm(
+  floor(inst_gq) ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    scale(I(urban_immig_pop / tot_pop)) +
+    scale(morill_acres_cum + homestead_acres_cum) +
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat %>% 
+    filter(year!=1910))
 
-exp(t1)
+welfgq_m2<-brm(
+  floor(inst_gq) ~ 
+    scale(I(blk_sharecrop / tot_pop)) +
+    I(blk / tot_pop > 0.3) +  
+    scale(I(urban_immig_pop / tot_pop)) +
+    I(urban_pop>2e5) + 
+    scale(morill_acres_cum + homestead_acres_cum) +
+    I((morill_acres_cum + homestead_acres_cum)>0) + 
+    log(children) + 
+    factor(year) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  data = mdat %>% 
+    filter(year!=1910))
+
+loo_compare(loo(welfgq_m1), loo(welfgq_m2))
+
+# boarding schools models use 1890 - 1920 time period
+# models need to be specified differntly because of census mismatch
+# may need to increase samples to up ESS
+# but for a simple descriptive, this looks fine IMO
+
+bs_mdat<-dist %>% 
+  left_join(bs) %>% 
+  left_join(agents %>% 
+              filter(year!=1910)) %>% # 1910 is incomplete data
+  filter(year>=1900, year<=1930) # 1900 - 1920 for enrollment / attendance, 1900 - 1930 for school count
+#### maybe run this zero inflated,
+
+######## BIVARIATES AND DIAGNOSTICS
+ggplot(bs_mdat %>% 
+         filter(boarding_schools>0),
+       aes(x = log(homestead_acres_cum + morill_acres_cum),
+           y = boarding_schools,
+           color = STUSPS)) + 
+  geom_point() + 
+  facet_wrap(~division)
+
+### stability in boarding_schools - increase in acres distributed, not much to model other than average
+### only action is decrease in time series at end, so that's a bit funky
+
+ggplot(bs_mdat %>% 
+         filter(boarding_schools>0,
+                morill_acres_cum>0),
+       aes(x = log(homestead_acres_cum + morill_acres_cum), 
+           y = boarding_n_acc,
+           color = STUSPS)) + 
+  geom_point() + 
+  facet_wrap(~division)
+
+### similar look here. some increases in bs pops, e.g. in SD, AZ, OK, but mostly stable over the period
+### some clear increases and maybe log-linear things going on, but don't want to overfit.
+
+ggplot(bs_mdat,
+       aes(x = boarding_schools,
+           fill = homestead_acres_cum>0))+ 
+  geom_density(alpha = 0.2)
+
+#### OK - zero inflation is way to go, but there's no 1900 - 1930 variation in the binary
+### boarding_schools>0 within states, so panel model makes no sense for binary outcome
+### subset to start of period condition, then conditional on nonzero in 1900, look at count
+
+#### LEANING TOWARD THE SCHOOL COUNT ONLY APPROACH. THE CAPACITY / ENROLLMENT NUMBERS NEED MORE WORK I THINK
+### START HERE TOMORROW. FOLD THE REVISED NO LOG(KIDS) MODEL IN, FOLD THE ZINFL MODEL IN
+
+bs_m1_z<-brm(boarding_schools>0 ~ 
+               scale(I(blk_sharecrop / tot_pop)) +
+               scale(I(urban_immig_pop / tot_pop)) +
+               scale(morill_acres_cum + homestead_acres_cum),
+             family = bernoulli(link = "logit"),
+             data = mdat %>% 
+               filter(year == 1900))
+
+bs_m1<-brm(boarding_schools ~ 
+             scale(I(blk_sharecrop / tot_pop)) +
+             scale(I(urban_immig_pop / tot_pop)) +
+             scale(morill_acres_cum + homestead_acres_cum) +
+             factor(year) +
+             (1|STUSPS),
+           family = negbinomial(),
+           data = mdat %>% 
+             filter(boarding_schools>0))
+
+### zero infl model for capacity
+
+cap_m1_z<-brm(boarding_n_acc>0 ~ 
+            scale(I(blk_sharecrop / tot_pop)) +
+            scale(I(urban_immig_pop / tot_pop)) +
+            scale(morill_acres_cum + homestead_acres_cum) ,
+            # factor(year) + 
+            # (1|STUSPS),
+          family = bernoulli(link = "logit"),
+          data = mdat %>% 
+            filter(year == 1900))
+
+cap_m1<-brm(boarding_n_acc ~ 
+            scale(I(blk_sharecrop / tot_pop)) +
+            scale(I(urban_immig_pop / tot_pop)) +
+            scale(morill_acres_cum + homestead_acres_cum) +
+            factor(year) +
+            (1|STUSPS),
+          family = negbinomial(),
+          data = mdat %>% 
+            filter(boarding_n_acc>0))
+
+#### model vis
+# bivar - looks like it's just a binary relationship, the linear component isn't adding much here
+ggplot(bs_mdat %>% 
+         filter(boarding_n_acc >0),
+       aes(x = morill_acres_cum + homestead_acres_cum,
+           y = boarding_n_acc,
+           color = STUSPS)) + 
+  geom_point()
+
+ggplot(bs_mdat %>% 
+         filter(boarding_n_acc >0),
+       aes(x = morill_acres_cum + homestead_acres_cum,
+           y = boarding_schools,
+           color = STUSPS)) + 
+  geom_point()
 

### 2024-03-06 ms.Rmd
random updates that I don't remember

@@
 title: "Settler colonial projects explain variation in US state formation in the early 20th c."
-author: "Frank Edwards"
+author: "Frank Edwards and Theresa Rocha Beardall"
 date: "last updated `r Sys.Date()`"
-output: bookdown::html_document2
+output: bookdown::pdf_document2
 ---
@@
 library(rstan)
-
+library(knitr)
+library(trackdown)
+# trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
+# trackdown::update_file(file = "ms.Rmd", 
+#                        hide_code = T,
+#                        rich_text = T)
 options(tigris_use_cache = T)
@@
 options(mc.cores = parallel::detectCores())
+# set hook for thousands comma
+knit_hooks$set(inline = function(x) {
+prettyNum(round(x,2), big.mark=",")
+})
 ```
@@
              "MP"))
-
 # read data from make_* scripts
@@
 agents<-read_csv("./data/temp_agents.csv")
-
 ## add region to all
@@
   left_join(bs) %>% 
-  left_join(agents)
-
+  left_join(agents %>% 
+              select(-state))
 dist<-dist %>%
@@
   left_join(bs) %>% 
-  left_join(agents)
+  left_join(agents %>% 
+              select(-state))
 ```
@@
 ```{r run-models, cache = T, include = F, message = F}
-#source("models.R")
+source("models.R")
 ```
@@
 In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers.
-White settlers and settler corporate interests flooded recently disposessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.
+White settlers and settler corporate interests flooded recently dispossessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.
 
@@
 
-Consistent with the historigraphy on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure.
-State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects.
-In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding shcools.
-In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies.
-Social control and reproduction instead relied on a coordinated formal and informal system of racial terror that developed alongside a relatively weak set of state social policy institutions.
-
 The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US.
@@
 We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation.
-Or, put more directly, that THE LOCAL SALIENCE OF PARTICULAR CONSITUENTS / STRUCTURES IS RELATED TO STATE FORMATION.
-
-These are three of the projects that were critical to settler colonialism in the early 20th c US.
-Not the only ones of course!
-Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILLIPINES COLONIZATION, AND OTHERS).
-*Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*
+In more general terms, we argue that the local salience and configuration of key social structures, like settler colonization, are closely related to local trajectories for state formation and institution building.
+
+Consistent with the historiography on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure.
+State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects.
+Our work bridges these sometimes disparate historical accounts by showing that the local interests of settler elites were closely tied to the kinds of social and racial projects that the early carceral and welfare states engaged in.
 
@@
 The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast.
-While Native land disposession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries.
+While Native land dispossession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries.
 In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.
 
-The Jim Crow system of apartheid was developed and becoming deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people.
-As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.
-
-And the Industrial Revolution, coupled with a wave of crises across Eastern and Southern Europe, created both strong push and pull forces for large populations of immigrants to travel to North America.
-These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts.
-Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US.
-As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becomingly increasingly urban and foreign born.
-
-Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'.
-During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels.
-Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.
-
-```{r three-maps, fig.cap = "Geographic and demographic contexts in 1910 for settler colonial governance"}
+```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1910"}
 ### ADD DISPOSESSION
@@
        subtitle = "1. Lands patented through\nMorill and Homestead acts")
-
 p2 <- st %>% 
@@
        subtitle = "2. Black population")
-
 p3 <- st %>% 
@@
        subtitle = "3. Urban immigrant population")
-
 grid.arrange(p1, p2, p3, nrow = 1)
@@
 
-## Settling the frontier
-
-OVERVIEW: BUILDING THE SETTLER NATION STATE, ERASING NATIVE NATIONS
-
-Dispossession
-
-Redistribution
-
-THE MORILL, HOMESTEAD AND RR ACTS
-
-Nation building
-
-Forced assimilation
-
-```{r fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
+The Jim Crow system of apartheid was developed and became deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people.
+As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.
+
+And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America.
+These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts.
+Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US.
+As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becoming increasingly urban and foreign born.
+
+Below, we briefly summarize these three social, economic, and demographic contexts for US state formation in the early 20th century, a period routinely referred to as the 'Progressive Era'.
+During this period, rapid increases in the complexity, capacity, and scope of domestic governance were occurring at the federal, state, and local levels.
+Following this brief discussion of historical contexts, we turn to a discussion and analysis of how these varying contexts for settler colonization related to heterogeneity in the kinds of governments that settlers built across places.
+
+## Building the settler nation-state
+
+KEY CHARARCTERISTICS: RECENT OR ONGOING DISPOSESSION / REMOVAL, RECENTLY ADMITTED STATES, RELATIVELY SPARSE WHITE POPS. ELIMINATION / REPLACEMENT AS KEY DYNAMIC.
+
+In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding schools.
+
+```{r cessions-map, fig.cap = "Land cessions from Native nations claimed by US government 1795 - 1893"}
 st<-states(cb = T) %>%
@@
              "MP"))
-
 st_acres<-data.frame(state = st$STUSPS,
                      total_acres_2021 = as.numeric(set_units(st_area(st), "acre")))
-
 ### royce map data
@@
 cessions<-read_sf("./data/us_forest_cessions_royce/S_USA.TRIBALCEDEDLANDS.shp")
-
 cessions_dat<-read_csv("./data/us_forest_cessions_royce/royce.csv") %>% 
@@
   select(CESSNUM, year)
-
 cessions<-cessions %>% 
   left_join(cessions_dat)
-
 cessions<-cessions %>% 
   st_transform(8528)
-
 cessions_diff<-st_difference(st, st_union(cessions))
-
 cessions_diff<-cessions_diff %>% 
   mutate(year = NA)
-
 ### layer state boundaries
@@
               mutate(year = NA))
-
 ggplot(cessions2,
@@
   labs(fill = "Year")
-```
-
-MANY OF THE LOCATIONS THESE FIRST GENERATION SETTLERS WERE GOING TO WERE RECENT DISPOSESSIONS, PARTICULARLY CIVIL WAR ERA.
-IN 1900, MANY OF THESE LANDS HAD BEEN CLAIMED BY US IN VERY RECENT PAST, WITHIN 50 YEARS.
-SECOND / THIRD WAVE OF WESTERN SETTLERS.
-BUILDING THE SETTLER COLONIAL POLITY / NATION STATE AS KEY AGENDAS.
-
-```{r fig.cap = "Disposession of Native lands in the 19th c."}
-p1<-ggplot(disp,
-       aes(x = year,
-           group = STUSPS)) + 
-  geom_line(aes(y = cumulative_ceded / 1e6)) + 
-  labs(x = "Year",
-       y = "Acres (millions)") + 
-  lims(y = c(0, 150)) + 
-  facet_wrap(region~division) + 
-  labs(subtitle = "1. Total area")
-
-p2<-ggplot(disp,
+
+```
+
+Figures \@ref(fig:cessions-map) and \@ref(fig:cessions-ts) show the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong).
+These maps show a clear difference in the timing of land cessions as a function of place.
+By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee.
+Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century.
+Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West.
+By 1895, nearly all lands in the contemporary lower 48 US states had been claimed by the federal government as US territory.
+
+```{r cessions-ts, fig.cap = "Cumulative disposession of Native lands in the 19th c. by US region and state as proportion of contemporary state area. Each line represents one state."}
+# p1<-ggplot(disp,
+#        aes(x = year,
+#            group = STUSPS)) + 
+#   geom_line(aes(y = cumulative_ceded / 1e6)) + 
+#   labs(x = "Year",
+#        y = "Acres (millions)") + 
+#   lims(y = c(0, 150)) + 
+#   facet_wrap(region~division) + 
+#   labs(subtitle = "1. Total area")
+p2<-ggplot(disp %>% 
+             filter(year > 1795),
        aes(x = year,
@@
   labs(x = "Year",
-       y = "Proportion") + 
-  facet_wrap(region~division) + 
-  labs(subtitle = "2. Proportion of state area")
-
-grid.arrange(p1, p2, ncol = 2)
-```
-
-This shows the timing and spatial distribution in cumulative land disposession by state and region.
-Panel 1 shows the lands claimed by the US in millions of acres by state from 1796 - 1894.
-Panel 2 shows these patterns of Native land loss as a proportion of the contemporary areas of US states.
-
-Clear regional patterns here.
-Land loss occurred earlier in states closer to the Atlantic coast.
-Dispossession accelerated in the North Central and Western regions throughout the middle portion of the 19th c.
-
-```{r fig.cap = "Lands patented under two legislative mechanisms for the distribution of disposessed Native Lands to settlers, cuulative total 1930"}
+       y = "Proportion of state area") + 
+  facet_wrap(region~division) 
+p2
+```
+
+The Federal government passed a series of laws in 1862 designed to rapidly redistribute dispossessed Native lands to white settlers and to large corporate and state institutions.
+The Homestead, Morill, and Pacific Railway Acts of 1862 provided legal frameworks through which expropriated Native lands could be rapidly redistributed to white settlers, large corporations, and Universities.
+Figure \@ref(fig:hs-morill-map) shows the total number of acres within a state that had been granted to settlers and settler institutions by 1930.
+More than 10 million acres of Native lands had been redistributed to US land grant Universities at little or no cost for their use or liquidation by 1930 under the Morill act.
+Under the Homestead Act, more than 255 million acres of Native land had been transferred to settlers by 1930.
+States in the Midwest, Great Plains, Mountain West, and West Coast saw the highest volumes of land redistribution under these laws.
+
+```{r hs-morill-map, fig.cap = "Lands patented under two legislative mechanisms for the redistribution of disposessed Native lands to settlers, cumulative total 1862 - 1930"}
 pdat<-disp %>% 
@@
   left_join(st)
-
 p1<-ggplot(pdat) + 
@@
   theme(legend.position = "bottom")
-
 grid.arrange(p1, p2, ncol = 2)
@@
 
-## Entrenching the exploitation of Black labor and marginalization of Black social and political power
-
-Between 1900 and 1930, settler elites in the the US South were busy entrenching Jim Crow and profiting from the deeply exploitative sharecropping system of agricultural production.
+## Jim Crow
+
+KEY CHARACTERISTICS: ANTI-BLACK SOCIAL, LEGAL, ECONOMIC REGIME. WHITE SETTLER HOMOGENEITY
+
+By 1877 efforts to build effective racial democracy through post-Civil War Reconstruction were abandoned by the federal government.
+By the turn of the century, white elites had legislated and entrenched a regime of brutal racial exploitation and apartheid across the US South (DU BOIS, FONER, FRANKLIN, CURTIN).
+
+DEFINE JIM CROW REGIME. APARTHEID, FORMAL / INFORMAL TERROR REGIME, BLACK LABOR EXPLOITATION, PRIMARILY AG through SHARECROPPING.
+
+READ APD CITATIONS FROM ANN REVIEWS ON JIM CROW, APD, RACISM. DOES APD SUPPORT THE WEAK STATE THESIS?
+
+In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies.
+Social control and reproduction instead relied on a coordinated formal and informal system of racial terrorism and apartheid that developed alongside a relatively weak set of state social policy institutions (QUADAGNO, KATZNELSON).
 
 Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period.
-In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agrictultural production.
+In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agricultural production.
 In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.
-
-### The sharecropping economy
-
-DEFINE SYSTEM
-
-SUMMARIZE LITERATURE
-
-### The white southern polity
-
-SUMMARIZE LIT ON SOUTHERN WHITENESS AND COHERENCE OF WHITE SETTLER POLITY IN US SOUTH
-
-### The early and mid-career of Jim Crow
-
-Racial order, law, and terrorism
-
-ESP THINK ON BUREAUCRACY, INFORMALITY OF SOCIAL CONTROL, COLLABORATION BETWEEN FORMAL AND INFORMAL, VIGILANTISM AND LAW
-
-EDUCATION AND JIM CROW
-
-SOCIAL WORK AND JIM CROW
 
@@
   filter(year == 1910)
-
 p1<-ggplot(pdat) + 
@@
   theme(legend.position = "bottom")
-
 p2<-ggplot(pdat) + 
@@
   labs(fill = "Proportion",
-        subtitle = "2. Proportion of white poulation foreign born") + 
+        subtitle = "2. Proportion of white population foreign born") + 
   theme(legend.position = "bottom")
-
 grid.arrange(p1, p2, nrow = 1)
-
-```
-
-## Incorporating urban white immigrants into the settler polity and industrial working class
+```
+
+## Regulating and incorporating urban white immigrants
+
+KEY CHARACTERISTICS: URBANIZATION, INDUSTRIALIZATION, MORAL PANICS OVER IMMIGRANT CULTURAL DIFFERENCE. (TWO ROUTES, REACTIONARY DISCIPLINE, LIBERAL WELFARE PATERNALISM, BOTH ACTIVE).
+
+In the rapidly urbanizing portions of the Northeast, Midwest, and West Coast, we see heavy investment in what has become the ideal type of the Progressive social state: a robust and professionalizing police force and a broad paternalistic welfare state.
+These institutions were developed by elites, in part, to regulate urban disorder discursively linked to growing populations of European immigrants and Black migrants from the US South, and with rapidly assimilating these groups into the norms and ideologies of industrial capitalism and the settler nation.
 
@@
   theme(legend.position = "bottom")
-
 grid.arrange(p1, p2, nrow = 1)
@@
 
-SUMMARIZE DEMOGRAPHY LIT
-
-### The rise of progressive domestic governance in Northern US cities
-
-social work
-
-policing
-
-## Patterns in the development of state infrastructure 1900 - 1930
-
-state capacity is hard to measure, but lets try!
-Here we are going to use employment measures for three roles that came to play critical roles in the developing practices of social welfare and social control: teachers, police, and social workers.
-
-We also think about physical infrastructure for similar sets of institutions.
-We look at various forms of institutional group quarters: correctional institutions, mental institutions, and institutions for the poor and disabled.
-
-Indian Boarding schools are another key piece of physical infrastructure. We look at those too, but I am still waiting on validating some of that data.
-
-### Employment in education, policing, and social welfare: 1900 - 1930
-
-There was dramatic growth in public sector employment in general (CHECK THIS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era.
-*These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (CITES).*
-
-In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930.
+Progressive reformers in the Northeast and Midwest were preoccupied with solving a battery of social problems associated with urban life, industrial capitalism, and their perceptions of the culture and behavior of Southern and Eastern European immigrants.
+
+HIGH LEVEL SUMMARY OF 1st WAVE WHITE FEMINISM AND SOCIAL WORK VIA GORDON, HARTMAN, OTHERS.
+
+The emerging fields of sociology, social work, and criminology also sought to apply scientific principles to the organization and practice of policing and corrections.
+HIGH LEVEL SUMMARY OF PROG ERA POLICING, JUV COURT, PRISON REFORM.
+LINK TO GOTTSCHALK ON CARCERAL STATE EMERGENCE FROM THESE MOVEMENTS.
+
+# Patterns in the development of social policy infrastructure 1900 - 1930
+
+We define social state capacity as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a detailed review of theory and measurement).
+Capacity is a function of both physical infrastructure and human resources available to state actors.
+We focus on a broad set of social policy systems including education, social welfare, and criminal justice.
+
+We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key street-level social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples.
+Taken together, these three sets of measures provide a diverse and broad set of insights into spending on the basic infrastructure of social governance during the early 20th century.
+
+## Employment in education, policing, and social welfare: 1900 - 1930
+
+There was dramatic growth in public sector employment in general (CLEAR FROM FED GOVT EMPLOYMENT, OTHER CITATIONS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era.
+*These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (GENERAL LIT ON GOVT COMPLEXITY AND ARTICULATION 1900 - 1930).*
+
+In Figure \@ref(fig:emp-ts), we display employment rates for the lower 48 US States between 1900 and 1930 in panel 1, and the proportional change in employment rates by category over time in panel 2.
 Employment of teachers increased from a rate of 5.5 per 1,000 population in 1900 to a rate of 8.6 per 1,000 in 1930, a 56 percent increase.
-Employment of police doubled between during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930.
+Employment of police doubled during this period, increasing from a rate of about 0.7 per 1,000 in 1900 to a rate of about 1.4 per 1,000 in 1930.
 Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900).
-However, by 1930, the Census recorded about 17,000 social workers nationally, and increase of over 40 fold in terms of employment rates per capita.
+However, by 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita.
+CHECK THESE NUMBERS
 
@@
   mutate(name = case_when(
-    name == "Teachers"~ "1. Teachers",
-    name == "Police" ~ "2. Police",
-    name == "social_workers" ~ "3. Social Workers")) %>% 
+    name == "Teachers"~ "Teachers",
+    name == "Police" ~ "Police",
+    name == "social_workers" ~ "Social Workers")) %>% 
   mutate(rate = value / pop * 1e3)
-
-emp_pdat  %>% 
-  ggplot(aes(x = year, y = rate)) + 
+### convert to proportional change
+emp_pdat00<-emp_pdat %>% 
+  filter(year == 1900) %>% 
+  rename(rate00 = rate) %>% 
+  select(-year, -value, -pop)
+emp_pdat_prop <- emp_pdat %>% 
+  left_join(emp_pdat00) %>% 
+  mutate(prop_change = rate / rate00) 
+p1<-emp_pdat_prop  %>% 
+  ggplot(aes(x = year, y = rate, color = name)) + 
+  geom_point() + 
   geom_line() + 
-  facet_wrap(~name, scales = "free_y") + 
-  labs(x = "Year", y = "Employment rate")
-
-```
-
-However, this expansion of the reach of the state did not occur uniformly across the country.
+  labs(x = "Year", y = "Rate per 1,000",
+       color = "",
+       subtitle = "1. Employment rate") + 
+  theme(legend.position = "bottom")
+p2<-emp_pdat_prop  %>% 
+  ggplot(aes(x = year, y = prop_change, color = name)) + 
+  geom_point() + 
+  geom_line() + 
+  labs(x = "Year", y = "Proportional change (log scale)",
+       color = "",
+       subtitle = "2. Change relative to 1900") + 
+  scale_y_log10() + 
+  theme(legend.position = "bottom")
+
+grid.arrange(p1, p2, nrow = 1)
+```
+
+However, this expansion of the reach of state and state-aligned street-level bureaucrats into social life did not occur uniformly across the country.
 Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States.
@@
 Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California.
-Police employment rates were relatively lower in places like Arkansas, Missisippi, and North Carolina.
-The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Missisippi and South Dakota recorded very few social workers in the 1930 Census.
+Police employment rates were relatively lower in places like Arkansas, Mississippi, and North Carolina.
+The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Mississippi and South Dakota recorded very few social workers in the 1930 Census.
 
@@
   filter(year == 1930)
-
 p1<-ggplot(pdat) + 
@@
   theme(legend.position = "bottom")
-
 p3<-ggplot(pdat) + 
@@
   theme(legend.position = "bottom")
-
 grid.arrange(p1, p2, p3, nrow = 1)
-```
-
-### Institutional capacity: prisons, jails, asylums, orphanages and poorhouses
-
-State capacity could also be measured through the ability of states to engage in therapeutic, correctional, or paternalistic confinement.
+
+```
+
+## Physical infrastructure: prisons, jails, asylums, orphanages and poorhouses
+
+The capacity of state welfare and carceral systems is also a function of their ability to engage in therapeutic, correctional, or paternalistic confinement of socially problematic populations.
 In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states.
 Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3.
-Using data from the Department of the Interior's ``Federal Indian Boarding School Initiative'' (2022), we display a count of Federal Indian boarding schools operating in 1930 in panel 4.
-I MAY ADD ENROLLMENT DATA FROM ARCIA REPORTS ONCE WE VALIDATE IT.
-
-Data on institutional group quarters populations are less reliable than employment data for constructing a full 1900 - 1930 Census time series.
-
-shows Prisons in NE, S, CA. Mental and poverty instutitons mostly in Urban NE, MW, CA
-
-```{r gq-map, fig.cap = "Institutional group quarters, US 1930"}
+Note that data quality issues in the 1910 Census limit the possibilities for time series analysis.
+
+Panel 3 displays the rate of confinement in institutions for the poor and disabled across US states in 1930.
+These institutions included poorhouses, settlement homes, orphan asylums, sanitariums, and similar institutions targeted at containing and/or supporting social groups understood to be 'dependent' on state or religious charity support.
+We find that rates of confinement (likely closely correlated with capacity) were lowest in states in the South, like Arkansas, Alabama, Mississippi and Florida.
+New England, the mid-Atlantic, and Great Lakes states had relatively high rates of confinement in these kinds of institutions.
+The East North Central Division, including states like Wisconsin, Illinois, and Ohio had the highest average rate of confinement, at 4.3 per 1,000 persons, while the West South Central division (Arkansas, Louisiana, Oklahoma, and Texas) had the lowest rate of confinement at 1.4 per 1,000 persons in 1930.
+
+Panel 2 displays regional patterns in confinement in institutions for the mentally ill across places in 1930.
+States in the mid Atlantic, New England, and West Coast had the highest rates of confinement, at over 3 per 1,000 on average.
+In New Hampshire and Massachusetts, confinement rates in 1930  exceeded 4 per 1,000 population.
+States in the East South Central Division (Mississippi, Alabama, Tennessee, and Kentucky) had the lowest average levels of confinement, at 1.4 per 1,000 persons.
+Levels of confinement ranged between 2 and 3 on average across the rest of the country.
+Of note, we observe noinmates in mental institutitons in North Dakota in the 1930 census data.
+
+Panel 1 of Figure \@ref{gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census.
+We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population.
+Confinement rates were highest on average in the West Coast states (2.5 per 1,000).
+Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population.
+Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.
+
+Whereas states in the South tended to have very low levels of confinement in institutions for the poor and disabled and relatively low levels of confinement in mental institutions, they had levels of confinement in correctional institutions that were near national averages.
+This suggests that Southern policy makers and philanthropic allies systematically invested fewer resources in physical infrastructure targeted at containing or supporting the poor, young, mentally ill, and disabled than did other states, while maintaining and developing prisons and jails at rates similar to the rest of the nation.
+
+```{r gq-map, fig.cap = "Population confined in prisons, mental institutions, and institutions for the poor and disabled, 1930"}
 pdat<-st %>% 
@@
   left_join(bs) %>% 
-  filter(year == 1930)
+  filter(year == 1930) %>% 
+  mutate(inst_correctional_pc = inst_correctional / tot_pop * 1e3,
+         inst_mental_pc = inst_mental / tot_pop * 1e3,
+         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3)
 
@@
   geom_sf(aes(geometry = geometry,
-              fill = inst_correctional / 1e3)) + 
+              fill = inst_correctional_pc)) +
   scale_fill_viridis_c() +
   theme_void() +
-  labs(fill = "Population\n(thousands)",
+  labs(fill = "Rate",
        subtitle = "1. Correctional institutions") +
   theme(legend.position = "bottom")
-
 p2<-ggplot(pdat) +
   geom_sf(aes(geometry = geometry,
-              fill = inst_mental / 1e3)) + 
+              fill = inst_mental_pc)) +
   scale_fill_viridis_c() +
   theme_void() +
-  labs(fill = "Population\n(thousands)",
+  labs(fill = "Rate",
         subtitle = "2. Mental institutions") +
   theme(legend.position = "bottom")
-
 p3<-ggplot(pdat) +
   geom_sf(aes(geometry = geometry,
-              fill = inst_poor_disabled / 1e3)) + 
+              fill = inst_poor_disabled_pc)) +
   scale_fill_viridis_c() +
   theme_void() +
-  labs(fill = "Population\n(thousands)",
+  labs(fill = "Rate",
         subtitle = "3. Institutions for the\npoor and disabled") +
   theme(legend.position = "bottom")
-
-grid.arrange(p1, p2, p3, nrow = 2)
-```
-
-### Institutional capacity: Indian boarding schools
-
-FROM DOI REPORTS
-
-Figure \@ref{fig:bsts} displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference.
+grid.arrange(p1, p2, p3, nrow = 1)
+
+# for in-text callouts
+pdat<-st %>% 
+  left_join(pop) %>% 
+  left_join(bs) %>% 
+  filter(year == 1930)%>% 
+  mutate(inst_correctional_pc = inst_correctional/tot_pop * 1e3,
+         inst_mental_pc = inst_mental / tot_pop * 1e3,
+         inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3)
+
+pdat_callout<-pdat %>% 
+  pivot_longer(cols = inst_correctional_pc:inst_poor_disabled_pc,
+               names_to = "inst_type",
+               values_to = "rate") %>% 
+  group_by(inst_type) %>% 
+  summarize(mean = mean(rate),
+            sd = sd(rate)) %>% 
+  as_tibble() %>% 
+  select(-geometry)
+```
+
+## Physical infrastructure: Indian boarding schools
+
+The Indian boarding school epitomizes the sustained effort by the US government and its allies in settler religious and philanthropic organizations to eliminate Native nations through the coercive assimilation of Indigenous children.
+Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934.
+Notably, these carceral and genocidal institutions represented an early, substantial, and sustained investment by the federal government in education infrastructure.
+
+Below, we provide a descriptive overview of the geography and timing of boarding school operations.
+Following initiatives by Secretary of the Interior Deb Haaland and leaders from across Indian Country, the US Department of the Interior has initiated its first comprehensive review of the scope and impacts of the Indian boarding school program.
+We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year.
+We pair these data with information coded by our research team from archived annual reports produced by the Commissioner on Indian Affairs between 1890 and 1920 (CITATIONS).
+
+Figure \@ref(fig:bsts) displays the number of Indian boarding schools operating annually between 1830 and 1930, with eras of Federal Indian law shaded for reference.
 The number of boarding schools grew dramatically during the Reservation and Assimilation eras, with most growth occurring between 1870 and 1905.
-After 1905, a slow decline in the number of operating schools began continued until the end of our focal period, 1930.
+After 1905, a slow decline in the number of operating schools began and continued until the end of our focal period, 1930.
 At the beginning of the Removal Era (1830), there were 18 boarding schools in operation, and 23 were operating at the end of the Removal era (1850).
@@
 ```{r bsts, fig.cap = "Indian boarding schools operating annually, US 1830 - 1930"}
-cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
-
+cbPalette <- c("#999999", "#E69F00", "#56B4E9")
 t<-bs %>% 
@@
          summarize(boarding_schools = sum(boarding_schools))
-
 ggplot(t) +
@@
            color = cbPalette[1:3],
-           label = c("Removal era", "Reservation era", "Assimilation era")) +
+           label = c("Removal era", "Reservation era", "Assimilation\nand allotment era")) +
   labs(x = "Year", y = "Number of facilities") + 
@@
 
-In Figure \@ref{fig:bsts2} we zoom in on our focal period. This uses two sources of data, p1 is DoI, p2 is ARCIA. EXPLAIN THAT THESE ARE LIKELY UNDERCOUNTS AGAIN.
-
-Expansion of program via enrollment even as N facilities is declining and capacity is declining. surprise surprise.
-
-```{r bsts2, fig.cap = "Boarding schools and recorded student enrollment and capacity, US totals 1890 - 1920"}
-## map out the enrollment and capacity data
+Figure \@ref(fig:bsmap) shows the geographic distribution of boarding schools in operation when enrollments were at a 1830 - 1930 period maximum (242 facilities documented as operating in 1907).
+The Department of the Interior has identified Indian Boarding schools operating in 1907 in most regions of the country, with the exception of New England and the Southeast.
+There were at least 64 boarding school facilities operating in Oklahama in 1907, by far the highest number in the US.
+There were 8 states with at least 10 facilities in operation in 1907: South Dakota (24), Minnesota (18), Arizona (14), Montana (14), New Mexico (11), Washington (11), Alaska (10, not displayed), and Wisconsin (10).
+There were at least 5 baording schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5).
+States with at least one confirmed facility in operation in 1907 included Nebraska (3), Nevada (3), Pennsylvania (3), Utah (3), Wyoming (3), Iowa (2), Michigan (2), Arkansas (1), Illinois (1), Missouri (1), New York (1), North Carolina (1), and Virginia (1).
+The data do not record facilities operating in 1907 in the remaining 21 states.
+
+```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
 bs_ts<-agents %>% 
@@
 
-t<-bs_ts %>% 
-  group_by(year) %>% 
-  summarize(Capacity = sum(boarding_n_acc),
-            Enrollment = sum(students_enroll)) %>% 
-  pivot_longer(cols = Capacity:Enrollment)
-
-### for in-text callouts
-p1<-ggplot(t, 
-           aes(x = year, y = value, color = name)) + 
-  geom_line() + 
-  labs(x = "Year", y = "Children",
-       color = "",
-       subtitle = "2. Reported capacity and enrollment (ARCIA)")
-
-p2<-ggplot(bs_ts %>% 
-  group_by(year) %>% 
-  summarize(boarding_schools = sum(boarding_schools)),
-           aes(x = year, y = boarding_schools)) + 
-  geom_line() + 
-  labs(x = "Year", y = "Count",
-       subtitle = "1. Facilities operating annually (Dept. of Interior)")
-
-grid.arrange(p2, p1, nrow = 1) 
-```
-
-Figure \@ref{fig:bsmap} shows the geographic distribution of boarding school infrastructure and utilization when enrollments were at a 1890 - 1920 period maximum in 1917. DISCUSS
-
-```{r bsmap, fig.cap = "Boarding schools in operation and reported enrollment and capacity at peak enrollment (1917)"}
-p3<-ggplot(bs_ts %>% 
-         filter(year == 1917),
-       aes(fill = students_enroll / 1e3,
-           geometry = geometry)) + 
-  geom_sf() + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Children\n(thousands)",
-       subtitle = "3. Reported enrollment") + 
-  theme(legend.position = "bottom")
-
-p2<-ggplot(bs_ts %>% 
-         filter(year == 1917),
-       aes(fill = boarding_n_acc / 1e3,
-           geometry = geometry)) + 
-  geom_sf() + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Children\n(thousands)",
-       subtitle = "3. Reported capacity") + 
-  theme(legend.position = "bottom")
-
-p1<-ggplot(bs_ts %>% 
-         filter(year == 1917),
+ggplot(bs_ts %>% 
+         filter(year == 1907),
        aes(fill = boarding_schools,
@@
   geom_sf() + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Facilities",
-       subtitle = "1. Boarding schools operating") + 
+  scale_fill_viridis_c(trans = "sqrt") + 
+  theme_void() + 
+  labs(fill = "Facilities\noperating") + 
   theme(legend.position = "bottom")
-
-grid.arrange(p1, p2, p3, nrow = 1)
 ```
@@
 
-The works of settler colonial and racial politics historical and theoretical scholars suggest a straightforward approach to selecting measures that are either a) direct, or b) proxy measures for elite white settler interests given a few assumptions about relationships between population distributions and political processes.
-
-To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of policy outcomes are related to a set of demographic and economic predictors at the state and year-levels.
+To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels.
 We conduct exploratory and inferential analysis using data on US states between 1900 and 1930.
 
-DESCRIBE THE MEASURES IN GENERAL TERMS HERE
-
-Generally our approach suggests that state formation in a particular place *s* and time *t* is a function of settler interests specific to that time and place.
+Generally our approach suggests that social policy infrastructure in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.
 
@@
 
-We proceed by producing and evaluating descriptive data visuals for our operational measures of 'settler interests' and comparing distributions of these measures across states and over time.
-We then produce and evaluate a similar set of visuals to evaluate the distribution of our 'infrastructure' measures.
+We proceed by producing and evaluating descriptive data visuals for our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time.
+We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures.
 Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national temporal trends, and basic characteristics of the population in the state and year.
+
+A replication package is available at github.com/f-edwards/racial_politics_path
 
@@
 We rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type.
-From these samples, we derive state-year measures of the population housed in correctional instiuttions, the population housed in mental institutions, and the population housed in institutions for the elderly, handicapped, and poor.
+From these samples, we derive state-year measures of the population housed in correctional institutions, the population housed in mental institutions, and the population housed in institutions for the elderly, handicapped, and poor.
 
 Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress (CITE).
-The distribution of Native lands to settlers is measured through data on the implementation of the Homestead Act and Morill Act of 1862.
-We use data on parcels of land transferred to land-grant universities under the Morill Act as documented by Robert Lee an High Country News as part of their 'Land Grab University' project.
+The distribution of Native lands to settlers is measured through data on the implementation of the Homestead Act and Morrill Act of 1862.
+We use data on parcels of land transferred to land-grant universities under the Morrill Act as documented by Robert Lee and High Country News as part of their 'Land Grab University' project.
 These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available.
@@
 
-geography itself endogenous to settler colonization, this is tricky!
-we are trying to understand the emergence of the state as bureaucratic entity in this period, so it makes sense to do this, but it is messy!
+geography itself is endogenous to settler colonization, this is tricky!
+We are trying to understand the emergence of the settler state as a bureaucratic entity in this period, so it makes sense to do this, but it is messy!
 we don't want to imply a teleological reading here.
 borders are unstable and contested.
-
-## statistical methods and models
-
-Nearly of our 'infrastructure' outcomes are measured as counts of persons.
+Here we are thinking about the human and physical infrastructures of the US carceral and welfare states, so we focus our attention on those spatial boundaries, while recognizing that the process of state formation is closely tied up with defining and reinforcing those borders while seeking to erase others.
+
+## Statistical methods and models
+
+We develop a series of regression models to explore the conditional relationships between outcomes measuring the capacity of settler state governments and predictors measuring the contexts for three focal settler colonial projects.
+These state-year models evaluate how three regionally specific settler projects -- Jim Crow, Native land redistribution, and white immigrant assimilation -- related to patterns in the development of core state infrastructure and capacity during a period of rapid increases in state capacity in the US.
+
+Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts.
+We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters.
+This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states.
+
+We discuss our process for model selection in detail in the appendix.
+
+Most of our infrastructure outcomes are measured as counts of persons.
 As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes.
-For facility counts (boarding schools), we employ a similar model that removes population offsets from the specification.
-
-DISCUSSIONS OF MODEL SELECTION IN APPX
-
-Model 1. is
+The population offset term on the left-hand side of equation XXXX allows us to adjust outcomes and estimated relationships for the size of state populations (as per capita rates).
+
+*Equation 1: Model for count outcomes*
 
@@
 
-\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} + \beta_4 \log(\textrm{child population}_{st}) \]
+\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} \]
 
@@
 
-Model 2 (for zero inflated outcomes) is
-
-XXXXX
+We also model the presence and number of Indian boarding schools in each state as a focal measure of state infrastructure in the Progressive era.
+Unlike other outcomes in our study, there are a substantial number of states that did not have Indian boarding schools during this period.
+We treat the presence of boarding schools as a zero-inflated process, and use a two-stage model to evaluate their distribution.
+One component of the model estimates the probability of having at least one boarding school during the focal period (a binary outcome).
+This binary measure of boarding schools is constant over the period 1900 - 1930 at the state-level, so we estimate this model using data only for the first observation in the period, 1900.
+The second component of the model estimates the count of boarding schools present in a state-year, conditional on there being at least one boarding school in the state, following equation 1, with the removal of the population offset term.
+
+*Model 2 (for zero inflated outcomes) is*
+
+\[ y_{s} > 0 \sim \textrm{Bernoulli}\left(\theta\right) \]
+\[ \textrm{logit} \left(\theta_s\right) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} \]
+\[\beta \sim N(0, 20)\]
+
+Across all models, we re-scale predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence and parameter interpretability in a context of widely differing scales across measures.
 
@@
 
+There are clear relationships between the geography of settler colonial projects and the local scale of welfare and carceral state capacity in the early 20th century US.
+In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators.
+In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between some of the physical infrastructure of the carceral and welfare state and settler colonial projects.
+And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between the locations of Federal Indian boarding schools and these measures of settler colonial project timing and location.
+
+In states where Native lands were being actively redistributed to settlers, there were higher levels of employment of teachers, and relatively lower levels of employment of police and social workers.
+These states had per capita rates of detention in correctional facilities, mental institutions, and institutions for the poor and disabled that were near the national average.
+They were dramatically more likely than other states to build and maintain Indian boarding schools.
+
+In states where there were large numbers of white immigrants residing in cities, teacher employment was around the national average, but police and social worker employment were dramatically higher than in other states.
+Rates of detention in correctional institutions were similar to all other states at the time, but the rate of detention in institutions for the mentally ill, poor, or disabled were much higher in these states than in other parts of the country.
+And while there were notable examples, such as the notorious Carlisle school in Pennsylvania, states with large urban white immigrant populations did not routinely build and maintain Indian boarding schools
+
+In states with large numbers of Black tenant farmers between 1900 and 1930, there was much lower employment of teachers, police officers, and social workers than in other US states.
+Despite these low levels of employment in front-line social policy positions, however, these states had similar rates of incarceration in correctional institutions compared to other states.
+They had relatively lower levels of detention in institutions for the mentally ill, the poor, or the disabled.
+These states were also the least likely to build and maintain Indian boarding schools when compared with all other US states.
+
+PUNCHY HIGH LEVEL SUMMARY. WHAT DO WE CALL THIS? FRONTIER DISPOSSESSION + NATION BUILDING, URBAN REGULATION + IMMIG INCORPORATION, JIM CROW HYBRID FORMAL+INFORMAL RACIAL TERROR + MALIGN NEGLECT.
+
 ## Employment models
 
-```{r}
-# # ### extract posterior draws for b1, b2, b3, set up for plotting
-# # pdat<-teachers_m1_bayes %>% 
-# #   gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
-# #                  b_scaleIurban_immig_popDtot_pop,             
-# #                  b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
-# #   mutate(model = "1. Teachers") %>% 
-# #   bind_rows(
-# #     cops_m1_bayes %>% 
-# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
-# #                      b_scaleIurban_immig_popDtot_pop,             
-# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>% 
-# #       mutate(model = "2. Police")) %>% 
-# #   bind_rows(
-# #     socwork_m1_bayes %>% 
-# #       gather_draws(c(b_scaleIblk_sharecropDtot_pop,       
-# #                      b_scaleIurban_immig_popDtot_pop,             
-# #                      b_scalemorill_acres_cumPhomestead_acres_cum)) %>%
-# #       mutate(model = "3. Social workers")) %>% 
-# #   mutate(.variable = case_when(
-# #     .variable == "b_scaleIblk_sharecropDtot_pop" ~ 
-# #       "β1: Black tenant\nfarmers per capita",
-# #     .variable == "b_scaleIurban_immig_popDtot_pop" ~ 
-# #       "β2: Urban white\nimmigrants\nper capita",
-# #     .variable == "b_scalemorill_acres_cumPhomestead_acres_cum" ~ 
-# #       "β3: Cumulative \nMorill +\nHomestead acres"))
-# 
-# 
-# ggplot(pdat,
-#        aes(x = .value, 
-#            y = .variable)) +
-#   stat_halfeye() + 
-#   geom_vline(aes(xintercept = 0), lty = 2) + 
-#   labs(x = "Estimate",
-#        y = "") + 
-#   facet_wrap(~model) + 
-#   theme_bw()
-```
+As shown in Figure \@ref(fig:emp-ts), there was a dramatic increase in the number of people employed as teachers between 1900, when there were `r emp_pdat %>% filter(year==1900, name=="Teachers") %>% select(value) %>% slice(1)` teachers recorded in the US Census, and 1930, when there was a reported total of `r emp_pdat %>% filter(year==1930, name=="Teachers") %>% select(value) %>% slice(1)` teachers.
+This amounts to a `r round((emp_pdat_prop %>% filter(year==1930, name=="Teachers") %>% select(prop_change) %>% slice(1)-1)*100)` percent growth in the per capita rate of teacher employment over a 30 year period.
+We model state-year teacher employment rates using the model described in Equation 1, and display the posterior distributions of focal parameters in Figure \@ref(fig:emp-m-vis) to evaluate how the geography of settler colonial projects related to patterns of teacher employment across places and over time.
+More details on model parameters and diagnostics are provided in Appendix Table (XXXXXX, full regression table).
+
+Panel 1 in Figure \@ref(fig:emp-m-vis) shows the distributions of the posterior parameter estimates of our relationship between three focal settler colonial contexts and teacher employment.
+Conditional on our data and model specifications, we find that states with higher levels of settler land redistribution under the Homestead and Morill acts employed more teachers per capita than did other states between 1900 and 1930.
+We find that states with higher levels of Black tenant farming employed fewer teachers per capita than did other states.
+States with high urban white immigrant populations employed teachers at rates near the national average.
+
+Panel 2 shows the distributions of focal posterior parameter estimates from our models of police employment.
+We find that states with higher levels of settler land distribution through the Morill and Homestead acts had lower levels of police employment on average.
+States with large urban white immigrant populations had significantly higher levels of police employment than did other states during this period.
+And states with large numbers of Black tenant farmers tended to have lower levels of police employment.
+Despite relatively low levels of police employment in many Southern states, there were incredible levels of both formal and informal white supremacist violence during this period (CITATIONS).
+
+The distribution of social work employment across states follows the same general pattern as police employment.
+In Panel 3, we show that states with large white urban immigrant populations had much higher levels of social worker employment than did other US states between 1900 and 1930.
+States with high levels of Black tenant farming and settler land redistribution tended to have slightly lower levels of social worker employment than other states, but these expected employment levels are not statistically significant from zero at conventional thresholds.
+
+```{r emp-m-vis, fig.cap = "Posterior distributions of focal model parameters by employment type, US States 1900 - 1930. Densities, expected values, and credible intervals, 0.5, 0.95)"}
+#### extract posterior draws for b1, b2, b3, set up for plotting
+pdat<-teachers_m1 %>%
+  gather_draws(c(b_scaleblk_sharecrop_pc,
+                 b_scaleurban_immig_pc,
+                 b_scalehomestead_plus_morill)) %>%
+  mutate(model = "1. Teachers") %>%
+  bind_rows(
+    cops_m1 %>%
+      gather_draws(c(b_scaleblk_sharecrop_pc,
+                     b_scaleurban_immig_pc,
+                     b_scalehomestead_plus_morill)) %>%
+      mutate(model = "2. Police")) %>%
+  bind_rows(
+    socwork_m1 %>%
+      gather_draws(c(b_scaleblk_sharecrop_pc,
+                     b_scaleurban_immig_pc,
+                     b_scalehomestead_plus_morill)) %>%
+      mutate(model = "3. Social workers")) %>%
+  mutate(.variable = case_when(
+    .variable == "b_scaleblk_sharecrop_pc" ~
+      "B1: Black tenant\nfarmers per capita",
+    .variable == "b_scaleurban_immig_pc" ~
+      "B2: Urban white\nimmigrants\nper capita",
+    .variable == "b_scalehomestead_plus_morill" ~
+      "B3: Cumulative \nMorill +\nHomestead acres"))
+ggplot(pdat,
+       aes(x = .value,
+           y = .variable)) +
+  stat_halfeye() +
+  geom_vline(aes(xintercept = 0), lty = 2) +
+  labs(x = "Estimate",
+       y = "") +
+  facet_wrap(~model) +
+  theme_bw()
+```
+
+We display the temporal components of these models, expected values for employment rates conditional on the range of observed settler colonial contexts, and the observed bivariate relationships between employment rates and context measures in Figures \@ref(fig:postpred-teachers), \@ref(fig:postpred-police), and \@ref(fig:postpred-socwork).
+As described in Figure \@~ref(fig:emp-ts), there were dramatic increases in the employment of teachers, social workers, and police nationally between 1900 and 1930.
+But these national trends obscure significant sub-national variation in employment rates and growth trajectories.
+
+Figure \@ref(fig:postpred-teachers) shows that while teacher employment grew substantially across all states between 1900 and 1930, states with high levels of Black tenant farming tended to have much lower levels of teacher employment across the full time period than did states that did not have high levels of Black tenant farming.
+Models suggest that the expected employment rate for teachers in states with high levels of Black tenant farming (25 percent of the population, similar to South Carolina in 1900 - 1930) would be about 3.4 with a 90 percent credible interval of [3.0, 3.8] per 1,000 population in 1900; about 3.9 [3.5, 4.4] per 1,000 persons in 1910; about 4.5 [4.0, 5.0] per 1,000 in 1920, and about 5.3 [4.7, 6.0] teachers per 1,000 in 1930.
+
+By contrast, states with high levels of settler land redistribution through the Morill and Homestead acts to have much higher levels of teacher employment than Jim Crow labor regime states.
+In 1900, a state with a high level of acreage redistributed to settlers in the recent past (10 million acres, similar to Minnesota, Nebraska, or Kansas) is expected to have had a teacher employment rate of about 6.7 [6.3, 7.2] per 1,000 population.
+In 1910, a state with high levels of Homestead and Morill act land distributions (13 million acres, similar to Kansas, North Dakota, and Oklahoma) is expected to have employed about 8.1 [7.5, 8.8] teachers per 1,000 persons.
+By 1930, states with high levels of contemporary settler land redistribution (about 20 million acres, similar to Colorado, North Dakota and South Dakota) had highest-in-the-nation levels of teacher employment, with expected rates of about 12.3 [11.3, 13.5] teachers per 1,000 persons.
+
+For each decade we evaluate, our models expect no meaningful differences in teacher employment across states with high and low levels of white urban immigrant populations, conditional on low levels of active settler land redistribution and low levels of Black tenant farming.
+There are, of course, high levels of expected increases in teacher employment over time, ranging from a low of 5.7 [5.2, 6.1] in 1900 to a high of 8.9 [8.1, 9.6] in 1930.
+
+```{r postpred-teachers, fig.cap = "Estimated and observed relationships between teacher edfmployment rates and focal settler colonial context variables, US States 1900 - 1930.
+Posterior prediction for expected teacher employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
+Dots and text indicate observed values."}
+### blk_sharecrop
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = 0,
+            urban_immig_pc = 0,
+            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
+            year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(teachers_m1,
+                  re_formula = NA)
+### credible intervals for text callouts
+t<-pred_dat %>% 
+  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p1<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) + 
+  geom_point(data = mdat,
+             aes(y = teachers / tot_pop * 1e3,
+                 x = blk_sharecrop_pc),
+             size = 1, alpha = 0.5) +
+  # geom_text(data = mdat %>% 
+  #             filter(blk_sharecrop_pc > 0.03),
+  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
+  #               x = blk_sharecrop_pc,
+  #               label = STUSPS),
+  #           alpha = 0.75,
+  #           size = 3) + 
+  facet_wrap(~year, nrow = 1) + 
+  guides(fill = "none") +
+  theme_bw() +
+  coord_cartesian(ylim = c(2.7,16.8)) + 
+  labs(y = "",
+       x = "Proportion of population Black tenant farmers")
+### homestead+morill
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = seq(0,
+                                                  max(mdat$homestead_plus_morill),
+                                                  length.out = 100),
+                      urban_immig_pc = 0,
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(teachers_m1,
+                  re_formula = NA)
+### text callouts
+t<-pred_dat %>% 
+  filter(homestead_plus_morill>20e6, homestead_plus_morill < 20.2e6) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p2<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) + 
+  geom_point(data = mdat,
+             aes(y = teachers / tot_pop * 1e3,
+                 x = homestead_plus_morill),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(homestead_plus_morill > 2e5),
+  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
+  #               x = homestead_plus_morill,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
+       y = "Teachers per 1,000 population") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(2.7,16.8)) + 
+  scale_x_continuous(labels = function(x)x / 1e6) + 
+  theme_bw() 
+### urban immig
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = 0,
+                      urban_immig_pc = seq(0,
+                                           max(mdat$urban_immig_pc),
+                                           length.out = 100),
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(teachers_m1,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p3<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = urban_immig_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) + 
+  geom_point(data = mdat,
+             aes(y = teachers / tot_pop * 1e3,
+                 x = urban_immig_pc),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(urban_immig_pc > 0.02),
+  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
+  #               x = urban_immig_pc,
+  #               label = STUSPS),
+            # alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Proportion of population urban white immigrants",
+       y = "") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(2.7,16.8)) + 
+  theme_bw() 
+grid.arrange(p1, p2, p3, nrow = 3)
+```
+
+We observe clear differences in police employment across settler colonial contexts.
+In Figure \@ref(fig:postpred-police) we show posterior expected values for police employment rates across the observed range of settler colonial contexts over time.
+Rates of police employment were relatively low in the Jim Crow South.
+Our models suggest that in a state with 25 percent of the population documented as Black tenant farmers in the Census, an expected police employment rate of 0.31 [0.26, 0.38] per 1,000 persons in 1900, a rate of 0.37 [0.30, 0.44] in 1910, a rate of 0.41 [0.34, 0.49] in 1920, and a rate of about 0.63 [0.52, 0.77] police employed per 1,000 persons in 1930.
+
+Employment rates of police officers were dramatically higher in states with large white urban immigrant populations.
+In states with 25 percent of their population recorded as white immigrants living in urban Census tracts (similar to New Jersey, New York, Massachusetts, and Rhode Island), our models expected states to have had about 1.2 [0.99, 1.4] police employed per 1,000 population in 1900, 1.4 [1.2, 1.6] per 1,000 population in 1910, 1.5 [1.3, 1.8] in 1920, and 2.3 [2.0, 2.8] police per 1,000 population in 1930.
+
+Police were employed at slightly lower levels in states with high levels of recent settler land distributions than in other states.
+In 1930, our models expect a state with no recent settler land distributions to have had an average of about 1.0 [0.90, 1.1] police per 1,000 population, compared to an expected police employment rate of 0.7 [0.6, 0.8] per 1,000 in a state with very high levels of recent settler land redistribution.
+
+```{r postpred-police, fig.cap = "Estimated and observed relationships between police employment rates and focal settler colonial context variables, US States 1900 - 1930.
+Posterior prediction for expected police employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
+Dots and text indicate observed values."}
+### blk_sharecrop
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = 0,
+            urban_immig_pc = 0,
+            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
+            year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(cops_m1,
+                  re_formula = NA)
+### credible intervals for text callouts
+t<-pred_dat %>% 
+  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p1<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6)  + 
+  geom_point(data = mdat,
+             aes(y = police / tot_pop * 1e3,
+                 x = blk_sharecrop_pc),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(blk_sharecrop_pc > 0.03),
+  #           aes(y = police / tot_pop * 1e3 + 0.1,
+  #               x = blk_sharecrop_pc,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  facet_wrap(~year, nrow = 1) + 
+  guides(fill = "none") +
+  theme_bw() +
+  coord_cartesian(ylim = c(0,3)) + 
+  labs(y = "",
+       x = "Proportion of population Black tenant farmers")
+### homestead+morill
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = seq(0,
+                                                  max(mdat$homestead_plus_morill),
+                                                  length.out = 100),
+                      urban_immig_pc = 0,
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(cops_m1,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  group_by(year, homestead_plus_morill) %>% 
+  summarize(med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3) %>% 
+  filter(year == 1930)
+# t[1,]
+# t[90,]
+p2<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
+ stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  geom_point(data = mdat,
+             aes(y = police / tot_pop * 1e3,
+                 x = homestead_plus_morill),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(homestead_plus_morill > 3e6),
+  #           aes(y = police / tot_pop * 1e3 + 0.1,
+  #               x = homestead_plus_morill,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
+       y = "Police per 1,000 population") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(0,3)) + 
+  scale_x_continuous(labels = function(x)x / 1e6) + 
+  theme_bw() 
+### urban immig
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = 0,
+                      urban_immig_pc = seq(0,
+                                           max(mdat$urban_immig_pc),
+                                           length.out = 100),
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(cops_m1,
+                  re_formula = NA)
+### text call out
+t<-pred_dat %>% 
+  filter(urban_immig_pc > 0.25, urban_immig_pc < 0.255) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p3<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = urban_immig_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  geom_point(data = mdat,
+             aes(y = police / tot_pop * 1e3,
+                 x = urban_immig_pc),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(urban_immig_pc > 0.03),
+  #           aes(y = police / tot_pop * 1e3 + 0.1,
+  #               x = urban_immig_pc,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Proportion of population urban white immigrants",
+       y = "") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(0,3)) + 
+  theme_bw() 
+grid.arrange(p1, p2, p3, nrow = 3)
+```
+
+Social work employment dramatically increased from near-zero levels in 1900 to becoming relatively common across all US states by 1930.
+Again, however, these levels of employment and trajectories varied substantially across contexts.
+For example, in a state with a large population of Black tenant farmers, the model expectations of social worker employment increased from 1.1 [0.7, 1.8] per million population in 1900 to 3.7 [2.6, 5.6] per 100,000 population in 1930.
+Expected employment for social workers followed similar trends in states with high levels of recent settler land redistribution, starting near zero in 1900 and rising to around 4 social workers per 100,000 by 1930 in a state with high levels of contemporary settler land redistribution.
+
+By contrast, social work employment started at relatively high levels in 1900 and rose dramatically in states with large urban immigrant populations.
+Our models expect a state with a large urban immigrant population (25 percent of population) to have had about 9.9 [7.0, 14.3] social workers per million population in 1900.
+This employment rate increased dramatically during this period.
+Median expected employment rates for social workers were 3.3 per 100,000 in 1910, 1.5 per 10,000 in 1920, and 3.2 per 10,000 in 1930.
+This represented a 32 fold increase between 1900 and 1930.
+Our models expect per capita social work employment to be about 8 times higher in states with high levels of white urban immigrants in 1930 than they were in states operating under a Jim Crow agricultural labor regime.
+
+```{r postpred-socwork, fig.cap = "Estimated and observed relationships between social worker employment rates and focal settler colonial context variables, US States 1900 - 1930.
+Posterior prediction for expected teacher employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
+Dots and text indicate observed values."}
+### blk_sharecrop
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = 0,
+            urban_immig_pc = 0,
+            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
+            year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(socwork_m1,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p1<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  geom_point(data = mdat,
+             aes(y = socwork / tot_pop * 1e3,
+                 x = blk_sharecrop_pc),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(blk_sharecrop_pc > 0.03),
+  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
+  #               x = blk_sharecrop_pc,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  facet_wrap(~year, nrow = 1) + 
+  guides(fill = "none") +
+  theme_bw() +
+  coord_cartesian(ylim = c(0,0.3)) + 
+  labs(y = "",
+       x = "Proportion of population Black tenant farmers")
+### homestead+morill
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = seq(0,
+                                                  max(mdat$homestead_plus_morill),
+                                                  length.out = 100),
+                      urban_immig_pc = 0,
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(socwork_m1,
+                  re_formula = NA)
+p2<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  geom_point(data = mdat,
+             aes(y = socwork / tot_pop * 1e3,
+                 x = homestead_plus_morill),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(homestead_plus_morill > 2e5),
+  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
+  #               x = homestead_plus_morill,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
+       y = "Social workers per 1,000 population") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(0,0.3)) + 
+  scale_x_continuous(labels = function(x)x / 1e6) + 
+  theme_bw() 
+### urban immig
+pred_dat<-expand_grid(tot_pop = 1e6,
+                      homestead_plus_morill = 0,
+                      urban_immig_pc = seq(0,
+                                           max(mdat$urban_immig_pc),
+                                           length.out = 100),
+                      blk_sharecrop_pc = 0,
+                      year = c(1900, 1910, 1920, 1930)) %>% 
+  add_epred_draws(socwork_m1,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  filter(urban_immig_pc > 0.25, urban_immig_pc < 0.255) %>% 
+  group_by(year) %>% 
+  summarize(
+    med = median(.epred) / 1e3,
+    lwr = quantile(.epred, 0.05) / 1e3,
+    upr = quantile(.epred, 0.95) / 1e3)
+p3<-ggplot(pred_dat,
+       aes(y = .epred / 1e3, x = urban_immig_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  geom_point(data = mdat,
+             aes(y = socwork / tot_pop * 1e3,
+                 x = urban_immig_pc),
+             size = 1, alpha = 0.5) + 
+  # geom_text(data = mdat %>% 
+  #             filter(urban_immig_pc > 0.02),
+  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
+  #               x = urban_immig_pc,
+  #               label = STUSPS),
+  #           alpha = 0.75, size = 3) + 
+  guides(fill = "none") +
+  labs(x = "Proportion of population urban white immigrants",
+       y = "") + 
+  facet_wrap(~year, nrow = 1) + 
+  coord_cartesian(ylim = c(0,0.3)) + 
+  theme_bw() 
+grid.arrange(p1, p2, p3, nrow = 3)
+```
+
+Across categories of employment, we find clear relationships between settler colonial contexts and levels of employment in welfare and carceral institutions.
+We argue that the rationally and temporally varying interests of settler elites determine the kinds of colonial projects they prioritize, reflected in the kinds of human infrastructure that they develop to pursue policy goals.
 
@@
 
-RAN BOARDING SCHOOL MODEL WITH DISCONTINUITY ON ACREAGE, IT WORKS WELL.
+Similar to employment, there are also clear relationships between settler colonial contexts and the magnitude of confinement in institutions across states between 1900 and 1930.
+Figure \@ref(fig:gq-m-vis) displays posterior distributions of focal parameters for models of the populations housed in institutional group quarters in the 1900, 1920, and 1930 US Census by state.
+We find that correctional populations were similar across contexts.
+By contrast, we find that populations housed in mental institutions, institutions for the poor, and institutions for the disabled were significantly higher in states with large urban white immigrant populations than in other states, and that these populations were significantly lower in states with large Black tenant farm populations than in other states.
+
+```{r gq-m-vis, fig.cap = "Relationships between focal settler colonial contexts and populations housed in group quarters per capita, US States 1900 - 1930.
+Densities of posterior parameter estimates with credible intervals (0.66, 0.95).
+Panel 1 displays results for populations in correctional institutions, Panel 2 displays results for populations in institutions for the poor and disabled and in mental institutions."}
+#### extract posterior draws for b1, b2, b3, set up for plotting
+pdat<-prison_m1 %>%
+  gather_draws(c(b_scaleblk_sharecrop_pc,
+                 b_scaleurban_immig_pc,
+                 b_scalehomestead_plus_morill)) %>%
+  mutate(model = "1. Correctional institutions") %>%
+  bind_rows(
+    inst_m1 %>%
+      gather_draws(c(b_scaleblk_sharecrop_pc,
+                     b_scaleurban_immig_pc,
+                     b_scalehomestead_plus_morill)) %>%
+      mutate(model = "2. Asylums and poorhouses")) %>%
+  mutate(.variable = case_when(
+    .variable == "b_scaleblk_sharecrop_pc" ~
+      "B1: Black tenant\nfarmers per capita",
+    .variable == "b_scaleurban_immig_pc" ~
+      "B2: Urban white\nimmigrants\nper capita",
+    .variable == "b_scalehomestead_plus_morill" ~
+      "B3: Cumulative \nMorill +\nHomestead acres"))
+ggplot(pdat,
+       aes(x = .value,
+           y = .variable)) +
+  stat_halfeye() +
+  geom_vline(aes(xintercept = 0), lty = 2) +
+  labs(x = "Estimate",
+       y = "") +
+  facet_wrap(~model) +
+  theme_bw()
+```
+
+CONSISTENT WITH HISTORICAL WORK ON PREVALENCE OF EARLY THERAPEUTIC / WELFARE STATES.
+CITE ALL THAT STUFF.
+These institutions reflected enlightenment ideals about the FOUCAULT AND STUFF.
+But they also operated within ideological contexts of rigid racist classification and eugenic policy priorities (METZL ETC).
+Where they existed, therapeutic and welfarist institutions routinely excluded Black people.
+These data further illustrate that the geographic distribution of these facilities followed similar racial logics.
+Welfare and mental institutions were relatively rare in the Jim Crow South, and very common in states with large numbers of white immigrants in cities.
+
+```{r gq-postpred-time, fig.cap = "Expected populations housed in group quarters per capita across varying settler colonial contexts, US States 1900 - 1930.
+Densities of posterior parameter estimates with credible intervals (0.66, 0.95)."}
+### set up sims for 1900, 1920, 1930 preds
+### for high on each context for prison_m1 and inst_m1
+t<-expand_grid(
+  year = c(1900, 1920, 1930),
+  blk_sharecrop_pc = c(0, 0.25),
+  urban_immig_pc = 0,
+  homestead_plus_morill = 0,
+  tot_pop = 1e6) %>% 
+  mutate(condition = 
+           ifelse(blk_sharecrop_pc == 0,
+                  "Low value", "High value"),
+         scen = "Black tenant farming") %>% 
+  bind_rows(
+    expand_grid(
+      year = c(1900, 1920, 1930),
+      blk_sharecrop_pc = c(0),
+      urban_immig_pc = c(0, 0.25),
+      homestead_plus_morill = 0,
+      tot_pop = 1e6) %>% 
+      mutate(condition = 
+               ifelse(urban_immig_pc == 0,
+                      "Low value", "High value"),
+             scen = "White urban immigrants")) %>% 
+    bind_rows(
+      expand_grid(
+      year = c(1900, 1920, 1930),
+      blk_sharecrop_pc = c(0),
+      urban_immig_pc = 0,
+      homestead_plus_morill = c(0, 10e6),
+      tot_pop = 1e6) %>% 
+      mutate(condition = 
+               ifelse(homestead_plus_morill == 0,
+                      "Low value", "High value"),
+             scen = "Homestead + Morill acres")) 
+t<-t%>% 
+  add_epred_draws(prison_m1,
+                  re_formula = NA) %>% 
+  mutate(model = "Prisons and jails") %>% 
+  bind_rows(t %>% 
+              add_epred_draws(inst_m1,
+                  re_formula = NA) %>% 
+              mutate(model = "Asylums and poorhouses"))
+
+ggplot(t,
+       aes(x = year, y = .epred/1e3,
+           color = condition)) + 
+  stat_pointinterval(position = position_dodge()) + 
+  labs(x = "Year", y = "Expected inmates per 1,000 pop.") + 
+  facet_wrap(model~scen) + 
+  theme_bw() + 
+  labs(color = "Context variable") + 
+  theme(legend.position = "bottom")
+```
+
+In Figure \@ref(fig:gq-postpred-time) we show how the physical confinement of persons in institutional group quarters shifted over time and across contexts between 1900 and 1930.
+Note that we exclude the 1910 US Census from group quarters analyses due to data quality concerns.
+We find that the confinement of people in institutions for the mentally ill, the poor, and the disabled was dramatically more common in states with large numbers of white urban immigrants than in other states.
+Our models show that confinement in these institutions was relatively less common in states with large numbers of Black tenant farmers than in states with few Black tenant farmers.
+And we see no statistically significant difference between states with high and low levels of Homestead and Morill act land redistribution.
+Rates of confinement in these kinds remained relatively stable between 1900 and 1920, then increasing substantially by 1930.
+
+Per capita levels of incarceration in correctional institutions were similar across contexts for all periods in our analyses, with relatively stable levels between 1900 and 1920, followed by an increase in incarceration rates in 1930.
+This homogeneity in carceral capacity is notable, given the differences in police employment described in Figure \@ref(fig:postpred-police).
+These findings illustrate that while states in the US South under Jim Crow and frontier states in the Midwest and West may have had relatively small police organizations compared to their peers in the urban Northeast and Midwest, the magnitude of their capacity for incarceration was similar.
+And while the absolute rate of incarceration was similar across states during this era, this should not diminish the importance of substantial differences in the conditions and contexts of confinement across places (CITE PRISON VARIATION LIT).
+Historical scholarship has clearly shown the salience of white supremacist and settler colonial logics for the development of early-modern policing and incarceration across US geographies (CITE CONVICT LEASE LIT, KLH ON California, NORTHEAST LIT, FRONTIER STUFF (MAYBE ROSS)).
+
+## Boarding school models
+
+As shown in Figure \@ref(fig:bsts), the count of boarding schools operating annually in the US fluctuated around 200 schools operating between 1900 and 1930.
+We estimate a two stage zero-inflated model to separately estimate 1) the likelihood that a state had at least one boarding school during the focal period, then conditional on operating at least one boarding school 2) the number of schools operating with the state during a Census year.
+The posterior distributions of focal parameters are shown in Figure \@ref(fig:bs-m-mvs).
+We visualize the probability of operating at least one boarding school across focal contexts in Figure \@ref(fig:pr-bs).
+
+```{r bs-m-mvs, fig.cap = "Relationships between focal settler colonial contexts and presence and count of Indian boarding shcools.
+Zero-inflated model described in equations 1 and 2.
+Panel 1 displays binary component of model (for presence of any boarding schools).
+Panel 2 displays count component of model for number of boarding schools, conditional on at least one boarding school in the state and year."}
+### zero component visual
+pdat<-bs_m1_z %>%
+  gather_draws(c(b_scaleblk_sharecrop_pc,
+                 b_scaleurban_immig_pc,
+                 b_scalehomestead_plus_morill)) %>%
+  mutate(model = "1. Boarding schools > 0") %>%
+  bind_rows(
+    bs_m1 %>%
+      gather_draws(c(b_scaleblk_sharecrop_pc,
+                     b_scaleurban_immig_pc,
+                     b_scalehomestead_plus_morill)) %>%
+      mutate(model = "2. Count when boarding schools > 0")) %>% 
+  mutate(.variable = case_when(
+    .variable == "b_scaleblk_sharecrop_pc" ~
+      "B1: Black tenant\nfarmers per capita",
+    .variable == "b_scaleurban_immig_pc" ~
+      "B2: Urban white\nimmigrants\nper capita",
+    .variable == "b_scalehomestead_plus_morill" ~
+      "B3: Cumulative \nMorill +\nHomestead acres"))
+ggplot(pdat,
+       aes(x = .value,
+           y = .variable)) +
+  stat_halfeye() +
+  geom_vline(aes(xintercept = 0), lty = 2) +
+  labs(x = "Estimate",
+       y = "") +
+  theme_bw() + 
+  facet_wrap(~model, 
+             scales = "free_x") 
+```
+
+We find that states with large redistributions of Native lands to settlers were far more likely to operate boarding schools during this period than were other states.
+We find that a (counterfactual) state that contained no lands redistributed under the Homestead or Morill acts in 1900 had an expected probability of containing boarding schools of about 0.51 [0.24, 0.77].
+All states with large numbers or contemporary settler land redistribution contained at least one boarding school during this period.
+The expected probability reaches one at about 10 million acres distributed under the Homestead and Morill acts.
+Conditional on operating at least one boarding school, however, these states were similar to each other in the expected count of operating schools across the period (posterior expected count 6.3 [2.7, 12.6]).
+We find no statistically significant difference in the count of schools operating across high settler redistribution lands during this period.
+
+The expected probability of operating an Indian boarding school was not significantly different across states with small and large white urban immigrant populations.
+Given the role of notorious facilities in the Northeast like Carlisle in Pennsylvania, this result is unsurprising.
+We do find that the expected count of facilities in states (conditional on having at least one facility) was generally lower in states with large urban immigrant populations than it was in states with small urban immigrant populations.
+
+Boarding schools were relatively rare in the Jim Crow South.
+We find that states with large populations of Black tenant farmers had a near zero probability of operating an Indian boarding school in 1900.
+There were boarding schools operating in states like North Carolina, Arkansas, and Virginia during this era, but the majority of states in the South did not have any operating boarding schools between 1900 and 1930.
+
+```{r pr-bs, fig.cap = "Probability of having at least one boarding school in 1900 across contexts, US states.
+Posterior expected values, dotted line (mean), credible intervals shaded (0.5, 0.8, 0.95)."}
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = seq(0, max(mdat[mdat$year==1900, "homestead_plus_morill"]),
+                                        length.out = 100),
+            urban_immig_pc = 0,
+            blk_sharecrop_pc = 0,
+            year = c(1900)) %>% 
+  add_epred_draws(bs_m1_z,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  group_by(year, homestead_plus_morill) %>% 
+  summarize(
+    med = median(.epred),
+    lwr = quantile(.epred, 0.05),
+    upr = quantile(.epred, 0.95))
+p1<-ggplot(pred_dat,
+       aes(y = .epred , x = homestead_plus_morill / 1e6)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6) +
+  guides(fill = "none") +
+  theme_bw() +
+  # scale_x_log10() + 
+  labs(y = "Probability",
+       x = "Acres (millions)",
+       subtitle = "1. Homestead + Morill") + 
+  coord_cartesian(ylim = c(0,1))
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = 0,
+            urban_immig_pc = seq(0, max(mdat$urban_immig_pc), length.out = 100),
+            blk_sharecrop_pc = 0,
+            year = c(1900)) %>% 
+  add_epred_draws(bs_m1_z,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  group_by(year, urban_immig_pc) %>% 
+  summarize(
+    med = median(.epred),
+    lwr = quantile(.epred, 0.05),
+    upr = quantile(.epred, 0.95))
+p2<-ggplot(pred_dat,
+       aes(y = .epred , x = urban_immig_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6)  +
+  guides(fill = "none") +
+  theme_bw() +
+  scale_x_log10() + 
+  labs(y = "",
+       x = "Proportion",
+       subtitle = "2. White urban immigrants")+ 
+  coord_cartesian(ylim = c(0,1))
+pred_dat<-expand_grid(tot_pop = 1e6,
+            homestead_plus_morill = 0,
+            urban_immig_pc = 0,
+            blk_sharecrop_pc =  seq(0, max(mdat$blk_sharecrop_pc), length.out = 100),
+            year = c(1900)) %>% 
+  add_epred_draws(bs_m1_z,
+                  re_formula = NA)
+t<-pred_dat %>% 
+  group_by(year, blk_sharecrop_pc) %>% 
+  summarize(
+    med = median(.epred),
+    lwr = quantile(.epred, 0.05),
+    upr = quantile(.epred, 0.95))
+p3<-ggplot(pred_dat,
+       aes(y = .epred , x = blk_sharecrop_pc)) +
+  stat_lineribbon(lty = 2,
+                  linewidth = 0.5,
+                  .width = 0.95,
+                  fill = "grey", alpha = 0.6)  +
+  guides(fill = "none") +
+  theme_bw() +
+  scale_x_log10() + 
+  labs(y = "",
+       x = "Proportion",
+       subtitle = "3. Black tenant farmers")+ 
+  coord_cartesian(ylim = c(0,1))
+grid.arrange(p1, p2, p3,
+             nrow = 1)
+```
+
+# Discussion
+
+The federal government, state governments, and local governments adopted and developed different configurations of welfare and carceral infrastructures in ways that closely relate to the geography of US settler colonial projects during the Progressive Era.
+HIGH LEVEL SUMMARY - NAME THE EMPIRICAL GROUPINGS!
+EXPLAIN HOW THEY RELATE THEORETICALLY TO SETTLER PROJECTS VIS A VIS A FUNCTIONALIST LOGIC (then apologize for being functionalist)
+
+States in the Jim Crow South, where settler white elites were deeply invested in the exploitation of Black labor and in a brutal regime of racial apartheid and terror, had relatively weak front-line social policy bureaucracies (reflected in low employment levels for teachers, police, and social workers) throughout this period.
+They also had less physical infrastructure characteristic of the early US welfare state, like poorhouses and mental institutions.
+However, these states tended to have incarceration levels near the national average.
+For all measures *except* incarceration, Jim Crow states exhibited relatively thin social policy states.
+These findings align with historical scholarship that has carefully documented the simultaneous brutality and ubiquity of a hybrid formal / informal regime of anti-Black racial terror with the routine underfunding and/or failure to develop core welfare state institutions across formerly slave states (CITE QUADAGNO, KATZNELSON, TOLNAY, BAKER, PERKINSON, OTHERS).
+
+Nation building and racial demographics were front-of-mind for settler elites in states where Native lands had been recently expropriated.
+A variety of initiatives had been undertaken by the federal government and by local policy makers to rapidly attract white settlers to these territories and to develop the social, cultural, and economic infrastructure that would fulfill and reinforce national settler colonial ideologies (REVISE).
+In states containing large amounts of Native lands granted to settlers and/or land-grant Universities, there were dramatically higher levels of teacher employment than in other states, while levels of police and social worker employment were comparable to other contemporary US states.
+The physical infrastructure of the carceral and welfare state was similar to other states, with one exception.
+While rates of confinement in correctional, mental, and proto-welfare institutions were similar to those in other states, these states were dramatically more likely than others to contain Indian boarding schools.
+Taken together, these findings suggest that states engaged in the early stages of Native land redistribution and white settlement were engaged in a trajectory of social state formation that was similar to other US states, with two exceptions.
+First, these states were heavily invested in education, suggesting aggressive efforts to build the social and cultural fabric of a settler nation in recently settled lands.
+Second, these states were heavily invested in the elimination of spatially proximate Native nations through the forced assimilation of their children.
+This dual approach, characteristic of settler colonial projects of elimination and replacement, was on full display in states in the Midwest, Mountain West, Southwest, Great Plains, and West Coast.
+
+While Herrenvolk democracy prevailed in Jim Crow and Frontier (IS THIS THE RIGHT TERM???
+I DON'T THINK SO, BUT WE NEED A CONCISE LABEL) states, these regimes developed distinct patterns of governance and policy infrastructure.
+CHECK OUT MCQUADE ON HERRENVOLK WELFARE STATE.
+THIS IS KEY TO NAIL
+
+States with large urban immigrant populations developed bureaucratically complex therapeutic-welfare states with significant physical infrastructure for the confinement of socially problematic populations (that was nevertheless intensely racially exclusive CITE SIMMONS HARTMAN ETC).
+This therapeutic and paternalistic model (which co-developed with nascent social and psychological disciplines) has become the ideal typical form of the modern welfare-carceral state in dominant historical and social scientific accounts (e.g.
+GARLAND, FOUCAULT, OTHERS).
+However, it was far from the only model articulated during the US Progressive era.
+As XXX HAS NOTED IN THEIR CRITIQUE OF FOUCAULT / GARLAND, ETC THE SOUTH WAS REALLY DIFFERENT, BUT STILL MODERN.
+RACIAL CAPITALISM CRITIQUES SUGGEST THAT UNIVERSALIST NARRATIVES ARE INCOMPLETE - RACIST LOGICS AND PROJECTS RESULTED IN HETEROGENEITY IN STATE BUILDING.
+BUT EVEN THESE CRITIQUES TYPICALLY FAIL TO ACCOUNT FOR THE MODEL THAT WAS DEVELOPED BY STATES SEEKING TO ELIMINATE NATIVE NATIONS AND BUILD NASCENT WHITE SETTLER NATIONS.
+
+FUTURE WORK SHOULD TRACE THESE PATTERNS MOVING FORWARD.
+DIRECT FOLLOW UP IS TO DO A TRAJECTORY ANALYSIS ON THESE OUTCOMES.
+To what extent did these foundations influence trajectories?
+To what extent do trajectories impact present patterns?
+How did they become durable?
+How did they converge?
+Role of federalism?
+Lots of questions here.
+But we know that there are HUGE differences across states today, and in many cases, they mirror the kinds of patterns we observe here.
+We should better appreciate the deeply complicated causal processes that have created variation in logics and infrastructures for carceral and welfare states across places today.
+
+These are three of the projects that were critical to settler colonialism in the early 20th c US.
+Not the only ones of course!
+Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILIPPINES COLONIZATION, AND OTHERS).
+*Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*
+
+AK AND HI EXCLUDED, EVEN THOUGH SETTLER COLONIZATION WAS V IMPORTANT THERE!
+THE DATA AND LAW ARE V DIFFERENT BECAUSE OF STATEHOOD TIMING.
+ALSO SHOULD CONSIDER OTHER COLONIES OF US LIKE PR, GUAM, USVI, PHILLIPINES, ETC

### 2024-03-06 model_fit_diagnostics.R

@@ -1,226 +1,0 @@
-## Diagnostics for model fit
-# focus on zero inflated processes
-# and parsimony / convergence trade offs
-# using default brms priors for ease
-
-mdat<-pop %>% 
-  left_join(dist) %>% 
-  left_join(bs)
-
-
-###### teachers - easiest outcome given power
-### baseline model as reported in eq (1) in ms
-t_m1<-brm(
-  teachers ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    scale(I(urban_immig_pop / tot_pop)) +
-    scale(morill_acres_cum + homestead_acres_cum) +
-    log(children) + 
-    factor(year) + 
-    (1|STUSPS) + 
-    offset(log(tot_pop)),
-  family = negbinomial(),
-  data = mdat
-) 
-
-### add regime zero inflation / threshold adjustments
-t_m2<-brm(
-  teachers ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    I(blk / tot_pop > 0.3) +  
-    scale(I(urban_immig_pop / tot_pop)) +
-    I(urban_pop>2e5) + 
-    scale(morill_acres_cum + homestead_acres_cum) +
-    I((morill_acres_cum + homestead_acres_cum)>0) + 
-    log(children) + 
-    factor(year) + 
-    (1|STUSPS) + 
-    offset(log(tot_pop)),
-  family = negbinomial(),
-  data = mdat
-) 
-
-loo_compare(loo(t_m1), loo(t_m2))
-
-#### cops
-
-p_m1<-brm(
-  police ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    scale(I(urban_immig_pop / tot_pop)) +
-    scale(morill_acres_cum + homestead_acres_cum) +
-    log(children) + 
-    factor(year) + 
-    (1|STUSPS) + 
-    offset(log(tot_pop)),
-  family = negbinomial(),
-  data = mdat
-) 
-
-### add regime zero inflation / threshold adjustments
-p_m2<-brm(
-  police ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    I(blk / tot_pop > 0.3) +  
-    scale(I(urban_immig_pop / tot_pop)) +
-    I(urban_pop>2e5) + 
-    scale(morill_acres_cum + homestead_acres_cum) +
-    I((morill_acres_cum + homestead_acres_cum)>0) + 
[…]

### 2024-03-06 models.R

@@ -2,11 +2,13 @@
 
 mdat<-pop %>% 
   left_join(dist) %>% 
-  left_join(bs)
+  left_join(bs) %>% 
+  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
+         urban_immig_pc = urban_immig_pop / tot_pop,
+         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)
 
 #### set weakly informative priors for model parameters
-### b ~ N(0, 20)
-### use brms defaults for variance params (PEG THEM DOWN)
+### for m1
 
 priors_m1<-c(
   prior(normal(0, 10), class = b),
@@ -14,15 +16,12 @@ priors_m1<-c(
   prior(student_t(3, 0, 2.5), class = sd),
   prior(gamma(0.01, 0.01), class = shape))
 
-
-
 ### fit HMC model with brms
-teachers_m1_bayes<-brm(
+teachers_m1<-brm(
   teachers ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    scale(I(urban_immig_pop / tot_pop)) +
-    scale(morill_acres_cum + homestead_acres_cum) +
-    log(children) + 
+    scale(blk_sharecrop_pc) +
+    scale(urban_immig_pc) +
+    scale(homestead_plus_morill) +
     factor(year) + 
     (1|STUSPS) + 
     offset(log(tot_pop)),
@@ -31,20 +30,11 @@ teachers_m1_bayes<-brm(
   data = mdat
 ) 
 
-
-### prior visual
-# teachers_m1_bayes %>% 
-#   prior_summary() %>% 
-#   parse_dist(prior) %>% 
-#   ggplot(aes(y = class, dist = .dist, args = .args)) + 
-#   stat_dist_halfeye()
-
-cops_m1_bayes<-brm(
+cops_m1<-brm(
   police ~ 
-    scale(I(blk_sharecrop / tot_pop)) +
-    scale(I(urban_immig_pop / tot_pop)) +
-    scale(morill_acres_cum + homestead_acres_cum) +
-    log(children) + 
+    scale(blk_sharecrop_pc) +
+    scale(urban_immig_pc) +
+    scale(homestead_plus_morill) +
     factor(year) + 
     (1|STUSPS) + 
     offset(log(tot_pop)),
@@ -52,12 +42,11 @@ cops_m1_bayes<-brm(
   data = mdat
 )
 
[…]

### 2024-03-06 read_bs.R

@@ -2,6 +2,7 @@ library(tidyverse)
 
 bs<-read_csv("./data/Boarding_Schools.csv")
 
+
 ### create row for each boarding-school-year
 ### using start end-pair
 
@@ -34,6 +35,8 @@ for(i in 1:nrow(bs)){
   }
 
 bs_ts<-bind_rows(dat_out)
+
+write_csv(bs_ts, "doi_school_index.csv")
 
 bs_state_ts<-bs_ts %>% 
   group_by(state, year) %>% 

### 2024-04-23 ms.Rmd
ff

@@
 library(trackdown)
-# trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
+trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
 # trackdown::update_file(file = "ms.Rmd",
@@
 #                        rich_text = T)
+trackdown::download_file(file = "ms.Rmd")
+
 options(tigris_use_cache = T)

### 2024-06-11 ms_backend.Rmd
revision ff

--- 2024-03-06 ms.Rmd
@@
 ---
-title: "Settler colonial projects explain variation in US state formation in the early 20th c."
+title: "Simultaneous Settler Projects and U.S. Racial State Formation
+METHODS AND FINDINGS"
 author: "Frank Edwards and Theresa Rocha Beardall"
@@
 library(knitr)
-library(trackdown)
-# trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
-# trackdown::update_file(file = "ms.Rmd", 
-#                        hide_code = T,
-#                        rich_text = T)
+# library(trackdown)
+# # trackdown::trackdown_auth_configure(path = "client_secret_243891325977-kgvbep6m6eb8t2feb8rvu9fneo8d6msa.apps.googleusercontent.com.json")
+# # trackdown::update_file(file = "ms.Rmd", 
+# #                        hide_code = T,
+# #                        rich_text = T)
 options(tigris_use_cache = T)
@@
 ```{r run-models, cache = T, include = F, message = F}
-source("models.R")
-```
-
-# Introduction
-
-Settler colonialism deeply and unevenly impacts the development of social policy systems.
-Settler colonial regimes engage in diverse forms of dispossession, extraction, exploitation, and social reproduction.
-We show that spatial and temporal variation in the salience of settler projects is closely related to the kinds of infrastructure that settler colonial states develop for social control and social reproduction.
-We show that the formation of the social state in the early 20th century in the US varied dramatically across places as a function of geographic variation in the salience of a set of interests of the settler colonial project: land dispossession, labor exploitation, and reproduction of the settler polity.
-
-We examine three deeply intertwined, but spatially and temporally variable, contexts for local policymaking in the Progressive Era (1900 - 1930): the Jim Crow agricultural economy; mass urbanization and industrialization; and the rapid distribution of Native lands to white settlers.
-Each of these contexts presented different sets of problems to local and national settler policy elites.
-
-Under sharecropping regimes in the US South, white settler elites sought to create and reproduce brutal and paternalistic regimes of informal and formal social regulation that kept Black workers and their families dependent on and subordinate to white interests.
-Industrial and social white elites in the North sought to discipline large numbers of European immigrants toward nascent ideals of a white American settler polity oriented toward a narrow gendered division of labor and subordination to the interests of industrial capital.
-In 1862, the federal government enacted the Morill, Homestead, and Pacific Railroad Acts, each of which provided mechanisms for the rapid distribution of dispossessed Native lands to settlers.
-White settlers and settler corporate interests flooded recently dispossessed Native lands with the promise of free or cheap title, seeking to rapidly develop the settler nation in recently admitted states.
-
-We suggest that each of these contexts yielded different trajectories for the formation of US state infrastructure.
-In the analysis that follows, we leverage data from a range of archival sources to map contexts for early 20th century settler colonization across the US, then build statistical models to evaluate whether these contexts were related to the kinds of capacities that local governments developed for contemporaneous social control and social reproduction.
-
-The foundations upon which states built out their infrastructure to control, educate, and care for their populations were closely related to regional variation in the political economy of settler colonialism across regions in the US.
-States developed more or less robust institutions in various domains to address the problems of governance posed by the local projects of settler colonization that both elites and ordinary settlers sought to advance.
-
-We focus our attention on three interconnected settler colonial structures and processes that had large impacts on social life in the early 20th century US: 1) the active dispossession and erasure of Native nations and the development of settler enterprises and communities; 2) the exploitation of Black labor for agricultural production and the fortification of an apartheid social order; and 3) the coercive and paternalistic assimilation of recent immigrants from Eastern and Southern Europe into an urban industrial settler colonial polity.
-We argue that these deeply related, but geographically uneven, processes were related to the development of state capacities to resolve emerging problems of domestic governance in a period of rapid social transformation.
-In more general terms, we argue that the local salience and configuration of key social structures, like settler colonization, are closely related to local trajectories for state formation and institution building.
-
-Consistent with the historiography on the Progressive Era (GORDON, FOX, OTHERS), we find clear relationships between place and policy infrastructure.
-State formation trajectories varied dramatically across the US, and this variation is closely related to the local salience of distinct racialized settler colonial projects.
-Our work bridges these sometimes disparate historical accounts by showing that the local interests of settler elites were closely tied to the kinds of social and racial projects that the early carceral and welfare states engaged in.
-
-# Three settler colonial projects in the early 20th c. US
-
-The geographic distribution of the local implementation of the three focal settler projects we focus on is crudely illustrated in Figure \@ref(fig:three-maps).
-The federal government was most heavily involved in actively redistributing Native lands to settlers in states in the Midwest, Great Plains, Mountain West, and on the West Coast.
-While Native land dispossession was still deeply salient in Eastern states, questions of settler land ownership and the establishment of coherent settler polities had been resolved (albeit incompletely) prior to the late 19th and early 20th centuries.
-In panel 1, we show the number of acres that had been dispossessed from Native nations and distributed to settlers and settler institutions through two key laws that heavily impacted patterns of colonial settlement in the late 19th and early 20th centuries, the Morill and Homestead acts of 1862.
-
-```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1910"}
-### ADD DISPOSESSION
-p1 <- dist %>% 
-  select(STUSPS, year,
-         morill_acres_cum, homestead_acres_cum,
-         total_acres_2021) %>% 
-  filter(year == 1910) %>% 
-  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
-  left_join(st) %>% 
-  ggplot(aes(fill = dist / 1e6,
-             geometry = geometry)) + 
-  geom_sf() + 
-  theme_void() + 
-  scale_fill_viridis_c() + 
-  theme(legend.position = "bottom") + 
-  labs(fill = "Acres\n(millions)",
-       subtitle = "1. Lands patented through\nMorill and Homestead acts")
-p2 <- st %>% 
-  left_join(pop %>% 
-              filter(year == 1910)) %>% 
-  ggplot(aes(fill = blk / tot_pop)) + 
-  geom_sf() + 
-  theme_void() + 
-  scale_fill_viridis_c() + 
-  theme(legend.position = "bottom") + 
-  labs(fill = "Proportion",
-       subtitle = "2. Black population")
-p3 <- st %>% 
-  left_join(pop %>% 
-              filter(year == 1910)) %>% 
-  ggplot(aes(fill = urban_immig_pop / tot_pop)) + 
-  geom_sf() + 
-  theme_void() + 
-  scale_fill_viridis_c() + 
-  theme(legend.position = "bottom") + 
-  labs(fill = "Proportion",
-       subtitle = "3. Urban immigrant population")
-grid.arrange(p1, p2, p3, nrow = 1)
-```
-
-The Jim Crow system of apartheid was developed and became deeply entrenched across the states in the US South that had built elaborate settler plantation economies through the exploitation and terrorization of formerly enslaved Black people.
-As of 1910, the first wave of the Great Migration of African Americans out of the US South was underway, but, as shown in Panel 2 of Figure \@ref(fig:three-maps), the Black population in the US was still heavily concentrated in the US Southeast.
-
-And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America.
-These immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts.
-Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US.
-As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast were becoming increasingly urban and foreign born.
+# source("models.R")
+mdat<-pop %>% 
+  left_join(dist) %>% 
+  left_join(bs) %>% 
+  mutate(blk_sharecrop_pc = blk_sharecrop / tot_pop,
+         urban_immig_pc = urban_immig_pop / tot_pop,
+         homestead_plus_morill = homestead_acres_cum + morill_acres_cum)
+
+#### set weakly informative priors for model parameters
+### for m1
+### try a growth model
+mdat<-mdat %>% 
+  mutate(year_c = (year - 1900)/10)
+
+teachers_m1<-readRDS("./models/teachers_m1.rds")
+cops_m1<-readRDS("./models/cops_m1.rds")
+socwork_m1<-readRDS("./models/socwork_m1.rds")
+prison_m1<-readRDS("./models/prison_m1.rds")
+inst_m1<-readRDS("./models/inst_m1.rds")
+bs_m1_z<-readRDS("./models/bs_m1_z.rds")
+bs_m1<-readRDS("./models/bs_m1.rds")
+```
+
+# Analytic strategy
+
+We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels.
+We use data on US states between 1900 and 1930 derived from US Census and archival sources.
+
+Our approach is theory-driven, informed by 1) RACIAL STATE THEORY and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD.
+We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts.
+
+Generally our approach suggests that social policy infrastructure in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.
+
+$E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})$
+
+We prioritize a descriptive approach that illustrates differences in macro-social contexts across places and their associations with state-formation outcomes and trajectories.
+We describe and evaluate the geographic and temporal co-occurrence of state formation outcomes and discrete settler projects.
+
+*While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (CITATIONS) we refrain from engaging in formal causal inference in this analysis.
+Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes.
+We suggest that these processes are visible at a relatively high-level in the historical data.
+However, settler colonialism is endogenous to the whole of US politics, economics, and demography.* TIGHTEN THIS
+
+We proceed by producing and evaluating descriptive data visuals and maps of our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time.
+We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures.
+We pay particular attention to regional patterns in these distributions, and situate our interpretations in dialogue with prior historical and theoretical research.
+
+Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national and regional temporal trends, and basic characteristics of the population in the state and year.
+These models formally evaluate associations between the geographies of settler colonial projects and a range of policy outcomes.
+
+A replication package is available at github.com/f-edwards/racial_politics_path
+
+## Data
+
+This study leverages historical data from a range of sources.
+We rely on data from the decennial US Census (1900 - 1930) for information on population features, employment, and group quarters populations.
+Information on the timing and location of Indigenous land disposession is obtained from shape files constructed from maps of Indigenous land dispossession through treaties and acts of congress compiled by the US Congress in 1899 (Royce Maps).
+Data on the distribution of Native lands to white settlers under the authority of the Homestead Act (1862) are obtained from federal archives of land patents.
+Data on the distribution of Native lands to US State governments for the construction of land-grant universities under the authority of the Morill Act (1862) was compiled by High Country News from federal archives (CITE).
+Finally, we leverage data on Indian Boarding Schools compiled by The Department of the Interior under their Federal Indian Boarding School Initiative (CITE 2022).
+
+### Population composition and employment
+
+All measures on key features of the US population are derived from the 1900, 1910, 1920, and 1930 decennial US Census 100% samples, accessed through IPUMS (Ruggles et al.).
+We use complete decennial census microdata for each decennial year to reduce the potential of measurement error for relatively small state-level sub-populations of interest.
+For each census year and for each of the lower 48 US states, we count the total population, the child population (as the population under 18 years old), the white non-immigrant population (as the total population identified as non-Hispanic white with both parents 'native-born'), the white immigrant population (as the population identified as non-Hispanic white either foreign born or with at least one foreign born parent), the white urban immigrant population (as the white immigrant population living in a household identified as urban), and the Black population as the total population identified as Black, and the Black tenant farm population as the total population identified as Black living in households identified as farms and where the ownership of the dwelling is identified as rented.
+
+State-year employment measures are also computed from the 1900 - 1930 Census 100 percent microdata files using the variable 'Occupation, 1950 basis' harmonized for earlier Census years by IPUMS staff.
+A census respondent is counted as a police employee if their occupation is coded as any of the following categories 'Marshals and constables'; 'Policemen and detectives'; or 'Sheriffs and bailiffs'.
+A respondent is counted as a social worker if their occupation is coded as 'Social and welfare worker'; and counted as a teacher if coded as 'Teachers (n.e.c)', which excludes music, art, and post-secondary teachers but includes other primary and secondary teachers.
+We compute counts for each of these employment categories for each state and decennial Census year.
+
+### Group quarters populations
+
+We also compute measures of populations in institutional group quarters, but rely on a different set of US Census samples.
+The 1910 census does not include information on group quarters type.
+For these data, we rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type.
+From these samples, we derive state-year measures of the population housed in correctional institutions; the population housed in mental institutions and the population housed in institutions for the elderly, handicapped, and poor (a single category in Census files).
+Because recorded populations housed in institutions for the elderly, handicapped, and poor are  sparse in some places and years, we combine these counts with populations hosed in mental institutions.
+This sum of populations housed in paternalistic institutions serves as a proxy measure of the development of what has been described as the 'therapeutic' state (CITE POLSKY) or the 'white parental state' (CITE WARD), or the precursors to the modern institutions of the welfare state (CITE KATZ).
+
+### Indigenous land disposession and settler redistribution
+
+Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress known as the 'Royce Maps', which detail lands claimed by the US Government from Native nations by either treaty or act of Congress  (CITE LOC U.S.
+Serial Set, Numbers 4014 and 4015).
+These maps provide information on the timing and location of Indigenous dispossession by the US government between 1784 and 1894, which we use to visualize the historical geography of settler westward expansion in the late 19th century.
+
+The redistribution of Native lands to settlers by the US federal government is measured through data on the implementation of the Homestead and Morrill Acts of 1862.
+We use data on parcels of land transferred to land-grant universities under the Morrill Act as documented by Robert Lee and High Country News as part of their 'Land Grab University' project.
+These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available (CITE LGU REPORT).
+Data on parcels of land distributed to settlers through the Homestead Act are obtained from the report ``Homesteads,'' by the Bureau of Land Management of the U.S.
+Department of the Interior, 1962, and digitized and distributed by Edwards, Friefeld and and Wingo (2017).
+For both Homestead and Morill act data, we measure cumulative acres of Native lands parceled since 1862 under each act within each state by year.
+
+## Statistical methods and models
+
+We develop a series of regression models to explore the conditional relationships between outcomes measuring the capacity of settler state governments and predictors measuring the contexts for three focal settler colonial projects.
+These state-year models evaluate how three regionally specific settler projects -- Jim Crow, Native land redistribution, and white immigrant assimilation -- related to patterns in the development of core state infrastructure and capacity during a period of rapid increases in state capacity in the US.
+
+Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts.
+We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters.
+This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states.
+
+We discuss our process for model selection in detail in the appendix.
+
+Most of our infrastructure outcomes are measured as counts of persons.
+As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes.
+The population offset term on the left-hand side of equation XXXX allows us to adjust outcomes and estimated relationships for the size of state populations (as per capita rates).
+
+**(1) Regression model for count outcomes**
+
+$y \sim \textrm{NegBin}(\lambda, \alpha)$
+
+$\lambda = \textrm{E}[y]$
+
+$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$
+
+$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$
+
+$\gamma \sim N(0, \sigma^2_\gamma)$
+
+With weakly informative priors
+
+$\beta \sim N(0, 20)$
+
+$\sigma \sim t(3, 0, 2.5)$
+
+$\alpha \sim \textrm{Gamma}(0.01, 0.01)$
+
+We also model the presence and number of Indian boarding schools in each state as a focal measure of state infrastructure in the Progressive era.
+Unlike other outcomes in our study, there are a substantial number of states that did not have Indian boarding schools during this period.
+We treat the presence of boarding schools as a zero-inflated process, and use a two-stage model to evaluate their distribution.
+One component of the model estimates the probability of having at least one boarding school during the focal period (a binary outcome).
+This binary measure of boarding schools is constant over the period 1900 - 1930 at the state-level, so we estimate this model using data only for the first observation in the period, 1900, following the model in equation (2).
+The second component of the model estimates the count of boarding schools present in a state-year, conditional on there being at least one boarding school in the state, following equation (1), excluding the population offset term.
+
+**(2) Regression model for binary outcomes**
+
+$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $
+$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $
+$\beta \sim N(0, 20)$
+
+For all regression models, we transform predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence in a context of widely differing scales across measures.
+
+# Findings
+
+We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states.
+In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity.
 
@@
 
-## Building the settler nation-state
-
-KEY CHARARCTERISTICS: RECENT OR ONGOING DISPOSESSION / REMOVAL, RECENTLY ADMITTED STATES, RELATIVELY SPARSE WHITE POPS. ELIMINATION / REPLACEMENT AS KEY DYNAMIC.
-
-In places where land speculation and homesteading fueled rapid white settlement, governments and churches hired large numbers of teachers for the intertwined goals of building a white settler polity on the 'frontier' and to continue the coerced assimilation of Indigenous children through a notorious and brutal system of boarding schools.
+## The geography of settler colonial projects in the late 19th and early 20th century US
+
+Figures \@ref(fig:cessions-map) shows the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong).
+These maps show a clear difference in the timing of land cessions as a function of place.
+By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee.
+Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century.
+Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West.
+By 1895, the overwhelming majority of lands in the contemporary lower 48 US states had been claimed by the US federal government.
 
@@
   labs(fill = "Year") 
-
-```
-
-Figures \@ref(fig:cessions-map) and \@ref(fig:cessions-ts) show the temporal and spatial process of the dispossession of lands from Native Nations by the US Government between 1795 and 1893 (DATA FROM USFS VIA Lib Of Cong).
-These maps show a clear difference in the timing of land cessions as a function of place.
-By 1795, the US Government claimed nearly all lands East of the Appalachian Mountains, including large parts of contemporary Kentucky and Tennessee.
-Land dispossession throughout the South and the Great Lakes regions accelerated quickly during the early 19th century.
-Throughout the middle of the 19th century, there were large-scale efforts by the federal government and vigilante settlers to dispossess tribal lands in California, the Midwest, Great Plains, and Mountain West.
-By 1895, nearly all lands in the contemporary lower 48 US states had been claimed by the federal government as US territory.
-
-```{r cessions-ts, fig.cap = "Cumulative disposession of Native lands in the 19th c. by US region and state as proportion of contemporary state area. Each line represents one state."}
-# p1<-ggplot(disp,
-#        aes(x = year,
-#            group = STUSPS)) + 
-#   geom_line(aes(y = cumulative_ceded / 1e6)) + 
-#   labs(x = "Year",
-#        y = "Acres (millions)") + 
-#   lims(y = c(0, 150)) + 
-#   facet_wrap(region~division) + 
-#   labs(subtitle = "1. Total area")
-p2<-ggplot(disp %>% 
-             filter(year > 1795),
-       aes(x = year,
-           group = STUSPS)) + 
-  geom_line(aes(y = prop_ceded)) + 
-  labs(x = "Year",
-       y = "Proportion of state area") + 
-  facet_wrap(region~division) 
-p2
-```
-
-The Federal government passed a series of laws in 1862 designed to rapidly redistribute dispossessed Native lands to white settlers and to large corporate and state institutions.
-The Homestead, Morill, and Pacific Railway Acts of 1862 provided legal frameworks through which expropriated Native lands could be rapidly redistributed to white settlers, large corporations, and Universities.
-Figure \@ref(fig:hs-morill-map) shows the total number of acres within a state that had been granted to settlers and settler institutions by 1930.
-More than 10 million acres of Native lands had been redistributed to US land grant Universities at little or no cost for their use or liquidation by 1930 under the Morill act.
-Under the Homestead Act, more than 255 million acres of Native land had been transferred to settlers by 1930.
-States in the Midwest, Great Plains, Mountain West, and West Coast saw the highest volumes of land redistribution under these laws.
-
-```{r hs-morill-map, fig.cap = "Lands patented under two legislative mechanisms for the redistribution of disposessed Native lands to settlers, cumulative total 1862 - 1930"}
-pdat<-disp %>% 
-  select(STUSPS, total_acres_2021) %>% 
-  distinct() %>% 
-  left_join(dist %>% 
+```
+
+The 1900 geographic distribution of the three focal settler projects we focus on is illustrated in Figure \@ref(fig:three-maps).
+This set of maps shows illustrates the clear differences in the settler political economy that were most salient in distinct regions of the early 20th century US.
+
+The extensive and methodical dispossession of Native lands that characterized the Assimilation and Allotment era was in full force during the Progressive Era.
+Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1900.
+Nearly 90 million acres of Native lands had been granted to settlers and states through the Morill and Homestead acts by 1900.
+By 1930, this total would reach over 265 million acres.
+By 1930, 47 percent of the contemporary area of Nebraska, and 44 percent of the contemporary area of North Dakota had been granted to settlers under these programs.
+More than 30 percent of the contemporary area of South Dakota, Oklahoma, Colorado, and Montana had been granted to settlers under these programs by 1930.
+These programs primarily impacted lands in the US West, Southwest, Mountain West, and Midwest during this period.
+
+```{r three-maps, fig.cap = "Three focal contexts for settler colonial governance, 1930"}
+### ADD DISPOSESSION
+p1 <- dist %>% 
+  select(STUSPS, year,
+         morill_acres_cum, homestead_acres_cum,
+         total_acres_2021) %>% 
   filter(year == 1930) %>% 
-              select(STUSPS, morill_acres_cum, homestead_acres_cum)) %>% 
-  replace_na(list(morill_acres_cum = 0, homestead_acres_cum = 0)) %>% 
-  left_join(st)
-p1<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = morill_acres_cum / 1e6)) + 
+  mutate(dist = morill_acres_cum + homestead_acres_cum) %>% 
+  left_join(st) %>% 
+  ggplot(aes(fill = dist / 1e6,
+             geometry = geometry)) + 
+  geom_sf() + 
+  theme_void() + 
   scale_fill_viridis_c() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "Acres\n(millions)",
+       subtitle = "1. Lands patented through\nMorill and Homestead acts")
+p2 <- st %>% 
+  left_join(pop %>% 
+              filter(year == 1930)) %>% 
+  ggplot(aes(fill = blk_sharecrop / tot_pop * 100)) + 
+  geom_sf() + 
   theme_void() + 
-  labs(fill = "Acres (millions)",
-       subtitle = "1. Morill Act") + 
-  theme(legend.position = "bottom")
-
-p2<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = homestead_acres_cum / 1e6)) + 
   scale_fill_viridis_c() + 
+  theme(legend.position = "bottom") + 
+  labs(fill = "% of\npop.",
+       subtitle = "2. Percent Black \ntenant farmers")
+p3 <- st %>% 
+  left_join(pop %>% 
+              filter(year == 1930)) %>% 
+  ggplot(aes(fill = urban_immig_pop / tot_pop* 100)) + 
+  geom_sf() + 
   theme_void() + 
-  labs(fill = "Acres (millions)",
-       subtitle = "2. Homestead Act") + 
-  theme(legend.position = "bottom")
-grid.arrange(p1, p2, ncol = 2)
-```
-
-## Jim Crow
-
-KEY CHARACTERISTICS: ANTI-BLACK SOCIAL, LEGAL, ECONOMIC REGIME. WHITE SETTLER HOMOGENEITY
-
-By 1877 efforts to build effective racial democracy through post-Civil War Reconstruction were abandoned by the federal government.
-By the turn of the century, white elites had legislated and entrenched a regime of brutal racial exploitation and apartheid across the US South (DU BOIS, FONER, FRANKLIN, CURTIN).
-
-DEFINE JIM CROW REGIME. APARTHEID, FORMAL / INFORMAL TERROR REGIME, BLACK LABOR EXPLOITATION, PRIMARILY AG through SHARECROPPING.
-
-READ APD CITATIONS FROM ANN REVIEWS ON JIM CROW, APD, RACISM. DOES APD SUPPORT THE WEAK STATE THESIS?
-
-In places where tenant farming by Black families was common, there was relatively little investment in state bureaucracies.
-Social control and reproduction instead relied on a coordinated formal and informal system of racial terrorism and apartheid that developed alongside a relatively weak set of state social policy institutions (QUADAGNO, KATZNELSON).
-
-Figure \@ref(fig:jim-crow-maps) shows two key demographic features of this period.
-In panel 1, we illustrate the entrenchment and ubiquity of the sharecropping system of racially exploitative agricultural production.
-In panel 2, we show that, unlike much of the rest of the United States at the time, the white population of the US South was composed overwhelmingly of white settlers who had been born in the United States.
-
-```{r jim-crow-maps, fig.cap = "Black tenant farming and immigrant status of white settlers, 1910"}
-pdat<-st %>% 
-  left_join(pop) %>% 
-  filter(year == 1910)
-p1<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = blk_sharecrop / 1e5)) + 
   scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Population\n(hundred thousands)",
-       subtitle = "1. Black population living on tenant farms") + 
-  theme(legend.position = "bottom")
-p2<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = white_immig / (white_immig + white_non_immig))) + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Proportion",
-        subtitle = "2. Proportion of white population foreign born") + 
-  theme(legend.position = "bottom")
-grid.arrange(p1, p2, nrow = 1)
-```
-
-## Regulating and incorporating urban white immigrants
-
-KEY CHARACTERISTICS: URBANIZATION, INDUSTRIALIZATION, MORAL PANICS OVER IMMIGRANT CULTURAL DIFFERENCE. (TWO ROUTES, REACTIONARY DISCIPLINE, LIBERAL WELFARE PATERNALISM, BOTH ACTIVE).
-
-In the rapidly urbanizing portions of the Northeast, Midwest, and West Coast, we see heavy investment in what has become the ideal type of the Progressive social state: a robust and professionalizing police force and a broad paternalistic welfare state.
-These institutions were developed by elites, in part, to regulate urban disorder discursively linked to growing populations of European immigrants and Black migrants from the US South, and with rapidly assimilating these groups into the norms and ideologies of industrial capitalism and the settler nation.
-
-The industrialization of the US Northeast and Midwest during the late 19th and early 20th centuries spurred two related demographic processes: rapid urbanization and mass European immigration.
-Figure \@ref(fig:immig-maps) displays the distribution of both the urban population and the white immigrant population in the US in 1910.
-States like New York, Massachusetts, Pennsylvania, Ohio, and Illinois saw dramatic growth in their populations, with much of that growth composed of European immigrants and concentrated in cities like Chicago, Boston, Cleveland, Philadelphia, and New York.
-
-```{r immig-maps, fig.cap = "White immigrant population and urban population, 1910"}
-pdat<-st %>% 
-  left_join(pop) %>% 
-  filter(year == 1910)
-p1<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = white_immig / 1e6)) + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Population\n(millions)",
-       subtitle = "1. White, foreign-born") + 
-  theme(legend.position = "bottom")
-p2<-ggplot(pdat) + 
-  geom_sf(aes(geometry = geometry,
-              fill = urban_pop / 1e6)) + 
-  scale_fill_viridis_c() + 
-  theme_void() + 
-  labs(fill = "Population\n(millions)",
-        subtitle = "2. Urban") + 
-  theme(legend.position = "bottom")
-grid.arrange(p1, p2, nrow = 1)
-```
-
-Progressive reformers in the Northeast and Midwest were preoccupied with solving a battery of social problems associated with urban life, industrial capitalism, and their perceptions of the culture and behavior of Southern and Eastern European immigrants.
-
-HIGH LEVEL SUMMARY OF 1st WAVE WHITE FEMINISM AND SOCIAL WORK VIA GORDON, HARTMAN, OTHERS.
-
-The emerging fields of sociology, social work, and criminology also sought to apply scientific principles to the organization and practice of policing and corrections.
-HIGH LEVEL SUMMARY OF PROG ERA POLICING, JUV COURT, PRISON REFORM.
-LINK TO GOTTSCHALK ON CARCERAL STATE EMERGENCE FROM THESE MOVEMENTS.
-
-# Patterns in the development of social policy infrastructure 1900 - 1930
-
-We define social state capacity as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a detailed review of theory and measurement).
+  theme(legend.position = "bottom") + 
+  labs(fill = "% of\npop.",
+       subtitle = "3. Percent white \nurban immigrant")
+
+grid.arrange(p1, p2, p3, nrow = 1)
+```
+
+The Jim Crow system of apartheid and its economic complement the sharecropping system were developed and became deeply rooted across the states in the US South during the Progressive era.
+Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, which was still a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest.
+In 1930, 644,000 Black residents of Mississippi were living on tenant farms, 32 percent of the state's total population.
+More than 400,000 Black people were recorded as residents of tenant famrs in Georgia, Alabama, and South Carolina in 1930.
+More than 300,000 were enumerated in North Carolina, Louisiana, and Texas.
+And more than 100,000 were enumerated in Arkansas, Tennessee, and Virginia.
+
+And the Industrial Revolution, coupled with a wave of crises across Europe, created both strong push and pull forces for large populations of immigrants to travel to North America.
+Southern and Eastern European immigrants settled across the industrial cities of the Northeast, Midwest, and West coasts.
+Many of them also joined the waves of white settlers pursuing the opportunity to claim recently dispossessed Native lands in the Central and Western portions of the US.
+As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast had become increasingly urban and foreign born.
+Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants.
+In Illinois, 1.1 million of the state's 7.6 million residents were first or second generation immigrants from Europe.
+In Massachusetts and Rhode Island, 23 percent of the each state's population were European immigrants in 1930.
+More than 10 percent of the population were European immigrants in New Jersey, Connecticut, Michigan, New Hampshire, California, and Pennsylvania.
+
+## The geography of social policy infrastructure 1900 - 1930
+
+We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general 'state capacity').
 Capacity is a function of both physical infrastructure and human resources available to state actors.
-We focus on a broad set of social policy systems including education, social welfare, and criminal justice.
-
-We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key street-level social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples.
+During the Progressive Era, federal, state, and local governments developed their abilities to intervene into and manage social life through a series of increasingly professionalized bureaucracies.
+We suggest that these nascent components of the carceral and welfare states developed in markedly different ways across places.
+To capture some of this heterogeneity, we focus our analysis on a employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice.
+
+We rely on three sets of measures associated with the capacity of states to accomplish social policy goals: 1) employment rates in key social policy positions as a measure of the relative size of bureaucracies and magnitude of human effort available to policy actors, 2) populations housed in institutional group quarters as a measure of the scope of particular kinds of physical infrastructure used by the welfare and carceral systems during the Progressive Era, and 3) counts of operating Indian boarding schools as indicators of physical infrastructure directed at coerced assimilation and erasure of Indigenous peoples.
 Taken together, these three sets of measures provide a diverse and broad set of insights into spending on the basic infrastructure of social governance during the early 20th century.
 
-## Employment in education, policing, and social welfare: 1900 - 1930
-
-There was dramatic growth in public sector employment in general (CLEAR FROM FED GOVT EMPLOYMENT, OTHER CITATIONS), and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era.
+### Employment in education, policing, and social welfare: 1900 - 1930
+
+There was dramatic growth in public sector employment in general, and in roles targeted at nation building, social control, and social welfare in particular during the Progressive Era (CITATIONS!
+CHECK NAIDU).
 *These dramatic increases in employment rates were reflected in the increasing complexity and reach of a wide range of public institutions described by a number of historians (GENERAL LIT ON GOVT COMPLEXITY AND ARTICULATION 1900 - 1930).*
@@
 Outside of a few cities in the northeast, social workers were almost never recorded in the Census in 1900 (only 361 recorded nationally in 1900).
-However, by 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita.
-CHECK THESE NUMBERS
+By 1930, the Census recorded about 17,000 social workers nationally, an increase of about 30 fold in terms of employment rates per capita.
+However, this expansion of the reach of state and state-aligned street-level bureaucrats into social life did not occur uniformly across the country.
+Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States.
 
@@
 
-However, this expansion of the reach of state and state-aligned street-level bureaucrats into social life did not occur uniformly across the country.
-Figure \@ref(fig:emp-maps) displays employment rates per 1,000 population in 1930 for teachers, police, and social workers for each of the lower 48 US States.
+Employment rates for teachers were highest in a cluster of Midwestern and Mountain West states.
+Thirteen states had teacher employment rates of over 10 per 1,000 population, all in the Midwest or West.
+North and South Dakota had highest-in-the-nation teacher employment rates, at over 14 per 1,000 persons.
+Teachers were instrumental in building human capital and national culture in settler homesteading communities.
+They also served as front-line agents of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS).
 Employment rates for teachers were lowest-in-the nation in the US South.
-Employment rates for teachers were highest in a cluster of Great Plains states, like the Dakotas, Wyoming, Nebraska, Iowa, and Kansas.
-Employment rates for police were at their peak in 1930 in Northeastern, Midwestern, and West Coast states like New York, New Jersey, Massachusetts, Illinois, and California.
-Police employment rates were relatively lower in places like Arkansas, Mississippi, and North Carolina.
-The distribution of social workers in 1930 shares some features with the distribution of police: states in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment, while states in the Great Plains and South, like Mississippi and South Dakota recorded very few social workers in the 1930 Census.
+There were more than twice as many teachers per capita in North Dakota in 1930 than there were in Alabama.
+In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer thatn 7 teacher per 1,000 residents.
+Southern states in general systematically underfunded or avoided funding public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE).
+
+Employment rates for police in 1930 were highest in Northeastern, Midwestern, and West Coast states with large urban hubs.
+New York had the highest numbers of recorded police, at nearly 30,000 identified in the 1930 Census (about 2.4 per 1,000 population).
+New Jersey and Massachusetts also employed more than 2 officers per 1000 population in 1930.
+The NYPD, and other big-city police agencies had undertook efforts at professionalization and bureaucratization during the Progressive Era (e.g.
+Guarigilia, OTHERS), resulting in substantial growth in many of the places with large police organizations.
+By contrast, the largely rural South under a public/private hybrid model of racial terror and segregation during Jim Crow employed relatively fewer police officers.
+In Arkansas, Mississippi, North Carolina, and Alabama, there were fewer than 0.6 per 1,000 police recorded in the 1930 Census.
 
@@
   filter(year == 1930)
+
 p1<-ggplot(pdat) + 
@@
   theme_void() + 
-  labs(fill = "Rate",
+  labs(fill = "Rate per\n 1000 pop.",
        subtitle = "1. Teachers") + 
@@
   theme_void() + 
-  labs(fill = "Rate",
+  labs(fill = "",
         subtitle = "2. Police") + 
@@
   theme_void() + 
-  labs(fill = "Rate",
+  labs(fill = "",
         subtitle = "3. Social Workers") + 
@@
 
-```
-
-## Physical infrastructure: prisons, jails, asylums, orphanages and poorhouses
-
-The capacity of state welfare and carceral systems is also a function of their ability to engage in therapeutic, correctional, or paternalistic confinement of socially problematic populations.
-In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states.
-Using the 1930 Census, we display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3.
+### text callouts
+temp<-pop %>% 
+  filter(year == 1930) %>% 
+  mutate(police_pc = police / tot_pop * 1e3,
+         teachers_pc = teachers / tot_pop * 1e3,
+         socwork_pc = socwork / tot_pop * 1e3) %>% 
+  select(STUSPS, police:socwork, police_pc:socwork_pc) %>% 
+  arrange(desc(teachers_pc))
+
+### make a prop change map? or delta rate map?
+
+```
+
+The distribution of social workers in 1930 shares some features with the distribution of police: states with large cities in the Northeast, Midwest, and West coast like New York, Ohio, Massachusetts, California, and Minnesota had high rates of social worker employment.
+These places were among the first to see the emergence of core Progressive social welfare institutions like settlement houses and juvenile courts, and saw rapid growth in social work employment.
+By 1930, five states were employing at least 0.2 social workers per 1,000 population: Massachusetts, New York, California, Rhode Island, and Ohio.
+As with education, the South exhbited very low levels of social worker employment during this period.
+In Mississippi, only 24 social workers were recorded in the 1930 Census, an employment rate of about 1.2 per 100,000 population.
+Social workers were also rare in many of the states with high levels of homesteading like the Dakotas and Oklahoma.
+
+### Physical infrastructure: prisons, jails, asylums, orphanages and poorhouses
+
+US welfare and carceral systems have historically prioritized developing physical infrastructure to confine and/or treat populations understood to be socially problematic (CITES ON CONFINEMENT).
+In Figure \@ref(fig:gq-map) we show the distribution of the population housed in institutional group quarters in 1930 across US states as recorded by the US Census to provide one measure of physical infrastructure for social welfare and social control.
+We display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3.
 Note that data quality issues in the 1910 Census limit the possibilities for time series analysis.
 
-Panel 3 displays the rate of confinement in institutions for the poor and disabled across US states in 1930.
-These institutions included poorhouses, settlement homes, orphan asylums, sanitariums, and similar institutions targeted at containing and/or supporting social groups understood to be 'dependent' on state or religious charity support.
-We find that rates of confinement (likely closely correlated with capacity) were lowest in states in the South, like Arkansas, Alabama, Mississippi and Florida.
-New England, the mid-Atlantic, and Great Lakes states had relatively high rates of confinement in these kinds of institutions.
-The East North Central Division, including states like Wisconsin, Illinois, and Ohio had the highest average rate of confinement, at 4.3 per 1,000 persons, while the West South Central division (Arkansas, Louisiana, Oklahoma, and Texas) had the lowest rate of confinement at 1.4 per 1,000 persons in 1930.
-
-Panel 2 displays regional patterns in confinement in institutions for the mentally ill across places in 1930.
-States in the mid Atlantic, New England, and West Coast had the highest rates of confinement, at over 3 per 1,000 on average.
-In New Hampshire and Massachusetts, confinement rates in 1930  exceeded 4 per 1,000 population.
-States in the East South Central Division (Mississippi, Alabama, Tennessee, and Kentucky) had the lowest average levels of confinement, at 1.4 per 1,000 persons.
-Levels of confinement ranged between 2 and 3 on average across the rest of the country.
-Of note, we observe noinmates in mental institutitons in North Dakota in the 1930 census data.
-
-Panel 1 of Figure \@ref{gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census.
-We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population.
-Confinement rates were highest on average in the West Coast states (2.5 per 1,000).
-Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population.
-Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.
-
-Whereas states in the South tended to have very low levels of confinement in institutions for the poor and disabled and relatively low levels of confinement in mental institutions, they had levels of confinement in correctional institutions that were near national averages.
-This suggests that Southern policy makers and philanthropic allies systematically invested fewer resources in physical infrastructure targeted at containing or supporting the poor, young, mentally ill, and disabled than did other states, while maintaining and developing prisons and jails at rates similar to the rest of the nation.
-
-```{r gq-map, fig.cap = "Population confined in prisons, mental institutions, and institutions for the poor and disabled, 1930"}
+```{r gq-map, fig.cap = "Population confined in prisons and jails, mental institutions, and institutions for the poor and disabled, 1930"}
 pdat<-st %>% 
@@
   theme_void() +
-  labs(fill = "Rate",
-       subtitle = "1. Correctional institutions") +
+  labs(fill = "Rate per\n1,000 pop.",
+       subtitle = "1. Correctional") +
   theme(legend.position = "bottom")
@@
   theme_void() +
-  labs(fill = "Rate",
-        subtitle = "2. Mental institutions") +
+  labs(fill = "",
+        subtitle = "2. Mental") +
   theme(legend.position = "bottom")
@@
   theme_void() +
-  labs(fill = "Rate",
-        subtitle = "3. Institutions for the\npoor and disabled") +
+  labs(fill = "",
+        subtitle = "3. Poor and disabled") +
   theme(legend.position = "bottom")
 grid.arrange(p1, p2, p3, nrow = 1)
-
 # for in-text callouts
@@
          inst_poor_disabled_pc = inst_poor_disabled / tot_pop * 1e3) 
-
 pdat_callout<-pdat %>% 
@@
   select(-geometry)
-```
-
-## Physical infrastructure: Indian boarding schools
-
-The Indian boarding school epitomizes the sustained effort by the US government and its allies in settler religious and philanthropic organizations to eliminate Native nations through the coercive assimilation of Indigenous children.
+
+temp<-pdat %>% 
+  ungroup() %>% 
+  select(-geometry) %>% 
+  select(STUSPS, inst_correctional_pc:inst_poor_disabled_pc)
+
+```
+
+Panel 3 displays the rate of confinement in institutions for the poor and disabled across US states in 1930.
+These institutions included poorhouses, settlement homes, orphan asylums, sanitariums, and similar institutions targeted at containing and/or supporting social groups understood to be 'dependent' on state or religious charity support.
+We find that rates of confinement were lowest in states in the South, like Arkansas, Alabama, Mississippi and Florida.
+New England, the mid-Atlantic, and Great Lakes states had relatively high rates of confinement in these kinds of institutions.
+Delaware, Wisconsin, and Massachusetts led the nation in populations housed in institutions for the poor and disabled, each recording at least 5 residents per 1,000 population in 1930.
+By contrast, these institutions were less developed in the South.
+Arkansas, Alabama, Florida, Georgia, Texas, and Mississippi each recorded about 1 resident in institutions for the poor and disabled per 1,000 population.
+
+Panel 2 displays regional patterns in confinement in institutions for the mentally ill across places in 1930.
+States in the mid Atlantic, New England, and West Coast had the highest rates of confinement, at over 3 per 1,000 on average.
+In Vermont, Massachusetts, New York, and New Hampshire confinement rates in 1930  exceeded 4 per 1,000 population.
+States in the East South Central Division (Mississippi, Alabama, Tennessee, and Kentucky) had the lowest average levels of confinement, at 1.4 per 1,000 persons.
+Levels of confinement ranged between 2 and 3 on average across the rest of the country.
+Of note, we observe no inmates in mental institutions in North Dakota in the 1930 census data.
+
+Panel 1 of Figure \@ref{gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census.
+We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population.
+Confinement rates were highest on average in the West Coast states (2.5 per 1,000).
+Average rates across the remaining regions varied between 1.3 and 2.0 per 1,000 population.
+Kansas, home to federal and military prisons in Leavenworth, had the highest incarceration rates in the nation, at 4.5 persons per 1,000 population.
+
+Whereas states in the South tended to have very low levels of confinement in institutions for the poor and disabled and relatively low levels of confinement in mental institutions, they had levels of confinement in correctional institutions that were near national averages.
+This suggests that Southern policy makers and philanthropic allies systematically invested fewer resources in physical infrastructure targeted at containing or supporting the poor, young, mentally ill, and disabled than did other states, while maintaining and developing prisons and jails at rates similar to the rest of the nation.
+DISCUSS WARD BRIEFLY HERE.
+
+### Physical infrastructure: Indian boarding schools
+
+Indian boarding schools were built by the US government and its allies in to eliminate Native nations through the coercive assimilation of Indigenous children (Adams 1995; Hoxie 2001).
 Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934.
-Notably, these carceral and genocidal institutions represented an early, substantial, and sustained investment by the federal government in education infrastructure.
-
+Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure.
 Below, we provide a descriptive overview of the geography and timing of boarding school operations.
@@
 We rely on data produced by the Department of the Interior (NEWLAND 2022) for historical counts of boarding schools in operation by state and year.
-We pair these data with information coded by our research team from archived annual reports produced by the Commissioner on Indian Affairs between 1890 and 1920 (CITATIONS).
 
@@
   left_join(st)
-
 ggplot(bs_ts %>% 
@@
 
-# Analytic strategy
-
-To explore the relationship between settler colonial political and economic contexts and state formation, we evaluate whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels.
-We conduct exploratory and inferential analysis using data on US states between 1900 and 1930.
-
-Generally our approach suggests that social policy infrastructure in a particular place *s* and time *t* is a function of settler elite interests specific to that time and place.
-
-\[E[\textrm{infrastructure}_{st}] = f(\textrm{settler interests}_{st})\]
-
-We proceed by producing and evaluating descriptive data visuals for our operational measures of settler colonial policy contexts and comparing distributions of these measures across states and over time.
-We then produce and evaluate a similar set of visuals to evaluate the distribution of our social policy infrastructure measures.
-Finally, we estimate regression models that evaluate the relationships between our measures of infrastructure and settler interests, conditional on time-stable place variation, national temporal trends, and basic characteristics of the population in the state and year.
-
-A replication package is available at github.com/f-edwards/racial_politics_path
-
-## Data
-
-All measures on features of the US population are derived from the 1900, 1910, 1920, and 1930 decennial US Census 100% samples, accessed through IPUMS (Ruggles et al.).
-For each census year and for each of the lower 48 US states, we count the total population, the child population (as the population under 18 years old), the white urban immigrant population (as the population identified as foreign born, white, and non-Hispanic living in an urban household), and the Black population living in tenant farm households.
-
-State-year employment measures are computed from the 1900 - 1930 Census.
-A census respondent is counted as a police employee if their code for the IPUMS variable OCC1950 is equal to one of {771 'Marshals and constables', 773 'Policemen and detectives', 782 'Sheriffs and bailiffs'}; is counted as a social worker if OCC1950 is coded 079 'Social and welfare worker'; and counted as a teacher if OCC1950 is equal to 093 'Teachers (n.e.c)', which excludes music and art teachers but includes other primary and secondary teachers.
-
-We also compute measures of populations in institutional group quarters, but rely on a slightly different set of samples.
-The 1910 census did not collect information on group quarters type.
-We rely on the 1900, 1920, and 1930 100% samples for the group quarters population by type, and the 1910 1.4% sample which contains IPUMS imputations for group quarters type.
-From these samples, we derive state-year measures of the population housed in correctional institutions, the population housed in mental institutions, and the population housed in institutions for the elderly, handicapped, and poor.
-
-Indigenous land dispossession over time is measured using shape files created by the US Forest Service from a set of 1899 maps detailing the boundaries of land cessions claimed by the US government from Native nations through treaties or acts of Congress (CITE).
-The distribution of Native lands to settlers is measured through data on the implementation of the Homestead Act and Morrill Act of 1862.
-We use data on parcels of land transferred to land-grant universities under the Morrill Act as documented by Robert Lee and High Country News as part of their 'Land Grab University' project.
-These data are compiled from land patents archived by the Bureau of Land Management and represent the most extensive archive of such land transfers currently available.
-Data on parcels of land distributed to settlers through the Homestead Act are obtained from  the report ``Homesteads,'' by the Bureau of Land Management of the U.S.
-Department of the Interior, 1962, and digitized and distributed by Edwards, Friefeld and and Wingo 2017.
-For both Homestead and Morill act data, we measure cumulative acres parceled since 1862 under each act within the state by year.
-
-## A note on settler colonial geography
-
-geography itself is endogenous to settler colonization, this is tricky!
-We are trying to understand the emergence of the settler state as a bureaucratic entity in this period, so it makes sense to do this, but it is messy!
-we don't want to imply a teleological reading here.
-borders are unstable and contested.
-Here we are thinking about the human and physical infrastructures of the US carceral and welfare states, so we focus our attention on those spatial boundaries, while recognizing that the process of state formation is closely tied up with defining and reinforcing those borders while seeking to erase others.
-
-## Statistical methods and models
-
-We develop a series of regression models to explore the conditional relationships between outcomes measuring the capacity of settler state governments and predictors measuring the contexts for three focal settler colonial projects.
-These state-year models evaluate how three regionally specific settler projects -- Jim Crow, Native land redistribution, and white immigrant assimilation -- related to patterns in the development of core state infrastructure and capacity during a period of rapid increases in state capacity in the US.
-
-Our first set of models (see equation XXX) predict infrastructure outcomes for each decennial Census year between 1900 and 1930, and include time-stable state-level random intercepts, a non-linear time component (a national decade-specific intercept), and parameters for per capita rates of Black tenant farming, per capita rates of white urban immigrants, and cumulative acreage redistributed to settlers under the Homestead and Morill acts.
-We use Bayesian Hamiltonian Monte Carlo methods to estimate all models with weakly informative priors on all model parameters.
-This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states.
-
-We discuss our process for model selection in detail in the appendix.
-
-Most of our infrastructure outcomes are measured as counts of persons.
-As such, we rely on Negative Binomial likelihoods that can flexibly handle over-dispersed count outcomes to model these employment and population processes.
-The population offset term on the left-hand side of equation XXXX allows us to adjust outcomes and estimated relationships for the size of state populations (as per capita rates).
-
-*Equation 1: Model for count outcomes*
-
-\[ y_{st} \sim \textrm{NegBin}(\lambda, \alpha)\]
-
-\[ \lambda = \textrm{E}[y_{st}] \]
-
-\[ \log \left( \frac{\textrm{E}[y_{st}]}{\textrm{population}_{st}} \right) = \gamma_s + \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} + \beta_2 \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} + \beta_3 \textrm{Morill and Homestead acres}_{st} \]
-
-\[\gamma_s \sim N(0, \sigma^2_\gamma)\]
-
-With weakly informative priors
-
-\[\beta \sim N(0, 20)\]
-\[\sigma \sim t(3, 0, 2.5)\]
-\[\alpha \sim \textrm{Gamma}(0.01, 0.01)\]
-
-We also model the presence and number of Indian boarding schools in each state as a focal measure of state infrastructure in the Progressive era.
-Unlike other outcomes in our study, there are a substantial number of states that did not have Indian boarding schools during this period.
-We treat the presence of boarding schools as a zero-inflated process, and use a two-stage model to evaluate their distribution.
-One component of the model estimates the probability of having at least one boarding school during the focal period (a binary outcome).
-This binary measure of boarding schools is constant over the period 1900 - 1930 at the state-level, so we estimate this model using data only for the first observation in the period, 1900.
-The second component of the model estimates the count of boarding schools present in a state-year, conditional on there being at least one boarding school in the state, following equation 1, with the removal of the population offset term.
-
-*Model 2 (for zero inflated outcomes) is*
-
-\[ y_{s} > 0 \sim \textrm{Bernoulli}\left(\theta\right) \]
-\[ \textrm{logit} \left(\theta_s\right) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} \]
-\[\beta \sim N(0, 20)\]
-
-Across all models, we re-scale predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence and parameter interpretability in a context of widely differing scales across measures.
-
-# Findings
-
-We suggest that regionally-specific political contexts, driven by spatially and temporally varying interests of white settler elites, caused variation in trajectories of welfare state formation across US states.
-In this study, we explore how three settler colonial political projects mattered for the emergence of the US welfare state: the economic exploitation and social exclusion of Black people from the settler polity, the consolidation of settler political, social, and cultural claims on Indigenous territories, and the incorporation of urban European immigrants into the settler polity.
-
-There are clear relationships between the geography of settler colonial projects and the local scale of welfare and carceral state capacity in the early 20th century US.
+## Model results: relationships between settler colonial projects and social policy infrastructure
+
 In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators.
@@
               mutate(model = "Asylums and poorhouses"))
-
 ggplot(t,

### 2024-06-11 make_pop_data.R

@@ -71,3 +71,6 @@ dat <- data.frame(state = state.abb,
   
 dat %>% 
   write_csv("./data/temp_population_ts.csv")
+
+
+

### 2024-06-11 models.R

@@ -9,6 +9,9 @@ mdat<-pop %>%
 
 #### set weakly informative priors for model parameters
 ### for m1
+### try a growth model
+mdat<-mdat %>% 
+  mutate(year_c = (year - 1900)/10)
 
 priors_m1<-c(
   prior(normal(0, 10), class = b),
@@ -22,13 +25,102 @@ teachers_m1<-brm(
     scale(blk_sharecrop_pc) +
     scale(urban_immig_pc) +
     scale(homestead_plus_morill) +
-    factor(year) + 
+    year_c + 
     (1|STUSPS) + 
     offset(log(tot_pop)),
   family = negbinomial(),
   prior = priors_m1,
   data = mdat
 ) 
+### try a growth model
+mdat<-mdat %>% 
+  mutate(year_c = (year - 1900)/10)
+
+teachers_m2<-brm(
+  teachers ~ 
+    (scale(blk_sharecrop_pc) +
+    scale(urban_immig_pc) +
+    scale(homestead_plus_morill)) *
+    factor(year_c) + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  prior = priors_m1,
+  data = mdat) 
+# context interaction (slopes by predictor type)
+teachers_m3<-brm(
+  teachers ~ 
+    (scale(blk_sharecrop_pc) +
+       scale(urban_immig_pc) +
+       scale(homestead_plus_morill)) *
+    year_c + 
+    (1|STUSPS) + 
+    offset(log(tot_pop)),
+  family = negbinomial(),
+  prior = priors_m1,
+  data = mdat) 
+
+## compare m1 / m3, clear performance gain for m3
+# loo_compare(loo(teachers_m1), loo(teachers_m3))
+### adding year RE doesn't help.
+
+loo_compare(loo(teachers_m3), loo(teachers_m4))
+
+
+#### START HERE!!!!
+### TIME VARYING SLOPES ON CONTEXT
+### DOES INTERACTION ON SHARECROPPING PERSIST?
+### simulate
+
+year_c = 0:7
+STUSPS <- "XX"
+tot_pop = 1e6
+
+pred_dat<-data.frame(
+  scen = c("South", "Northeast", "West"),
+  blk_sharecrop_pc = quantile(mdat$blk_sharecrop_pc, c(0.9, 0, 0)),
+  urban_immig_pc = quantile(mdat$urban_immig_pc, c(0, 0.9, 0)),
[…]

### 2024-06-28 ms_backend.Rmd
fast forward before merge with front end

@@
 # read data from make_* scripts
-pop<-read_csv("./data/temp_population_ts.csv")%>%
+pop<-read_csv("./data/population_ts_full.csv")%>%
   filter(STUSPS!="AK")
@@
 bs<-read_csv("./data/temp_bs_ts.csv")
-agents<-read_csv("./data/temp_agents.csv")
 ## add region to all
@@
   distinct()) %>% 
-  left_join(bs) %>% 
-  left_join(agents %>% 
-              select(-state))
+  left_join(bs) 
 dist<-dist %>%
@@
   distinct()) %>% 
-  left_join(bs) %>% 
-  left_join(agents %>% 
-              select(-state))
+  left_join(bs) 
 ```
@@
 ```{r run-models, cache = T, include = F, message = F}
-# source("models.R")
+
 mdat<-pop %>% 
@@
 
-#### set weakly informative priors for model parameters
-### for m1
-### try a growth model
+# rescale year
 mdat<-mdat %>% 
   mutate(year_c = (year - 1900)/10)
+
+# source("models.R")
 
@@
 prison_m1<-readRDS("./models/prison_m1.rds")
-inst_m1<-readRDS("./models/inst_m1.rds")
+mental_m1<-readRDS("./models/inst_m1.rds")
+poor_m1<-readRDS("./models/poor_m1.rds")
 bs_m1_z<-readRDS("./models/bs_m1_z.rds")
-bs_m1<-readRDS("./models/bs_m1.rds")
 ```
@@
 
-We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures of policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels.
+We describe the relationship between settler colonial political and economic contexts and state formation by evaluating whether a set of measures capturing social policy infrastructure are related to a set of measures that are related to locally-specific settler colonial policy contexts at the state and year-levels.
 We use data on US states between 1900 and 1930 derived from US Census and archival sources.
 
-Our approach is theory-driven, informed by 1) RACIAL STATE THEORY and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD.
+While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (ACHARYA) we refrain from engaging in formal causal inference in this analysis.
+Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes.
+We suggest that these processes are visible at a relatively high-level in the historical data.
+Settler colonialism is endogenous to the whole of US politics, economics, and demography.
+Our interest here is not to identify causal effects, but rather to descriptively chart how the historical geography of settler colonialism in the US overlaps with the historical geography of US state formation.
+Typical approaches in causal inference seek to eliminate stable cross-sectional differences in units.
+Our approach prioritizes understanding both these stable cross-sectional differences and dynamics of change over time.
+Toward this end, we pursue a descriptive and interpretive macro-sociological approach to understanding how settler contexts related to heterogeneity in state formation in the early 20th century.
+
+Our approach is theory-driven, informed by insights derived from 1) RACIAL STATE and SETTLER COLONIAL THEORIES OF US STATE FORMATION and 2) the HISTORIOGRAPHY OF EACH REGION IN THIS PERIOD.
 We suggest that RACIAL STATE THEORY provides a coherent framework for understanding heterogeneity in state formation during this period, by applying the simultaneous settler projects framework to these distinct contexts.
@@
 We describe and evaluate the geographic and temporal co-occurrence of state formation outcomes and discrete settler projects.
-
-*While there have been a number of compelling analyses of the long-run effects of variation in colonial systems on colonized societies (ACEMOGLU ETC) and of chattel slavery and racial terror on contemporary politics (CITATIONS) we refrain from engaging in formal causal inference in this analysis.
-Rich theoretical and historical work suggests that racial and settler colonial processes deeply structure policy outcomes through complex and long-running processes that become obscured and entangled with a battery of social processes.
-We suggest that these processes are visible at a relatively high-level in the historical data.
-However, settler colonialism is endogenous to the whole of US politics, economics, and demography.* TIGHTEN THIS
 
@@
 This approach adjusts for national time trends (secular increases in employment across places), as well as adjusting for idiosyncratic differences across states.
+Note that the log link function of our negative binomial likelihoods implies multiplicative relationships between additive focal predictors and outcomes, resulting in rates of linear change that vary depending on starting values (intercepts).
 
@@
 
-$y \sim \textrm{NegBin}(\lambda, \alpha)$
-
-$\lambda = \textrm{E}[y]$
-
-$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$
-
-$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$
-
-$\gamma \sim N(0, \sigma^2_\gamma)$
+$$y \sim \textrm{NegBin}(\lambda, \alpha)$$
+
+$$\lambda = \textrm{E}[y]$$
+
+$$\log \left( \frac{\textrm{E}[y]}{\textrm{population}} \right) = \gamma_s + \beta \textrm{X}_{st}$$
+
+$$x_{st} = t \times \left(\frac{\textrm{Black tenant farmers}_{st}}{\textrm{population}_{st}} +  \frac{\textrm{White urban immigrants}_{st}}{\textrm{population}_{st}} +\textrm{Morill and Homestead acres}_{st}\right)$$
+
+$$\gamma \sim N(0, \sigma^2_\gamma)$$
 
@@
 
-$\beta \sim N(0, 20)$
-
-$\sigma \sim t(3, 0, 2.5)$
-
-$\alpha \sim \textrm{Gamma}(0.01, 0.01)$
+$$\beta \sim N(0, 20)$$
+
+$$\sigma \sim t(3, 0, 2.5)$$
+
+$$\alpha \sim \textrm{Gamma}(0.01, 0.01)$$
 
@@
 
-$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $
-$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $
-$\beta \sim N(0, 20)$
+$$ y > 0 \sim \textrm{Bernoulli}\left(\theta\right) $$
+$$ \textrm{logit} (\theta) =  \beta_{0t} + \beta_1 \frac{\textrm{Black tenant farmers}_{s}}{\textrm{population}_{s}} + \beta_2 \frac{\textrm{White urban immigrants}_{s}}{\textrm{population}_{s}} + \beta_3 \textrm{Morill and Homestead acres}_{s} $$
+$$\beta \sim N(0, 20)$$
 
 For all regression models, we transform predictors to a mean-centered standard deviation scale (*z* scores) to improve model convergence in a context of widely differing scales across measures.
+More details on model selection are provided in the supplemental materials.
 
@@
 The extensive and methodical dispossession of Native lands that characterized the Assimilation and Allotment era was in full force during the Progressive Era.
-Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1900.
+Panel 1 of Figure \@ref(fig:three-maps) shows the cumulative area of dispossessed Native lands that had been redistributed to settlers and US states at low or no cost under the Homestead and Morill Acts of 1862 by 1930.
 Nearly 90 million acres of Native lands had been granted to settlers and states through the Morill and Homestead acts by 1900.
@@
 
-The Jim Crow system of apartheid and its economic complement the sharecropping system were developed and became deeply rooted across the states in the US South during the Progressive era.
-Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, which was still a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest.
+The Jim Crow system of apartheid and its economic complement the sharecropping system were deeply rooted across the states in the US South during the Progressive era.
+Panel 2 of Figure \@ref(fig:three-maps) shows the entrenchment of the sharecropping system in 1930, a dominant mode of racialized political economy in the US South several decades into the first Great Migration of African Americans out of the US South to industrial cities in the Northeast and Midwest.
 In 1930, 644,000 Black residents of Mississippi were living on tenant farms, 32 percent of the state's total population.
-More than 400,000 Black people were recorded as residents of tenant famrs in Georgia, Alabama, and South Carolina in 1930.
+More than 400,000 Black people were recorded as residents of tenant farms in Georgia, Alabama, and South Carolina in 1930.
 More than 300,000 were enumerated in North Carolina, Louisiana, and Texas.
@@
 As shown in Panel 3 of Figure \@ref(fig:three-maps), the populations of many states in the Northeast, Midwest, and West Coast had become increasingly urban and foreign born.
-Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants.
+Nearly 3 million residents of New York state's 12.5 million residents in 1930 were either first or second generation European immigrants (CONFIRM THIS IS URBAN).
 In Illinois, 1.1 million of the state's 7.6 million residents were first or second generation immigrants from Europe.
@@
 
-We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general 'state capacity').
+We define 'social state capacity' as the resources available to governments to direct and intervene in processes of social control and social reproduction (see Cingolani 2013 for a review of theory and measurement of the more general concept 'state capacity').
 Capacity is a function of both physical infrastructure and human resources available to state actors.
@@
 We suggest that these nascent components of the carceral and welfare states developed in markedly different ways across places.
-To capture some of this heterogeneity, we focus our analysis on a employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice.
+To capture some of this heterogeneity, we focus our analysis on employment and infrastructure outcomes in a broad set of social policy systems including education, social welfare, and criminal justice.
 
@@
 Teachers were instrumental in building human capital and national culture in settler homesteading communities.
-They also served as front-line agents of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS).
+They also served as front-line %>% of the assimilationist programs throughout the region's vast network of Indian boarding schools (Marquez 2024, OTHERS).
 Employment rates for teachers were lowest-in-the nation in the US South.
 There were more than twice as many teachers per capita in North Dakota in 1930 than there were in Alabama.
-In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer thatn 7 teacher per 1,000 residents.
-Southern states in general systematically underfunded or avoided funding public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE).
+In Alabama, Arkansas, Kentucky, Missisippi, and Georgia, the 1930 Census recorded fewer than 7 teacher per 1,000 residents.
+Southern states in general systematically underfunded or failed to develop public institutions that could benefit African Americans, and the Jim Crow school system was an exemplar of this practice (CITE).
 
@@
 We display the population housed in prisons in jails in panel 1, the population housed in mental institutions in panel 2, and the population housed in institutions for the poor and/or disabled in figure 3.
-Note that data quality issues in the 1910 Census limit the possibilities for time series analysis.
+Note that alternative measures of group quarters populations are available, but choice of measures does not have a meaningful impact on our substantive conclusions (for a detailed discussion, see Appendix).
 
@@
 
-Panel 1 of Figure \@ref{gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census.
+Panel 1 of Figure \@ref{fig:gq-map} shows state-level variation in confinement in correctional group quarters reported in the 1930 US Census.
 We find average confinement rates to be lowest in the Mid-Atlantic and New England states, at 1.2 per 1,000 population.
@@
 Investment in Indian boarding schools peaked during the Assimilation and Allotment era of Federal Indian Law, roughly spanning 1887 - 1934.
-Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure.
+Notably, these carceral and genocidal institutions represented one of the earliest sustained investments by the federal government in education infrastructure (alongside the Morill act of 1862).
 Below, we provide a descriptive overview of the geography and timing of boarding school operations.
@@
 During the subsequent Reservation era (1850 - 1887), the number of operating boarding schools increased from 23 to 173.
-The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as had existed at the beginning of the period.
+The number of operating facilities continued to increase during the Assimilation era to a maximum of 242 in 1908, followed by a decline to 174 facilities recorded as operating in 1930, a similar level of facilities operating as at the beginning of the period.
 
@@
 There were 8 states with at least 10 facilities in operation in 1907: South Dakota (24), Minnesota (18), Arizona (14), Montana (14), New Mexico (11), Washington (11), Alaska (10, not displayed), and Wisconsin (10).
-There were at least 5 baording schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5).
+There were at least 5 boarding schools in operation in 7 states: California (7), North Dakota (7), Oregon (7), Colorado (5), Hawaii (5, not displayed), Idaho (5), and Kansas (5).
 States with at least one confirmed facility in operation in 1907 included Nebraska (3), Nevada (3), Pennsylvania (3), Utah (3), Wyoming (3), Iowa (2), Michigan (2), Arkansas (1), Illinois (1), Missouri (1), New York (1), North Carolina (1), and Virginia (1).
@@
 ```{r bsmap, fig.cap = "Boarding schools in operation in 1907 by state"}
-bs_ts<-agents %>% 
-  left_join(bs) %>% 
+bs_ts<-bs %>% 
   left_join(st)
@@
 
+We describe the results of our regression models in terms of expected values of each infrastructure outcome described above.
+We compute mean expected values and 95 percent credible intervals for each decennial year 1900 - 1930 from model posterior distributions for each outcome across three scenarios that closely correspond to the regional geography of the US during this period.
+We describe these scenarios in Table \@ref(tab:scen).
+All models include state-specific intercepts and slopes over time to account for other unmeasured features of states that impact both levels of employment and rates of change in employment.
+
+The "High land redistribution" scenario is intended to describe states with a high volume of lands patented under the Morill and Homestead acts.
+These states typically had relatively low levels of Black tenant farmers per capita and white urban immigrants per capita (see Figure \@ref(fig:three-maps).
+For this scenario, we fix the sum of cumulative Homestead Act and Morill Act acres at the 90th percentile of the observed data (12.5 million acres), Black tenant farm residents per capita at the 5th percentile of the observed data (about 2 per 100,000 population), and white urban immigrants per capita at the 5th percentile of the observed data (about 2.6 per 1,000 population).
+This scenario is similar to the observed values of North Dakota, South Dakota, Montana, Colorado, Wyoming, and Kansas.
+
+The "High sharecropping scenario" is intended to describe states with a high volume of Black residents on tenant farms.
+These states, all in the US South, typically had very small urban populations and did not see a high volume of lands parceled under the Homestead and Morill acts.
+For this scenario, we fix Black tenant farm residents per capita at the 90th percentile of the observed data (143 per 1,000 population), white urban immigrant population per capita at the 5th percentile of the observed data (2.6 per 1,000 population), and the sum of Homestead and Morill acres patented at the 5th percentile of the observed data (0 acres).
+This scenario is most similar to Georgia and South Carolina during this period.
+The scenario is also similar to Louisiana, Alabama and Mississippi, though these states did have substantial acres of land patented under the Homestead and Morill acts prior to 1900.
+
+```{r scen, results = "asis", tab.cap= "Scenarios specified for model predictions"}
+year_c = 0:3
+STUSPS <- "XX"
+tot_pop = 1e3
+
+pred_dat<-data.frame(
+  scen = c("High sharecropping", "High urban immigrants", "High land redistribution"),
+  blk_sharecrop_pc = quantile(mdat$blk_sharecrop_pc, c(0.9, 0.05, 0.05)),
+  urban_immig_pc = quantile(mdat$urban_immig_pc, c(0.05, 0.9, 0.05)),
+  homestead_plus_morill = quantile(mdat$homestead_plus_morill, c(0.05, 0.05, 0.9))) %>% 
+  expand_grid(year_c, tot_pop)
+
+tab_dat<-pred_dat %>% 
+  filter(year_c == 0) %>% 
+  distinct() %>% 
+  mutate(blk_sharecrop_pc = blk_sharecrop_pc * 1e3,
+         urban_immig_pc = urban_immig_pc * 1e3,
+         homestead_plus_morill = homestead_plus_morill / 1e6) %>% 
+  select(-year_c, -tot_pop) %>% 
+  rename("Scenario" = scen,
+         "Black tenant\nfarmers per 1,000 pop." = blk_sharecrop_pc,
+         "White urban\nimmigrants per 1,000 pop." = urban_immig_pc,
+         "Homestead + Morill\nacres patented" = homestead_plus_morill)
+
+kable(tab_dat, digits = 2)
+```
+
+The "High urban immigrants" scenario corresponds to state with large urban industrial centers that experienced a high volume of European immigration between 1900 and 1930.
+These states typically had very low numbers of Black tenant farmers during this period.
+States in the Northeast did not see lands patented to settlers under the Homestead and Morill acts.
+However, states on the West Coast and Midwest, some of which had large urban immigrant populations, were exposed to high-levels of homesteading and settler land redistribution during this period.
+For ease of comparison, we set the "High urban immigrant" scenario at the 90th percentile of observed white urban immigrants per capita (176 per 1,000 residents), and the 5th percentile of observed Black tenant farm residents per capita and cumulative Homestead and Morill acres patented.
+This scenario is similar to conditions in Connecticut, Illinois, Massachusetts, New Jersey, New York, and Rhode Island.
+
 In Figure \@ref(fig:emp-m-vis) we display our estimates of the spatial and temporal relationships between employment in key social policy professions and settler colonial project indicators.
-In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between some of the physical infrastructure of the carceral and welfare state and settler colonial projects.
-And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between the locations of Federal Indian boarding schools and these measures of settler colonial project timing and location.
+In Figure \@ref(fig:gq-m-vis) we show our estimates for the relationship between institutional group quarters populations and settler colonial project indicators.
+And in Figure \@ref(fig:bs-m-mvs) we show our estimates of the spatial and temporal relationships between operating Federal Indian boarding schools and these measures of settler colonial projects.
 
@@
 
-PUNCHY HIGH LEVEL SUMMARY. WHAT DO WE CALL THIS? FRONTIER DISPOSSESSION + NATION BUILDING, URBAN REGULATION + IMMIG INCORPORATION, JIM CROW HYBRID FORMAL+INFORMAL RACIAL TERROR + MALIGN NEGLECT.
-
-## Employment models
-
-As shown in Figure \@ref(fig:emp-ts), there was a dramatic increase in the number of people employed as teachers between 1900, when there were `r emp_pdat %>% filter(year==1900, name=="Teachers") %>% select(value) %>% slice(1)` teachers recorded in the US Census, and 1930, when there was a reported total of `r emp_pdat %>% filter(year==1930, name=="Teachers") %>% select(value) %>% slice(1)` teachers.
-This amounts to a `r round((emp_pdat_prop %>% filter(year==1930, name=="Teachers") %>% select(prop_change) %>% slice(1)-1)*100)` percent growth in the per capita rate of teacher employment over a 30 year period.
-We model state-year teacher employment rates using the model described in Equation 1, and display the posterior distributions of focal parameters in Figure \@ref(fig:emp-m-vis) to evaluate how the geography of settler colonial projects related to patterns of teacher employment across places and over time.
-More details on model parameters and diagnostics are provided in Appendix Table (XXXXXX, full regression table).
-
-Panel 1 in Figure \@ref(fig:emp-m-vis) shows the distributions of the posterior parameter estimates of our relationship between three focal settler colonial contexts and teacher employment.
-Conditional on our data and model specifications, we find that states with higher levels of settler land redistribution under the Homestead and Morill acts employed more teachers per capita than did other states between 1900 and 1930.
-We find that states with higher levels of Black tenant farming employed fewer teachers per capita than did other states.
-States with high urban white immigrant populations employed teachers at rates near the national average.
-
-Panel 2 shows the distributions of focal posterior parameter estimates from our models of police employment.
-We find that states with higher levels of settler land distribution through the Morill and Homestead acts had lower levels of police employment on average.
-States with large urban white immigrant populations had significantly higher levels of police employment than did other states during this period.
-And states with large numbers of Black tenant farmers tended to have lower levels of police employment.
-Despite relatively low levels of police employment in many Southern states, there were incredible levels of both formal and informal white supremacist violence during this period (CITATIONS).
-
-The distribution of social work employment across states follows the same general pattern as police employment.
-In Panel 3, we show that states with large white urban immigrant populations had much higher levels of social worker employment than did other US states between 1900 and 1930.
-States with high levels of Black tenant farming and settler land redistribution tended to have slightly lower levels of social worker employment than other states, but these expected employment levels are not statistically significant from zero at conventional thresholds.
-
-```{r emp-m-vis, fig.cap = "Posterior distributions of focal model parameters by employment type, US States 1900 - 1930. Densities, expected values, and credible intervals, 0.5, 0.95)"}
-#### extract posterior draws for b1, b2, b3, set up for plotting
-pdat<-teachers_m1 %>%
+### Employment
+
+There was a dramatic increase in employment in education, policing, and social welfare between 1900 and 1930 nationally (see Figure \@ref(fig:emp-ts)), with substantial differences across regions of the US (see Figure \@ref(fig:emp-maps)).
+We construct a series of regression models to explore how the geography of settler projects related to both levels of employment and growth in employment across states.
+We display key results from these models as posterior expected values for focal scenarios in Figure \@ref(fig:emp-m-vis).
+We find that there was substantial growth in employment across all settler project contexts between 1900 and 1930, initial levels of employment and rates of change in employment varied dramatically across contexts.
+
+```{r emp-m-vis, fig.cap = "Posterior expected employment rates per 1,000 population by employment type across settler project contexts, US States 1900 - 1930.
+Mean posterior expected value (line) and 95 percent credible interval (shaded)."}
+preds_teach<-teachers_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat) %>% 
+  mutate(type = "Teachers")
+
+preds_cops<-cops_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat)%>% 
+  mutate(type = "Police")
+
+preds_socwork<-socwork_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat)%>% 
+  mutate(type = "Social workers")
+
+p1<-ggplot(preds_teach,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
+                  .width = 0.95, 
+                  fill = "gray50",
+                  linewidth = 0.5) +
+  theme_bw() + 
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
+  labs(y = "",
+       x = "",
+       subtitle = "1. Teachers")
+
+p2<-ggplot(preds_cops,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
+                  .width = 0.95, 
+                  fill = "gray50",
+                  linewidth = 0.5) +
+  theme_bw() + 
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
+  labs(y = "Employment rate",
+       x = "",
+       subtitle = "2. Police")
+
+p3<-ggplot(preds_socwork,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
+                  .width = 0.95, 
+                  fill = "gray50",
+                  linewidth = 0.5) +
+  theme_bw() + 
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
+  labs(y = "",
+       x = "Year",
+       subtitle = "3. Social workers")
+
+grid.arrange(p1, p2, p3, nrow = 3)
+```
+
+Across categories of employment, we find clear relationships between settler colonial contexts and levels of employment in welfare and carceral institutions.
+We argue that the rationally and temporally varying interests of settler elites determine the kinds of colonial projects they prioritize, reflected in the kinds of human infrastructure that they develop to pursue policy goals.
+
+#### Teachers
+
+```{r callouts1}
+temp <- preds_teach %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred), 
+            yhat_lwr = quantile(.epred, 0.025), 
+            yhat_upr = quantile(.epred, 0.975)) %>% 
+  arrange(scen, year_c)
+```
+
+Panel 1 of Figure \@ref(fig:emp-m-vis) shows posterior expected values and 95 percent credible intervals for teacher employment across contexts of high settler land redistribution, high levels of sharecropping, and high levels of white urban immigrant population composition for each decennial Census year 1900 - 1930.
+Across all contexts, our models show substantial heterogeneity in baseline state employment levels ($\sigma_\gamma = 0.23 [0.19, 0.29]$) and in growth trajectories ($\sigma_\delta = 0.06 [0.04, 0.08]$).
+
+We find that states with higher levels of settler land redistribution under the Homestead and Morill acts employed more dramatically more teachers per capita than did other states between 1900 and 1930.
+Model results suggest that states undergoing high volumes of settler land redistribution in 1900 had expected teacher employment rates near 7.2 [CI 6.3, 8.1] per 1,000 persons.
+By 1930, expected teacher employment would increase to 10.8 per 1,000 [10.2, 11.4].
+This represented a nearly 50 percent increase in teacher employment over this period of 30 years.
+By 1930, we estimate that high land redistribution states employed teachers at a per capita rate about 40 percent higher than high sharecropping states and about 30 percent higher than high white urban immigrant states.
+
+States with high levels of Black tenant farm residency had lower expected teacher employment than did other states.
+Our model suggests that a high sharecropping state could be expected to have about 3.9 [3.5, 4.4] teachers per 1,000 population in 1900, growing to about 7.5 [6.9, 8.0] teachers per 1,000 in 1930, about a 90 percent increase.
+We find evidence that growth in teacher employment in high sharecropping states was higher than in other contexts.
+Despite this higher-than-average growth, however, expected teacher employment in high sharecropping states in 1930 remained lower on average than in states that did not have widespread use of the sharecropping system.
+Our models suggest that by 1930, high sharecropping states had just caught up to teacher employment levels common in high land redistribution states in 1900.
+
+In states with a high proportion of their population composed of white urban immigrants, our models suggest an expected teacher employment rate of 5.8 [5.2, 6.4] per 1,000 in 1900 and a rate of 8.1 [7.5, 8.7] in 1930, an increase of about 40 percent.
+Our models suggest that these states had slightly lower rates of growth in teacher employment than the national average rate of growth.
+
+#### Police
+
+```{r callouts2}
+temp <- preds_cops %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred), 
+            yhat_lwr = quantile(.epred, 0.025), 
+            yhat_upr = quantile(.epred, 0.975)) %>% 
+  arrange(scen, year_c)
+```
+
+Panel 2 of Figure \@ref(fig:emp-m-vis) shows expected values and credible intervals from our model of police employment across contexts.
+As with our teacher employment model, we find considerable variation in baseline state values ($\sigma_\gamma = 0.33 [0.26, 0.43]$) and in rates of growth ($\sigma_\delta = 0.05 [0.02, 0.07]$).
+
+State with large populations of white urban immigrants had much higher police employment levels than did other states during this period.
+In 1900, our model suggests an expected police employment rate of about 8.0 [6.6, 9.5] per 10,000 population in these states.
+This employment rate increased to about 15.5 [13.7, 17.5] per 10,000 by 1930, about a 94 percent growth in the expected employment rate over these three decades.
+We find no clear evidence that the proportional growth rates in police employment in these states were higher than the national rate of growth.
+Our results suggest that by 1930, these states would employ police at a rate nearly double the rate of police employment in high sharecropping states and high land redistribution states.
+
+In states with high levels of settler land redistribution, our models suggest average police employment rates of about 3.4 per 10,000 in 1900 [2.7, 4.2], growing to about 8.7 [7.9, 9.7] per 10,000 in 1930.
+Police employment grew faster [2.5 percent, 12.2 percent] in these states than the national average.
+Our models suggest that a typical high settler land redistribution state had lower levels of police employment that other kinds of states in 1900, but by 1930, had higher levels of police employment than high sharecropping states, but still much lower employment levels than high urban immigrant states.
+
+In states with large populations of Black tenant farmers, we find comparatively low levels of police employment.
+In 1900, our model expects a high sharecropping state to employ 4.0 [3.4, 4.9] police per 10,000 population.
+In 1930, expected police employment increases to 7.1 [6.2, 8.0] per 10,000.
+We do not find evidence that the rate of change of police employment in these contexts is differs from other contexts during this period.
+This finding is somewhat surprising given the prominent role of law enforcement in cooperating with and upholding the regime of racial terror in force throughout the US South during this period, an issue we explore at greater length in the discussion.
+
+#### Social workers
+
+In Panel 3 of Figure \@ref(fig:emp-m-vis), we show model expectations for social work employment across contexts and over time.
+Social workers were very rarely documented in the Census in 1900.
+However, by the end of the period, there were substantial numbers of social workers across the US.
+As shown in Figure \@ref(fig:emp-ts), there was explosive growth in social worker employment during the Progressive Era, which varied dramatically across places ($\sigma_\gamma = 0.54, \sigma_\delta = 0.1$).
+We estimate a national expected growth rate of between 300 and 356 percent per decade.
+We do not detect meaningful differences in the growth rate of social workers across contexts, though there are dramatic differences in baseline levels, which have profound impacts on employment rate trajectories.
+
+```{r callouts3}
+temp <- preds_socwork %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred) * 1e2, 
+            yhat_lwr = quantile(.epred, 0.025)* 1e2, 
+            yhat_upr = quantile(.epred, 0.975) * 1e2) %>% 
+  arrange(scen, year_c)
+```
+
+Historians have long argued that states with large populations of white urban immigrants were the locus of the development of social work as a field (e.g.
+GORDON, WARD, POLSKY, ETC), and our findings strongly support this claim.
+We find that states with a large urban immigrant population had expected levels of social work employment (about 0.7 [0.4, 1.0] per 100,000 population) that were about 60 [27, 103] percent higher than other states in 1900.
+Our estimate of expected social work employment rises to 2.3 [1.7, 3.0] in 1910, 7.9 [6.0, 10.1] in 1920, and 27.2 [19.7, 36.6] in 1930, a forty-fold increase over 30 years.
+
+In line with the observed geography of employment shown in \ref(@fig:emp-maps), we find levels of social work employment were relatively low in states with high levels of settler land redistribution.
+Expected social work employment rise from about 0.2 [0.1, 0.3] per 100,000 in 1900 to about 6.1 [4.6, 7.9] per 100,000 in 1930.
+
+In states with large numbers of Black tenant farmers, social work employment rates began very near zero in 1900.
+Our expected social work employment rate in 1900 for a high sharecropping state is about 0.2 [0.1, 0.3] social workers per 100,000 population.
+By 1930, our analysis expects that a high sharecropping state would have about 5.4 [4.0, 7.3] social workers per 100,000 population, about a 30-fold increase.
+By 1930, we estimate that states with high numbers of urban white immigrants would have had between `r round(temp[12, 4] / temp[8, 5],1)` and `r round(temp[12, 5] / temp[8, 4],1)` times more social workers per capita than states with high levels of sharecropping.
+
+#### Indian boarding schools
+
+As shown in Figure \@ref(fig:bsts), the count of boarding schools operating annually in the US fluctuated around 200 schools operating between 1900 and 1930.
+We estimate a logistic regression model to estimate the likelihood that a state had at least one boarding school in operation for each decade during the focal period.
+The posterior point estimates and credible intervals for focal parameters are shown in Figure \@ref(fig:bs-m).
+
+```{r bs-m, fig.cap = "Relationships between settler colonial contexts and presence of Indian boarding schools.
+Logistic regression with operating boarding school binary outcome.
+Posterior mean and credible intervals (0.66, 0.95) for focal Beta parameters"}
+
+pdat<-bs_m1_z %>%
   gather_draws(c(b_scaleblk_sharecrop_pc,
@@
                  b_scalehomestead_plus_morill)) %>% 
-  mutate(model = "1. Teachers") %>%
-  bind_rows(
-    cops_m1 %>%
-      gather_draws(c(b_scaleblk_sharecrop_pc,
-                     b_scaleurban_immig_pc,
-                     b_scalehomestead_plus_morill)) %>%
-      mutate(model = "2. Police")) %>%
-  bind_rows(
-    socwork_m1 %>%
-      gather_draws(c(b_scaleblk_sharecrop_pc,
-                     b_scaleurban_immig_pc,
-                     b_scalehomestead_plus_morill)) %>%
-      mutate(model = "3. Social workers")) %>%
   mutate(.variable = case_when(
-    .variable == "b_scaleblk_sharecrop_pc" ~
-      "B1: Black tenant\nfarmers per capita",
-    .variable == "b_scaleurban_immig_pc" ~
-      "B2: Urban white\nimmigrants\nper capita",
-    .variable == "b_scalehomestead_plus_morill" ~
-      "B3: Cumulative \nMorill +\nHomestead acres"))
+    .variable == "b_scaleblk_sharecrop_pc" ~ "Sharecropping",
+    .variable == "b_scaleurban_immig_pc" ~ "Urban immigrants",
+    .variable == "b_scalehomestead_plus_morill" ~ "Homestead + Morill acres"
+  ))
+
 ggplot(pdat,
@@
            y = .variable)) +
-  stat_halfeye() +
+  stat_pointinterval() +
   geom_vline(aes(xintercept = 0), lty = 2) +
@@
        y = "") +
-  facet_wrap(~model) +
   theme_bw()
-```
-
-We display the temporal components of these models, expected values for employment rates conditional on the range of observed settler colonial contexts, and the observed bivariate relationships between employment rates and context measures in Figures \@ref(fig:postpred-teachers), \@ref(fig:postpred-police), and \@ref(fig:postpred-socwork).
-As described in Figure \@~ref(fig:emp-ts), there were dramatic increases in the employment of teachers, social workers, and police nationally between 1900 and 1930.
-But these national trends obscure significant sub-national variation in employment rates and growth trajectories.
-
-Figure \@ref(fig:postpred-teachers) shows that while teacher employment grew substantially across all states between 1900 and 1930, states with high levels of Black tenant farming tended to have much lower levels of teacher employment across the full time period than did states that did not have high levels of Black tenant farming.
-Models suggest that the expected employment rate for teachers in states with high levels of Black tenant farming (25 percent of the population, similar to South Carolina in 1900 - 1930) would be about 3.4 with a 90 percent credible interval of [3.0, 3.8] per 1,000 population in 1900; about 3.9 [3.5, 4.4] per 1,000 persons in 1910; about 4.5 [4.0, 5.0] per 1,000 in 1920, and about 5.3 [4.7, 6.0] teachers per 1,000 in 1930.
-
-By contrast, states with high levels of settler land redistribution through the Morill and Homestead acts to have much higher levels of teacher employment than Jim Crow labor regime states.
-In 1900, a state with a high level of acreage redistributed to settlers in the recent past (10 million acres, similar to Minnesota, Nebraska, or Kansas) is expected to have had a teacher employment rate of about 6.7 [6.3, 7.2] per 1,000 population.
-In 1910, a state with high levels of Homestead and Morill act land distributions (13 million acres, similar to Kansas, North Dakota, and Oklahoma) is expected to have employed about 8.1 [7.5, 8.8] teachers per 1,000 persons.
-By 1930, states with high levels of contemporary settler land redistribution (about 20 million acres, similar to Colorado, North Dakota and South Dakota) had highest-in-the-nation levels of teacher employment, with expected rates of about 12.3 [11.3, 13.5] teachers per 1,000 persons.
-
-For each decade we evaluate, our models expect no meaningful differences in teacher employment across states with high and low levels of white urban immigrant populations, conditional on low levels of active settler land redistribution and low levels of Black tenant farming.
-There are, of course, high levels of expected increases in teacher employment over time, ranging from a low of 5.7 [5.2, 6.1] in 1900 to a high of 8.9 [8.1, 9.6] in 1930.
-
-```{r postpred-teachers, fig.cap = "Estimated and observed relationships between teacher edfmployment rates and focal settler colonial context variables, US States 1900 - 1930.
-Posterior prediction for expected teacher employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
-Dots and text indicate observed values."}
-### blk_sharecrop
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = 0,
-            urban_immig_pc = 0,
-            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
-            year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(teachers_m1,
-                  re_formula = NA)
-### credible intervals for text callouts
-t<-pred_dat %>% 
-  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p1<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) + 
-  geom_point(data = mdat,
-             aes(y = teachers / tot_pop * 1e3,
-                 x = blk_sharecrop_pc),
-             size = 1, alpha = 0.5) +
-  # geom_text(data = mdat %>% 
-  #             filter(blk_sharecrop_pc > 0.03),
-  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
-  #               x = blk_sharecrop_pc,
-  #               label = STUSPS),
-  #           alpha = 0.75,
-  #           size = 3) + 
-  facet_wrap(~year, nrow = 1) + 
-  guides(fill = "none") +
-  theme_bw() +
-  coord_cartesian(ylim = c(2.7,16.8)) + 
-  labs(y = "",
-       x = "Proportion of population Black tenant farmers")
-### homestead+morill
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = seq(0,
-                                                  max(mdat$homestead_plus_morill),
-                                                  length.out = 100),
-                      urban_immig_pc = 0,
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(teachers_m1,
-                  re_formula = NA)
-### text callouts
-t<-pred_dat %>% 
-  filter(homestead_plus_morill>20e6, homestead_plus_morill < 20.2e6) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p2<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) + 
-  geom_point(data = mdat,
-             aes(y = teachers / tot_pop * 1e3,
-                 x = homestead_plus_morill),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(homestead_plus_morill > 2e5),
-  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
-  #               x = homestead_plus_morill,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
-       y = "Teachers per 1,000 population") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(2.7,16.8)) + 
-  scale_x_continuous(labels = function(x)x / 1e6) + 
-  theme_bw() 
-### urban immig
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = 0,
-                      urban_immig_pc = seq(0,
-                                           max(mdat$urban_immig_pc),
-                                           length.out = 100),
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(teachers_m1,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p3<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = urban_immig_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) + 
-  geom_point(data = mdat,
-             aes(y = teachers / tot_pop * 1e3,
-                 x = urban_immig_pc),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(urban_immig_pc > 0.02),
-  #           aes(y = teachers / tot_pop * 1e3 + 0.5,
-  #               x = urban_immig_pc,
-  #               label = STUSPS),
-            # alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Proportion of population urban white immigrants",
-       y = "") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(2.7,16.8)) + 
-  theme_bw() 
-grid.arrange(p1, p2, p3, nrow = 3)
-```
-
-We observe clear differences in police employment across settler colonial contexts.
-In Figure \@ref(fig:postpred-police) we show posterior expected values for police employment rates across the observed range of settler colonial contexts over time.
-Rates of police employment were relatively low in the Jim Crow South.
-Our models suggest that in a state with 25 percent of the population documented as Black tenant farmers in the Census, an expected police employment rate of 0.31 [0.26, 0.38] per 1,000 persons in 1900, a rate of 0.37 [0.30, 0.44] in 1910, a rate of 0.41 [0.34, 0.49] in 1920, and a rate of about 0.63 [0.52, 0.77] police employed per 1,000 persons in 1930.
-
-Employment rates of police officers were dramatically higher in states with large white urban immigrant populations.
-In states with 25 percent of their population recorded as white immigrants living in urban Census tracts (similar to New Jersey, New York, Massachusetts, and Rhode Island), our models expected states to have had about 1.2 [0.99, 1.4] police employed per 1,000 population in 1900, 1.4 [1.2, 1.6] per 1,000 population in 1910, 1.5 [1.3, 1.8] in 1920, and 2.3 [2.0, 2.8] police per 1,000 population in 1930.
-
-Police were employed at slightly lower levels in states with high levels of recent settler land distributions than in other states.
-In 1930, our models expect a state with no recent settler land distributions to have had an average of about 1.0 [0.90, 1.1] police per 1,000 population, compared to an expected police employment rate of 0.7 [0.6, 0.8] per 1,000 in a state with very high levels of recent settler land redistribution.
-
-```{r postpred-police, fig.cap = "Estimated and observed relationships between police employment rates and focal settler colonial context variables, US States 1900 - 1930.
-Posterior prediction for expected police employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
-Dots and text indicate observed values."}
-### blk_sharecrop
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = 0,
-            urban_immig_pc = 0,
-            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
-            year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(cops_m1,
-                  re_formula = NA)
-### credible intervals for text callouts
-t<-pred_dat %>% 
-  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p1<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6)  + 
-  geom_point(data = mdat,
-             aes(y = police / tot_pop * 1e3,
-                 x = blk_sharecrop_pc),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(blk_sharecrop_pc > 0.03),
-  #           aes(y = police / tot_pop * 1e3 + 0.1,
-  #               x = blk_sharecrop_pc,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  facet_wrap(~year, nrow = 1) + 
-  guides(fill = "none") +
-  theme_bw() +
-  coord_cartesian(ylim = c(0,3)) + 
-  labs(y = "",
-       x = "Proportion of population Black tenant farmers")
-### homestead+morill
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = seq(0,
-                                                  max(mdat$homestead_plus_morill),
-                                                  length.out = 100),
-                      urban_immig_pc = 0,
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(cops_m1,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  group_by(year, homestead_plus_morill) %>% 
-  summarize(med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3) %>% 
-  filter(year == 1930)
-# t[1,]
-# t[90,]
-p2<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
- stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) +
-  geom_point(data = mdat,
-             aes(y = police / tot_pop * 1e3,
-                 x = homestead_plus_morill),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(homestead_plus_morill > 3e6),
-  #           aes(y = police / tot_pop * 1e3 + 0.1,
-  #               x = homestead_plus_morill,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
-       y = "Police per 1,000 population") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(0,3)) + 
-  scale_x_continuous(labels = function(x)x / 1e6) + 
-  theme_bw() 
-### urban immig
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = 0,
-                      urban_immig_pc = seq(0,
-                                           max(mdat$urban_immig_pc),
-                                           length.out = 100),
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(cops_m1,
-                  re_formula = NA)
-### text call out
-t<-pred_dat %>% 
-  filter(urban_immig_pc > 0.25, urban_immig_pc < 0.255) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p3<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = urban_immig_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) +
-  geom_point(data = mdat,
-             aes(y = police / tot_pop * 1e3,
-                 x = urban_immig_pc),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(urban_immig_pc > 0.03),
-  #           aes(y = police / tot_pop * 1e3 + 0.1,
-  #               x = urban_immig_pc,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Proportion of population urban white immigrants",
-       y = "") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(0,3)) + 
-  theme_bw() 
-grid.arrange(p1, p2, p3, nrow = 3)
-```
-
-Social work employment dramatically increased from near-zero levels in 1900 to becoming relatively common across all US states by 1930.
-Again, however, these levels of employment and trajectories varied substantially across contexts.
-For example, in a state with a large population of Black tenant farmers, the model expectations of social worker employment increased from 1.1 [0.7, 1.8] per million population in 1900 to 3.7 [2.6, 5.6] per 100,000 population in 1930.
-Expected employment for social workers followed similar trends in states with high levels of recent settler land redistribution, starting near zero in 1900 and rising to around 4 social workers per 100,000 by 1930 in a state with high levels of contemporary settler land redistribution.
-
-By contrast, social work employment started at relatively high levels in 1900 and rose dramatically in states with large urban immigrant populations.
-Our models expect a state with a large urban immigrant population (25 percent of population) to have had about 9.9 [7.0, 14.3] social workers per million population in 1900.
-This employment rate increased dramatically during this period.
-Median expected employment rates for social workers were 3.3 per 100,000 in 1910, 1.5 per 10,000 in 1920, and 3.2 per 10,000 in 1930.
-This represented a 32 fold increase between 1900 and 1930.
-Our models expect per capita social work employment to be about 8 times higher in states with high levels of white urban immigrants in 1930 than they were in states operating under a Jim Crow agricultural labor regime.
-
-```{r postpred-socwork, fig.cap = "Estimated and observed relationships between social worker employment rates and focal settler colonial context variables, US States 1900 - 1930.
-Posterior prediction for expected teacher employment rate as dotted line (mean), with credible intervals shaded (0.5, 0.8, 0.95).
-Dots and text indicate observed values."}
-### blk_sharecrop
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = 0,
-            urban_immig_pc = 0,
-            blk_sharecrop_pc = seq(0,max(mdat$blk_sharecrop_pc), length.out = 100),
-            year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(socwork_m1,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  filter(blk_sharecrop_pc < 0.251, blk_sharecrop_pc > 0.250) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p1<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = blk_sharecrop_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) +
-  geom_point(data = mdat,
-             aes(y = socwork / tot_pop * 1e3,
-                 x = blk_sharecrop_pc),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(blk_sharecrop_pc > 0.03),
-  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
-  #               x = blk_sharecrop_pc,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  facet_wrap(~year, nrow = 1) + 
-  guides(fill = "none") +
-  theme_bw() +
-  coord_cartesian(ylim = c(0,0.3)) + 
-  labs(y = "",
-       x = "Proportion of population Black tenant farmers")
-### homestead+morill
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = seq(0,
-                                                  max(mdat$homestead_plus_morill),
-                                                  length.out = 100),
-                      urban_immig_pc = 0,
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(socwork_m1,
-                  re_formula = NA)
-p2<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = homestead_plus_morill)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) +
-  geom_point(data = mdat,
-             aes(y = socwork / tot_pop * 1e3,
-                 x = homestead_plus_morill),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(homestead_plus_morill > 2e5),
-  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
-  #               x = homestead_plus_morill,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Homestead Acres Plus Morill Acres (millions) Distributed to Settlers",
-       y = "Social workers per 1,000 population") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(0,0.3)) + 
-  scale_x_continuous(labels = function(x)x / 1e6) + 
-  theme_bw() 
-### urban immig
-pred_dat<-expand_grid(tot_pop = 1e6,
-                      homestead_plus_morill = 0,
-                      urban_immig_pc = seq(0,
-                                           max(mdat$urban_immig_pc),
-                                           length.out = 100),
-                      blk_sharecrop_pc = 0,
-                      year = c(1900, 1910, 1920, 1930)) %>% 
-  add_epred_draws(socwork_m1,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  filter(urban_immig_pc > 0.25, urban_immig_pc < 0.255) %>% 
-  group_by(year) %>% 
-  summarize(
-    med = median(.epred) / 1e3,
-    lwr = quantile(.epred, 0.05) / 1e3,
-    upr = quantile(.epred, 0.95) / 1e3)
-p3<-ggplot(pred_dat,
-       aes(y = .epred / 1e3, x = urban_immig_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6) +
-  geom_point(data = mdat,
-             aes(y = socwork / tot_pop * 1e3,
-                 x = urban_immig_pc),
-             size = 1, alpha = 0.5) + 
-  # geom_text(data = mdat %>% 
-  #             filter(urban_immig_pc > 0.02),
-  #           aes(y = socwork / tot_pop * 1e3 + 0.004,
-  #               x = urban_immig_pc,
-  #               label = STUSPS),
-  #           alpha = 0.75, size = 3) + 
-  guides(fill = "none") +
-  labs(x = "Proportion of population urban white immigrants",
-       y = "") + 
-  facet_wrap(~year, nrow = 1) + 
-  coord_cartesian(ylim = c(0,0.3)) + 
-  theme_bw() 
-grid.arrange(p1, p2, p3, nrow = 3)
-```
-
-Across categories of employment, we find clear relationships between settler colonial contexts and levels of employment in welfare and carceral institutions.
-We argue that the rationally and temporally varying interests of settler elites determine the kinds of colonial projects they prioritize, reflected in the kinds of human infrastructure that they develop to pursue policy goals.
-
-## Group quarters models
-
-Similar to employment, there are also clear relationships between settler colonial contexts and the magnitude of confinement in institutions across states between 1900 and 1930.
-Figure \@ref(fig:gq-m-vis) displays posterior distributions of focal parameters for models of the populations housed in institutional group quarters in the 1900, 1920, and 1930 US Census by state.
-We find that correctional populations were similar across contexts.
-By contrast, we find that populations housed in mental institutions, institutions for the poor, and institutions for the disabled were significantly higher in states with large urban white immigrant populations than in other states, and that these populations were significantly lower in states with large Black tenant farm populations than in other states.
-
-```{r gq-m-vis, fig.cap = "Relationships between focal settler colonial contexts and populations housed in group quarters per capita, US States 1900 - 1930.
-Densities of posterior parameter estimates with credible intervals (0.66, 0.95).
-Panel 1 displays results for populations in correctional institutions, Panel 2 displays results for populations in institutions for the poor and disabled and in mental institutions."}
-#### extract posterior draws for b1, b2, b3, set up for plotting
-pdat<-prison_m1 %>%
-  gather_draws(c(b_scaleblk_sharecrop_pc,
-                 b_scaleurban_immig_pc,
-                 b_scalehomestead_plus_morill)) %>%
-  mutate(model = "1. Correctional institutions") %>%
-  bind_rows(
-    inst_m1 %>%
-      gather_draws(c(b_scaleblk_sharecrop_pc,
-                     b_scaleurban_immig_pc,
-                     b_scalehomestead_plus_morill)) %>%
-      mutate(model = "2. Asylums and poorhouses")) %>%
-  mutate(.variable = case_when(
-    .variable == "b_scaleblk_sharecrop_pc" ~
-      "B1: Black tenant\nfarmers per capita",
-    .variable == "b_scaleurban_immig_pc" ~
-      "B2: Urban white\nimmigrants\nper capita",
-    .variable == "b_scalehomestead_plus_morill" ~
-      "B3: Cumulative \nMorill +\nHomestead acres"))
-ggplot(pdat,
-       aes(x = .value,
-           y = .variable)) +
-  stat_halfeye() +
-  geom_vline(aes(xintercept = 0), lty = 2) +
-  labs(x = "Estimate",
-       y = "") +
-  facet_wrap(~model) +
-  theme_bw()
-```
-
-CONSISTENT WITH HISTORICAL WORK ON PREVALENCE OF EARLY THERAPEUTIC / WELFARE STATES.
+
+```
+
+```{r include = F}
+# callouts
+temp<-mdat %>% 
+  filter(boarding_schools>0) %>% 
+  select(STUSPS, year, homestead_plus_morill, boarding_schools) %>% arrange(homestead_plus_morill)
+
+temp %>% filter(homestead_plus_morill==0)
+
+# how many no morill+hs acre states
+mdat %>% filter(homestead_plus_morill==0) %>% select(STUSPS) %>% distinct() %>% nrow()
+### how many ANY morill + hs acre states
+mdat %>% filter(homestead_plus_morill>0) %>% select(STUSPS) %>% distinct() %>% nrow()
+# how many no bs with any morill + hs acres
+mdat %>% filter(homestead_plus_morill>0 & boarding_schools ==0) %>% select(STUSPS) %>% distinct()
+
+```
+
+We find that states actively redistributing Native lands to settlers during the Progressive Era were dramatically more likely to operate boarding schools during than were other states.
+Among the 19 states that did not have lands parceled to settlers under the Homestead or Morill acts (though these states were unarguably beneficiaries of the acts), only New York, North Carolina, Pennsylvania, and Virginia housed Indian boarding schools during the Progressive Era.
+Among the 29 states within which Native lands were redistributed to settlers under the Morill or Homestead Acts, federal data suggest that only 6 did not have an operating Indian boarding school during this period.
+These patterns our clearly reflected in our regression results.
+States with high levels of sharecropping were less likely than states in other settler colonial contexts to operate a boarding school during the Progressive Era.
+White urban immigrant population does not have a meaningful statistical relationship to boarding school operation that we detect conditional on our data and model specifications.
+A large number of states with large urban populations operated no boarding schools, while one of the most notorious facilities, the Carlisle Industrial School, operated in Pennsylvania, among the leaders in white urban immigrant population during this period.
+
+### Physical institutions
+
+There are also clear relationships between settler colonial contexts and the magnitude of confinement in carceral and welfare institutions across states between 1900 and 1930.
+Figure \@ref(fig:gq-m-vis) displays posterior distributions of focal parameters for models of the populations housed in institutional group quarters between 1900 and 1930 by settler context scenario.
+
+By 1930, we find that expected rates of incarceration in prisons and jails were similar across each settler colonial context we specify.
+By contrast, we find lower expected levels of confinement in mental institutions in states with high levels of sharecropping compared to other contexts, and very high rates of mental institutionalization in places with a high volume of white urban immigrants.
+Similarly, we see lower rates of residence in institutions for the poor, elderly and disabled in high sharecropping states than in other contexts and much higher capacities in these institutions in high urban immigrant contexts.
+
+NOTE FLAT / slight DOWN TRENDS IN NE STATES. INDICATES BUILDOUT HAPPENED EARLIER. BUILDOUT OCCURRING DURING PERIOD FOR OTHER STATES.
+
+<!-- CONSISTENT WITH HISTORICAL WORK ON PREVALENCE OF EARLY THERAPEUTIC / WELFARE STATES.
 CITE ALL THAT STUFF.
@@
 Welfare and mental institutions were relatively rare in the Jim Crow South, and very common in states with large numbers of white immigrants in cities.
-
-```{r gq-postpred-time, fig.cap = "Expected populations housed in group quarters per capita across varying settler colonial contexts, US States 1900 - 1930.
-Densities of posterior parameter estimates with credible intervals (0.66, 0.95)."}
-### set up sims for 1900, 1920, 1930 preds
-### for high on each context for prison_m1 and inst_m1
-t<-expand_grid(
-  year = c(1900, 1920, 1930),
-  blk_sharecrop_pc = c(0, 0.25),
-  urban_immig_pc = 0,
-  homestead_plus_morill = 0,
-  tot_pop = 1e6) %>% 
-  mutate(condition = 
-           ifelse(blk_sharecrop_pc == 0,
-                  "Low value", "High value"),
-         scen = "Black tenant farming") %>% 
-  bind_rows(
-    expand_grid(
-      year = c(1900, 1920, 1930),
-      blk_sharecrop_pc = c(0),
-      urban_immig_pc = c(0, 0.25),
-      homestead_plus_morill = 0,
-      tot_pop = 1e6) %>% 
-      mutate(condition = 
-               ifelse(urban_immig_pc == 0,
-                      "Low value", "High value"),
-             scen = "White urban immigrants")) %>% 
-    bind_rows(
-      expand_grid(
-      year = c(1900, 1920, 1930),
-      blk_sharecrop_pc = c(0),
-      urban_immig_pc = 0,
-      homestead_plus_morill = c(0, 10e6),
-      tot_pop = 1e6) %>% 
-      mutate(condition = 
-               ifelse(homestead_plus_morill == 0,
-                      "Low value", "High value"),
-             scen = "Homestead + Morill acres")) 
-t<-t%>% 
-  add_epred_draws(prison_m1,
-                  re_formula = NA) %>% 
-  mutate(model = "Prisons and jails") %>% 
-  bind_rows(t %>% 
-              add_epred_draws(inst_m1,
-                  re_formula = NA) %>% 
-              mutate(model = "Asylums and poorhouses"))
-ggplot(t,
-       aes(x = year, y = .epred/1e3,
-           color = condition)) + 
-  stat_pointinterval(position = position_dodge()) + 
-  labs(x = "Year", y = "Expected inmates per 1,000 pop.") + 
-  facet_wrap(model~scen) + 
-  theme_bw() + 
-  labs(color = "Context variable") + 
-  theme(legend.position = "bottom")
-```
-
-In Figure \@ref(fig:gq-postpred-time) we show how the physical confinement of persons in institutional group quarters shifted over time and across contexts between 1900 and 1930.
-Note that we exclude the 1910 US Census from group quarters analyses due to data quality concerns.
-We find that the confinement of people in institutions for the mentally ill, the poor, and the disabled was dramatically more common in states with large numbers of white urban immigrants than in other states.
-Our models show that confinement in these institutions was relatively less common in states with large numbers of Black tenant farmers than in states with few Black tenant farmers.
-And we see no statistically significant difference between states with high and low levels of Homestead and Morill act land redistribution.
-Rates of confinement in these kinds remained relatively stable between 1900 and 1920, then increasing substantially by 1930.
-
-Per capita levels of incarceration in correctional institutions were similar across contexts for all periods in our analyses, with relatively stable levels between 1900 and 1920, followed by an increase in incarceration rates in 1930.
+-->
+
+<!-- Per capita levels of incarceration in correctional institutions were similar across contexts for all periods in our analyses, with relatively stable levels between 1900 and 1920, followed by an increase in incarceration rates in 1930.
 This homogeneity in carceral capacity is notable, given the differences in police employment described in Figure \@ref(fig:postpred-police).
@@
 Historical scholarship has clearly shown the salience of white supremacist and settler colonial logics for the development of early-modern policing and incarceration across US geographies (CITE CONVICT LEASE LIT, KLH ON California, NORTHEAST LIT, FRONTIER STUFF (MAYBE ROSS)).
-
-## Boarding school models
-
-As shown in Figure \@ref(fig:bsts), the count of boarding schools operating annually in the US fluctuated around 200 schools operating between 1900 and 1930.
-We estimate a two stage zero-inflated model to separately estimate 1) the likelihood that a state had at least one boarding school during the focal period, then conditional on operating at least one boarding school 2) the number of schools operating with the state during a Census year.
-The posterior distributions of focal parameters are shown in Figure \@ref(fig:bs-m-mvs).
-We visualize the probability of operating at least one boarding school across focal contexts in Figure \@ref(fig:pr-bs).
-
-```{r bs-m-mvs, fig.cap = "Relationships between focal settler colonial contexts and presence and count of Indian boarding shcools.
-Zero-inflated model described in equations 1 and 2.
-Panel 1 displays binary component of model (for presence of any boarding schools).
-Panel 2 displays count component of model for number of boarding schools, conditional on at least one boarding school in the state and year."}
-### zero component visual
-pdat<-bs_m1_z %>%
-  gather_draws(c(b_scaleblk_sharecrop_pc,
-                 b_scaleurban_immig_pc,
-                 b_scalehomestead_plus_morill)) %>%
-  mutate(model = "1. Boarding schools > 0") %>%
-  bind_rows(
-    bs_m1 %>%
-      gather_draws(c(b_scaleblk_sharecrop_pc,
-                     b_scaleurban_immig_pc,
-                     b_scalehomestead_plus_morill)) %>%
-      mutate(model = "2. Count when boarding schools > 0")) %>% 
-  mutate(.variable = case_when(
-    .variable == "b_scaleblk_sharecrop_pc" ~
-      "B1: Black tenant\nfarmers per capita",
-    .variable == "b_scaleurban_immig_pc" ~
-      "B2: Urban white\nimmigrants\nper capita",
-    .variable == "b_scalehomestead_plus_morill" ~
-      "B3: Cumulative \nMorill +\nHomestead acres"))
-ggplot(pdat,
-       aes(x = .value,
-           y = .variable)) +
-  stat_halfeye() +
-  geom_vline(aes(xintercept = 0), lty = 2) +
-  labs(x = "Estimate",
-       y = "") +
+-->
+
+```{r gq-vis, fig.cap = "Posterior expected employment rates per 1,000 population by employment type across settler project contexts, US States 1900 - 1930.
+Mean posterior expected value (line) and 95 percent credible interval (shaded)."}
+preds_prison<-prison_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat) %>% 
+  mutate(type = "Prison")
+
+preds_mental<-mental_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat)%>% 
+  mutate(type = "Mental inst.")
+
+preds_poor<-poor_m1 %>% 
+  add_epred_draws(re_formula = NA,
+                  newdata = pred_dat)%>% 
+  mutate(type = "Poverty inst.")
+
+p1<-ggplot(preds_prison,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
+                  .width = 0.95, 
+                  fill = "gray50",
+                  linewidth = 0.5) +
   theme_bw() + 
-  facet_wrap(~model, 
-             scales = "free_x") 
-```
-
-We find that states with large redistributions of Native lands to settlers were far more likely to operate boarding schools during this period than were other states.
-We find that a (counterfactual) state that contained no lands redistributed under the Homestead or Morill acts in 1900 had an expected probability of containing boarding schools of about 0.51 [0.24, 0.77].
-All states with large numbers or contemporary settler land redistribution contained at least one boarding school during this period.
-The expected probability reaches one at about 10 million acres distributed under the Homestead and Morill acts.
-Conditional on operating at least one boarding school, however, these states were similar to each other in the expected count of operating schools across the period (posterior expected count 6.3 [2.7, 12.6]).
-We find no statistically significant difference in the count of schools operating across high settler redistribution lands during this period.
-
-The expected probability of operating an Indian boarding school was not significantly different across states with small and large white urban immigrant populations.
-Given the role of notorious facilities in the Northeast like Carlisle in Pennsylvania, this result is unsurprising.
-We do find that the expected count of facilities in states (conditional on having at least one facility) was generally lower in states with large urban immigrant populations than it was in states with small urban immigrant populations.
-
-Boarding schools were relatively rare in the Jim Crow South.
-We find that states with large populations of Black tenant farmers had a near zero probability of operating an Indian boarding school in 1900.
-There were boarding schools operating in states like North Carolina, Arkansas, and Virginia during this era, but the majority of states in the South did not have any operating boarding schools between 1900 and 1930.
-
-```{r pr-bs, fig.cap = "Probability of having at least one boarding school in 1900 across contexts, US states.
-Posterior expected values, dotted line (mean), credible intervals shaded (0.5, 0.8, 0.95)."}
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = seq(0, max(mdat[mdat$year==1900, "homestead_plus_morill"]),
-                                        length.out = 100),
-            urban_immig_pc = 0,
-            blk_sharecrop_pc = 0,
-            year = c(1900)) %>% 
-  add_epred_draws(bs_m1_z,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  group_by(year, homestead_plus_morill) %>% 
-  summarize(
-    med = median(.epred),
-    lwr = quantile(.epred, 0.05),
-    upr = quantile(.epred, 0.95))
-p1<-ggplot(pred_dat,
-       aes(y = .epred , x = homestead_plus_morill / 1e6)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
+  labs(y = "",
+       x = "",
+       subtitle = "1. Correctional institutions")
+
+p2<-ggplot(preds_mental,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
                   .width = 0.95, 
-                  fill = "grey", alpha = 0.6) +
-  guides(fill = "none") +
+                  fill = "gray50",
+                  linewidth = 0.5) +
   theme_bw() + 
-  # scale_x_log10() + 
-  labs(y = "Probability",
-       x = "Acres (millions)",
-       subtitle = "1. Homestead + Morill") + 
-  coord_cartesian(ylim = c(0,1))
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = 0,
-            urban_immig_pc = seq(0, max(mdat$urban_immig_pc), length.out = 100),
-            blk_sharecrop_pc = 0,
-            year = c(1900)) %>% 
-  add_epred_draws(bs_m1_z,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  group_by(year, urban_immig_pc) %>% 
-  summarize(
-    med = median(.epred),
-    lwr = quantile(.epred, 0.05),
-    upr = quantile(.epred, 0.95))
-p2<-ggplot(pred_dat,
-       aes(y = .epred , x = urban_immig_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
+  labs(y = "Rate",
+       x = "",
+       subtitle = "2. Mental institutions")
+
+p3<-ggplot(preds_poor,
+       aes(x = year_c*10 + 1900, y = .epred)) + 
+  stat_lineribbon(show.legend = F, 
                   .width = 0.95, 
-                  fill = "grey", alpha = 0.6)  +
-  guides(fill = "none") +
+                  fill = "gray50",
+                  linewidth = 0.5) +
   theme_bw() + 
-  scale_x_log10() + 
+  theme(panel.spacing = unit(0.4, "cm")) +
+  facet_wrap(~scen) + 
   labs(y = "",
-       x = "Proportion",
-       subtitle = "2. White urban immigrants")+ 
-  coord_cartesian(ylim = c(0,1))
-pred_dat<-expand_grid(tot_pop = 1e6,
-            homestead_plus_morill = 0,
-            urban_immig_pc = 0,
-            blk_sharecrop_pc =  seq(0, max(mdat$blk_sharecrop_pc), length.out = 100),
-            year = c(1900)) %>% 
-  add_epred_draws(bs_m1_z,
-                  re_formula = NA)
-t<-pred_dat %>% 
-  group_by(year, blk_sharecrop_pc) %>% 
-  summarize(
-    med = median(.epred),
-    lwr = quantile(.epred, 0.05),
-    upr = quantile(.epred, 0.95))
-p3<-ggplot(pred_dat,
-       aes(y = .epred , x = blk_sharecrop_pc)) +
-  stat_lineribbon(lty = 2,
-                  linewidth = 0.5,
-                  .width = 0.95,
-                  fill = "grey", alpha = 0.6)  +
-  guides(fill = "none") +
-  theme_bw() +
-  scale_x_log10() + 
-  labs(y = "",
-       x = "Proportion",
-       subtitle = "3. Black tenant farmers")+ 
-  coord_cartesian(ylim = c(0,1))
-grid.arrange(p1, p2, p3,
-             nrow = 1)
-```
-
-# Discussion
-
-The federal government, state governments, and local governments adopted and developed different configurations of welfare and carceral infrastructures in ways that closely relate to the geography of US settler colonial projects during the Progressive Era.
-HIGH LEVEL SUMMARY - NAME THE EMPIRICAL GROUPINGS!
-EXPLAIN HOW THEY RELATE THEORETICALLY TO SETTLER PROJECTS VIS A VIS A FUNCTIONALIST LOGIC (then apologize for being functionalist)
-
-States in the Jim Crow South, where settler white elites were deeply invested in the exploitation of Black labor and in a brutal regime of racial apartheid and terror, had relatively weak front-line social policy bureaucracies (reflected in low employment levels for teachers, police, and social workers) throughout this period.
-They also had less physical infrastructure characteristic of the early US welfare state, like poorhouses and mental institutions.
-However, these states tended to have incarceration levels near the national average.
-For all measures *except* incarceration, Jim Crow states exhibited relatively thin social policy states.
-These findings align with historical scholarship that has carefully documented the simultaneous brutality and ubiquity of a hybrid formal / informal regime of anti-Black racial terror with the routine underfunding and/or failure to develop core welfare state institutions across formerly slave states (CITE QUADAGNO, KATZNELSON, TOLNAY, BAKER, PERKINSON, OTHERS).
-
-Nation building and racial demographics were front-of-mind for settler elites in states where Native lands had been recently expropriated.
-A variety of initiatives had been undertaken by the federal government and by local policy makers to rapidly attract white settlers to these territories and to develop the social, cultural, and economic infrastructure that would fulfill and reinforce national settler colonial ideologies (REVISE).
-In states containing large amounts of Native lands granted to settlers and/or land-grant Universities, there were dramatically higher levels of teacher employment than in other states, while levels of police and social worker employment were comparable to other contemporary US states.
-The physical infrastructure of the carceral and welfare state was similar to other states, with one exception.
-While rates of confinement in correctional, mental, and proto-welfare institutions were similar to those in other states, these states were dramatically more likely than others to contain Indian boarding schools.
-Taken together, these findings suggest that states engaged in the early stages of Native land redistribution and white settlement were engaged in a trajectory of social state formation that was similar to other US states, with two exceptions.
-First, these states were heavily invested in education, suggesting aggressive efforts to build the social and cultural fabric of a settler nation in recently settled lands.
-Second, these states were heavily invested in the elimination of spatially proximate Native nations through the forced assimilation of their children.
-This dual approach, characteristic of settler colonial projects of elimination and replacement, was on full display in states in the Midwest, Mountain West, Southwest, Great Plains, and West Coast.
-
-While Herrenvolk democracy prevailed in Jim Crow and Frontier (IS THIS THE RIGHT TERM???
-I DON'T THINK SO, BUT WE NEED A CONCISE LABEL) states, these regimes developed distinct patterns of governance and policy infrastructure.
-CHECK OUT MCQUADE ON HERRENVOLK WELFARE STATE.
-THIS IS KEY TO NAIL
-
-States with large urban immigrant populations developed bureaucratically complex therapeutic-welfare states with significant physical infrastructure for the confinement of socially problematic populations (that was nevertheless intensely racially exclusive CITE SIMMONS HARTMAN ETC).
-This therapeutic and paternalistic model (which co-developed with nascent social and psychological disciplines) has become the ideal typical form of the modern welfare-carceral state in dominant historical and social scientific accounts (e.g.
-GARLAND, FOUCAULT, OTHERS).
-However, it was far from the only model articulated during the US Progressive era.
-As XXX HAS NOTED IN THEIR CRITIQUE OF FOUCAULT / GARLAND, ETC THE SOUTH WAS REALLY DIFFERENT, BUT STILL MODERN.
-RACIAL CAPITALISM CRITIQUES SUGGEST THAT UNIVERSALIST NARRATIVES ARE INCOMPLETE - RACIST LOGICS AND PROJECTS RESULTED IN HETEROGENEITY IN STATE BUILDING.
-BUT EVEN THESE CRITIQUES TYPICALLY FAIL TO ACCOUNT FOR THE MODEL THAT WAS DEVELOPED BY STATES SEEKING TO ELIMINATE NATIVE NATIONS AND BUILD NASCENT WHITE SETTLER NATIONS.
-
-FUTURE WORK SHOULD TRACE THESE PATTERNS MOVING FORWARD.
-DIRECT FOLLOW UP IS TO DO A TRAJECTORY ANALYSIS ON THESE OUTCOMES.
-To what extent did these foundations influence trajectories?
-To what extent do trajectories impact present patterns?
-How did they become durable?
-How did they converge?
-Role of federalism?
-Lots of questions here.
-But we know that there are HUGE differences across states today, and in many cases, they mirror the kinds of patterns we observe here.
-We should better appreciate the deeply complicated causal processes that have created variation in logics and infrastructures for carceral and welfare states across places today.
-
-These are three of the projects that were critical to settler colonialism in the early 20th c US.
-Not the only ones of course!
-Colonial processes deeply impacted the experiences of a broad range of groups and produced a range of distinct racialized social structures (CITES FOR MEXICO BORDER, MIGRANT WORKERS, CHINESE EXCLUSION, JAPANESE INTERNMENT, PHILIPPINES COLONIZATION, AND OTHERS).
-*Here, we narrowly focus on three settler colonial projects to illustrate how a macro-comparative approach can be helpful in validating core findings in historical scholarship and in incorporating the insights of critical scholarship on settler colonization and racial capitalism into some of the core methods of political, economic, and comparative historical sociology REVISE.*
-
-AK AND HI EXCLUDED, EVEN THOUGH SETTLER COLONIZATION WAS V IMPORTANT THERE!
-THE DATA AND LAW ARE V DIFFERENT BECAUSE OF STATEHOOD TIMING.
-ALSO SHOULD CONSIDER OTHER COLONIES OF US LIKE PR, GUAM, USVI, PHILLIPINES, ETC
+       x = "Year",
+       subtitle = "3. Institutions for poor, disabled, and elderly")
+
+grid.arrange(p1, p2, p3)
+```
+
+#### Prisons and jails
+
+There was considerable variation in populations detained in correctional institutions across places and over time during the Progressive Era.
+We estimate large differences in average incarceration rates across places ($\sigma_\gamma = 0.31 [0.09, 0.54]$), but have a high level of uncertainty about any meaningful differences in state-level growth trajectories ($\sigma_\delta = 0.05 [0, 0.14]$).
+As shown in Figure \@ref(fig:gq-vis), incarceration rates trended upward in high sharecropping and high land redistribution contexts during this period.
+In contexts of large white urban immigrant populations, incarceration rates remained at relatively high levels throughout the period, despite a small decline in expected incarceration over these four decades.
+
+In state with large urban immigrant populations, our models suggest that incarceration rates could be expected to be in the range of about 1-2 per 1,000 population between 1900 and 1930.
+We estimate an expected incarceration rate of 1.7 [1.2, 2.3] for this scenario in 1900, much higher than expected incarceration rates in other contexts.
+This rate remains relatively stable over the full period with a small expected decline to a rate of 1.2 [0.9, 1.7] by 1930.
+
+In states with high levels of Native land redistribution, we see a substantial rise in expected incarceration.
+In 1900, we expect a state with high levels of Morill and Homestead acres patented to have an incarceration rate of about 0.6 [0.3, 1.0] per 1,000 population.
+This expected value rises rapidly to 0.85 [0.6, 1.2] in 1910, 1.2 [0.9, 1.6] in 1920 and 1.8 [1.3, 2.4] in 1930.
+This growth represents about 300 percent growth in expected incarceration rates.
+By 1930, these states had higher expected incarceration rates than other scenarios, though there is a substantial amount of uncertainty in the rank order of scenarios for 1930.
+
+States with a large population of Black tenant farmers also saw substantial growth in incarceration rates during this period.
+These states also started with significant prison and jail capacities.
+We estimate an expected incarceration rate of 0.8 [0.6, 1.1] per 1,000 population in high sharecropping states in 1900.
+By 1930, expected incarceration rates rose by about 90 percent by 1930 to a rate of 1.5 [1.1, 2.2].
+Despite relatively thin infrastructure on other measures, states in the Jim Crow South had incarceration rates that were similar to the rest of the country.
+Carceral investments appear to be an exception to the trend of under-development in other areas of social infrastructure in high sharecropping states.
+
+```{r callouts4}
+temp_prison <- preds_prison %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred) , 
+            yhat_lwr = quantile(.epred, 0.025), 
+            yhat_upr = quantile(.epred, 0.975)) %>% 
+  arrange(scen, year_c)
+
+temp_mental <- preds_mental %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred) , 
+            yhat_lwr = quantile(.epred, 0.025), 
+            yhat_upr = quantile(.epred, 0.975)) %>% 
+  arrange(scen, year_c)
+
+temp_poor <- preds_poor %>% 
+  group_by(year_c, scen) %>% 
+  summarize(yhat_mn = mean(.epred) , 
+            yhat_lwr = quantile(.epred, 0.025), 
+            yhat_upr = quantile(.epred, 0.975)) %>% 
+  arrange(scen, year_c)
+```
+
+#### Mental institutions
+
+There are clear and striking regional patterns in the distribution of the US population housed in mental institutions across states during the Progressive Era.
+As shown in Figure \@ref(fig:gq-vis) Panel 2, states with large white urban immigrant populations had much higher levels of confinement in mental institutions than did other states.
+However, these levels appear to have remained flat during this 40-year period.
+Mental institutionalization grew rapidly in states with high levels of settler land redistribution.
+Capacity in mental institutions appears to have grown, albeit at a slow rate, in high sharecropping states.
+We find substantial variation in individual state intercepts ($\sigma_\gamma = 0.32 [0.07, 0.56]$) but thin evidence for heterogeneity in state-specific growth trajectories ($\sigma_\delta = 0.05[0.00,0.14]$).
+
+We find that states actively redistributing Native lands to settlers have an expected mental institutionalization rate of about 1.5 [0.9, 2.4] per 1,000 population in 1900.
+This rate rose dramatically during the Progressive era.
+Mean expected values rose by about 290 percent between 1900 and 1930, ending the period at a rate of about 4.5 [3.6, 5.6] per 1,000.
+This rapid growth suggests substantial investments in expanding capacity for mental institutionalization in states undergoing rapid settler land redistribution in the early 20th century.
+
+States in the sharecropping South tended to have lower levels of mental institutionalization than other regions.
+In 1900, rates of institutionalization were relatively low, at 1.89 [1.4, 2.5] per 1,000 population, similar to levels observed in high land redistribution states.
+However, rates of institutionalization grew much less quickly in high sharecropping states than in high land redistribution states.
+By 1930, these states had an expected rate of mental institutionalization of 3.2 [2.5, 4.0], somewhat lower than high sharecropping states and substantially lower than high urban immigrant population states.
+Like other paternalistic welfare and public health institutions, states in the US South appear to have invested far fewer resources in infrastructure when compared with peer states (UNDERDEVELOPMENT CITES?).
+
+#### Institutions for the poor, disabled and elderly
+
+As with mental institutions, we find large expected increases in populations housed in institutions for the poor, elderly, and disabled in states with high levels of settler land redistribution between 1900 and 1930, relatively small increases in size in high sharecropping states, and small decreases in already-high levels in states with large numbers of white urban immigrants.
+We find substantial heterogeneity in both intercepts ($\sigma_\gamma = 0.6[0.39, 0.83]$ and in their growth trajectories $[\sigma_\delta = 0.09[0.01.
+0.18]$).
+
+High sharecropping states in the South tended to have very low levels of populations in these kinds of institutions across the period.
+In 1900, we estimate an expected rate of 0.96 [0.6, 1.4] residents per 1,000 population housed in institutions for the poor, disabled and elderly.
+This rate grew slowly over the 40 year period, to and avergage of 1.1 in 1910, 1.3 in 1920, and 1.5 [1.1, 2.1] in 1930.
+Again, these results indicate generally lower levels of investment in social institutions in the Jim Crow South than in other settler contexts.
+
+We find that states with large populations of urban immigrants had relatively large numbers of people housed in poverty institutions like nursing homes, poorhouses, and orphanages.
+In 1900, these states have an expected rate of 4.3 [2.7, 6.3] per 1,000 housed in poverty institutions.
+By 1930, this rate had fallen slightly to 3.7 [2.6, 5.1] per 1,000.
+Despite this decline, these states had higher levels of housing in poverty institutions in 1930 than did states in our other two focal scenarios.
+
+As with other infrastructure outcomes, we see a rapid growth in the population of people housed in poverty institutions in states with high levels of settler land redistribution.
+We expect a rate of about 0.79 [0.4, 1.4] per 1,000 population in these facilities in 1900, rising to about 2.6 [1.9, 3.4] per 1,000 in 1930.
+These states had a rate of growth that was significantly faster than states enmeshed in other settler colonial projects during the Progressive Era.

### 2024-06-28 make_pop_data.R

@@ -1,76 +1,76 @@
-## ----setup, include = F, message = F----------------------------------
-library(tidyverse)
-library(maps)
-
-## ----read IPUMS tables--------------------------------------------------
-# ### occupations 
-dat_occ<-read_csv("./data/usa_00071.csv") %>%
-  rename_all(tolower) %>%
-  mutate(occ1950 =
-           case_when(
-             occ1950 == 79 ~ "socwork",
-             occ1950 == 773 ~ "police",
-             occ1950 == 771 ~ "police",
-             occ1950 == 782 ~ "police",
-             occ1950 == 93 ~ "teachers"
-           )) %>%
-  group_by(year, statefip, occ1950) %>%
-  summarize(emp = sum(perwt)) %>%
-  pivot_wider(names_from = occ1950, values_from = emp,
-              values_fill = 0)
-
-# pop (total, age, race, nativitiy, urban, farm, landown)
-dat_immig<-read_csv("./data/usa_00076.csv") %>%
-  rename_all(tolower) %>%
-  group_by(year, statefip) %>%
-  summarize(tot_pop = sum(perwt),
-            children = sum((age<18) * perwt),
-            white_immig =
-              sum((nativity > 1) * (race == 1) * (hispan == 0) * perwt),
-            white_non_immig = 
-              sum((nativity == 1) * (race == 1) * (hispan == 0) * perwt),
-            blk =
-              sum((race == 2) * perwt),
-            blk_sharecrop = 
-              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
-            blk_farms = sum((race==2) * (farm==2) * perwt),
-            urban_pop = sum(hhwt * (urban ==2)),
-            urban_immig_pop = 
-              sum((urban==2) * 
-                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
-            tot_hh = sum(hhwt))
-### group quarters, 100%
-gq<-read_csv("./data/usa_00069.csv") %>%
-  rename_all(tolower) %>%
-  filter(year!=1910) %>% 
-  group_by(year, statefip, gqtype) %>%
-  summarize(gq_pop = sum(perwt)) %>%
-  mutate(gqtype = case_when(
-    gqtype == 2 ~ "inst_correctional",
-    gqtype == 3 ~ "inst_mental",
-    gqtype == 4 ~ "inst_poor_disabled")) %>% 
-  pivot_wider(names_from = gqtype, values_from = gq_pop,
[…]

### 2024-06-28 models.R

@@ -1,5 +1,8 @@
 ## sourced by ms.Rmd
 
+
+### CONSIDER SPLITTING MENTAL INST AND POVERTY INST????
+
 mdat<-pop %>% 
   left_join(dist) %>% 
   left_join(bs) %>% 
@@ -10,6 +13,8 @@ mdat<-pop %>%
 #### set weakly informative priors for model parameters
 ### for m1
 ### try a growth model
+### try a growth model
+
 mdat<-mdat %>% 
   mutate(year_c = (year - 1900)/10)
 
@@ -22,200 +27,106 @@ priors_m1<-c(
 ### fit HMC model with brms
 teachers_m1<-brm(
   teachers ~ 
-    scale(blk_sharecrop_pc) +
-    scale(urban_immig_pc) +
-    scale(homestead_plus_morill) +
-    year_c + 
-    (1|STUSPS) + 
-    offset(log(tot_pop)),
-  family = negbinomial(),
-  prior = priors_m1,
-  data = mdat
-) 
-### try a growth model
-mdat<-mdat %>% 
-  mutate(year_c = (year - 1900)/10)
-
-teachers_m2<-brm(
-  teachers ~ 
-    (scale(blk_sharecrop_pc) +
-    scale(urban_immig_pc) +
-    scale(homestead_plus_morill)) *
-    factor(year_c) + 
-    (1|STUSPS) + 
-    offset(log(tot_pop)),
-  family = negbinomial(),
-  prior = priors_m1,
-  data = mdat) 
-# context interaction (slopes by predictor type)
-teachers_m3<-brm(
-  teachers ~ 
     (scale(blk_sharecrop_pc) +
        scale(urban_immig_pc) +
        scale(homestead_plus_morill)) *
     year_c + 
-    (1|STUSPS) + 
+    (year_c|STUSPS) + 
     offset(log(tot_pop)),
   family = negbinomial(),
   prior = priors_m1,
   data = mdat) 
 
-## compare m1 / m3, clear performance gain for m3
-# loo_compare(loo(teachers_m1), loo(teachers_m3))
-### adding year RE doesn't help.
-
-loo_compare(loo(teachers_m3), loo(teachers_m4))
-
-
-#### START HERE!!!!
-### TIME VARYING SLOPES ON CONTEXT
-### DOES INTERACTION ON SHARECROPPING PERSIST?
-### simulate
-
-year_c = 0:7
-STUSPS <- "XX"
[…]
